### 2022-06-27 src/tex/ms.tex
[showyourwork] first commit

% Define document class
\documentclass[twocolumn]{aastex631}
\usepackage{showyourwork}

% Begin!
\begin{document}

% Title
\title{An open source scientific article}

% Author list
\author{@tedwards2412}

% Abstract with filler text
\begin{abstract}
    Lorem ipsum dolor sit amet, consectetuer adipiscing elit. 
    Ut purus elit, vestibulum ut, placerat ac, adipiscing vitae, felis. 
    Curabitur dictum gravida mauris, consectetuer id, vulputate a, magna. 
    Donec vehicula augue eu neque, morbi tristique senectus et netus et. 
    Mauris ut leo, cras viverra metus rhoncus sem, nulla et lectus vestibulum. 
    Phasellus eu tellus sit amet tortor gravida placerat. 
    Integer sapien est, iaculis in, pretium quis, viverra ac, nunc. 
    Praesent eget sem vel leo ultrices bibendum. 
    Aenean faucibus, morbi dolor nulla, malesuada eu, pulvinar at, mollis ac. 
    Curabitur auctor semper nulla donec varius orci eget risus. 
    Duis nibh mi, congue eu, accumsan eleifend, sagittis quis, diam. 
    Duis eget orci sit amet orci dignissim rutrum.
\end{abstract}

% Main body with filler text
\section{Introduction}
\label{sec:intro}

Lorem ipsum dolor sit amet, consectetuer adipiscing elit. 
Ut purus elit, vestibulum ut, placerat ac, adipiscing vitae, felis. 
Curabitur dictum gravida mauris, consectetuer id, vulputate a, magna. 
Donec vehicula augue eu neque, morbi tristique senectus et netus et. 
Mauris ut leo, cras viverra metus rhoncus sem, nulla et lectus vestibulum. 
Phasellus eu tellus sit amet tortor gravida placerat. 
Integer sapien est, iaculis in, pretium quis, viverra ac, nunc. 
Praesent eget sem vel leo ultrices bibendum. 
Aenean faucibus, morbi dolor nulla, malesuada eu, pulvinar at, mollis ac. 
Curabitur auctor semper nulla donec varius orci eget risus. 
Duis nibh mi, congue eu, accumsan eleifend, sagittis quis, diam. 
Duis eget orci sit amet orci dignissim rutrum.

Nam dui ligula, fringilla a, euismod sodales, sollici- tudin vel, wisi. 
Morbi auctor lorem non justo, nam lacus libero, pretium at, lobortis vitae. 
Donec aliquet, tortor sed accumsan bibendum, erat ligula aliquet magna. 
Morbi ac orci et nisl hendrerit mollis, suspendisse ut massa, cras nec ante. 
Pellentesque a nulla cum sociis natoque penatibus et magnis dis parturient. 
Aliquam tincidunt urna, nulla ullamcorper vestibulum turpis. 
Pellentesque cursus luctus mauris \citep{Luger2021}.

\bibliography{bib}

\end{document}

### 2022-06-27 src/tex/ms.tex
modified author list
set up structure for paper

@@
 % Title
-\title{An open source scientific article}
+\title{Differentiable Waveforms for Gravitational Wave Data Analysis}
 
 % Author list
-\author{@tedwards2412}
+\author{Adam Coogan}
+\author{Thomas D.~P.~Edwards}
+\author{Daniel Foreman-Mackey}
+\author{Maximiliano Isi}
+\author{Kaze W.~K.~Wong}
+\author{Aaron Zimmerman}
 
@@
 \begin{abstract}
-    Lorem ipsum dolor sit amet, consectetuer adipiscing elit. 
-    Ut purus elit, vestibulum ut, placerat ac, adipiscing vitae, felis. 
-    Curabitur dictum gravida mauris, consectetuer id, vulputate a, magna. 
-    Donec vehicula augue eu neque, morbi tristique senectus et netus et. 
-    Mauris ut leo, cras viverra metus rhoncus sem, nulla et lectus vestibulum. 
-    Phasellus eu tellus sit amet tortor gravida placerat. 
-    Integer sapien est, iaculis in, pretium quis, viverra ac, nunc. 
-    Praesent eget sem vel leo ultrices bibendum. 
-    Aenean faucibus, morbi dolor nulla, malesuada eu, pulvinar at, mollis ac. 
-    Curabitur auctor semper nulla donec varius orci eget risus. 
-    Duis nibh mi, congue eu, accumsan eleifend, sagittis quis, diam. 
-    Duis eget orci sit amet orci dignissim rutrum.
+    Here we will discuss our implementation of differentiable waveforms and demonstrate their benefits for a variety of data analysis tasks.
 \end{abstract}
 
-% Main body with filler text
 \section{Introduction}
@@
 
-Lorem ipsum dolor sit amet, consectetuer adipiscing elit. 
-Ut purus elit, vestibulum ut, placerat ac, adipiscing vitae, felis. 
-Curabitur dictum gravida mauris, consectetuer id, vulputate a, magna. 
-Donec vehicula augue eu neque, morbi tristique senectus et netus et. 
-Mauris ut leo, cras viverra metus rhoncus sem, nulla et lectus vestibulum. 
-Phasellus eu tellus sit amet tortor gravida placerat. 
-Integer sapien est, iaculis in, pretium quis, viverra ac, nunc. 
-Praesent eget sem vel leo ultrices bibendum. 
-Aenean faucibus, morbi dolor nulla, malesuada eu, pulvinar at, mollis ac. 
-Curabitur auctor semper nulla donec varius orci eget risus. 
-Duis nibh mi, congue eu, accumsan eleifend, sagittis quis, diam. 
-Duis eget orci sit amet orci dignissim rutrum.
+Derivatives are ubiqitously useful throughout data analysis tasks.
+However, in the field of gravitational wave (GW) data analysis, they have historically been difficult to obtain analytically.
+Numerical derivatives also suffer from accuracy issues stemming from rounding or truncation errors.
+Recent progress in automatic differentiation (AD) has.... \citep{Coogan:2022qxs}
 
-Nam dui ligula, fringilla a, euismod sodales, sollici- tudin vel, wisi. 
-Morbi auctor lorem non justo, nam lacus libero, pretium at, lobortis vitae. 
-Donec aliquet, tortor sed accumsan bibendum, erat ligula aliquet magna. 
-Morbi ac orci et nisl hendrerit mollis, suspendisse ut massa, cras nec ante. 
-Pellentesque a nulla cum sociis natoque penatibus et magnis dis parturient. 
-Aliquam tincidunt urna, nulla ullamcorper vestibulum turpis. 
-Pellentesque cursus luctus mauris \citep{Luger2021}.
+\section{Differentiable Waveforms}
+\label{sec:waveforms}
+
+\subsection{Benchmarks}
+\label{subsec:benchmarks}
+
+\section{Applications}
+\label{sec:applications}
+
+\subsection{Effectualness Calculations}
+\label{sec:effectualness}
+
+\subsection{Fisher Analysis}
+\label{subsec:fisher}
+
+\subsection{Derivative Based Samplers - Hamiltonian Monte Carlo}
+\label{subsec:hmc}
 

### 2022-07-01 src/tex/ms.tex
Started to add text plus first effectualness calculation using ripple

@@
 
+The discovery of gravitational waves (GWs) from inspiraling and mergering compact objects (COs) has revolutionized our understanding of both fundamental physics and astronomy.
+Although the data volumes coming from GW detectors are relatively small, analyzing the data is a computationally demanding task.
+In addition, this computational cost will substatially increase when next generation detectors come online.
+
 Derivatives are ubiqitously useful throughout data analysis tasks.
-However, in the field of gravitational wave (GW) data analysis, they have historically been difficult to obtain analytically.
+However, in the field of GW data analysis, analytic derivatives of the necessary quantities (such as the likelihood) have historically been difficult to obtain.
 Numerical derivatives also suffer from accuracy issues stemming from rounding or truncation errors.
@@
 
+In this paper we argue that differentiable waveforms will be a vital component for the future of GW data analysis.
+In addition, we present \texttt{ripple}, a small GW \texttt{python} package containing differentiable implementations of some of the main waveforms currently used in LIGO and Virgo analyses.
+
 \section{Differentiable Waveforms}
 \label{sec:waveforms}
+
+A variety of waveform families have been developed to accurately model the GW emission from COs. 
+When the COs are relatively well separated, the dynamics of the system can be well approximated with post-Newtonian corrections to the... .
+However, close to merger, numerical relatively simulations are required to accurately model the GW emission.
+Unfortunately, these numerical simulations are computationally expensive and cannot be run in conjunction with data analysis.
 
@@
 
+\subsection{Fitting Waveform Coefficients}
+\label{subsec:hmc}
+
 \bibliography{bib}
 
+\section{Appendix}
+\label{sec:appendix}
+
 \end{document}

### 2022-07-01 src/scripts/effectualness_demonstrator.py

@@ -1,0 +1,121 @@
+import jax.numpy as jnp
+import jax.scipy.optimize as jopt
+import scipy.optimize as opt
+import jax
+from jax.config import config
+config.update("jax_enable_x64", True)
+
+# from tqdm.auto import tqdm, trange
+from typing import Tuple
+from ripple.waveforms import IMRPhenomD
+from ripple import ms_to_Mc_eta
+import timeit
+
+
+Array = jnp.ndarray
+
+def get_eff_pads(fs: Array) -> Tuple[Array, Array]:
+    r"""
+    Gets arrays of zeros to pad a function evaluated on a frequency grid so the
+    function values can be passed to ``jax.numpy.fft.ifft``.
+    Args:
+        fs: uniformly-spaced grid of frequencies. It is assumed that the first
+            element in the grid must be an integer multiple of the grid spacing
+            (i.e., ``fs[0] % df == 0``, where ``df`` is the grid spacing).
+    Returns:
+        The padding arrays of zeros. The first is of length ``fs[0] / df`` and
+        the second is of length ``fs[-1] / df - 2``.
+    """
+    df = fs[1] - fs[0]
+    N = 2 * jnp.array(fs[-1] / df - 1).astype(int)
+    pad_low = jnp.zeros(jnp.array(fs[0] / df).astype(int))
+    pad_high = jnp.zeros(N - jnp.array(fs[-1] / df).astype(int))
+    return pad_low, pad_high
+
+@jax.jit
+def get_match_arr(
+    h1: Array, h2: Array, Sns: Array, pad_low: Array, pad_high: Array
+) -> Array:
+    """
+    Calculates the match between two frequency-domain complex strains. The maximizations
+    over the difference in time and phase at coalescence are performed by taking
+    the absolute value of the inverse Fourier transform.
+    Args:
+        h1: the first set of strains
+        h2: the second set of strains
+        Sns: the noise power spectral densities
+        fs: frequencies at which the strains and noise PSDs were evaluated
+        pad_low: array of zeros to pad the left side of the integrand before it
+            is passed to ``jax.numpy.fft.ifft``
+        pad_right: array of zeros to pad the right side of the integrand before
[…]

### 2022-07-01 src/tex/ms.tex
Added affiliations and acknowledgements

@@
 
+\newcommand{\OKC}{\affiliation{The Oskar Klein Centre, Department of Physics, Stockholm University, AlbaNova, SE-106 91 Stockholm, Sweden}} 
+\newcommand{\NORDITA}{\affiliation{Nordic Institute for Theoretical Physics (NORDITA), 106 91 Stockholm, Sweden}}
+\newcommand{\UdeM}{\affiliation{Département de Physique, Université de Montréal, 1375 Avenue Thérèse-Lavoie-Roux, Montréal, QC H2V 0B3, Canada}} 
+\newcommand{\Mila}{\affiliation{Mila -- Quebec AI Institute, 6666 St-Urbain, \#200, Montreal, QC, H2S 3H1}} 
+\newcommand{\CGP}{\affiliation{Center for Gravitational Physics, University of Texas at Austin, Austin, TX 78712, USA}}
+\newcommand{\CCA}{\affiliation{Center for Computational Astrophysics, Flatiron Institute, New York, NY 10010, USA}}
+
 % Author list
-\author{Adam Coogan}
-\author{Thomas D.~P.~Edwards}
-\author{Daniel Foreman-Mackey}
-\author{Maximiliano Isi}
-\author{Kaze W.~K.~Wong}
-\author{Aaron Zimmerman}
+\author{Adam Coogan} \UdeM \Mila
+\author{Thomas D.~P.~Edwards} \OKC \NORDITA
+\author{Daniel Foreman-Mackey} \CCA
+\author{Maximiliano Isi} \CCA
+\author{Kaze W.~K.~Wong} \CCA
+\author{Aaron Zimmerman} \CGP
 
@@
 
+\section{Acknowledgments}
+This work was supported by collaborative visits funded by the Cosmology and Astroparticle Student and Postdoc Exchange Network (CASPEN).
+
 \bibliography{bib}

### 2022-07-18 src/tex/ms.tex
Adding first draft of intro

@@
 % Title
-\title{Differentiable Waveforms for Gravitational Wave Data Analysis}
+\title{\texttt{ripple}: Differentiable Waveforms for Gravitational Wave Data Analysis}
 
@@
 
+% Intro outline
+% - What are gravitational waves and what are data analysis tasks?
+% - What are derivatives, and how can they be used?
+% - What is automatic differentiation, and why has it become particularly useful now? JAX?
+% - Discussion of what waveforms are most ameanable to being differentiable
+% - In this paper
+
 The discovery of gravitational waves (GWs) from inspiraling and mergering compact objects (COs) has revolutionized our understanding of both fundamental physics and astronomy.
-Although the data volumes coming from GW detectors are relatively small, analyzing the data is a computationally demanding task.
+Although the data volumes from GW detectors are relatively small, analyzing the data is a computationally demanding task.
 In addition, this computational cost will substatially increase when next generation detectors come online.
+The complexity begins before data gathering where one is required to generate banks of template waveforms which will be used for a matched-filter search~\citep{Owen:1998dk, Owen:1995tm}. 
+Once potential candidates are found, parameter estimation (PE) is performed to extract the detailed source properties of each event.
+In the most minimial example, this requires one to perform MCMC over a 15 dimensional parameter space for each event.
+Beyond this simple scenario, more complex waveform models with many additional parameters may be used to test for deviations from General Relativity.
+Finally, using the results of the PE, population synthesis models are used to constrain the progenitors systems from which the black holes we see merging today began their journey~\citep{Wong:2022flg}.
+All of these tasks require significant compute.
+In this paper, we will argue that differentiable waveforms (and more generally differentiable pipelines) can play a significant role in alleaviating this computational demand.
 
 Derivatives are ubiqitously useful throughout data analysis tasks.
-However, in the field of GW data analysis, analytic derivatives of the necessary quantities (such as the likelihood) have historically been difficult to obtain.
+For instance, during PE, derivative information can be used to guide the sampling towards regions with higher likelihood values (e.g.
+in Hamiltonian Monte Carlo~\citep{2017arXiv170102434B} or Gradient Decsent).
+This use of the gradients is particularly useful for high dimensional spaces.
+Unfortunately, in the field of GW data analysis, analytic derivatives of the necessary quantities (such as the likelihood) have historically been difficult to obtain.
 Numerical derivatives also suffer from accuracy issues stemming from rounding or truncation errors.
-Recent progress in automatic differentiation (AD) has.... \citep{Coogan:2022qxs}
+However, recent progress in automatic differentiation (AD) has shown promise in allowing general, fast derivative calcuations for gravitational waveforms~\citep{Coogan:2022qxs}.
+
+Automatic Differentiaton is a family of methods used to compute the machine precision derivatives with little computational ovearhead. 
+Its has become popular due to its requirement in NNs...
+Core idea is chain rule.
+Why JAX?
+
+There are a variety of gravitational waveforms currently used in analysis pipelines.
+They are generally structured into different families, the most common of which are: the effective one body, inspiral-merger-ringerdown phenom, and numerical GW....
+Unfortunately, not all are well suited for easy implementation into
 
@@
 In addition, we present \texttt{ripple}, a small GW \texttt{python} package containing differentiable implementations of some of the main waveforms currently used in LIGO and Virgo analyses. 
+The remainder of this paper is structured as follows. 
+In Sec.~\ref{sec:waveforms} we discuss the differentiable waveforms implemented in \texttt{ripple} and perform some benchmarks to demonstrate the speed and accruacy of the derivative calculations. 
+In Sec.~\ref{sec:applications} we discuss four distinct applications using differentiable waveforms. 
+First, a simple demonstration of accelerated effectualness calculations using gradient descent.
+Second, we show that Fisher analyses are quickly enabled using.
+Third, we implement differnetiable detector response functions and show an illutrative example Hamiltonian Monte Carlo
+Finally, we illustrate how the fit coefficients that form part of the IMRPhenom waveform models could be improved by high dimensional fitting, enabled by a differentiable waveform. 
+The differentiable waveforms can be found at \texttt{ripple}~\cite{ripple}.
 

### 2022-07-19 src/tex/ms.tex
Updated intro

@@
 \usepackage{showyourwork}
+
+% \usepackage[dvipsnames]{xcolor}
+% \usepackage[colorlinks]{hyperref}
+% \usepackage{colortbl}
+\definecolor{pyBlue}{RGB}{31, 119, 180}
+\definecolor{pyRed}{RGB}{214, 39, 40}
+\definecolor{pyGreen}{RGB}{44, 160, 44}
+
+% Link setup
+% \newcommand\myshade{80}
+% \hypersetup{
+%   linkcolor  = pyGreen!\myshade!black,
+%   citecolor  = pyBlue!\myshade!black,
+%   urlcolor   = pyRed!\myshade!black,
+%   colorlinks = true
+% }
+
+\newcommand{\TE}[1]{{\color{pyGreen} #1}}
+\newcommand{\te}[1]{\textbf{\color{pyGreen}(TE: #1)}}
 
@@
 in Hamiltonian Monte Carlo~\citep{2017arXiv170102434B} or Gradient Decsent).
-This use of the gradients is particularly useful for high dimensional spaces.
+This use of gradients is particularly useful for high dimensional spaces.
 Unfortunately, in the field of GW data analysis, analytic derivatives of the necessary quantities (such as the likelihood) have historically been difficult to obtain.
@@
 
-Automatic Differentiaton is a family of methods used to compute the machine precision derivatives with little computational ovearhead. 
-Its has become popular due to its requirement in NNs...
-Core idea is chain rule.
-Why JAX?
+Automatic differentiaton is a family of methods used to compute the machine precision derivatives with little computational ovearhead. 
+AD's recent ascendance is primarily driven by its use in machine learning, particularly for derivative computations of neural networks which require use gradient descent during training.
+The core idea of AD is that each any any mathematical function can be broken down into a small set of basic operations, each with a known differentiation rule.\footnote{
+    Of course, non-differentiable functions exist and care must be taken when treating these special cases.
+    }
+The full derivative can then be contructed using the chain rule.
+There are now a variety of AD implementations, most notably in deep learning frameworks such as \texttt{pytorch}~\citep{pytorch} and \texttt{tensorflow}~\citep{tensorflow2015-whitepaper}.
+More general frameworks exist in \texttt{julia}~\citep{zygote, forwarddiff}, although \texttt{julia}'s limited use in GW physics precludes there general use.
+Here we make use of \texttt{jax}~\citep{jax2018github} due to its easy integration with \texttt{python} libraries, ability to just-in-time compile, and ability to utilize a variety of computational resources (i.e.
+CPUs and GPUs) without changes to the code.
+\te{Do we need a longer intro to AD here?}
 
 There are a variety of gravitational waveforms currently used in analysis pipelines.
-They are generally structured into different families, the most common of which are: the effective one body, inspiral-merger-ringerdown phenom, and numerical GW....
-Unfortunately, not all are well suited for easy implementation into
+They are generally structured into different families, the most common of which are: the effective one body, the phenomenological inspiral-merger-ringerdown (IMR), and numerical relativity surrogate.
+Of these, the phenomenological IMR family is most well suited for an AD implementation in JAX.
+In particular, its closed form expression, ... \TE{Why else?}
+\te{Do we need more of an intro for why IMRPhenom?}
 
@@
 First, a simple demonstration of accelerated effectualness calculations using gradient descent.
-Second, we show that Fisher analyses are quickly enabled using.
-Third, we implement differnetiable detector response functions and show an illutrative example Hamiltonian Monte Carlo
+Second, we show that Fisher analyses are quickly enabled using AD.
+Third, we implement differnetiable detector response functions and show an illutrative example using Hamiltonian Monte Carlo.
 Finally, we illustrate how the fit coefficients that form part of the IMRPhenom waveform models could be improved by high dimensional fitting, enabled by a differentiable waveform. 
-The differentiable waveforms can be found at \texttt{ripple}~\cite{ripple}.
+The associated code can be found at \texttt{ripple}~\citep{ripple}.
 

### 2022-09-26 src/tex/ms.tex
Making progress on section 2. Finished first draft of intro

@@
 
-Automatic differentiaton is a family of methods used to compute the machine precision derivatives with little computational ovearhead. 
-AD's recent ascendance is primarily driven by its use in machine learning, particularly for derivative computations of neural networks which require use gradient descent during training.
-The core idea of AD is that each any any mathematical function can be broken down into a small set of basic operations, each with a known differentiation rule.\footnote{
+Automatic differentiaton is a family of methods used to compute machine precision derivatives with little computational ovearhead. 
+AD's recent ascendance is primarily driven by its use in machine learning, particularly for derivative computations of neural networks which use gradient descent during training.
+The core idea of AD is that each any mathematical function can be broken down into a small set of basic operations, each with a known differentiation rule.\footnote{
     Of course, non-differentiable functions exist and care must be taken when treating these special cases.
@@
 There are now a variety of AD implementations, most notably in deep learning frameworks such as \texttt{pytorch}~\citep{pytorch} and \texttt{tensorflow}~\citep{tensorflow2015-whitepaper}.
-More general frameworks exist in \texttt{julia}~\citep{zygote, forwarddiff}, although \texttt{julia}'s limited use in GW physics precludes there general use.
+More general frameworks exist in \texttt{julia}~\citep{zygote, forwarddiff}, although \texttt{julia}'s limited use in GW physics precludes its general use.
 Here we make use of \texttt{jax}~\citep{jax2018github} due to its easy integration with \texttt{python} libraries, ability to just-in-time compile, and ability to utilize a variety of computational resources (i.e.
@@
 Unfortunately, these numerical simulations are computationally expensive and cannot be run in conjunction with data analysis.
+Approximate phenomenological waveforms have therefore been constructued to enable relatively fast waveform generation at sufficient accuracy.
+
+The three major waveform families that have been developed to date are: the Effective-one-body (EOB), numerical relativity (NR) surrogate, and phenomenological inspiral-merger-ringdown (PhenomIMR).
+Note that unlike NR surrogate, both the EOB and PhenomIMR waveforms are calibrated to NR waveforms.
+EOB waveforms require one to model the binary system using a Hamiltonian and are typically slow to evaluate where as PhenomIMR waveforms are constructed with simple closed-form expressions.
+PhenomIMR waveforms are therefore ideally suited for AD, espeically using \texttt{jax}. 
+In this paper we focus mainly on the aligned spin circular orbit model IMRPhenomD~\citep{Husa:2015iqa, 2015jqa}.
+
+The implementation of IMRPhenomD in lalsuite is mainly in C, and therefore needs to be rewritten natively into \texttt{pyhton} to be compatible with \texttt{jax}.
+We have re-written IMRPhenomD from scratch using a combination of pure \texttt{pyhton} and \texttt{jax} derivatives.
+In addition, we have restructed the code for readability and evaluation speed as well as exposing the internal fitting coefficients to the user (which we will use later in Sec.~\ref{sec:applications}).
+
+To ensure our implempation of IMRPhenomD is faithful to the lalsuite implementation, we start by defining the noise weighted inner product:
+\begin{equation}
+    \left(h_{\theta_1}|h_{\theta_2}\right) \equiv 4 \, \mathrm{Re} \int^{\infty}_{0} \mathrm{d} f \, \frac{ h_{\theta_1}(f) h^*_{\theta_2}(f)}{S_n(f)}\, ,
+\end{equation}
+where $S_n$ is the (one-sided) noise power spectral density (PSD) and $h_{\theta_1}$ is the frequency domain waveform evaluated at the instrinsic parameter point $\theta_1$.
+We can then normalize the inner product through
+\begin{equation}
+    \left[h_{\theta_1}|h_{\theta_2}\right] = \frac{\left(h_{\theta_1}|h_{\theta_2}\right)}{\sqrt{\left(h_{\theta_1}|h_{\theta_1}\right)\left(h_{\theta_2}|h_{\theta_2}\right)}}\, .
+\end{equation}
+Now we are ready to define the \textit{match}
+
+\begin{figure}[ht!]
+    \script{random_numbers.py}
+    \begin{centering}
+        \includegraphics[width=\linewidth]{figures/random_numbers.pdf}
+        \caption{
+            This plot should show a match colour bar.
+        }
+        \label{fig:random_numbers}
+    \end{centering}
+\end{figure}
 

### 2022-09-26 src/scripts/random_numbers.py

@@ -1,0 +1,44 @@
+import matplotlib.pyplot as plt
+import numpy as np
+import paths
+import matplotlib as mpl
+
+params = {
+    "font.size": 18,
+    "legend.fontsize": 18,
+    "legend.frameon": False,
+    "axes.labelsize": 18,
+    "axes.titlesize": 18,
+    "xtick.labelsize": 18,
+    "ytick.labelsize": 18,
+    "figure.figsize": (7, 5),
+    "xtick.top": True,
+    "ytick.right": True,
+    "xtick.bottom": True,
+    "ytick.left": True,
+    "xtick.major.size": 8,
+    "xtick.minor.size": 4,
+    "ytick.major.size": 8,
+    "ytick.minor.size": 4,
+    "xtick.direction": "in",
+    "ytick.direction": "in",
+    "text.usetex": True,
+    "font.family": "serif",
+    "axes.linewidth": 1.5,
+    "mathtext.fontset": "dejavuserif",
+}
+
+
+mpl.rcParams["text.latex.preamble"] = [r"\usepackage{amsmath}"]  # for \text command
+
+mpl.rcParams.update(params)
+
+# Generate some data
+random_numbers = np.random.randn(100, 10)
+
+# Plot and save
+fig = plt.figure(figsize=(7, 6))
+plt.plot(random_numbers)
+plt.xlabel("x")
+plt.ylabel("y")
+fig.savefig(paths.figures / "random_numbers.pdf", bbox_inches="tight", dpi=300)

### 2022-09-29 src/tex/ms.tex
Added first figure and nearly finised section 2

@@
 They are generally structured into different families, the most common of which are: the effective one body, the phenomenological inspiral-merger-ringerdown (IMR), and numerical relativity surrogate.
-Of these, the phenomenological IMR family is most well suited for an AD implementation in JAX.
+Of these, the phenomenological IMR family is most well suited for an AD implementation in \texttt{jax}.
 In particular, its closed form expression, ... \TE{Why else?}
@@
 \end{equation}
-Now we are ready to define the \textit{match}
+Now we are ready to define the \textit{match} which is given by 
+\begin{equation}
+    \mathrm{m}(h_{\theta_1}|h_{\theta_2}) \equiv \max_{t_c,\, \phi_c} \left[h_{\theta_1} \, \middle| \, h_{\theta_2} \right]\,
+\end{equation}
+where $t_c$ and $\phi_c$ are, respectively, the time and phase of coalesence.
 
+Since the match is a measure of the difference between two waveforms, we can use it to demonstrate that the implementation of IMRPhenomD in \texttt{ripple} accurately matches the \texttt{lalsuite} implementation.
+This is shown in Fig.~\ref{fig:match}, where we have calculated the $\mathrm{m}(h_{\theta_1}|h^{*}_{\theta_1})$ across the entire parameter space.\footnote{
+    Specifically, we use 2000 points in the ranges $m_{1,2} = (1,50)\,M\_{\odot}$ and $\chi_{1,2} = (-1,1)$.
+In addition, we evaulated the waveforms on a frequency grid from $32\,$Hz to $1024\,$Hz with frequency spacing $\Delta f = 0.0125 \,$Hz.
+}
+Here, $h_{\theta_1}$ corresponds to the \texttt{ripple} waveform implementation and $h^{*}_{\theta_1}$ is the \texttt{lalsuite} implementation. 
+From Fig.~\ref{fig:match}, it is clear that the \texttt{ripple} waveform matches with the \texttt{lalsuite} waveform to numerical precision across the majority of the parameter space.
+In particular, the blue points indicate parameter points where the $1-\mathrm{m}(h_{\theta_1}|h^{*}_{\theta_1})=0$ and would therefore not appear in the log scale colorbar.
 \begin{figure}[ht!]
-    \script{random_numbers.py}
+    \script{random_matches.py}
     \begin{centering}
-        \includegraphics[width=\linewidth]{figures/random_numbers.pdf}
+        \includegraphics[width=\linewidth]{figures/random_matches.pdf}
         \caption{
@@
         }
-        \label{fig:random_numbers}
+        \label{fig:match}
     \end{centering}
@@
 
-\subsection{Benchmarks}
-\label{subsec:benchmarks}
+There remains some slight deviation between the two waveform implementations at higher masses. 
+This is likely due to the fact that cubic interpolators, which are used within IMRPhenomD to calculate the ringdown and damping frequencies, are not currently supported in \texttt{jax}.
+Instead, we initially use \texttt{scipy}'s cubic interpolator to create a fine grid of XXX values, which we then linearly interpret during waveform generation.
+Unfortunately, we cannot make the initial cubic interpolation arbitrarily fine as this would add additional computational overhead when loading the data during waveform evaluation.
+The remaining error in the waveform at high masses is therefore primarily down to the coarseness in this interpolation.
+Note however, that the differences are well below the accuracy requirements of the waveforms and will have no impact on any data analysis tasks.
+
+In order to be usuable in most data analysis tasks, a waveform needs to be fast to evaluate.
+Fortunately, the Phenom IMR waveforms are constructed from simple closed form expressions which are therefore computationally efficient.
+The primary arguement of this paper is that waveform derivatives will also be invauable to data analysis tasks.
 
@@
 \label{sec:effectualness}
-
-\subsection{Fisher Analysis}
-\label{subsec:fisher}
 

### 2022-09-29 src/scripts/random_matches.py

@@ -1,0 +1,56 @@
+import matplotlib.pyplot as plt
+import numpy as np
+import paths
+import matplotlib as mpl
+
+params = {
+    "font.size": 18,
+    "legend.fontsize": 18,
+    "legend.frameon": False,
+    "axes.labelsize": 18,
+    "axes.titlesize": 18,
+    "xtick.labelsize": 18,
+    "ytick.labelsize": 18,
+    "figure.figsize": (7, 5),
+    "xtick.top": True,
+    "ytick.right": True,
+    "xtick.bottom": True,
+    "ytick.left": True,
+    "xtick.major.size": 8,
+    "xtick.minor.size": 4,
+    "ytick.major.size": 8,
+    "ytick.minor.size": 4,
+    "xtick.direction": "in",
+    "ytick.direction": "in",
+    "text.usetex": True,
+    "font.family": "serif",
+    "axes.linewidth": 1.5,
+    "mathtext.fontset": "dejavuserif",
+}
+
+# mpl.rcParams["text.latex.preamble"] = [r"\usepackage{amsmath}"]  # for \text command
+
+mpl.rcParams.update(params)
+
+data = np.loadtxt(paths.data / "ripple_phenomD_matches.txt")
+exactmatch_indices = (1.0 - data[:, -1]) <= 0.0
+other_indices = (1.0 - data[:, -1]) > 0.0
+
+# Plot and save
+plt.figure(figsize=(7, 5))
+cm = plt.cm.get_cmap("inferno")
+sc = plt.scatter(
+    data[other_indices, 0],
+    data[other_indices, 1],
+    c=np.log10(1.0 - data[other_indices, -1]),
+    cmap=cm,
+)
+plt.colorbar(sc, label=r"$\log_{10}(1-\mathrm{Match})$")
+
+plt.scatter(data[exactmatch_indices, 0], data[exactmatch_indices, 1], color="C0")
+plt.xlabel(r"$m_1 \,\left[M_{\odot}\right]$")
+plt.ylabel(r"$m_2 \,\left[M_{\odot}\right]$")
+plt.xlim(1, 50)
+plt.ylim(1, 50)
+
+plt.savefig(paths.figures / "random_matches.pdf", bbox_inches="tight")

### 2022-09-29 src/scripts/random_numbers.py

@@ -1,44 +1,0 @@
-import matplotlib.pyplot as plt
-import numpy as np
-import paths
-import matplotlib as mpl
-
-params = {
-    "font.size": 18,
-    "legend.fontsize": 18,
-    "legend.frameon": False,
-    "axes.labelsize": 18,
-    "axes.titlesize": 18,
-    "xtick.labelsize": 18,
-    "ytick.labelsize": 18,
-    "figure.figsize": (7, 5),
-    "xtick.top": True,
-    "ytick.right": True,
-    "xtick.bottom": True,
-    "ytick.left": True,
-    "xtick.major.size": 8,
-    "xtick.minor.size": 4,
-    "ytick.major.size": 8,
-    "ytick.minor.size": 4,
-    "xtick.direction": "in",
-    "ytick.direction": "in",
-    "text.usetex": True,
-    "font.family": "serif",
-    "axes.linewidth": 1.5,
-    "mathtext.fontset": "dejavuserif",
-}
-
-
-mpl.rcParams["text.latex.preamble"] = [r"\usepackage{amsmath}"]  # for \text command
-
-mpl.rcParams.update(params)
-
-# Generate some data
-random_numbers = np.random.randn(100, 10)
-
-# Plot and save
-fig = plt.figure(figsize=(7, 6))
-plt.plot(random_numbers)
-plt.xlabel("x")
-plt.ylabel("y")
-fig.savefig(paths.figures / "random_numbers.pdf", bbox_inches="tight", dpi=300)

### 2022-10-07 src/tex/ms.tex
Wrote more of section 2

@@
 
+\newcommand{\jax}{\texttt{JAX}\xspace}
+\newcommand{\ripple}{\texttt{ripple}\xspace}
+\newcommand{\lalsuite}{\texttt{lalsuite}\xspace}
+
 \newcommand{\TE}[1]{{\color{pyGreen} #1}}
@@
 % Title
-\title{\texttt{ripple}: Differentiable Waveforms for Gravitational Wave Data Analysis}
+\title{\ripple: Differentiable Waveforms for Gravitational Wave Data Analysis}
 
@@
 More general frameworks exist in \texttt{julia}~\citep{zygote, forwarddiff}, although \texttt{julia}'s limited use in GW physics precludes its general use.
-Here we make use of \texttt{jax}~\citep{jax2018github} due to its easy integration with \texttt{python} libraries, ability to just-in-time compile, and ability to utilize a variety of computational resources (i.e.
+Here we make use of \jax~\citep{jax2018github} due to its easy integration with \texttt{python} libraries, ability to just-in-time compile, and ability to utilize a variety of computational resources (i.e.
 CPUs and GPUs) without changes to the code.
@@
 They are generally structured into different families, the most common of which are: the effective one body, the phenomenological inspiral-merger-ringerdown (IMR), and numerical relativity surrogate.
-Of these, the phenomenological IMR family is most well suited for an AD implementation in \texttt{jax}.
+Of these, the phenomenological IMR family is most well suited for an AD implementation in \jax.
 In particular, its closed form expression, ... \TE{Why else?}
@@
 In this paper we argue that differentiable waveforms will be a vital component for the future of GW data analysis.
-In addition, we present \texttt{ripple}, a small GW \texttt{python} package containing differentiable implementations of some of the main waveforms currently used in LIGO and Virgo analyses. 
+In addition, we present \ripple, a small GW \texttt{python} package containing differentiable implementations of some of the main waveforms currently used in LIGO and Virgo analyses. 
 The remainder of this paper is structured as follows. 
-In Sec.~\ref{sec:waveforms} we discuss the differentiable waveforms implemented in \texttt{ripple} and perform some benchmarks to demonstrate the speed and accruacy of the derivative calculations. 
+In Sec.~\ref{sec:waveforms} we discuss the differentiable waveforms implemented in \ripple and perform some benchmarks to demonstrate the speed and accruacy of the derivative calculations. 
 In Sec.~\ref{sec:applications} we discuss four distinct applications using differentiable waveforms. 
@@
 Finally, we illustrate how the fit coefficients that form part of the IMRPhenom waveform models could be improved by high dimensional fitting, enabled by a differentiable waveform. 
-The associated code can be found at \texttt{ripple}~\citep{ripple}.
+The associated code can be found at \ripple~\citep{ripple}.
 
@@
 EOB waveforms require one to model the binary system using a Hamiltonian and are typically slow to evaluate where as PhenomIMR waveforms are constructed with simple closed-form expressions.
-PhenomIMR waveforms are therefore ideally suited for AD, espeically using \texttt{jax}. 
+PhenomIMR waveforms are therefore ideally suited for AD, espeically using \jax. 
 In this paper we focus mainly on the aligned spin circular orbit model IMRPhenomD~\citep{Husa:2015iqa, 2015jqa}.
 
-The implementation of IMRPhenomD in lalsuite is mainly in C, and therefore needs to be rewritten natively into \texttt{pyhton} to be compatible with \texttt{jax}.
-We have re-written IMRPhenomD from scratch using a combination of pure \texttt{pyhton} and \texttt{jax} derivatives.
+The implementation of IMRPhenomD in \lalsuite is mainly in C, and therefore needs to be rewritten natively into \texttt{pyhton} to be compatible with \jax.
+We have re-written IMRPhenomD from scratch using a combination of pure \texttt{pyhton} and \jax derivatives.
 In addition, we have restructed the code for readability and evaluation speed as well as exposing the internal fitting coefficients to the user (which we will use later in Sec.~\ref{sec:applications}).
 
-To ensure our implempation of IMRPhenomD is faithful to the lalsuite implementation, we start by defining the noise weighted inner product:
+To ensure our implempation of IMRPhenomD is faithful to the \lalsuite implementation, we start by defining the noise weighted inner product:
 \begin{equation}
@@
 
-Since the match is a measure of the difference between two waveforms, we can use it to demonstrate that the implementation of IMRPhenomD in \texttt{ripple} accurately matches the \texttt{lalsuite} implementation.
+Since the match is a measure of the difference between two waveforms, we can use it to demonstrate that the implementation of IMRPhenomD in \ripple accurately matches the \lalsuite implementation.
 This is shown in Fig.~\ref{fig:match}, where we have calculated the $\mathrm{m}(h_{\theta_1}|h^{*}_{\theta_1})$ across the entire parameter space.\footnote{
@@
 }
-Here, $h_{\theta_1}$ corresponds to the \texttt{ripple} waveform implementation and $h^{*}_{\theta_1}$ is the \texttt{lalsuite} implementation. 
-From Fig.~\ref{fig:match}, it is clear that the \texttt{ripple} waveform matches with the \texttt{lalsuite} waveform to numerical precision across the majority of the parameter space.
+Here, $h_{\theta_1}$ corresponds to the \ripple waveform implementation and $h^{*}_{\theta_1}$ is the \lalsuite implementation. 
+From Fig.~\ref{fig:match}, it is clear that the \ripple waveform matches with the \lalsuite waveform to numerical precision across the majority of the parameter space.
 In particular, the blue points indicate parameter points where the $1-\mathrm{m}(h_{\theta_1}|h^{*}_{\theta_1})=0$ and would therefore not appear in the log scale colorbar.
@@
 There remains some slight deviation between the two waveform implementations at higher masses. 
-This is likely due to the fact that cubic interpolators, which are used within IMRPhenomD to calculate the ringdown and damping frequencies, are not currently supported in \texttt{jax}.
+This is likely due to the fact that cubic interpolators, which are used within IMRPhenomD to calculate the ringdown and damping frequencies, are not currently supported in \jax.
 Instead, we initially use \texttt{scipy}'s cubic interpolator to create a fine grid of XXX values, which we then linearly interpret during waveform generation.
@@
 Fortunately, the Phenom IMR waveforms are constructed from simple closed form expressions which are therefore computationally efficient.
+Since the \lalsuite implementation of IMRPhenomD is written in \texttt{C}, the evaluation of the waveform itself is faster than \ripple.
+Benchmarking on a MacBook Pro with an $2.4$ GHz 8-Core Intel Core i9 processor, we find that a single waveform evaluation takes $\sim 1\,\mathrm{ms}$ for \lalsuite (interfaced with python) and $\sim 38\,\mathrm{s}$ for \ripple.\footnote{
+    For this benchmark we used $32\,\mathrm{Hz}$ and $1024\,\mathrm{Hz}$ for the lowest and highest frequencies respectively. In addition we used a frequency spacing of $0.25\,\mathrm{Hz}$.
+}
+Although slower on a CPU benchmark, a key advantage of using \jax is its native ability to run on a GPU.
+Waveform calls can therefore be evaluated in parallel, massively improving performance. \te{I couldn't get these to be similar on google colab. @Kaze any ideas how to make this fast?}
+
 The primary arguement of this paper is that waveform derivatives will also be invauable to data analysis tasks. 
+AD provides two big advantages when it comes to evaluating derivatives compared to numerical differentiation.
+Firstly, the accuracy of derivatives from AD are signficantly more stable than finite difference methods.
+In particular, finite differences suffer from both rounding and truncation errors, meaning that the user is required to tune the width over which the difference is taken.
+On the other hand, AD produces machine precision derivatives with no tuninig.
+Secondly, AD scales favourably with the dimensionality of the function.
+In particular, for every input dimension, $D$, added, one would need to evaluate the function at least $2D$ times to calcuate finite difference derivatives for all input parameters.
+For reverse mode AD, one only needs two function calls to evaluate the derivative of all input parameters, regardless of dimenison.\footnote{
+    Note that although the number of function calls is small, reverse mode AD does add memory overhead.
+    We've not found this to be limiting in any of the situations tested so far.
+}
+Since the parameter space of GWs in General Relativity (GR) has $\mathcal{O}(10)$ dimensons, the speed of derivative evaluation is less crucial than the stability.
+However, this might change for waveforms in beyond GR models where many more parameters are added.
+
+% Emphasise the stability of gradients.
+% Not just a faster way of evaluating PhenomD.
+% Change plot to be q vs chieff.
+% Note that lalsuite cannot be made massively parallel on GPUs.
+% Don't need a full section on Fisher, but make sure to mention. People love Fisher.
+% Use acceptance rate as a good explanation of speed ups
+% Wall time a messy comparison. Based too much off of hardware
+% Emphasise that gradients are beneficial
 

### 2022-10-07 src/scripts/random_matches.py

@@ -36,21 +36,27 @@ data = np.loadtxt(paths.data / "ripple_phenomD_matches.txt")
 exactmatch_indices = (1.0 - data[:, -1]) <= 0.0
 other_indices = (1.0 - data[:, -1]) > 0.0
 
+q = data[:, 1] / data[:, 0]
+chieff = (data[:, 0] * data[:, 2] + data[:, 1] * data[:, 3]) / (data[:, 0] + data[:, 1])
+Mtot = data[:, 1] + data[:, 0]
+
 # Plot and save
 plt.figure(figsize=(7, 5))
 cm = plt.cm.get_cmap("inferno")
 sc = plt.scatter(
-    data[other_indices, 0],
-    data[other_indices, 1],
+    q[other_indices],
+    chieff[other_indices],
     c=np.log10(1.0 - data[other_indices, -1]),
     cmap=cm,
 )
 plt.colorbar(sc, label=r"$\log_{10}(1-\mathrm{Match})$")
 
-plt.scatter(data[exactmatch_indices, 0], data[exactmatch_indices, 1], color="C0")
-plt.xlabel(r"$m_1 \,\left[M_{\odot}\right]$")
-plt.ylabel(r"$m_2 \,\left[M_{\odot}\right]$")
-plt.xlim(1, 50)
-plt.ylim(1, 50)
+plt.scatter(q[exactmatch_indices], chieff[exactmatch_indices], color="C0")
+# plt.xlabel(r"Total Mass, $M$")
+# plt.ylabel(r"Mass Ratio, $q$")
+plt.xlabel(r"Mass Ratio, $q$")
+plt.ylabel(r"Effective Spin, $\chi_{\rm eff}$")
+# plt.xlim(1, 50)
+# plt.ylim(1, 50)
 
 plt.savefig(paths.figures / "random_matches.pdf", bbox_inches="tight")

### 2022-10-13 src/tex/ms.tex
Finished first draft of section 2

@@
 Note that unlike NR surrogate, both the EOB and PhenomIMR waveforms are calibrated to NR waveforms.
-EOB waveforms require one to model the binary system using a Hamiltonian and are typically slow to evaluate where as PhenomIMR waveforms are constructed with simple closed-form expressions.
+EOB waveforms require one to model the binary system using a Hamiltonian and are typically slow to evaluate whereas PhenomIMR waveforms are constructed with simple closed-form expressions.
 PhenomIMR waveforms are therefore ideally suited for AD, espeically using \jax. 
@@
 This is shown in Fig.~\ref{fig:match}, where we have calculated the $\mathrm{m}(h_{\theta_1}|h^{*}_{\theta_1})$ across the entire parameter space.\footnote{
-    Specifically, we use 2000 points in the ranges $m_{1,2} = (1,50)\,M\_{\odot}$ and $\chi_{1,2} = (-1,1)$.
+    Specifically, we use 2000 points in the ranges $m_{1,2} = (1,50)\,M_{\odot}$ and $\chi_{1,2} = (-1,1)$.
 In addition, we evaulated the waveforms on a frequency grid from $32\,$Hz to $1024\,$Hz with frequency spacing $\Delta f = 0.0125 \,$Hz.
@@
 This is likely due to the fact that cubic interpolators, which are used within IMRPhenomD to calculate the ringdown and damping frequencies, are not currently supported in \jax.
-Instead, we initially use \texttt{scipy}'s cubic interpolator to create a fine grid of XXX values, which we then linearly interpret during waveform generation.
+Instead, we initially use \texttt{scipy}'s cubic interpolator to create a fine grid of XXX values, which we then linearly interpolate during waveform generation.
 Unfortunately, we cannot make the initial cubic interpolation arbitrarily fine as this would add additional computational overhead when loading the data during waveform evaluation.
 The remaining error in the waveform at high masses is therefore primarily down to the coarseness in this interpolation.
-Note however, that the differences are well below the accuracy requirements of the waveforms and will have no impact on any data analysis tasks.
+Note however, that the differences are well below the accuracy requirements of the waveforms and will have no noticeable for realistic data analysis tasks.
 
-In order to be usuable in most data analysis tasks, a waveform needs to be fast to evaluate.
+For maximum utility, a waveform needs to be fast to evaluate.
 Fortunately, the Phenom IMR waveforms are constructed from simple closed form expressions which are therefore computationally efficient.
 Since the \lalsuite implementation of IMRPhenomD is written in \texttt{C}, the evaluation of the waveform itself is faster than \ripple.
-Benchmarking on a MacBook Pro with an $2.4$ GHz 8-Core Intel Core i9 processor, we find that a single waveform evaluation takes $\sim 1\,\mathrm{ms}$ for \lalsuite (interfaced with python) and $\sim 38\,\mathrm{s}$ for \ripple.\footnote{
-    For this benchmark we used $32\,\mathrm{Hz}$ and $1024\,\mathrm{Hz}$ for the lowest and highest frequencies respectively. In addition we used a frequency spacing of $0.25\,\mathrm{Hz}$.
+Benchmarking on a MacBook Pro with an M1 Max Apple Silicon processor, we find that a single waveform evaluation takes $\sim 4\,\mathrm{ms}$ for \lalsuite (interfaced with python) and $\sim 10\,\mathrm{s}$ for \ripple.\footnote{
+    For this benchmark we used $16\,\mathrm{Hz}$ and $1024\,\mathrm{Hz}$ for the lowest and highest frequencies respectively. 
+    In addition, we used a frequency spacing of $0.0125\,\mathrm{Hz}$. 
+    We performed this benchmark by evaluating the waveform $5000$ times and taking the average evaluation time.
 }
@@
 Waveform calls can therefore be evaluated in parallel, massively improving performance. \te{I couldn't get these to be similar on google colab. @Kaze any ideas how to make this fast?}
+Generalizing \lalsuite waveforms to run on a GPU would be a significant undertaking.
 
@@
 Secondly, AD scales favourably with the dimensionality of the function.
-In particular, for every input dimension, $D$, added, one would need to evaluate the function at least $2D$ times to calcuate finite difference derivatives for all input parameters.
+In particular, for every input dimension, $D$, added one would need to evaluate the function at least $2D$ times to calcuate finite difference derivatives for all input parameters.
 For reverse mode AD, one only needs two function calls to evaluate the derivative of all input parameters, regardless of dimenison.\footnote{
@@
 
-% Emphasise the stability of gradients.
-% Not just a faster way of evaluating PhenomD.
-% Change plot to be q vs chieff.
-% Note that lalsuite cannot be made massively parallel on GPUs.
+Overall, we have demonstrated that the PhenomIMR waveform family is ideally suited for AD.
+Moreover, we have shown that our implementation of IMRPhenomD in \ripple is accurate and quite to evaluate.
+In the next section, we will discuss a variety of potential use cases of differentiable waveforms.
+
+% Emphasise the stability of gradients. $Done$
+% Not just a faster way of evaluating PhenomD. $Done$
+% Change plot to be q vs chieff. $Done$
+% Note that lalsuite cannot be made massively parallel on GPUs. $Done$
 % Don't need a full section on Fisher, but make sure to mention. People love Fisher.
@@
 
+Here, we will illustrate how three core tasks in GW science can be substantially improved through the use of differentiable waveforms.
+In the paper we will primarily look at toy examples, leaving a more careful analysis to future work. 
+The three tasks discussed here cover a wide range of GW science, starting with waveform development all the way to parameter estimation.
+
+\subsection{Fitting Waveform Coefficients}
+\label{subsec:hmc}
+
 \subsection{Effectualness Calculations}
@@
 \subsection{Derivative Based Samplers - Hamiltonian Monte Carlo}
-\label{subsec:hmc}
-
-\subsection{Fitting Waveform Coefficients}
 \label{subsec:hmc}

### 2022-10-15 src/tex/ms.tex
Adding a bit more text

@@
 \author{Maximiliano Isi} \CCA
+\author{Kelvin Lam} \CCA
 \author{Kaze W.~K.~Wong} \CCA
@@
 
-\subsection{Effectualness Calculations}
-\label{sec:effectualness}
+\te{Kelvin should add stuff here}
+
+\subsection{Fisher Forecasting}
+\label{subsec:hmc}
+
+Forecasting the sensitivity of future experiments is an essential task in GW science.
+Due to its theoretical simplicity and its evaluation speed, the Fisher matrix formalism is commonly deployed to estimate how well the a particular binary system's parameters can be measure.
+The Fisher matrix approach is built around the assumption of a Gaussian likelihood.
+Although in practice this assumption is often violated in for real GW datasets, the results obtained using a Fisher analysis can provide quick and useful diagnotistics in evaluating sensitivities for a variety of models.
+
+Computing the Fisher matrix requires one to evaluate derivatives of the likelihhood...
 

### 2022-11-10 src/tex/ms.tex
Wrote some test for the Fisher forecasting

--- 2022-10-13 src/tex/ms.tex
@@
 PhenomIMR waveforms are therefore ideally suited for AD, espeically using \jax. 
-In this paper we focus mainly on the aligned spin circular orbit model IMRPhenomD~\citep{Husa:2015iqa, 2015jqa}.
+In this paper we focus mainly on the aligned spin circular orbit model IMRPhenomD~\citep{Husa:2015iqa, Khan:2015jqa}.
 
@@
 Here, we will illustrate how three core tasks in GW science can be substantially improved through the use of differentiable waveforms.
-In the paper we will primarily look at toy examples, leaving a more careful analysis to future work. 
-The three tasks discussed here cover a wide range of GW science, starting with waveform development all the way to parameter estimation.
+In the paper we will primarily look at toy examples, leaving a more careful analyses to future work. 
+The three tasks discussed here cover a wide range of GW science, starting with waveform development all the way to forecasting and parameter estimation.
 
@@
 
-\subsection{Effectualness Calculations}
-\label{sec:effectualness}
+\te{Kelvin should add stuff here}
+
+\subsection{Fisher Forecasting}
+\label{subsec:hmc}
+
+Forecasting the sensitivity of future experiments is an essential task in GW science.
+Due to its theoretical simplicity and its evaluation speed, the Fisher matrix formalism is commonly deployed to estimate how well a binary system's parameters can be measured.
+The Fisher matrix approach is built around the assumption of a Gaussian likelihood.
+Although in practice this assumption is often violated in real GW datasets, the results obtained using a Fisher analysis can provide quick and useful diagnotistics in evaluating sensitivities for a variety of models and detector configurations.
+
+Computing the Fisher matrix requires one to evaluate derivatives of the likelihhood which in turn involves derivatives of the waveform model and detector projection functions.
+AD is therefore perfectly suited for computing Fisher matrices accurately and efficiently. 
+Forecasting with Fisher matrices for third generation detectors has already been extensively explored in \citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.
+Here we purely want to illustrate the simplicitly and speed of AD for forecasting rather than providing new physics insights.
+We therefore consider a simple, three detector setup corresponding to the two LIGO detectors in addition to Virgo.
+
+The Fisher information matrix for a single detector is typically given by 
+\begin{equation}
+    F^{k}_{ij} = (\partial_i h_0 | \partial_j h_0) \, ,
+\end{equation}
+where $k$ indicates the detector, $\partial_i = \partial/\partial \theta_i$, and $h_0$ is the strain measured by the detector which is given by,
+\begin{equation}
+    h_0(\theta) = F_+(\lambda) h_{+}(\phi) + F_\times(\lambda) h_{\times}(\phi) \, .
+\end{equation}
+Here we have separated out the intrinsic ($\phi$) and extrinsic ($\lambda$) variables.
+Since we are considering a three detector setup we simply add the Fisher matrices from the individual detectors to get the combined Fisher matrix:
+\begin{equation}
+    F_{ij} =  F^{\mathrm{Hanford}}_{ij} + F^{\mathrm{Livingston}}_{ij} + F^{\mathrm{Virgo}}_{ij}   \, .
+\end{equation}
+Finally, we invert the Fisher matrix to calculate the covariance matrix.
+
+To illustrate the computational speed of computing Fisher matrices with AD, we consider a population of binaries and compute the sky localtization error following Eq.~(28) in \citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.
+Since the Fisher matrix approach is known to have both theoretical issues as well as numerical instabilities for low signal-to-noise events, we restrict our population to only nearby systems.
+A full list of the distributions used to generate the various parameters in our population are given in Tab...
+
+The distribution of sky localization errors from a a population of $10^3$ binaries can be seen in Fig... 
+We have verified that our errors agree with a separate dedicated Fisher forecasting code~\citep{Borhanian:2020ypi} to within X percent.
+This demonstrates that AD can be used to accurately produce population level forecasts.
+
+Moreover, each error calculation (including computing the Fisher matrices for each detector and the inversion procress) is substantially faster.
+In particular, we find that after compilation each error calculation takes approximately one second on a single computing core.
+This can be compared to \texttt{GWbench}~\citep{Borhanian:2020ypi} where each Fisher calculation takes $\mathcal{O}$(minutes) for a similar detector setup and frequency grid.
+The factor of almost 60 speed up is despite the fact that a single core evaluation of the \ripple waveform is significantly slower that the \lalsuite call.
+As discussed above, performance can be further improved by utilising harware acceleration such as parallel GPU processing.
+AD therefore represents a fast and accurate way of performing population level analyses, and should be utilized for testing the capabilities of next generation detectors.
 
@@
 \label{subsec:hmc}
+
+\section{Discussion}
+\label{subsec:discussion}
 

### 2022-11-17 src/tex/ms.tex
Added plots in fitting waveform coefficients section

@@
 
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=\linewidth]{../static/amplitude_loss.pdf}
+    \caption{
+        Relative error of amplitudes of IMRPhenomD waveforms. Blue: Error between original Phenom waveform and NR waveform; Orange: Error between optimized waveform and NR waveform.
+    }
+\end{figure}
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=\linewidth]{../static/MM_logdiff_q148.pdf.pdf}
+    \caption{
+        Distribution of the log difference of mismatches between optimized and original model. The mean shows a 36\% decrease in mismatch. 
+    }
+\end{figure}
+
 \subsection{Fisher Forecasting}

### 2022-11-18 src/tex/ms.tex
Updates throughout

--- 2022-11-10 src/tex/ms.tex
@@
 \begin{equation}
-    F^{k}_{ij} = (\partial_i h_0 | \partial_j h_0) \, ,
+    \mathcal{I}^{k}_{ij} = (\partial_i h_0 | \partial_j h_0) \, ,
 \end{equation}
@@
 \end{equation}
-Here we have separated out the intrinsic ($\phi$) and extrinsic ($\lambda$) variables.
+Note that here we have separated out the intrinsic ($\phi$) and extrinsic ($\lambda$) variables as well as introducing the detector projection functions for the plus and cross polarizations as $F_+$ and $F_\times$ respectively.
 Since we are considering a three detector setup we simply add the Fisher matrices from the individual detectors to get the combined Fisher matrix:
 \begin{equation}
-    F_{ij} =  F^{\mathrm{Hanford}}_{ij} + F^{\mathrm{Livingston}}_{ij} + F^{\mathrm{Virgo}}_{ij}   \, .
+    \mathcal{I}_{ij} =  \mathcal{I}^{\mathrm{Hanford}}_{ij} + \mathcal{I}^{\mathrm{Livingston}}_{ij} + \mathcal{I}^{\mathrm{Virgo}}_{ij}   \, .
 \end{equation}
@@
 A full list of the distributions used to generate the various parameters in our population are given in Tab...
+The resulting population produces binaries with signal-to-noise ratios ranging from $\mathcal{O}(10-10^2)$.
 
-The distribution of sky localization errors from a a population of $10^3$ binaries can be seen in Fig... 
+\begin{figure}[ht!]
+    \script{sky_localization.py}
+    \begin{centering}
+        \includegraphics[width=\linewidth]{figures/sky_localization.pdf}
+        \caption{
+            This plot should show a sky localization error bar.
+        }
+        \label{fig:sky_localization}
+    \end{centering}
+\end{figure}
+
+The distribution of sky localization errors from a a population of $10^3$ binaries can be seen in Fig.~\ref{fig:sky_localization}.
 We have verified that our errors agree with a separate dedicated Fisher forecasting code~\citep{Borhanian:2020ypi} to within X percent.

### 2022-11-18 src/scripts/sky_localization.py

@@ -33,31 +33,31 @@ params = {
 
 mpl.rcParams.update(params)
 
-data = np.loadtxt(paths.data / "ripple_phenomD_matches.txt")
-exactmatch_indices = (1.0 - data[:, -1]) <= 0.0
-other_indices = (1.0 - data[:, -1]) > 0.0
+(
+    Mc,
+    eta,
+    chi1,
+    chi2,
+    dist,
+    tc,
+    phic,
+    inclination,
+    polarization_angle,
+    ra,
+    dec,
+    sky_error_list,
+    SNR_list,
+) = np.loadtxt(paths.data / "sky_localization.txt", unpack=True)
 
-q = data[:, 1] / data[:, 0]
-chieff = (data[:, 0] * data[:, 2] + data[:, 1] * data[:, 3]) / (data[:, 0] + data[:, 1])
-Mtot = data[:, 1] + data[:, 0]
-
-# Plot and save
 plt.figure(figsize=(7, 5))
-cm = plt.cm.get_cmap("inferno")
-sc = plt.scatter(
-    Mtot[other_indices],
-    chieff[other_indices],
-    c=np.log10(1.0 - data[other_indices, -1]),
-    cmap=cm,
-)
-plt.colorbar(sc, label=r"$\log_{10}(1-\mathrm{Match})$")
+bins = np.geomspace(0.1, np.max(sky_error_list), 50)
+# bins = np.linspace(0.0, np.max(sky_error_list), 80)
+plt.hist(sky_error_list, bins=bins, density=True, alpha=0.3, color="C0")
+plt.hist(sky_error_list, bins=bins, density=True, histtype="step", color="C0")
 
-plt.scatter(Mtot[exactmatch_indices], chieff[exactmatch_indices], color="C0")
-# plt.xlabel(r"Total Mass, $M$")
-# plt.ylabel(r"Mass Ratio, $q$")
-plt.xlabel(r"Total Mass, $M$")
-plt.ylabel(r"Effective Spin, $\chi_{\rm eff}$")
-# plt.xlim(1, 50)
-# plt.ylim(1, 50)
+plt.xscale("log")
+plt.xlim(0.1, 300)
+plt.xlabel("Sky Localization Error [deg$^2$]")
+plt.ylabel("Probability")
+plt.savefig(paths.figures / "sky_localization.pdf", bbox_inches="tight")
 
-plt.savefig(paths.figures / "random_matches.pdf", bbox_inches="tight")

### 2022-11-21 src/tex/ms.tex
Added table to section 3

@@
 % \usepackage{colortbl}
+\usepackage{multirow}
 \definecolor{pyBlue}{RGB}{31, 119, 180}
@@
 \subsection{Fitting Waveform Coefficients}
-\label{subsec:hmc}
+\label{subsec:coeffs}
 
@@
 \subsection{Fisher Forecasting}
-\label{subsec:hmc}
+\label{subsec:fisher}
 
@@
 Finally, we invert the Fisher matrix to calculate the covariance matrix.
+
+% \begin{table*}[ht!]
+%     \begin{center}
+%     \begin{tabular}{c|c|c|c|c|c|c|c|c}
+%     \hline\hline
+%     \hfil $\boldsymbol{m_1, m_2}$ &
+%     \hfil $\boldsymbol{\chi_1, \chi_2}$ &
+%     \hfil $\boldsymbol{D}$ &
+%     \hfil $\boldsymbol{t_c}$ &
+%     \hfil $\boldsymbol{\phi_c}$ &
+%     \hfil $\boldsymbol{\iota}$ &
+%     \hfil $\boldsymbol{\psi}$ &
+%     \hfil $\boldsymbol{RA}$ &
+%     \hfil $\boldsymbol{DEC}$
+%     \\
+%     \hline\hline
+%     \hfil [0.1, 3] &
+%     \hfil [0, 1] &
+%     \hfil [0, 1] &
+%     \hfil [0.7, 1.3] &
+%     \hfil [-3, 0] &
+%     \hfil [0, 3] &
+%     \hfil [-3, 3] &
+%     \hfil [-3, 3] &
+%     \hfil [-3, 3]
+%     \\
+%     \hline\hline
+%     \end{tabular}
+%     \end{center}
+%     \caption{caption here}
+%     \label{tab:priors}
+% \end{table*}
+
+\begin{table}[t!]
+    % \begin{center}
+    \centering
+    \begin{tabular}{c|c}
+    \hline\hline
+    $m_1, m_2$ & $\boldsymbol{U}[1.0, 50]\, \mathrm{M}_\odot$ \\ \hline
+    $\chi_1, \chi_2$ & [-0.99, 0.99] \\ \hline
+    $D$ & [500, 1000]\, Mpc \\ \hline
+    $t_c$ & 0.0 \\ \hline
+    $\phi_c$ & 0.0 \\ \hline
+    Inclincation Angle, $\iota$ & $\boldsymbol{U}[0, 1]$ \\ \hline
+    Polarization Angle, $\psi$ & $\boldsymbol{U}[0, 1]$ \\ \hline
+    Right Ascension, $\delta$ & $\boldsymbol{U}[0, 1]$ \\ \hline
+    Declination, $\alpha$ & $\boldsymbol{U}[0, 1]$ \\ 
+    \hline\hline
+    \end{tabular}
+    % \end{center}
+    \caption{Priors for the 11 dimensional parameter space used for the Fisher forecasting population analysis in Sec.~\ref{subsec:fisher}. 
+    $\boldsymbol{U}$ indicates a uniform distribution between the two variables in the brackets.}
+    \label{tab:priors}
+\end{table}
 

### 2022-11-21 src/scripts/random_matches.py
Trying to get online compilation to work

@@ -27,6 +27,7 @@ params = {
     "font.family": "serif",
     "axes.linewidth": 1.5,
     "mathtext.fontset": "dejavuserif",
+    "text.usetex": True,
 }
 
 # mpl.rcParams["text.latex.preamble"] = [r"\usepackage{amsmath}"]  # for \text command

### 2022-11-21 src/tex/ms.tex
Fixing typos in Fisher section

@@
 Forecasting the sensitivity of future experiments is an essential task in GW science.
-Due to its theoretical simplicity and its evaluation speed, the Fisher matrix formalism is commonly deployed to estimate how well a binary system's parameters can be measured.
+Due to its theoretical simplicity and evaluation speed, the Fisher matrix formalism is commonly deployed to estimate how well a binary system's parameters can be measured.
 The Fisher matrix approach is built around the assumption of a Gaussian likelihood.
@@
 Since the Fisher matrix approach is known to have both theoretical issues as well as numerical instabilities for low signal-to-noise events, we restrict our population to only nearby systems.
-A full list of the distributions used to generate the various parameters in our population are given in Tab...
+A full list of the distributions used to generate the various parameters in our population are given in Tab.~\ref{tab:priors}.
 The resulting population produces binaries with signal-to-noise ratios ranging from $\mathcal{O}(10-10^2)$.
@@
 
-The distribution of sky localization errors from a a population of $10^3$ binaries can be seen in Fig.~\ref{fig:sky_localization}.
+The distribution of sky localization errors from a population of $10^3$ binaries can be seen in Fig.~\ref{fig:sky_localization}.
 We have verified that our errors agree with a separate dedicated Fisher forecasting code~\citep{Borhanian:2020ypi} to within X percent.
@@
 Moreover, each error calculation (including computing the Fisher matrices for each detector and the inversion procress) is substantially faster.
-In particular, we find that after compilation each error calculation takes approximately one second on a single computing core.
-This can be compared to \texttt{GWbench}~\citep{Borhanian:2020ypi} where each Fisher calculation takes $\mathcal{O}$(minutes) for a similar detector setup and frequency grid.
-The factor of almost 60 speed up is despite the fact that a single core evaluation of the \ripple waveform is significantly slower that the \lalsuite call.
-As discussed above, performance can be further improved by utilising harware acceleration such as parallel GPU processing.
+In particular, we find that after compilation each error calculation takes approximately half a second on a single computing core.
+\texttt{GWbench}~\citep{Borhanian:2020ypi}, on the other hand, takes $\mathcal{O}$(minutes) for each Fisher calculation using the same detector setup and frequency grid.
+This factor of over 100 speed up is substantial considering the fact that a single core evaluation of the \ripple waveform is significantly slower that the \lalsuite call.
+As discussed above, performance can be further improved by utilising hardware acceleration such as parallel GPU processing.
 AD therefore represents a fast and accurate way of performing population level analyses, and should be utilized for testing the capabilities of next generation detectors.

### 2022-11-22 src/tex/ms.tex
add some words to fitting paragraph

@@
 \newcommand{\te}[1]{\textbf{\color{pyGreen}(TE: #1)}}
+\definecolor{rb4}{HTML}{27408B}
+\newcommand{\kw}[1]{{\color{rb4}[KW: #1 ]}}
 
@@
 
+Having an accurate waveform model is essential for many data analysis tasks in
+GW, such as searching for signals and estimating source parameters given data.
+While waveforms generated using numerical relativity simulations are commonly
+regarded as the most accurate, they are too computationally expensive to be used
+in any practical data analysis tasks. To create waveform models that can be used
+in data analysis, there are a number of groups creating "approximants" of the
+full numerical relativity waveforms, which are essentially simpler ansatzes that
+are not as accurate but can be evaluated much faster. There are a few families
+of approximants, such as the IMRPhenom family, SEOB family, and the NRSurrogate
+model. \kw{cite and add a couple more sentences to introduce the waveform
+families.}
+
+Every waveform approximants have some fitting parameters that are fixed during
+the process of calibrating the ansatz to NR waveforms. In the case of
+IMRPhenomD, there are $\sim 220$ \kw{Kelvin check number} fitting parameters
+used to capture the behavior of the amplitude and phase as a function of the
+mass ratio and spins.
+
 \begin{figure}[t]
@@
     \includegraphics[width=\linewidth]{../static/amplitude_loss.pdf}
-    \caption{
-        Relative error of amplitudes of IMRPhenomD waveforms. Blue: Error between original Phenom waveform and NR waveform; Orange: Error between optimized waveform and NR waveform.
-    }
+    \caption{ Relative error of amplitudes of IMRPhenomD waveforms. Blue: Error
+        between original Phenom waveform and NR waveform; Orange: Error between
+        optimized waveform and NR waveform. }
 \end{figure}
@@
     \centering
-    \includegraphics[width=\linewidth]{../static/MM_logdiff_q148.pdf.pdf}
-    \caption{
-        Distribution of the log difference of mismatches between optimized and original model. The mean shows a 36\% decrease in mismatch. 
-    }
+    \includegraphics[width=\linewidth]{../static/MM_logdiff_q148.pdf}
+    \caption{ Distribution of the log difference of mismatches between optimized
+        and original model. The mean shows a 36\% decrease in mismatch. }
 \end{figure}

### 2022-11-22 src/tex/ms.tex
Add some headers

@@
 used to capture the behavior of the amplitude and phase as a function of the
-mass ratio and spins.
+mass ratio and spins. \kw{Say something about how the fitting parameters affect the accuracy of the waveform model}
+
+% Motivation
+\kw{Motivate the need for improving the fitting process by joint fit.}
+
+% How the fitting process is done now.
+\kw{Mention how these parameters are often hand tuned.}
+
+Since we can differentiate through the entire waveform model against the fitting
+parameters, one can use gradient descent to find the optimal fitting parameters
+all at once instead of fitting them in subsets.
+
+\kw{Describe figure loss_compare}
 
@@
     optimized waveform and NR waveform. }
+\label{fig:loss_compare}
 \end{figure}
+
+\kw{Describe figure logdiff}
 
@@
     and original model. The mean shows a 36\% decrease in mismatch. }
+\label{fig:MM_logdiff}
 \end{figure}
+
+\kw{Describe implication}
+
+% Implication
 

### 2022-11-22 src/tex/ms.tex
added some words in fitting coefficients

@@
 
-\te{Kelvin should add stuff here}
+% \te{Kelvin should add stuff here}
 
@@
 
-Every waveform approximants have some fitting parameters that are fixed during
+Every waveform approximants have some fitting coefficients that are fixed during
 the process of calibrating the ansatz to NR waveforms. In the case of 
-IMRPhenomD, there are $\sim 220$ \kw{Kelvin check number} fitting parameters
-used to capture the behavior of the amplitude and phase as a function of the
-mass ratio and spins. \kw{Say something about how the fitting parameters affect the accuracy of the waveform model}
+IMRPhenomD, there are 209 fitting coefficients used to capture the separate behavior of 
+the amplitude and phase as a function of the mass ratio and spins. 
+\kw{Say something about how the fitting parameters affect the accuracy of the waveform model}
 
@@
 
+Previously in the construction of IMRPhenomD \citep{Khan:2015jqa}, waveform coefficients are fitted in segments with amplitude and phase separated. 
+Individual segments are then connected together such that they are differentiable. 
+Using this method, the generated waveforms are expected to be erroneous, as the connections would affect the originally fitted segments. 
+Hence, coefficients of the current IMRPhenomD implementation are often hand tuned in the final procedure. 
+The set of coefficients are both is expensive and time-costly to calculate, and not the best-fitted solution. 
+Instead, by joint fit, all coefficients are simultaneously fitted without having issues discussed above. 
+Therefore, it is essential to fit the waveform jointly.
+
 Since we can differentiate through the entire waveform model against the fitting
@@
 \kw{Describe figure loss_compare}
+Here, we define the mismatch as
+\begin{equation}
+	\mathcal{M}=1-\mathrm{m}(h_{\mathrm{IMR}}|h_{\mathrm{NR}}), 
+\end{equation} 
+In our optimization procedure, we took NR waveforms stated in \citep{Khan:2015jqa} as training waveforms to fit waveform coefficients. 
+By defining the mean of individual mismatches as the loss function for optimization, we can calibrate the set of coefficients using gradient descent.
 
@@
 \includegraphics[width=\linewidth]{../static/amplitude_loss.pdf}
-\caption{ Relative error of amplitudes of IMRPhenomD waveforms. Blue: Error
+\caption{Relative error of amplitudes of IMRPhenomD waveforms. Blue: Error
     between original Phenom waveform and NR waveform; Orange: Error between
@@
 \kw{Describe figure logdiff}
+As demonstration, Fig. \ref{fig:loss_compare} shows the relative error of original and optimized waveform at different frequencies. 
+In general, the error of optimized waveform is lower than that of the original waveform. 
+The error in the early inspiral region is substantially decreased by half and other regions also show good improvement in accuracy.
 
@@
 \centering
-\includegraphics[width=\linewidth]{../static/MM_logdiff_q148.pdf}
-\caption{ Distribution of the log difference of mismatches between optimized
+\includegraphics[width=\linewidth]{../static/MM_logdiff_q148_LIGO.pdf}
+\caption{Distribution of the log difference of mismatches between optimized
     and original model. The mean shows a 36\% decrease in mismatch. }
@@
 
+Instead of examining individual waveforms, we investigate the general behavior of mismatches for a set of testing waveforms. 
+In Fig. \ref{fig:logdiff_q148}, the distribution of log difference of mismatches are shown. 
+The distribution mean lies towards the negative side, which indicates a improvement in the model. 
+Despite there are testing waveforms that performed worse after optimization, it is verified that the ansatz of IMRPhenomD does not fit well with most of these waveforms, hence cannot be improved by optimization.
+Therefore, the new set of waveform coefficients can produce better waveforms for GW analyses.
+
 \kw{Describe implication}
+% Implication
 
-% Implication
+With further modifications to the optimization process, the set of waveform coefficients can be retuned to give almost an order-of-magnitude improvement. 
+Utilizing AD, differentiable waveform models can be further improved easily and quickly using similar computational schemes. 
+Moreover, in the construction of new waveform models, AD shows significant advantages during their calibrating process.
 

### 2022-11-26 src/tex/ms.tex
some update

@@
 
-\subsection{Fitting Waveform Coefficients}
+\subsection{Finetuning Waveform Coefficients}
 \label{subsec:coeffs}
@@
 the process of calibrating the ansatz to NR waveforms. In the case of
-IMRPhenomD, there are 209 fitting coefficients used to capture the separate behavior of 
-the amplitude and phase as a function of the mass ratio and spins. 
+IMRPhenomD, there are 209 fitting coefficients used to capture the separate
+behavior of the amplitude and phase as a function of the mass ratio and spins.
+The accuracy of fitting coefficients determines how well the approximant
+reproduces the NR waveform. The inaccuracy in obtaining the fitting coefficients
+means misrepresentation of the NR waveform, meaning the use of approximant will
+introduce some systematic error in downstream data analysis task. For example,
+this would mean the recovered source parameters will be biased in the case of PE.
 \kw{Say something about how the fitting parameters affect the accuracy of the waveform model}
@@
 
-% How the fitting process is done now.
-\kw{Mention how these parameters are often hand tuned.}
-
-Previously in the construction of IMRPhenomD \citep{Khan:2015jqa}, waveform coefficients are fitted in segments with amplitude and phase separated. 
-Individual segments are then connected together such that they are differentiable. 
-Using this method, the generated waveforms are expected to be erroneous, as the connections would affect the originally fitted segments. 
-Hence, coefficients of the current IMRPhenomD implementation are often hand tuned in the final procedure. 
-The set of coefficients are both is expensive and time-costly to calculate, and not the best-fitted solution. 
-Instead, by joint fit, all coefficients are simultaneously fitted without having issues discussed above. 
-Therefore, it is essential to fit the waveform jointly.
+Previously in the construction of IMRPhenomD \citep{Khan:2015jqa}, waveform
+coefficients are fitted in segments with amplitude and phase separated.
+Individual segments are then connected together such that they are
+differentiable. Using this method, the generated waveforms are expected to be
+erroneous, as the connections would affect the originally fitted segments. The
+coefficients of the current IMRPhenomD implementation are often hand tuned, and
+are tuned in sub-sets of parameters instead of all together. This means the
+tuning process ignore the correlation between different sub-sets of parameters,
+so the best-fitted solution obtained by fitting sub-sets of parameters may not
+be the global optimum compared to fitting all coefficients at once. Instead, by
+jointly fitting all coefficients, the inaccuracy in fitting individual and
+connecting different segments can be accounted during the optimization.
+Therefore, there is potential accuracy one can gain by fitting all coefficients
+at once.
 
@@
 
-\kw{Describe figure loss_compare}
 Here, we define the mismatch as
@@
 By defining the mean of individual mismatches as the loss function for optimization, we can calibrate the set of coefficients using gradient descent.
+
+\kw{Describe figure losscompare}
 

### 2022-11-26 src/tex/ms.tex
Finishing Kaze's edit

@@
 this would mean the recovered source parameters will be biased in the case of PE.
-\kw{Say something about how the fitting parameters affect the accuracy of the waveform model}
 
 % Motivation
-\kw{Motivate the need for improving the fitting process by joint fit.}
-
 Previously in the construction of IMRPhenomD \citep{Khan:2015jqa}, waveform
@@
 
+\kw{Kelvin Fill}
 Since we can differentiate through the entire waveform model against the fitting
@@
 
-With further modifications to the optimization process, the set of waveform coefficients can be retuned to give almost an order-of-magnitude improvement. 
-Utilizing AD, differentiable waveform models can be further improved easily and quickly using similar computational schemes. 
-Moreover, in the construction of new waveform models, AD shows significant advantages during their calibrating process.
-
+This fine-tuning step is rather simple and general. Given a different
+differentiable waveform model, one can apply the same procedure to optimize the
+coefficients in the new model, i.e. applying gradient descent to optimize all
+the fitting coefficients all together. This provides a more flexible control
+during the calibration process, and potentially better final result.
+With further modifications to the optimization process, the set of waveform
+coefficients can be retuned to give almost an order-of-magnitude improvement.
+\kw{Where is the proof for this?} 
 \subsection{Fisher Forecasting}

### 2022-12-01 src/tex/ms.tex
Correcting github compilation

@@
 \newcommand{\lalsuite}{\texttt{lalsuite}\xspace}
+\newcommand{\CC}{C\nolinebreak\hspace{-.05em}\raisebox{.4ex}{\tiny\bf +}\nolinebreak\hspace{-.10em}\raisebox{.4ex}{\tiny\bf +}}
 
@@
 
+\newcommand{\JHU}{\affiliation{William H. Miller III Department of Physics and Astronomy, Johns Hopkins University, Baltimore, Maryland 21218, USA}} 
 \newcommand{\OKC}{\affiliation{The Oskar Klein Centre, Department of Physics, Stockholm University, AlbaNova, SE-106 91 Stockholm, Sweden}} 
@@
 \author{Adam Coogan} \UdeM \Mila
-\author{Thomas D.~P.~Edwards} \OKC \NORDITA
+\author{Thomas D.~P.~Edwards} \JHU \OKC \NORDITA
 \author{Daniel Foreman-Mackey} \CCA
@@
 \begin{abstract}
-    Here we will discuss our implementation of differentiable waveforms and demonstrate their benefits for a variety of data analysis tasks.
+    Since their discovery in 2015, gravitaitonal waves (GWs) have developed into a unique observational probe used throughout fundamental and astro-phyics.
+    As more detections continue, data analysis taks are becoming increasingly computationally demanding.
+    Here we propose the use of automatic differentiation through the programming framework \jax for accelerating a variety of these analsyis tasks.
+    Firstly, we demonstrate that highly accurate waveforms (IMRPhenomD) can be written in \jax and demonstrate that the evaluation speed of the waveform (and its derivative) is not substantially slower than the \lalsuite implenetation in \CC.
+    We then focus on three applications where speed and differentiable waveforms are essential.
+    Firstly, we demonstrate that stochastic gradient descent can be used to optimize the $\sim 200$ coefficients that make up the waveform. 
+    In particular, we demonstrate that the typical \textit{match} with numerical relativiely waveforms can be improved by more than 30\% without any additional overhead.
+    Secondly, we show that Fisher forecasting calculations can be sped up by $\sim 100\times$ with no loss in accuracy.
+    This increased speed makes population forecasting substantially simpler, as discussed extensively in \citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.
+    Finally, we show that gradient based samplers like Hamiltonian Monte Carlo can improve acceptance rates of XXX. 
+    Since differentiable waveforms appear to have substatial advantages for a variety of tasks throughtout GW science (with little downside), we propose that waveform developers use \jax to build new waveforms moving forward.
+    Our waveform code, called \ripple, can be found at XXX, and will continue to be updated with new waveforms as they are formulated.
+
 \end{abstract}
@@
 Once potential candidates are found, parameter estimation (PE) is performed to extract the detailed source properties of each event.
-In the most minimial example, this requires one to perform MCMC over a 15 dimensional parameter space for each event.
+In the most minimial example, this requires one to perform an MCMC on an 11 dimensional parameter space for each event.
 Beyond this simple scenario, more complex waveform models with many additional parameters may be used to test for deviations from General Relativity.
@@
 in Hamiltonian Monte Carlo~\citep{2017arXiv170102434B} or Gradient Decsent).
-This use of gradients is particularly useful for high dimensional spaces.
+Gradients are particularly valuable for high dimensional spaces.
 Unfortunately, in the field of GW data analysis, analytic derivatives of the necessary quantities (such as the likelihood) have historically been difficult to obtain.
@@
 There are now a variety of AD implementations, most notably in deep learning frameworks such as \texttt{pytorch}~\citep{pytorch} and \texttt{tensorflow}~\citep{tensorflow2015-whitepaper}.
-More general frameworks exist in \texttt{julia}~\citep{zygote, forwarddiff}, although \texttt{julia}'s limited use in GW physics precludes its general use.
+More general frameworks exist in \texttt{julia}~\citep{zygote, forwarddiff}, although \texttt{julia}'s limited support in GW physics precludes its general use.
 Here we make use of \jax~\citep{jax2018github} due to its easy integration with \texttt{python} libraries, ability to just-in-time compile, and ability to utilize a variety of computational resources (i.e.
@@
 In this paper we argue that differentiable waveforms will be a vital component for the future of GW data analysis.
-In addition, we present \ripple, a small GW \texttt{python} package containing differentiable implementations of some of the main waveforms currently used in LIGO and Virgo analyses. 
+In addition, we present \ripple, a small GW \texttt{python} package which can serve as a respository for differentiable implementations of some of the main waveforms currently used in LIGO and Virgo analyses.
 The remainder of this paper is structured as follows. 
 In Sec.~\ref{sec:waveforms} we discuss the differentiable waveforms implemented in \ripple and perform some benchmarks to demonstrate the speed and accruacy of the derivative calculations. 
-In Sec.~\ref{sec:applications} we discuss four distinct applications using differentiable waveforms. 
-First, a simple demonstration of accelerated effectualness calculations using gradient descent.
-Second, we show that Fisher analyses are quickly enabled using AD.
-Third, we implement differnetiable detector response functions and show an illutrative example using Hamiltonian Monte Carlo.
-Finally, we illustrate how the fit coefficients that form part of the IMRPhenom waveform models could be improved by high dimensional fitting, enabled by a differentiable waveform. 
+In Sec.~\ref{sec:applications} we discuss three distinct applications using differentiable waveforms. 
+Firstly, we illustrate how the fit coefficients that form part of the IMRPhenom waveform models could be improved by high dimensional fitting, enabled by a differentiable waveform. 
+Secondly, we implement differnetiable detector response functions and show that the speed of Fisher matrix calculations can be substantially improved using AD.
+Finally, we run an illustrative injection example using Hamiltonian Monte Carlo to demonstrate that the acceptance rate of derivative samplers is substantially larger than traditional MCMCs.
 The associated code can be found at \ripple~\citep{ripple}.
@@
 In particular, the blue points indicate parameter points where the $1-\mathrm{m}(h_{\theta_1}|h^{*}_{\theta_1})=0$ and would therefore not appear in the log scale colorbar.
-\begin{figure}[ht!]
+\begin{figure}[t]
     \script{random_matches.py}
@@
 Since the \lalsuite implementation of IMRPhenomD is written in \texttt{C}, the evaluation of the waveform itself is faster than \ripple.
-Benchmarking on a MacBook Pro with an M1 Max Apple Silicon processor, we find that a single waveform evaluation takes $\sim 4\,\mathrm{ms}$ for \lalsuite (interfaced with python) and $\sim 10\,\mathrm{s}$ for \ripple.\footnote{
+Benchmarking on a MacBook Pro with an M1 Max Apple Silicon processor, we find that a single waveform evaluation takes $\sim 2-5\,\mathrm{ms}$ for \lalsuite (interfaced with python) and $\sim 5-8\,\mathrm{ms}$ for \ripple.\footnote{
     For this benchmark we used $16\,\mathrm{Hz}$ and $1024\,\mathrm{Hz}$ for the lowest and highest frequencies respectively. 
@@
 Although slower on a CPU benchmark, a key advantage of using \jax is its native ability to run on a GPU.
-Waveform calls can therefore be evaluated in parallel, massively improving performance. \te{I couldn't get these to be similar on google colab. @Kaze any ideas how to make this fast?}
+Waveform calls can therefore be evaluated in parallel, massively improving performance.
+Performing the same benchmark as above on a google colab notebook with the freely provided GPU, we find that on average waveform evaluations take $\sim 0.00005\mathrm{ms}$, nearly five orders of magnitude faster.
 Generalizing \lalsuite waveforms to run on a GPU would be a significant undertaking.
@@
 model. \kw{cite and add a couple more sentences to introduce the waveform
-families.}
+families.} \te{I think a full dicussion of the various waveform families just belongs in section 2, so don't need too much more here I'd say.}
 
-Every waveform approximants have some fitting coefficients that are fixed during
+Every waveform approximant has some fitting coefficients that are fixed during
 the process of calibrating the ansatz to NR waveforms. In the case of
@@
 means misrepresentation of the NR waveform, meaning the use of approximant will
-introduce some systematic error in downstream data analysis task. For example,
-this would mean the recovered source parameters will be biased in the case of PE.
+introduce some systematic error in downstream data analysis task. For example, sufficiently large systematic errors 
+would mean cause the recovered source parameters to be biased in the case of PE.
 
@@
     between original Phenom waveform and NR waveform; Orange: Error between
-    optimized waveform and NR waveform. }
+    optimized waveform and NR waveform. \te{Can you add the data and the plotting script to the repo so that we can generate the plots using the same matplotlib settings?}}
 \label{fig:loss_compare}
@@
 \kw{Where is the proof for this?}
+
 \subsection{Fisher Forecasting}
 \label{subsec:fisher}
+
+\begin{figure}[t!]
+    \script{sky_localization.py}
+    \centering
+    \includegraphics[width=\linewidth]{figures/sky_localization.pdf}
+    \caption{
+        This plot should show a sky localization error bar.
+    }
+    \label{fig:sky_localization}
+\end{figure}
 
@@
 
-\begin{table}[t!]
+\begin{table}[t]
     % \begin{center}
@@
 
-\begin{figure}[ht!]
-    \script{sky_localization.py}
-    \begin{centering}
-        \includegraphics[width=\linewidth]{figures/sky_localization.pdf}
-        \caption{
-            This plot should show a sky localization error bar.
-        }
-        \label{fig:sky_localization}
-    \end{centering}
-\end{figure}
-
 The distribution of sky localization errors from a population of $10^3$ binaries can be seen in Fig.~\ref{fig:sky_localization}.
@@
 This work was supported by collaborative visits funded by the Cosmology and Astroparticle Student and Postdoc Exchange Network (CASPEN). 
+T.E. is supported by the Horizon Postdoctoral Fellowship.
 
@@
 
-\section{Appendix}
-\label{sec:appendix}
-
 \end{document}

### 2022-12-01 src/scripts/effectualness_demonstrator.py

@@ -1,121 +1,0 @@
-import jax.numpy as jnp
-import jax.scipy.optimize as jopt
-import scipy.optimize as opt
-import jax
-from jax.config import config
-config.update("jax_enable_x64", True)
-
-# from tqdm.auto import tqdm, trange
-from typing import Tuple
-from ripple.waveforms import IMRPhenomD
-from ripple import ms_to_Mc_eta
-import timeit
-
-
-Array = jnp.ndarray
-
-def get_eff_pads(fs: Array) -> Tuple[Array, Array]:
-    r"""
-    Gets arrays of zeros to pad a function evaluated on a frequency grid so the
-    function values can be passed to ``jax.numpy.fft.ifft``.
-    Args:
-        fs: uniformly-spaced grid of frequencies. It is assumed that the first
-            element in the grid must be an integer multiple of the grid spacing
-            (i.e., ``fs[0] % df == 0``, where ``df`` is the grid spacing).
-    Returns:
-        The padding arrays of zeros. The first is of length ``fs[0] / df`` and
-        the second is of length ``fs[-1] / df - 2``.
-    """
-    df = fs[1] - fs[0]
-    N = 2 * jnp.array(fs[-1] / df - 1).astype(int)
-    pad_low = jnp.zeros(jnp.array(fs[0] / df).astype(int))
-    pad_high = jnp.zeros(N - jnp.array(fs[-1] / df).astype(int))
-    return pad_low, pad_high
-
-@jax.jit
-def get_match_arr(
-    h1: Array, h2: Array, Sns: Array, pad_low: Array, pad_high: Array
-) -> Array:
-    """
-    Calculates the match between two frequency-domain complex strains. The maximizations
-    over the difference in time and phase at coalescence are performed by taking
-    the absolute value of the inverse Fourier transform.
-    Args:
-        h1: the first set of strains
-        h2: the second set of strains
-        Sns: the noise power spectral densities
-        fs: frequencies at which the strains and noise PSDs were evaluated
-        pad_low: array of zeros to pad the left side of the integrand before it
-            is passed to ``jax.numpy.fft.ifft``
-        pad_right: array of zeros to pad the right side of the integrand before
[…]

### 2022-12-01 src/scripts/random_matches.py

@@ -23,12 +23,39 @@ params = {
     "ytick.minor.size": 4,
     "xtick.direction": "in",
     "ytick.direction": "in",
-    "text.usetex": True,
-    "font.family": "serif",
     "axes.linewidth": 1.5,
+    "text.usetex": False,
     "mathtext.fontset": "dejavuserif",
-    "text.usetex": True,
 }
+
+# mpl.rcParams["text.latex.preamble"] = [r"\usepackage{amsmath}"]  # for \text command
+
+# params = {
+#     "font.size": 18,
+#     "legend.fontsize": 18,
+#     "legend.frameon": False,
+#     "axes.labelsize": 18,
+#     "axes.titlesize": 18,
+#     "xtick.labelsize": 18,
+#     "ytick.labelsize": 18,
+#     "figure.figsize": (7, 5),
+#     "xtick.top": True,
+#     "ytick.right": True,
+#     "xtick.bottom": True,
+#     "ytick.left": True,
+#     "xtick.major.pad": 8,
+#     "xtick.major.size": 8,
+#     "xtick.minor.size": 4,
+#     "ytick.major.size": 8,
+#     "ytick.minor.size": 4,
+#     "xtick.direction": "in",
+#     "ytick.direction": "in",
+#     "text.usetex": True,
+#     "font.family": "serif",
+#     "axes.linewidth": 1.5,
+#     "mathtext.fontset": "dejavuserif",
+#     "text.usetex": True,
+# }
 
 # mpl.rcParams["text.latex.preamble"] = [r"\usepackage{amsmath}"]  # for \text command
 
@@ -58,7 +85,8 @@ plt.scatter(Mtot[exactmatch_indices], chieff[exactmatch_indices], color="C0")
 # plt.ylabel(r"Mass Ratio, $q$")
 plt.xlabel(r"Total Mass, $M$")
 plt.ylabel(r"Effective Spin, $\chi_{\rm eff}$")
-# plt.xlim(1, 50)
+plt.xlim(0, 200)
+plt.ylim(-1, 1.0)
 # plt.ylim(1, 50)
 
 plt.savefig(paths.figures / "random_matches.pdf", bbox_inches="tight")

### 2022-12-01 src/scripts/sky_localization.py

@@ -23,12 +23,40 @@ params = {
     "ytick.minor.size": 4,
     "xtick.direction": "in",
     "ytick.direction": "in",
-    "text.usetex": True,
-    "font.family": "serif",
     "axes.linewidth": 1.5,
+    "text.usetex": False,
     "mathtext.fontset": "dejavuserif",
-    "text.usetex": True,
 }
+
+
+# mpl.rcParams["text.latex.preamble"] = [r"\usepackage{amsmath}"]  # for \text command
+
+# params = {
+#     "font.size": 18,
+#     "legend.fontsize": 18,
+#     "legend.frameon": False,
+#     "axes.labelsize": 18,
+#     "axes.titlesize": 18,
+#     "xtick.labelsize": 18,
+#     "ytick.labelsize": 18,
+#     "figure.figsize": (7, 5),
+#     "xtick.top": True,
+#     "ytick.right": True,
+#     "xtick.bottom": True,
+#     "ytick.left": True,
+#     "xtick.major.pad": 8,
+#     "xtick.major.size": 8,
+#     "xtick.minor.size": 4,
+#     "ytick.major.size": 8,
+#     "ytick.minor.size": 4,
+#     "xtick.direction": "in",
+#     "ytick.direction": "in",
+#     "text.usetex": True,
+#     "font.family": "serif",
+#     "axes.linewidth": 1.5,
+#     "mathtext.fontset": "dejavuserif",
+#     "text.usetex": True,
+# }
 
 # mpl.rcParams["text.latex.preamble"] = [r"\usepackage{amsmath}"]  # for \text command
 
@@ -52,7 +80,6 @@ mpl.rcParams.update(params)
 
 plt.figure(figsize=(7, 5))
 bins = np.geomspace(0.1, np.max(sky_error_list), 50)
-# bins = np.linspace(0.0, np.max(sky_error_list), 80)
 plt.hist(sky_error_list, bins=bins, density=True, alpha=0.3, color="C0")
 plt.hist(sky_error_list, bins=bins, density=True, histtype="step", color="C0")
 

### 2022-12-15 src/tex/ms.tex
Started adding text to HMC and first part of discussion

@@
 
-\section{Discussion}
+After the search algorithms have constructed a list of confidently detected binaries, the next step is to sample from the posterior of each sources parameters - so called parameter estimation (PE).
+To do this, one typically uses a Markov Chain Monte Carlo (MCMC) sampler or nested sampler~\citep{multinest, dynesty}.
+Although robust, both MCMC and nested sampling are slow to converge and are known to perform poorly in high dimensional parameter spaces.
+For example, sampling the 15 dimensional parameter space for a BBH system can take $\mathcal{O}(10)$hours, whilst BNS systems can take up to weeks.
+Dedicated fast samplers have been designed to get approximate posteriors on the sky localization to faciliate follow-up electromagnetic observations.
+Nevertheless, these do not present the whole picture; fast, general parameter estimation therefore remains a primary goal of GW data analysis.
+
+A primary issue with both MCMC and nested sampling is that neither utlizes information about the likelihood's derivative and must therefore randomly walk towards areas of highest likelihood.
+Derivative based samplers, on the other hand, have been shown to extrapolate well to higher dimensions although they sometimes come with their own drawbacks.
+Here we simply aim to demonstrate the utility of a derivative based sampler and their efficiency on a small test problem.
+In particular, we will show that proposals made using an Hamiltonian Monte Carlo (MC) sampler are accepted at a much higher rate than a traditional MCMC algorithm.
+In a follow up paper we will demonstrate that minute scale PE can be achieved by exploiting normalizing flows~\citep{2022arXiv221106397W, Gabrie:2021tlu} together with a derivative based sampler~\citep{PEpaper}.
+
+\section{Discussion and Conclusion}
 \label{subsec:discussion}
+
+In this paper we introduced and discussed the various benefits of differentiable waveforms in \jax for GW data analysis.
+First, we demonstrated the speed and accuracy of our implementation of the aligned spin IMRPhenomD waveform.
+In particular, we showed that it matches the \lalsuite implementation to near machine precision and can be easily parallelized on a GPU, making waveform calls up to five orders of magnitude faster.
+Second, we discussed three data analysis tasks which can all be substantially improved by utlizing derivative information of the waveform.
+Although we primarily discuss toy examples in this paper, each can be extended to the full data analysis task. 
+Differentiable waveforms therefore represent a crucial tool to perform efficient GW science.
+
+Currently the biggest constraint to adopting differentiable waveforms is the need to re-write the most commonly used waveforms into \jax.
+At the time of writing...
 

### 2022-12-16 src/tex/ms.tex
Started editing and added to discussion

@@
 
-The discovery of gravitational waves (GWs) from inspiraling and mergering compact objects (COs) has revolutionized our understanding of both fundamental physics and astronomy.
+The discovery of gravitational waves (GWs) from inspiraling and mergering compact objects (COs) has revolutionized our understanding of both fundamental physics and astronomy~\citep{LIGOScientific:2021djp}.
 Although the data volumes from GW detectors are relatively small, analyzing the data is a computationally demanding task.
@@
 Once potential candidates are found, parameter estimation (PE) is performed to extract the detailed source properties of each event.
-In the most minimial example, this requires one to perform an MCMC on an 11 dimensional parameter space for each event.
+In the most minimial example, this requires a Markov Chain Monte Carlo (MCMC) on an 11 dimensional parameter space for each event.
 Beyond this simple scenario, more complex waveform models with many additional parameters may be used to test for deviations from General Relativity.
-Finally, using the results of the PE, population synthesis models are used to constrain the progenitors systems from which the black holes we see merging today began their journey~\citep{Wong:2022flg}.
+Finally, using the results of the PE, population synthesis models constrain the progenitors systems from which the black holes we see merging today began their journey~\citep{Wong:2022flg}.
 All of these tasks require significant compute.
@@
 AD's recent ascendance is primarily driven by its use in machine learning, particularly for derivative computations of neural networks which use gradient descent during training.
-The core idea of AD is that each any mathematical function can be broken down into a small set of basic operations, each with a known differentiation rule.\footnote{
+The core idea of AD is that any mathematical function can be broken down into a small set of basic operations, each with a known differentiation rule.\footnote{
     Of course, non-differentiable functions exist and care must be taken when treating these special cases.
@@
 More general frameworks exist in \texttt{julia}~\citep{zygote, forwarddiff}, although \texttt{julia}'s limited support in GW physics precludes its general use.
-Here we make use of \jax~\citep{jax2018github} due to its easy integration with \texttt{python} libraries, ability to just-in-time compile, and ability to utilize a variety of computational resources (i.e.
-CPUs and GPUs) without changes to the code.
-\te{Do we need a longer intro to AD here?}
+Here we make use of \jax~\citep{jax2018github} due to its easy integration with \texttt{python} libraries and its just-in-time (JIT) compiler which allows the same code to run on a variety of hardware accelerators (i.e.
+GPUs).
 
 There are a variety of gravitational waveforms currently used in analysis pipelines.
-They are generally structured into different families, the most common of which are: the effective one body, the phenomenological inspiral-merger-ringerdown (IMR), and numerical relativity surrogate.
-Of these, the phenomenological IMR family is most well suited for an AD implementation in \jax.
-In particular, its closed form expression, ... \TE{Why else?}
-\te{Do we need more of an intro for why IMRPhenom?}
+They are generally structured into different families, the most common of which are: the effective-one-body (EOB), the phenomenological inspiral-merger-ringerdown (IMRPhenom), and numerical relativity surrogate (NRsurrogate).
+Of these, the IMRPhenom family is most well suited for an AD implementation in \jax.
+In particular, its closed form expression and fixed size frequency grid make a \jax implementation that complies with the constraints of JIT compilation simple.
+EOB waveforms on the other hand require... \te{Max or Kaze, could one of you write something about why its hard? I don't know enough about EOB.}
 
@@
 A variety of waveform families have been developed to accurately model the GW emission from COs. 
-When the COs are relatively well separated, the dynamics of the system can be well approximated with post-Newtonian corrections to the... .
-However, close to merger, numerical relatively simulations are required to accurately model the GW emission.
+When the COs are relatively well separated, the dynamics of the system can be well approximated with post-Newtonian expansion.
+However, close to merger, numerical relatively simulations are required to accurately model the binary.
 Unfortunately, these numerical simulations are computationally expensive and cannot be run in conjunction with data analysis.
-Approximate phenomenological waveforms have therefore been constructued to enable relatively fast waveform generation at sufficient accuracy.
+Approximate, phenomenological waveforms have therefore been constructued to enable relatively fast waveform generation at sufficient accuracy.
 
-The three major waveform families that have been developed to date are: the Effective-one-body (EOB), numerical relativity (NR) surrogate, and phenomenological inspiral-merger-ringdown (PhenomIMR).
-Note that unlike NR surrogate, both the EOB and PhenomIMR waveforms are calibrated to NR waveforms.
-EOB waveforms require one to model the binary system using a Hamiltonian and are typically slow to evaluate whereas PhenomIMR waveforms are constructed with simple closed-form expressions.
-PhenomIMR waveforms are therefore ideally suited for AD, espeically using \jax. 
+As mentioned in the previous section, the three major waveform families that have been developed to date are:  EOB,  NRsurrogate, and IMRPhenom.
+Note that unlike NRsurrogate, both the EOB and PhenomIMR waveforms are calibrated to numerical relativity waveforms.
+EOB waveforms require one to model the binary system using a Hamiltonian and are typically slow to evaluate whereas IMRPhenom waveforms are constructed with simple closed-form expressions.
+IMRPhenom waveforms are therefore ideally suited for AD, espeically using \jax. 
 In this paper we focus mainly on the aligned spin circular orbit model IMRPhenomD~\citep{Husa:2015iqa, Khan:2015jqa}.
 
-The implementation of IMRPhenomD in \lalsuite is mainly in C, and therefore needs to be rewritten natively into \texttt{pyhton} to be compatible with \jax.
+The implementation of IMRPhenomD in \lalsuite is in C, and therefore needs to be rewritten natively into \texttt{pyhton} to be compatible with \jax.
 We have re-written IMRPhenomD from scratch using a combination of pure \texttt{pyhton} and \jax derivatives.
@@
 
-To ensure our implempation of IMRPhenomD is faithful to the \lalsuite implementation, we start by defining the noise weighted inner product:
+To demonstrate our implempation of IMRPhenomD is faithful to the \lalsuite implementation, we start by defining the noise weighted inner product:
 \begin{equation}
@@
 Since the match is a measure of the difference between two waveforms, we can use it to demonstrate that the implementation of IMRPhenomD in \ripple accurately matches the \lalsuite implementation.
-This is shown in Fig.~\ref{fig:match}, where we have calculated the $\mathrm{m}(h_{\theta_1}|h^{*}_{\theta_1})$ across the entire parameter space.\footnote{
+This is shown in Fig.~\ref{fig:match}, where we have calculated the $\mathrm{m}(h_{\theta_1}|\bar{h}_{\theta_1})$ across the entire parameter space.\footnote{
     Specifically, we use 2000 points in the ranges $m_{1,2} = (1,50)\,M_{\odot}$ and $\chi_{1,2} = (-1,1)$.
@@
 }
-Here, $h_{\theta_1}$ corresponds to the \ripple waveform implementation and $h^{*}_{\theta_1}$ is the \lalsuite implementation. 
+Here, $h_{\theta_1}$ corresponds to the \ripple waveform implementation and $\bar{h}_{\theta_1}$ is the \lalsuite implementation. 
 From Fig.~\ref{fig:match}, it is clear that the \ripple waveform matches with the \lalsuite waveform to numerical precision across the majority of the parameter space.
@@
         \caption{
-            This plot should show a match colour bar.
+            Match between the \ripple and \lalsuite implementations of the IMRPhenomD waveform as a function of total mass and effective spin.
+            \te{Need to investigate more?}
         }
@@
 
-There remains some slight deviation between the two waveform implementations at higher masses. 
+There remains some slight deviation between the two waveform implementations at low total mass. \te{Do I need to look more into this...? The discussion below looks odd with this plot}
 This is likely due to the fact that cubic interpolators, which are used within IMRPhenomD to calculate the ringdown and damping frequencies, are not currently supported in \jax.
@@
 For maximum utility, a waveform needs to be fast to evaluate.
-Fortunately, the Phenom IMR waveforms are constructed from simple closed form expressions which are therefore computationally efficient.
+Fortunately, the IMRPhenom waveforms are constructed from simple closed-form expressions which are computationally efficient.
 Since the \lalsuite implementation of IMRPhenomD is written in \texttt{C}, the evaluation of the waveform itself is faster than \ripple.
@@
 
-The primary arguement of this paper is that waveform derivatives will also be invauable to data analysis tasks. 
+The primary arguement of this paper is that waveform derivatives will also be highly vauable to data analysis tasks. 
 AD provides two big advantages when it comes to evaluating derivatives compared to numerical differentiation.
-Firstly, the accuracy of derivatives from AD are signficantly more stable than finite difference methods.
-In particular, finite differences suffer from both rounding and truncation errors, meaning that the user is required to tune the width over which the difference is taken.
+First, the accuracy of derivatives from AD are signficantly more stable than finite difference methods.
+In particular, finite differences suffer from both rounding and truncation errors, meaning that the user is required to \textit{tune} the width over which the difference is taken.
 On the other hand, AD produces machine precision derivatives with no tuninig.
-Secondly, AD scales favourably with the dimensionality of the function.
+Second, AD scales favourably with the dimensionality of the function.
 In particular, for every input dimension, $D$, added one would need to evaluate the function at least $2D$ times to calcuate finite difference derivatives for all input parameters.
@@
 
-Overall, we have demonstrated that the PhenomIMR waveform family is ideally suited for AD.
-Moreover, we have shown that our implementation of IMRPhenomD in \ripple is accurate and quite to evaluate.
+Overall, we have demonstrated that the IMRPhenom waveform family is ideally suited for AD.
+Moreover, we have shown that our implementation of IMRPhenomD in \ripple is accurate and quick to evaluate, especially when hardware acceleration is available.
 In the next section, we will discuss a variety of potential use cases of differentiable waveforms.
@@
 Second, we discussed three data analysis tasks which can all be substantially improved by utlizing derivative information of the waveform.
-Although we primarily discuss toy examples in this paper, each can be extended to the full data analysis task. 
+Although we primarily discuss toy examples in this paper, each can be extended to the full data analysis task, some of which will be shown in upcoming papers~\citep{PEpaper}.
 Differentiable waveforms therefore represent a crucial tool to perform efficient GW science.
 
-Currently the biggest constraint to adopting differentiable waveforms is the need to re-write the most commonly used waveforms into \jax.
-At the time of writing...
+Currently the biggest constraint to adopting differentiable waveforms is the need to re-write the most commonly used waveforms into \jax (or pure python).
+In order to showcase the benefits of differentiable waveforms as quickly as possible, at the time of writing, we have only implemented an aligned spin GW model (IMRPhenomD).
+We plan on adding a variety of different waveforms to \ripple in the near future with the primary goal of reaching a \jax version of a fully precessing, higher order mode waveform such as IMRPhenomXPHM.
+Ideally, future waveforms should be implemented either in pure python or using \jax directly.
+This would ensure that the community can easily utilize differentiability and hardware acceleration in the future.
+
+In this paper, we have primarily focussed on the IMRPhenom family of waveforms as their closed form expression is perfectly suited for implementation in \jax.
+Two other waveform families are commonly used in GW data analysis: the effective-one-body (EOB) and numerical relativity surrogate (NRSurrogate). 
+Some progress has been made towards the construction of a differentiable NRsurrogate, currently it seems difficult to implement EOB waveforms in \jax.
+In particular, the evolution of the Hamiltonian required to evaluate an EOB waveform is both inherently slow to differentiate and difficult to implement in \jax which requires fixed length arrays for its JIT compilation.
+Since EOB waveforms are often regarded as the state-of-the-art, more work is required to see if a fast, differentiable implementation is possible.
+
+\te{End on a final good thing?}
 

### 2022-12-17 src/tex/ms.tex
added scripts for figures

--- 2022-12-01 src/tex/ms.tex
@@
 \newcommand{\lalsuite}{\texttt{lalsuite}\xspace}
-\newcommand{\CC}{C\nolinebreak\hspace{-.05em}\raisebox{.4ex}{\tiny\bf +}\nolinebreak\hspace{-.10em}\raisebox{.4ex}{\tiny\bf +}}
 
@@
 
-\newcommand{\JHU}{\affiliation{William H. Miller III Department of Physics and Astronomy, Johns Hopkins University, Baltimore, Maryland 21218, USA}} 
 \newcommand{\OKC}{\affiliation{The Oskar Klein Centre, Department of Physics, Stockholm University, AlbaNova, SE-106 91 Stockholm, Sweden}} 
@@
 \newcommand{\CCA}{\affiliation{Center for Computational Astrophysics, Flatiron Institute, New York, NY 10010, USA}}
+\newcommand{\CUHK}{\affiliation{Department of Physics, The Chinese University of Hong Kong, Shatin, N.T., Hong Kong}}
 
@@
 \author{Adam Coogan} \UdeM \Mila
-\author{Thomas D.~P.~Edwards} \JHU \OKC \NORDITA
+\author{Thomas D.~P.~Edwards} \OKC \NORDITA
 \author{Daniel Foreman-Mackey} \CCA
 \author{Maximiliano Isi} \CCA
-\author{Kelvin Lam} \CCA
+\author{Kelvin K.~H.~Lam} \CCA \CUHK
 \author{Kaze W.~K.~Wong} \CCA
@@
 \begin{abstract}
-    Since their discovery in 2015, gravitaitonal waves (GWs) have developed into a unique observational probe used throughout fundamental and astro-phyics.
-    As more detections continue, data analysis taks are becoming increasingly computationally demanding.
-    Here we propose the use of automatic differentiation through the programming framework \jax for accelerating a variety of these analsyis tasks.
-    Firstly, we demonstrate that highly accurate waveforms (IMRPhenomD) can be written in \jax and demonstrate that the evaluation speed of the waveform (and its derivative) is not substantially slower than the \lalsuite implenetation in \CC.
-    We then focus on three applications where speed and differentiable waveforms are essential.
-    Firstly, we demonstrate that stochastic gradient descent can be used to optimize the $\sim 200$ coefficients that make up the waveform. 
-    In particular, we demonstrate that the typical \textit{match} with numerical relativiely waveforms can be improved by more than 30\% without any additional overhead.
-    Secondly, we show that Fisher forecasting calculations can be sped up by $\sim 100\times$ with no loss in accuracy.
-    This increased speed makes population forecasting substantially simpler, as discussed extensively in \citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.
-    Finally, we show that gradient based samplers like Hamiltonian Monte Carlo can improve acceptance rates of XXX. 
-    Since differentiable waveforms appear to have substatial advantages for a variety of tasks throughtout GW science (with little downside), we propose that waveform developers use \jax to build new waveforms moving forward.
-    Our waveform code, called \ripple, can be found at XXX, and will continue to be updated with new waveforms as they are formulated.
-
+    Here we will discuss our implementation of differentiable waveforms and demonstrate their benefits for a variety of data analysis tasks.
 \end{abstract}
@@
 Once potential candidates are found, parameter estimation (PE) is performed to extract the detailed source properties of each event.
-In the most minimial example, this requires one to perform an MCMC on an 11 dimensional parameter space for each event.
+In the most minimial example, this requires one to perform MCMC over a 15 dimensional parameter space for each event.
 Beyond this simple scenario, more complex waveform models with many additional parameters may be used to test for deviations from General Relativity.
@@
 in Hamiltonian Monte Carlo~\citep{2017arXiv170102434B} or Gradient Decsent).
-Gradients are particularly valuable for high dimensional spaces.
+This use of gradients is particularly useful for high dimensional spaces.
 Unfortunately, in the field of GW data analysis, analytic derivatives of the necessary quantities (such as the likelihood) have historically been difficult to obtain.
@@
 There are now a variety of AD implementations, most notably in deep learning frameworks such as \texttt{pytorch}~\citep{pytorch} and \texttt{tensorflow}~\citep{tensorflow2015-whitepaper}.
-More general frameworks exist in \texttt{julia}~\citep{zygote, forwarddiff}, although \texttt{julia}'s limited support in GW physics precludes its general use.
+More general frameworks exist in \texttt{julia}~\citep{zygote, forwarddiff}, although \texttt{julia}'s limited use in GW physics precludes its general use.
 Here we make use of \jax~\citep{jax2018github} due to its easy integration with \texttt{python} libraries, ability to just-in-time compile, and ability to utilize a variety of computational resources (i.e.
@@
 In this paper we argue that differentiable waveforms will be a vital component for the future of GW data analysis.
-In addition, we present \ripple, a small GW \texttt{python} package which can serve as a respository for differentiable implementations of some of the main waveforms currently used in LIGO and Virgo analyses.
+In addition, we present \ripple, a small GW \texttt{python} package containing differentiable implementations of some of the main waveforms currently used in LIGO and Virgo analyses. 
 The remainder of this paper is structured as follows. 
 In Sec.~\ref{sec:waveforms} we discuss the differentiable waveforms implemented in \ripple and perform some benchmarks to demonstrate the speed and accruacy of the derivative calculations. 
-In Sec.~\ref{sec:applications} we discuss three distinct applications using differentiable waveforms. 
-Firstly, we illustrate how the fit coefficients that form part of the IMRPhenom waveform models could be improved by high dimensional fitting, enabled by a differentiable waveform. 
-Secondly, we implement differnetiable detector response functions and show that the speed of Fisher matrix calculations can be substantially improved using AD.
-Finally, we run an illustrative injection example using Hamiltonian Monte Carlo to demonstrate that the acceptance rate of derivative samplers is substantially larger than traditional MCMCs.
+In Sec.~\ref{sec:applications} we discuss four distinct applications using differentiable waveforms. 
+First, a simple demonstration of accelerated effectualness calculations using gradient descent.
+Second, we show that Fisher analyses are quickly enabled using AD.
+Third, we implement differnetiable detector response functions and show an illutrative example using Hamiltonian Monte Carlo.
+Finally, we illustrate how the fit coefficients that form part of the IMRPhenom waveform models could be improved by high dimensional fitting, enabled by a differentiable waveform. 
 The associated code can be found at \ripple~\citep{ripple}.
@@
 In particular, the blue points indicate parameter points where the $1-\mathrm{m}(h_{\theta_1}|h^{*}_{\theta_1})=0$ and would therefore not appear in the log scale colorbar.
-\begin{figure}[t]
+\begin{figure}[ht!]
     \script{random_matches.py}
@@
 Since the \lalsuite implementation of IMRPhenomD is written in \texttt{C}, the evaluation of the waveform itself is faster than \ripple.
-Benchmarking on a MacBook Pro with an M1 Max Apple Silicon processor, we find that a single waveform evaluation takes $\sim 2-5\,\mathrm{ms}$ for \lalsuite (interfaced with python) and $\sim 5-8\,\mathrm{ms}$ for \ripple.\footnote{
+Benchmarking on a MacBook Pro with an M1 Max Apple Silicon processor, we find that a single waveform evaluation takes $\sim 4\,\mathrm{ms}$ for \lalsuite (interfaced with python) and $\sim 10\,\mathrm{s}$ for \ripple.\footnote{
     For this benchmark we used $16\,\mathrm{Hz}$ and $1024\,\mathrm{Hz}$ for the lowest and highest frequencies respectively. 
@@
 Although slower on a CPU benchmark, a key advantage of using \jax is its native ability to run on a GPU.
-Waveform calls can therefore be evaluated in parallel, massively improving performance.
-Performing the same benchmark as above on a google colab notebook with the freely provided GPU, we find that on average waveform evaluations take $\sim 0.00005\mathrm{ms}$, nearly five orders of magnitude faster.
+Waveform calls can therefore be evaluated in parallel, massively improving performance. \te{I couldn't get these to be similar on google colab. @Kaze any ideas how to make this fast?}
 Generalizing \lalsuite waveforms to run on a GPU would be a significant undertaking.
@@
 
-% \te{Kelvin should add stuff here}
-
 Having an accurate waveform model is essential for many data analysis tasks in
@@
 full numerical relativity waveforms, which are essentially simpler ansatzes that
-are not as accurate but can be evaluated much faster. There are a few families
-of approximants, such as the IMRPhenom family, SEOB family, and the NRSurrogate
-model. \kw{cite and add a couple more sentences to introduce the waveform
-families.} \te{I think a full dicussion of the various waveform families just belongs in section 2, so don't need too much more here I'd say.}
+are not as accurate but can be evaluated much faster. 
+There are a few families of approximants, such as the IMRPhenom family, SEOB family, and the NRSurrogate model. 
+\kw{cite and add a couple more sentences to introduce the waveform families.} 
+\te{I think a full dicussion of the various waveform families just belongs in section 2, so don't need too much more here I'd say.}
 
-Every waveform approximant has some fitting coefficients that are fixed during
+Every waveform approximants have some fitting coefficients that are fixed during
 the process of calibrating the ansatz to NR waveforms. In the case of
@@
 means misrepresentation of the NR waveform, meaning the use of approximant will
-introduce some systematic error in downstream data analysis task. For example, sufficiently large systematic errors 
-would mean cause the recovered source parameters to be biased in the case of PE.
+introduce some systematic error in downstream data analysis task. For example,
+this would mean the recovered source parameters will be biased in the case of PE.
 
-% Motivation
 Previously in the construction of IMRPhenomD \citep{Khan:2015jqa}, waveform
@@
 
-\kw{Kelvin Fill}
 Since we can differentiate through the entire waveform model against the fitting
@@
 
-Here, we define the mismatch as
+Here, we define the mismatch and loss function as
 \begin{equation}
-	\mathcal{M}=1-\mathrm{m}(h_{\mathrm{IMR}}|h_{\mathrm{NR}}), 
+	\mathcal{M}&=1-\mathrm{m}(h_{\mathrm{IMR}}|h_{\mathrm{NR}}).
 \end{equation}
-In our optimization procedure, we took NR waveforms stated in \citep{Khan:2015jqa} as training waveforms to fit waveform coefficients. 
-By defining the mean of individual mismatches as the loss function for optimization, we can calibrate the set of coefficients using gradient descent.
-
-\kw{Describe figure losscompare}
+In the optimization procedure, we took NR waveforms $h_{\mathrm{NR}}$ stated in \citep{Khan:2015jqa} as training waveforms to fit waveform coefficients.  
+Then, combining individual mismatches, a loss function is constructed as follows:
+\begin{equation}
+	\mathcal{L}&=\frac{1}{N}\sum_{i=1}^{N}\mathcal{M}_i, 
+\end{equation}
+where $\mathcal{M}_i$ is the mismatch of individual training waveforms and $N$ is the total number of training waveforms used in optimization. 
+Then, one can proceed to perform optimization on waveform coefficients.
 
 \begin{figure}[t]
-\centering
-\includegraphics[width=\linewidth]{../static/amplitude_loss.pdf}
-\caption{Relative error of amplitudes of IMRPhenomD waveforms. Blue: Error
+	\script{loss_compare.py}
+	\begin{centering}
+		\includegraphics[width=\linewidth]{figures/loss_compare.pdf}
+		\caption{
+			Relative error of IMRPhenomD waveform amplitudes. Blue: Error
 			between original Phenom waveform and NR waveform; Orange: Error between
-    optimized waveform and NR waveform. \te{Can you add the data and the plotting script to the repo so that we can generate the plots using the same matplotlib settings?}}
+			optimized waveform and NR waveform.
+		}
 		\label{fig:loss_compare}
+	\end{centering}
 \end{figure}
 
-\kw{Describe figure logdiff}
 As demonstration, Fig. \ref{fig:loss_compare} shows the relative error of original and optimized waveform at different frequencies. 
 In general, the error of optimized waveform is lower than that of the original waveform. 
-The error in the early inspiral region is substantially decreased by half and other regions also show good improvement in accuracy.
+The error in the early inspiral region is decreased by half while other regions also show good improvement in accuracy.
 
 \begin{figure}[t]
-\centering
-\includegraphics[width=\linewidth]{../static/MM_logdiff_q148_LIGO.pdf}
-\caption{Distribution of the log difference of mismatches between optimized
-    and original model. The mean shows a 36\% decrease in mismatch. }
-\label{fig:MM_logdiff}
+	\script{mismatch_hist.py}
+	\begin{centering}
+		\includegraphics[width=\linewidth]{figures/mismatch_hist.pdf}
+		\caption{
+			Distributions of log mismatches of training waveforms. Blue: Original waveforms; Orange: Optimized waveforms.
+		}
+		\label{fig:mismatch_hist}
+	\end{centering}
 \end{figure}
@@
 Instead of examining individual waveforms, we investigate the general behavior of mismatches for a set of testing waveforms. 
-In Fig. \ref{fig:logdiff_q148}, the distribution of log difference of mismatches are shown. 
-The distribution mean lies towards the negative side, which indicates a improvement in the model. 
-Despite there are testing waveforms that performed worse after optimization, it is verified that the ansatz of IMRPhenomD does not fit well with most of these waveforms, hence cannot be improved by optimization.
+In Fig. \ref{fig:mismatch_hist}, the distribution of log mismatches are shown. 
+The entire optimized waveform distribution is slightly shifted to the negative side. 
+Waveforms originally with moderate mismatch moved down to the left end of the distribution. 
+Both indicate that the optimization procedure improved the model. 
 Therefore, the new set of waveform coefficients can produce better waveforms for GW analyses.
-
-\kw{Describe implication}
-% Implication
 
@@
 differentiable waveform model, one can apply the same procedure to optimize the
-coefficients in the new model, i.e. applying gradient descent to optimize all
+coefficients in a new model, i.e. applying gradient descent to optimize all
 the fitting coefficients all together. This provides a more flexible control
@@
 With further modifications to the optimization process, the set of waveform
-coefficients can be retuned to give almost an order-of-magnitude improvement.
-\kw{Where is the proof for this?}
+coefficients can be retuned to give almost $50\%$ improvement in mismatch for well-behaved waveforms.
+%\kw{Where is the proof for this?}
 
@@
 \label{subsec:fisher}
-
-\begin{figure}[t!]
-    \script{sky_localization.py}
-    \centering
-    \includegraphics[width=\linewidth]{figures/sky_localization.pdf}
-    \caption{
-        This plot should show a sky localization error bar.
-    }
-    \label{fig:sky_localization}
-\end{figure}
 
@@
 
-\begin{table}[t]
+\begin{table}[t!]
     % \begin{center}
@@
 
+\begin{figure}[ht!]
+    \script{sky_localization.py}
+    \begin{centering}
+        \includegraphics[width=\linewidth]{figures/sky_localization.pdf}
+        \caption{
+            This plot should show a sky localization error bar.
+        }
+        \label{fig:sky_localization}
+    \end{centering}
+\end{figure}
+
 The distribution of sky localization errors from a population of $10^3$ binaries can be seen in Fig.~\ref{fig:sky_localization}.
@@
 This work was supported by collaborative visits funded by the Cosmology and Astroparticle Student and Postdoc Exchange Network (CASPEN).
-T.E. is supported by the Horizon Postdoctoral Fellowship.
 
@@
 
+\section{Appendix}
+\label{sec:appendix}
+
 \end{document}

### 2022-12-17 src/scripts/loss_compare.py

@@ -1,0 +1,12 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import paths
+
+(freq, ori_error, opt_error) = np.loadtxt(paths.data / "loss_compare.txt", unpack=True)
+
+plt.plot(freq, ori_error, label='original')
+plt.plot(freq, opt_error, label='optimized')
+plt.legend()
+plt.xlabel(r'$Mf$')
+plt.ylabel(r'$\Delta|\tilde{h}(f)|$')
+plt.savefig(paths.figures / "loss_compare.pdf", bbox_inches="tight")

### 2022-12-17 src/tex/ms.tex
fixed some minor errors

@@
 \begin{equation}
-	\mathcal{M}&=1-\mathrm{m}(h_{\mathrm{IMR}}|h_{\mathrm{NR}}).
+	\mathcal{M}=1-\mathrm{m}(h_{\mathrm{IMR}}|h_{\mathrm{NR}}).
 \end{equation}
@@
 \begin{equation}
-	\mathcal{L}&=\frac{1}{N}\sum_{i=1}^{N}\mathcal{M}_i, 
+	\mathcal{L}=\frac{1}{N}\sum_{i=1}^{N}\mathcal{M}_i, 
 \end{equation}
@@
 In Fig. \ref{fig:mismatch_hist}, the distribution of log mismatches are shown. 
-The entire optimized waveform distribution is slightly shifted to the negative side. 
-Waveforms originally with moderate mismatch moved down to the left end of the distribution. 
+The entire optimized waveform distribution slightly shifted to the negative side. 
+The peak of the original waveform distribution moved to the left end of the distribution. 
 Both indicate that the optimization procedure improved the model. 

### 2022-12-19 src/scripts/random_matches.py
Adding a bunch of stuff

@@ -11,6 +11,7 @@ params = {
     "axes.titlesize": 18,
     "xtick.labelsize": 18,
     "ytick.labelsize": 18,
+    "axes.unicode_minus": False,
     "figure.figsize": (7, 5),
     "xtick.top": True,
     "ytick.right": True,
@@ -25,7 +26,10 @@ params = {
     "ytick.direction": "in",
     "axes.linewidth": 1.5,
     "text.usetex": False,
-    "mathtext.fontset": "dejavuserif",
+    "font.family": "serif",
+    "font.serif": "cmr10",
+    "mathtext.fontset": "cm",
+    "axes.formatter.use_mathtext": True,  # needed when using cm=cmr10 for normal text
 }
 
 # mpl.rcParams["text.latex.preamble"] = [r"\usepackage{amsmath}"]  # for \text command

### 2022-12-19 src/scripts/sky_localization.py

@@ -13,6 +13,7 @@ params = {
     "ytick.labelsize": 18,
     "figure.figsize": (7, 5),
     "xtick.top": True,
+    "axes.unicode_minus": False,
     "ytick.right": True,
     "xtick.bottom": True,
     "ytick.left": True,
@@ -25,7 +26,10 @@ params = {
     "ytick.direction": "in",
     "axes.linewidth": 1.5,
     "text.usetex": False,
-    "mathtext.fontset": "dejavuserif",
+    "font.family": "serif",
+    "font.serif": "cmr10",
+    "mathtext.fontset": "cm",
+    "axes.formatter.use_mathtext": True,  # needed when using cm=cmr10 for normal text
 }
 
 

### 2022-12-19 src/tex/ms.tex
Adding a bunch of stuff
Added more text to HMC section

--- 2022-12-16 src/tex/ms.tex
@@
 \newcommand{\lalsuite}{\texttt{lalsuite}\xspace}
+\newcommand{\flowMC}{\texttt{flowMC}\xspace}
 \newcommand{\CC}{C\nolinebreak\hspace{-.05em}\raisebox{.4ex}{\tiny\bf +}\nolinebreak\hspace{-.10em}\raisebox{.4ex}{\tiny\bf +}}
@@
 
+\begin{figure}[t!]
+    \centering
+    \includegraphics[width=\linewidth]{../static/amplitude_loss.pdf}
+    \caption{Relative error of amplitudes of IMRPhenomD waveforms. Blue: Error
+        between original Phenom waveform and NR waveform; Orange: Error between
+        optimized waveform and NR waveform. \te{Can you add the data and the plotting script to the repo so that we can generate the plots using the same matplotlib settings?}}
+    \label{fig:loss_compare}
+    \end{figure}
+
 Having an accurate waveform model is essential for many data analysis tasks in
@@
 
-\begin{figure}[t]
-\centering
-\includegraphics[width=\linewidth]{../static/amplitude_loss.pdf}
-\caption{Relative error of amplitudes of IMRPhenomD waveforms. Blue: Error
-    between original Phenom waveform and NR waveform; Orange: Error between
-    optimized waveform and NR waveform. \te{Can you add the data and the plotting script to the repo so that we can generate the plots using the same matplotlib settings?}}
-\label{fig:loss_compare}
-\end{figure}
-
 \kw{Describe figure logdiff}
@@
 
-\begin{figure}[t]
+\begin{figure}[t!]
 \centering
@@
 \texttt{GWbench}~\citep{Borhanian:2020ypi}, on the other hand, takes $\mathcal{O}$(minutes) for each Fisher calculation using the same detector setup and frequency grid.
-This factor of over 100 speed up is substantial considering the fact that a single core evaluation of the \ripple waveform is significantly slower that the \lalsuite call.
+This factor of over 100 speed up is substantial considering the fact that a single core evaluation of the \ripple waveform is slower than the \lalsuite call.
 As discussed above, performance can be further improved by utilising hardware acceleration such as parallel GPU processing.
@@
 
+\begin{figure*}[t]
+    \script{corner_plot.py}
+    \centering
+    \includegraphics[width=\linewidth]{figures/PE.pdf}
+    \caption{
+        Corner plot
+    }
+    \label{fig:corner}
+\end{figure*}
+
 After the search algorithms have constructed a list of confidently detected binaries, the next step is to sample from the posterior of each sources parameters - so called parameter estimation (PE).
-To do this, one typically uses a Markov Chain Monte Carlo (MCMC) sampler or nested sampler~\citep{multinest, dynesty}.
+To do this, one typically uses a Markov Chain Monte Carlo (MCMC) or nested sampler~\citep{multinest, dynesty}.
 Although robust, both MCMC and nested sampling are slow to converge and are known to perform poorly in high dimensional parameter spaces.
-For example, sampling the 15 dimensional parameter space for a BBH system can take $\mathcal{O}(10)$hours, whilst BNS systems can take up to weeks.
+For example, sampling the 15 dimensional parameter space for a BBH system can take $\mathcal{O}(10)$ hours, while BNS systems can take up to weeks.
 Dedicated fast samplers have been designed to get approximate posteriors on the sky localization to faciliate follow-up electromagnetic observations.
@@
 In particular, we will show that proposals made using an Hamiltonian Monte Carlo (MC) sampler are accepted at a much higher rate than a traditional MCMC algorithm.
-In a follow up paper we will demonstrate that minute scale PE can be achieved by exploiting normalizing flows~\citep{2022arXiv221106397W, Gabrie:2021tlu} together with a derivative based sampler~\citep{PEpaper}.
+
+For our basic example we perform an injection recovery test on a seven dimensional parameter space with the same three detector network setup considered in \S~\ref{subsec:fisher}.
+More specifically, we generate Gaussian noise consistent with the measured PSDs for each detector and then inject a BBH signal with parameters: $M_c = 23.82\,\mathrm{M_\odot}$,  $\eta = 0.248$, $\chi_1=0.3$, $\chi_1=-0.4$, $D = 1.6\,\mathrm{Gpc}$, $t_c = 0.0$, and $\phi_c = 0.0$.
+Using a standard Gaussian likelihood, we then run the HMC sampler implemented in \flowMC for XX steps. 
+\te{Describe more carefully what we do}
+
+In Fig.~\ref{fig:corner}, the grey contours show the posterior recovered from our injected signal and the orange represents the injected parameters.
+From the one dimensional histograms along the diagonal, it is clear that we successfully recover all seven parameters.
+This is expected since the injected binary is relatively nearby.
+
+To illustrate the benefit of a gradient based sampler, we run the same sampling as above but instead use a Gaussian random walk (GRW) sampler. 
+We sample for the same number of steps as used in the HMC with identical initializations to explore acceptance rates of new points.
+For the HMC example, find that the mean local acceptance rate across the 15 chains is around $\sim0.28$, where as the local acceptance rate for the equivalent GRW sampler is XXX.
+Since we are running the chains from the same initialization, this enhanced acceptance for HMC can be mainly associated with use of gradient information to guide sampling.
+We therefore expect gradient based samplers to converge significantly faster than typical samplers, especially in higher dimensions.
+In a follow up paper we will demonstrate that minute scale PE can be achieved by combining normalizing flows~\citep{2022arXiv221106397W, Gabrie:2021tlu}, GPU acceleration, and a derivative based sampler~\citep{PEpaper}.
 

### 2022-12-21 src/scripts/random_matches.py
Fix styling params in the scripts related to finetuning

@@ -23,11 +23,10 @@ params = {
     "ytick.minor.size": 4,
     "xtick.direction": "in",
     "ytick.direction": "in",
-    "text.usetex": True,
+    "text.usetex": False,
     "font.family": "serif",
     "axes.linewidth": 1.5,
     "mathtext.fontset": "dejavuserif",
-    "text.usetex": True,
 }
 
 # mpl.rcParams["text.latex.preamble"] = [r"\usepackage{amsmath}"]  # for \text command

### 2022-12-28 src/tex/ms.tex

@@
 \author{Maximiliano Isi} \CCA
-\author{Kelvin Lam} \CCA
+\author{Kelvin K.~H.~Lam} \CCA \CUHK
 \author{Kaze W.~K.~Wong} \CCA
@@
 
-% \te{Kelvin should add stuff here}
-
-\begin{figure}[t!]
-    \centering
-    \includegraphics[width=\linewidth]{../static/amplitude_loss.pdf}
-    \caption{Relative error of amplitudes of IMRPhenomD waveforms. Blue: Error
-        between original Phenom waveform and NR waveform; Orange: Error between
-        optimized waveform and NR waveform. \te{Can you add the data and the plotting script to the repo so that we can generate the plots using the same matplotlib settings?}}
-    \label{fig:loss_compare}
-    \end{figure}
-
 Having an accurate waveform model is essential for many data analysis tasks in
@@
 full numerical relativity waveforms, which are essentially simpler ansatzes that
-are not as accurate but can be evaluated much faster. There are a few families
-of approximants, such as the IMRPhenom family, SEOB family, and the NRSurrogate
-model. \kw{cite and add a couple more sentences to introduce the waveform
-families.} \te{I think a full dicussion of the various waveform families just belongs in section 2, so don't need too much more here I'd say.}
+
+are not as accurate but can be evaluated much faster. 
+There are a few families of approximants, such as the IMRPhenom family, SEOB family, and the NRSurrogate model. 
+\kw{cite and add a couple more sentences to introduce the waveform families.} 
+\te{I think a full dicussion of the various waveform families just belongs in section 2, so don't need too much more here I'd say.}
 
@@
 
-% Motivation
 Previously in the construction of IMRPhenomD \citep{Khan:2015jqa}, waveform
@@
 
-Here, we define the mismatch as
+Here, we define the mismatch and loss function as
 \begin{equation}
-	\mathcal{M}=1-\mathrm{m}(h_{\mathrm{IMR}}|h_{\mathrm{NR}}), 
+	\mathcal{M}=1-\mathrm{m}(h_{\mathrm{IMR}}|h_{\mathrm{NR}}).
 \end{equation}
-In our optimization procedure, we took NR waveforms stated in \citep{Khan:2015jqa} as training waveforms to fit waveform coefficients. 
-By defining the mean of individual mismatches as the loss function for optimization, we can calibrate the set of coefficients using gradient descent.
+In the optimization procedure, we took NR waveforms $h_{\mathrm{NR}}$ stated in \citep{Khan:2015jqa} as training waveforms to fit waveform coefficients.  
+Then, combining individual mismatches, a loss function is constructed as follows:
+\begin{equation}
+	\mathcal{L}=\frac{1}{N}\sum_{i=1}^{N}\mathcal{M}_i, 
+\end{equation}
+where $\mathcal{M}_i$ is the mismatch of individual training waveforms and $N$ is the total number of training waveforms used in optimization. 
+Then, one can proceed to perform optimization on waveform coefficients.
 
-\kw{Describe figure losscompare}
+\begin{figure}[t]
+	\script{loss_compare.py}
+	\begin{centering}
+		\includegraphics[width=\linewidth]{figures/loss_compare.pdf}
+		\caption{
+			Relative error of IMRPhenomD waveform amplitudes. Blue: Error
+			between original Phenom waveform and NR waveform; Orange: Error between
+			optimized waveform and NR waveform.
+		}
+		\label{fig:loss_compare}
+	\end{centering}
+\end{figure}
 
-\kw{Describe figure logdiff}
 As demonstration, Fig. \ref{fig:loss_compare} shows the relative error of original and optimized waveform at different frequencies. 
 In general, the error of optimized waveform is lower than that of the original waveform. 
-The error in the early inspiral region is substantially decreased by half and other regions also show good improvement in accuracy.
+The error in the early inspiral region is decreased by half while other regions also show good improvement in accuracy.
 
-\begin{figure}[t!]
-\centering
-\includegraphics[width=\linewidth]{../static/MM_logdiff_q148_LIGO.pdf}
-\caption{Distribution of the log difference of mismatches between optimized
-    and original model. The mean shows a 36\% decrease in mismatch. }
-\label{fig:MM_logdiff}
+\begin{figure}[t]
+	\script{mismatch_hist.py}
+	\begin{centering}
+		\includegraphics[width=\linewidth]{figures/mismatch_hist.pdf}
+		\caption{
+			Distributions of log mismatches of training waveforms. Blue: Original waveforms; Orange: Optimized waveforms.
+		}
+		\label{fig:mismatch_hist}
+	\end{centering}
 \end{figure}
@@
 Instead of examining individual waveforms, we investigate the general behavior of mismatches for a set of testing waveforms. 
-In Fig. \ref{fig:logdiff_q148}, the distribution of log difference of mismatches are shown. 
-The distribution mean lies towards the negative side, which indicates a improvement in the model. 
-Despite there are testing waveforms that performed worse after optimization, it is verified that the ansatz of IMRPhenomD does not fit well with most of these waveforms, hence cannot be improved by optimization.
+In Fig. \ref{fig:mismatch_hist}, the distribution of log mismatches are shown. 
+The entire optimized waveform distribution slightly shifted to the negative side. 
+The peak of the original waveform distribution moved to the left end of the distribution. 
+Both indicate that the optimization procedure improved the model. 
 Therefore, the new set of waveform coefficients can produce better waveforms for GW analyses.
-
-\kw{Describe implication}
-% Implication
 
@@
 With further modifications to the optimization process, the set of waveform
-coefficients can be retuned to give almost an order-of-magnitude improvement.
-\kw{Where is the proof for this?}
+
+coefficients can be retuned to give almost $50\%$ improvement in mismatch for well-behaved waveforms.
+%\kw{Where is the proof for this?}
 

### 2022-12-28 src/tex/ms.tex
Fix tuning section

@@
 full numerical relativity waveforms, which are essentially simpler ansatzes that
+are only approximations but can be evaluated much faster.
 
-are not as accurate but can be evaluated much faster. 
-There are a few families of approximants, such as the IMRPhenom family, SEOB family, and the NRSurrogate model. 
-\kw{cite and add a couple more sentences to introduce the waveform families.} 
-\te{I think a full dicussion of the various waveform families just belongs in section 2, so don't need too much more here I'd say.}
-
-Every waveform approximant has some fitting coefficients that are fixed during
+Every waveform approximant has some fitting coefficients that are tuned during
 the process of calibrating the ansatz to NR waveforms. In the case of
@@
 reproduces the NR waveform. The inaccuracy in obtaining the fitting coefficients
-means misrepresentation of the NR waveform, meaning the use of approximant will
-introduce some systematic error in downstream data analysis task. For example, sufficiently large systematic errors 
-would mean cause the recovered source parameters to be biased in the case of PE.
+means misrepresentation of the NR waveform, which can translate to systematic
+error in downstream data analysis task. For example, sufficiently large
+systematic errors in the waveform would cause the recovered source parameters to be biased
+in the case of PE.
 
@@
 coefficients are fitted in segments with amplitude and phase separated.
-Individual segments are then connected together such that they are
-differentiable. Using this method, the generated waveforms are expected to be
-erroneous, as the connections would affect the originally fitted segments. The
-coefficients of the current IMRPhenomD implementation are often hand tuned, and
-are tuned in sub-sets of parameters instead of all together. This means the
-tuning process ignore the correlation between different sub-sets of parameters,
-so the best-fitted solution obtained by fitting sub-sets of parameters may not
-be the global optimum compared to fitting all coefficients at once. Instead, by
-jointly fitting all coefficients, the inaccuracy in fitting individual and
-connecting different segments can be accounted during the optimization.
-Therefore, there is potential accuracy one can gain by fitting all coefficients
-at once.
+Furthermore, IMRPhenomD is divided in three fitting segments, inpsiral, merger and ringdown.
+Each of these segments has their own set of fitting coefficients.
+After obtaining the fitting coefficient for individual segments, they are then
+"stitched" together such that they are continuous in the first derivative.
+The process of stitching introduces some addition
+inaccuracy in the waveform model, as the connections would affect the originally
+fitted segments.
+The coefficients of the current IMRPhenomD implementation are tuned in sub-sets
+of parameters instead of all together. This means the tuning process ignore the
+correlation between different sub-sets of parameters, so the best-fitted
+solution obtained by fitting sub-sets of parameters may not be the global
+optimum compared to fitting all coefficients at once. By jointly
+fitting all coefficients, the correlation in fitting individual and connecting
+different segments can be accounted during the optimization. Therefore, there is
+potential improvement of the accuracy by fitting all coefficients at once.
 
-Since we can differentiate through the entire waveform model against the fitting
-parameters, one can use gradient descent to find the optimal fitting parameters
-all at once instead of fitting them in subsets.
+In general, optimization problems in a high dimensional space benefit from
+having access to the gradient of the objective function. Since we can
+differentiate through the entire waveform model against the fitting parameters,
+one can use gradient descent to find the optimal fitting parameters all at once
+instead of fitting them in subsets.
 
-Here, we define the mismatch and loss function as
+The first step in fitting all coefficients is to define a loss function that
+measures the goodness-of-fit of the optimization procedure, which we choose it
+to be the mismatch between the NR waveform and the approximant waveform.
 \begin{equation}
-	\mathcal{M}=1-\mathrm{m}(h_{\mathrm{IMR}}|h_{\mathrm{NR}}).
+	\mathcal{M}(\lambda)=1-\mathrm{m}(h_{\mathrm{IMR}}(\lambda)|h_{\mathrm{NR}(\lambda)}),
 \end{equation}
-In the optimization procedure, we took NR waveforms $h_{\mathrm{NR}}$ stated in \citep{Khan:2015jqa} as training waveforms to fit waveform coefficients.  
-Then, combining individual mismatches, a loss function is constructed as follows:
+where $\lambda$ is a vector of the fitting coefficients, $h_{\mathrm{IMR}}$ is
+the waveform generated by the IMRPhenomD and $h_{\mathrm{NR}}$ is the waveform
+generated by the numerical relativity simulation. 
+Given the loss function, we use gradient descent to find the optimal fitting
+coefficients.
+% Add description of the optimization procedure.
+
+As a demonstration, Fig. \ref{fig:loss_compare} shows the relative error of
+original and optimized waveform at different frequencies. In this optimization
+procedure, we took NR waveforms $h_{\mathrm{NR}}$ stated in \citep{Khan:2015jqa}
+as training waveforms to fit waveform coefficients.   
+We can see the error of optimized waveform is lower than that of the original waveform for most of the domain.
+In particular, the error in the early inspiral region is decreased by half while other regions also show good improvement in accuracy.
+
+We can generalize the fitting procedure described above to a collection of waveforms.
+In that case, the loss function is defined as the average of the mismatch of
+individual waveforms, given by
 \begin{equation}
 	\mathcal{L}=\frac{1}{N}\sum_{i=1}^{N}\mathcal{M}_i,
+    \label{eq:multi_loss}
 \end{equation}
-where $\mathcal{M}_i$ is the mismatch of individual training waveforms and $N$ is the total number of training waveforms used in optimization. 
-Then, one can proceed to perform optimization on waveform coefficients.
+where $\mathcal{M}_i$ is the mismatch of individual training waveforms and $N$
+is the total number of training waveforms used in optimization. This
+optimization is more difficult since we are applying the same set of
+coefficients to waveforms with different intrinsic parameters, such as mass
+ratio and spins. From eq. \ref{eq:multi_loss}, we can see the averaging between
+waveforms with different intrinsic parameters implies there are trade-off in
+performance in different regions of the intrinsic parameter space.
 
@@
 
-As demonstration, Fig. \ref{fig:loss_compare} shows the relative error of original and optimized waveform at different frequencies. 
-In general, the error of optimized waveform is lower than that of the original waveform. 
-The error in the early inspiral region is decreased by half while other regions also show good improvement in accuracy.
-
 \begin{figure}[t]
@@
 
-Instead of examining individual waveforms, we investigate the general behavior of mismatches for a set of testing waveforms. 
-In Fig. \ref{fig:mismatch_hist}, the distribution of log mismatches are shown. 
-The entire optimized waveform distribution slightly shifted to the negative side. 
+In Fig. \ref{fig:mismatch_hist}, we show the distribution of log mismatches.
+One can see the distribution of mismatch after optimization has smaller mismatch compared to before optimization,
+showing the optimization improves the accuracy of the waveform model.
+% However, it is not improving for every waveforms. Say why briefly, and say there is another paper coming out.
 The peak of the original waveform distribution moved to the left end of the distribution. 
@@
 Therefore, the new set of waveform coefficients can produce better waveforms for GW analyses.
-
-This fine-tuning step is rather simple and general. Given a different
-differentiable waveform model, one can apply the same procedure to optimize the
-coefficients in a new model, i.e. applying gradient descent to optimize all
-the fitting coefficients all together. This provides a more flexible control
-during the calibration process, and potentially better final result.
-With further modifications to the optimization process, the set of waveform
-
-coefficients can be retuned to give almost $50\%$ improvement in mismatch for well-behaved waveforms.
-%\kw{Where is the proof for this?}
 
@@
 
-Computing the Fisher matrix requires one to evaluate derivatives of the likelihhood which in turn involves derivatives of the waveform model and detector projection functions.
+Computing the Fisher matrix requires one to evaluate derivatives of the likelihood which in turn involves derivatives of the waveform model and detector projection functions.
 AD is therefore perfectly suited for computing Fisher matrices accurately and efficiently. 

### 2023-01-03 src/scripts/Gaussian_vs_HMC_ACL.py
Add Gaussian HMC comparison scripts

@@ -1,0 +1,172 @@
+import numpy as np
+import arviz as az
+import matplotlib.pyplot as plt
+import corner
+%matplotlib inline
+
+data_HMC = np.load('../data/chains_HMC.npz')
+data_gaussian = np.load('../data/chains_gaussian.npz')
+
+HMC_bestchain = data_HMC['chains'][np.argmax(data_HMC['log_prob'][:,-1])]
+gaussian_bestchain = data_gaussian['chains'][np.argmax(data_gaussian['log_prob'][:,-1])]
+import matplotlib as mpl
+import corner
+from matplotlib.ticker import MultipleLocator, FormatStrFormatter, AutoMinorLocator
+from matplotlib.lines import Line2D
+params = {
+    "axes.labelsize": 32,
+    "font.family": "serif",
+    "font.size": 32,
+    "axes.linewidth": 2,
+    "legend.fontsize": 20,
+    "xtick.labelsize": 20,
+    "xtick.top": True,
+    "xtick.direction": "in",
+    "ytick.labelsize": 20,
+    "ytick.right": True,
+    "ytick.direction": "in",
+    "axes.grid": False,
+    "text.usetex": False,
+    "savefig.dpi": 100,
+    "lines.markersize": 14,
+    "axes.formatter.limits": (-3, 3),
+    "mathtext.fontset": "dejavuserif",
+}
+
+
+mpl.rcParams["text.latex.preamble"] = [r"\usepackage{amsmath}"]  # for \text command
+
+mpl.rcParams.update(params)
+
+
+def fill_corner(
+    fig,
+    data,
+    ndim,
+    axis,
+    color="C0",
+    true_val=None,
+    zorder=1,
+    lw=3,
+    style="-",
+    smooth1d=0.01,
+    smooth2d=0.01,
+    no_fill_contours=False,
+    fill_contours=True,
+    alpha=1,
+    levels=[0.68, 0.95],
+):
+    for i in range(ndim):
+        for j in range(i + 1):
+            ax = fig.axes[np.sum(np.arange(i + 1)) + j]
+            if i == j:
+                ax.hist(
+                    data[i],
+                    bins=axis[i],
+                    histtype="step",
+                    lw=lw,
+                    color=color,
+                    density=True,
+                    zorder=zorder,
+                )
+                ylim = ax.get_ylim()
+                ax.set_xlim(axis[i][0], axis[i][-1])
+                ax.set_ylim(ylim[0], ylim[1])
[…]

### 2023-01-03 src/scripts/Gaussian_vs_HMC_corner.py

@@ -1,0 +1,23 @@
+import numpy as np
+import arviz as az
+import matplotlib.pyplot as plt
+import corner
+%matplotlib inline
+
+data_HMC = np.load('../data/chains_HMC.npz')
+data_gaussian = np.load('../data/chains_gaussian.npz')
+
+HMC_bestchain = data_HMC['chains'][np.argmax(data_HMC['log_prob'][:,-1])]
+gaussian_bestchain = data_gaussian['chains'][np.argmax(data_gaussian['log_prob'][:,-1])]
+
+acl_HMC = az.autocorr(HMC_bestchain.T).T
+acl_gaussian = az.autocorr(gaussian_bestchain.T).T
+print(acl_HMC.shape, acl_gaussian.shape)
+plt.figure(figsize=(10,9))
+plt.plot(acl_HMC[::100,0], label='HMC')
+plt.plot(acl_gaussian[::100,0], label='Gaussian')
+plt.xlim(0,1000)
+plt.xlabel('Lag')
+plt.ylabel('Autocorrelation')
+for i in range(7):
+    print(az.ess(HMC_bestchain[::100,i]), az.ess(gaussian_bestchain[::100,i]))

### 2023-01-04 src/tex/ms.tex
added more to tuning section

@@
 coefficients.
+\begin{equation}
+	\lambda\leftarrow\lambda-\alpha\nabla\mathcal{M}, 
+\end{equation}
+where $\alpha$ is the learning rate.
+We set $\alpha$ to be a very small number, i.e.
+$10^{-6}$, since the initial guess already lies close to the desired optimum.
+We terminate the program after 30000 steps, in which $\lambda$ has converged to the optimum.
 % Add description of the optimization procedure.
@@
 coefficients to waveforms with different intrinsic parameters, such as mass
-ratio and spins. From eq. \ref{eq:multi_loss}, we can see the averaging between
+ratio and spins. From Eq. \ref{eq:multi_loss}, we can see the averaging between
 waveforms with different intrinsic parameters implies there are trade-off in
@@
 		\caption{
-			Distributions of log mismatches of training waveforms. Blue: Original waveforms; Orange: Optimized waveforms.
+			Distributions of log mismatches of testing waveforms. Blue: Original waveforms; Orange: Optimized waveforms.
 		}
@@
 % However, it is not improving for every waveforms. Say why briefly, and say there is another paper coming out.
-The peak of the original waveform distribution moved to the left end of the distribution. 
-Both indicate that the optimization procedure improved the model. 
+While many waveforms improved their accuracy, some remain to have similar mismatches.
+This phenomenon is mainly due to the reduced spin approximation used in IMRPhenomD, where spin degeneracy is introduced, hence restricting further improvements.
+In an upcoming paper, we will give a detailed discussion of how such an approximation scheme affects the optimization procedure.
+Nevertheless, the peak of the original waveform distribution moved to the left end of the distribution, indicating that the optimization procedure improved the model. 
 Therefore, the new set of waveform coefficients can produce better waveforms for GW analyses.

### 2023-01-04 src/tex/ms.tex
Edited/added plots and added some new text

@@
 Here we simply aim to demonstrate the utility of a derivative based sampler and their efficiency on a small test problem.
-In particular, we will show that proposals made using an Hamiltonian Monte Carlo (MC) sampler are accepted at a much higher rate than a traditional MCMC algorithm.
+% In particular, we will show that proposals made using an Hamiltonian Monte Carlo (MC) sampler are accepted at a much higher rate than a traditional MCMC algorithm.
+In particular, we will show that the autocorrelation of a Hamiltonian Monte Carlo (HMC) sampler is significantly lower than a traditional MCMC algorithm.
 
 For our basic example we perform an injection recovery test on a seven dimensional parameter space with the same three detector network setup considered in \S~\ref{subsec:fisher}.
-More specifically, we generate Gaussian noise consistent with the measured PSDs for each detector and then inject a BBH signal with parameters: $M_c = 23.82\,\mathrm{M_\odot}$,  $\eta = 0.248$, $\chi_1=0.3$, $\chi_1=-0.4$, $D = 1.6\,\mathrm{Gpc}$, $t_c = 0.0$, and $\phi_c = 0.0$.
-Using a standard Gaussian likelihood, we then run the HMC sampler implemented in \flowMC for XX steps. 
+More specifically, we generate Gaussian noise consistent with the measured PSDs for each detector and then inject a BBH signal with parameters: $M_c = 23.82\,\mathrm{M_\odot}$,  $\eta = 0.248$, $\chi_1=0.3$, $\chi_1=-0.4$, $D = 1.6\,\mathrm{Gpc}$, $t_c = 0.0$, and $\phi_c = 0.0$.\footnote{
+    The remaining parameters are set to $\pi/3$.
+}
+Using a standard Gaussian likelihood, we then run the HMC sampler implemented in \flowMC for 200000 steps and the Gaussian random walk sampler for 1500000 steps (each with four randomly initialized independent chains).
+The number of steps and and mass matrix used for each example was hand tuned to give good performance for the specific sampler.
 \te{Describe more carefully what we do}
@@
 This is expected since the injected binary is relatively nearby.
+
+\begin{figure}[t]
+	\script{autocorrelation.py}
+	\begin{centering}
+		\includegraphics[width=\linewidth]{figures/autocorrelation.pdf}
+		\caption{
+			Autocorrelation
+		}
+		\label{fig:autocorrelation}
+	\end{centering}
+\end{figure}
 

### 2023-01-04 src/scripts/autocorrelation.py

@@ -2,6 +2,7 @@ import numpy as np
 import matplotlib.pyplot as plt
 import matplotlib as mpl
 import paths
+import arviz as az
 
 params = {
     "font.size": 18,
@@ -11,9 +12,9 @@ params = {
     "axes.titlesize": 18,
     "xtick.labelsize": 18,
     "ytick.labelsize": 18,
-    "axes.unicode_minus": False,
     "figure.figsize": (7, 5),
     "xtick.top": True,
+    "axes.unicode_minus": False,
     "ytick.right": True,
     "xtick.bottom": True,
     "ytick.left": True,
@@ -35,17 +36,22 @@ params = {
 
 mpl.rcParams.update(params)
 
-(freq, ori_error, opt_error) = np.loadtxt(paths.data / "loss_compare.txt", unpack=True)
+data_HMC = np.load(paths.data / "chains_HMC.npz")
+data_gaussian = np.load(paths.data / "chains_gaussian.npz")
 
+HMC_bestchain = data_HMC["chains"][np.argmax(data_HMC["log_prob"][:, -1])]
+gaussian_bestchain = data_gaussian["chains"][
+    np.argmax(data_gaussian["log_prob"][:, -1])
+]
 
-plt.axvline(x=0.018, color="k", alpha=0.5, ls="--")
-# fRD =
-# 0.5fRD
-plt.axvline(x=0.08, color="k", alpha=0.4, ls="--")
-plt.plot(freq, ori_error, label="Original")
-plt.plot(freq, opt_error, label="Optimized")
+acl_HMC = az.autocorr(HMC_bestchain.T).T
+acl_gaussian = az.autocorr(gaussian_bestchain.T).T
+
+plt.plot(acl_HMC[:, 0], label="HMC")
+plt.plot(acl_gaussian[:, 0], label="Gaussian")
+plt.xlim(0, 1000)
+plt.ylim(-0.1, 1.0)
+plt.xlabel("Lag")
+plt.ylabel("Autocorrelation")
 plt.legend()
-plt.xlabel(r"$Mf$")
-plt.ylabel(r"$\Delta|\tilde{h}(f)|$")
-plt.xlim(0.01, 0.1)
-plt.savefig(paths.figures / "loss_compare.pdf", bbox_inches="tight")
+plt.savefig(paths.figures / "autocorrelation.pdf", bbox_inches="tight")

### 2023-01-04 src/scripts/corner_plot.py

@@ -35,7 +35,10 @@ params = {
 
 
 mpl.rcParams.update(params)
-file = np.load(paths.data / "ripple_flowMC_results.npz")
+
+# file = np.load(paths.data / "ripple_flowMC_results.npz")
+data_HMC = np.load(paths.data / "chains_HMC.npz")
+HMC_bestchain = data_HMC["chains"][np.argmax(data_HMC["log_prob"][:, -1])]
 
 
 def ms_to_Mc_eta(m):
@@ -64,7 +67,7 @@ true_params = np.array(
 )
 
 # Plot all chains
-figure = corner.corner(file["chains"].reshape(-1, n_dim), labels=labels, labelpad=0.1)
+figure = corner.corner(HMC_bestchain, labels=labels, labelpad=0.1)
 
 # Extract the axes
 axes = np.array(figure.axes).reshape((n_dim, n_dim))

### 2023-01-04 src/scripts/mismatch_hist.py

@@ -1,13 +1,51 @@
 import numpy as np
 import matplotlib.pyplot as plt
+import matplotlib as mpl
 import paths
 
-(catalog_num, ori_loss, opt_loss) = np.loadtxt(paths.data / "mismatch_hist.txt", unpack=True)
+params = {
+    "font.size": 18,
+    "legend.fontsize": 18,
+    "legend.frameon": False,
+    "axes.labelsize": 18,
+    "axes.titlesize": 18,
+    "xtick.labelsize": 18,
+    "ytick.labelsize": 18,
+    "figure.figsize": (7, 5),
+    "xtick.top": True,
+    "axes.unicode_minus": False,
+    "ytick.right": True,
+    "xtick.bottom": True,
+    "ytick.left": True,
+    "xtick.major.pad": 8,
+    "xtick.major.size": 8,
+    "xtick.minor.size": 4,
+    "ytick.major.size": 8,
+    "ytick.minor.size": 4,
+    "xtick.direction": "in",
+    "ytick.direction": "in",
+    "axes.linewidth": 1.5,
+    "text.usetex": False,
+    "font.family": "serif",
+    "font.serif": "cmr10",
+    "mathtext.fontset": "cm",
+    "axes.formatter.use_mathtext": True,  # needed when using cm=cmr10 for normal text
+}
+
+
+mpl.rcParams.update(params)
+
+(catalog_num, ori_loss, opt_loss) = np.loadtxt(
+    paths.data / "mismatch_hist.txt", unpack=True
+)
 bins = np.arange(-5, -1, 0.125)
 
-plt.hist(np.log10(ori_loss), bins=bins, alpha=0.5, label='original')
-plt.hist(np.log10(opt_loss), bins=bins, alpha=0.5, label='optimized')
+plt.hist(np.log10(ori_loss), bins=bins, alpha=0.3, color="C0")
+plt.hist(np.log10(ori_loss), bins=bins, label="Original", histtype="step", color="C0")
+plt.hist(np.log10(opt_loss), bins=bins, alpha=0.3, color="C1")
+plt.hist(np.log10(opt_loss), bins=bins, label="Optimized", histtype="step", color="C1")
 plt.legend()
-plt.xlabel('$\log(\mathcal{M})$')
-plt.ylabel('Count')
+plt.xlim(1e-5, 0.1)
+plt.xlabel("$\log(\mathcal{M})$")
+plt.ylabel("Count")
 plt.savefig(paths.figures / "mismatch_hist.pdf", bbox_inches="tight")

### 2023-01-05 src/tex/ms.tex
Updates thoughout

@@
     As more detections continue, data analysis taks are becoming increasingly computationally demanding.
-    Here we propose the use of automatic differentiation through the programming framework \jax for accelerating a variety of these analsyis tasks.
-    Firstly, we demonstrate that highly accurate waveforms (IMRPhenomD) can be written in \jax and demonstrate that the evaluation speed of the waveform (and its derivative) is not substantially slower than the \lalsuite implenetation in \CC.
-    We then focus on three applications where speed and differentiable waveforms are essential.
-    Firstly, we demonstrate that stochastic gradient descent can be used to optimize the $\sim 200$ coefficients that make up the waveform. 
-    In particular, we demonstrate that the typical \textit{match} with numerical relativiely waveforms can be improved by more than 30\% without any additional overhead.
-    Secondly, we show that Fisher forecasting calculations can be sped up by $\sim 100\times$ with no loss in accuracy.
+    Here, we propose the use of automatic differentiation through the programming framework \jax for accelerating a variety of these analsyis tasks.
+    Firstly, we demonstrate that highly accurate waveforms (IMRPhenomD) can be written in \jax and demonstrate that the serial evaluation speed of the waveform (and its derivative) is similar to the \lalsuite implementation in \CC.
+    Moreover, \jax allows for GPU accelerated waveform calls which can be over five orders of magnitude faster than serial evaluation on a CPU.
+    We then focus on three applications where efficient and differentiable waveforms are essential.
+    Firstly, we demonstrate how stochastic gradient descent can be used to optimize the $\sim 200$ coefficients that make up the waveform. 
+    In particular, we demonstrate that the typical \textit{match} with numerical relativiely waveforms can be improved by more than 10\% without any additional overhead.
+    Secondly, we show that Fisher forecasting calculations can be sped up by $\sim 100\times$ (on CPU) with no loss in accuracy.
     This increased speed makes population forecasting substantially simpler, as discussed extensively in \citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.
-    Finally, we show that gradient based samplers like Hamiltonian Monte Carlo can improve acceptance rates of XXX. 
-    Since differentiable waveforms appear to have substatial advantages for a variety of tasks throughtout GW science (with little downside), we propose that waveform developers use \jax to build new waveforms moving forward.
-    Our waveform code, called \ripple, can be found at XXX, and will continue to be updated with new waveforms as they are formulated.
+    Finally, we show that gradient based samplers like Hamiltonian Monte Carlo lead to significantly reduced autocorrelation values when compared to tradiaitional Monte Carlo methods.
+    Since differentiable waveforms appear to have substantial advantages for a variety of tasks throughtout GW science, we propose that waveform developers use \jax to build new waveforms moving forward.
+    Our waveform code, \ripple, can be found at \href{https://github.com/tedwards2412/ripple}{github.com/tedwards2412/ripple}, and will continue to be updated with new waveforms as they are formulated.
 
 \end{abstract}
+
+% \section{Todo}
+% \begin{itemize}
+%     \item Finish intro \te{Done}
+%     \item Finish section 2 
+%     \item Write captions for all plots \te{Done}
+%     \item Finish discussion
+% \end{itemize}
 
@@
 In addition, this computational cost will substatially increase when next generation detectors come online.
+
 The complexity begins before data gathering where one is required to generate banks of template waveforms which will be used for a matched-filter search~\citep{Owen:1998dk, Owen:1995tm}. 
@@
 Beyond this simple scenario, more complex waveform models with many additional parameters may be used to test for deviations from General Relativity.
-Finally, using the results of the PE, population synthesis models constrain the progenitors systems from which the black holes we see merging today began their journey~\citep{Wong:2022flg}.
+Finally, using the results of PE, population synthesis models constrain the progenitors systems from which the black holes we see merging today began their journey~\citep{Wong:2022flg}.
 All of these tasks require significant compute.
@@
     }
-The full derivative can then be contructed using the chain rule.
+The full derivative can then be constructed using the chain rule.
 There are now a variety of AD implementations, most notably in deep learning frameworks such as \texttt{pytorch}~\citep{pytorch} and \texttt{tensorflow}~\citep{tensorflow2015-whitepaper}.
@@
 Of these, the IMRPhenom family is most well suited for an AD implementation in \jax.
-In particular, its closed form expression and fixed size frequency grid make a \jax implementation that complies with the constraints of JIT compilation simple.
-EOB waveforms on the other hand require... \te{Max or Kaze, could one of you write something about why its hard? I don't know enough about EOB.}
+In particular, their closed form expressions and fixed size frequency grids make a \jax implementation that complies with the constraints of JIT compilation simple.
+EOB waveforms on the other hand require dynamic frequency grids and are therefore tricky to implement in \jax. 
+In addition, EOB waveforms use hypergeometric functions which have limited support in \jax.
+NRsurrogate waveforms have no fundamental issue but require the user to download relatively large data files, making their use cubersome in practice.\te{Max/Kaze, please add to/change this if you think its incorrect}
 
@@
 The remainder of this paper is structured as follows. 
-In Sec.~\ref{sec:waveforms} we discuss the differentiable waveforms implemented in \ripple and perform some benchmarks to demonstrate the speed and accruacy of the derivative calculations. 
+In Sec.~\ref{sec:waveforms} we discuss the differentiable IMRPhenomD waveform implemented in \ripple and perform some benchmarks to demonstrate its speed and accruacy. 
 In Sec.~\ref{sec:applications} we discuss three distinct applications using differentiable waveforms. 
-Firstly, we illustrate how the fit coefficients that form part of the IMRPhenom waveform models could be improved by high dimensional fitting, enabled by a differentiable waveform. 
+Firstly, we illustrate how the fit coefficients that form part of the IMRPhenom waveform models could be improved by high dimensional fitting enabled by a differentiable waveform. 
 Secondly, we implement differnetiable detector response functions and show that the speed of Fisher matrix calculations can be substantially improved using AD.
-Finally, we run an illustrative injection example using Hamiltonian Monte Carlo to demonstrate that the acceptance rate of derivative samplers is substantially larger than traditional MCMCs.
+Finally, we run an illustrative injection example using Hamiltonian Monte Carlo to demonstrate that the autocorrelation of derivative samplers is substantially smaller than that of traditional MCMCs.
 The associated code can be found at \ripple~\citep{ripple}.
@@
 This is shown in Fig.~\ref{fig:match}, where we have calculated the $\mathrm{m}(h_{\theta_1}|\bar{h}_{\theta_1})$ across the entire parameter space.\footnote{
-    Specifically, we use 2000 points in the ranges $m_{1,2} = (1,50)\,M_{\odot}$ and $\chi_{1,2} = (-1,1)$.
+    Specifically, we use $10^3$ points in the ranges $m_{1,2} = (1,100)\,M_{\odot}$ and $\chi_{1,2} = (-1,1)$.
 In addition, we evaulated the waveforms on a frequency grid from $32\,$Hz to $1024\,$Hz with frequency spacing $\Delta f = 0.0125 \,$Hz.
@@
             Match between the \ripple and \lalsuite implementations of the IMRPhenomD waveform as a function of total mass and effective spin. 
-            \te{Need to investigate more?}
+            % The blue points indicate where the match is exactly 1.0 and would therefore not appear in the log scale. 
+            It is clear that the \ripple waveform matches the \lalsuite implementation many of orders of magnitude more than necessary for all data analysis tasks across the entire parameter space.
         }
@@
 in any practical data analysis tasks. To create waveform models that can be used
-in data analysis, there are a number of groups creating "approximants" of the
+in data analysis, there are a number of groups creating ``approximants'' of the
 full numerical relativity waveforms, which are essentially simpler ansatzes that
@@
 \begin{equation}
+    \label{eq:mismatch}
 	\mathcal{M}(\lambda)=1-\mathrm{m}(h_{\mathrm{IMR}}(\lambda)|h_{\mathrm{NR}(\lambda)}),
@@
 		\caption{
-			Relative error of IMRPhenomD waveform amplitudes. Blue: Error
-			between original Phenom waveform and NR waveform; Orange: Error between
-			optimized waveform and NR waveform.
+			Relative error between the numerical relatively and IMRPhenomD waveform amplitudes. 
+            The blue line shows the relative error of the IMRPhenomD model with the original model coefficients where as the orange line uses the IMRPhenomD model after optimization using stochastic gradient descent.
+            Interestingly, the error is reduced across the frequency range, demonstrating that high dimensional optimization is useful for all parts of the waveform.
 		}
@@
 		\caption{
-			Distributions of log mismatches of testing waveforms. Blue: Original waveforms; Orange: Optimized waveforms.
+			Distribution of 536 log mismatches (see Eq.~\ref{eq:mismatch}) for the original (blue) and optimized (orange) IMRPhenomD. 
+            Overall, the optimized distribution shifts to lower mismatches and therefore a better waveform model. 
+            More quantitatively, the mean mismatch (shown as vertical dashed lines) is reduced by $\sim11$\%.
 		}
@@
     \caption{
-        This plot should show a sky localization error bar.
+        Distribution of Fisher information sky localization error for a population of nearby binaries.
+        Automatic differentiation (AD) was used to compute the derivatives in Eq.~\ref{eq:fisher}.
+        As empahsized in the text, after JIT compilation, each error calculation was around 100 times faster than a similar forecasting code \texttt{GWbench}~\citep{Borhanian:2020ypi}.
     }
@@
 \begin{equation}
+    \label{eq:fisher}
     \mathcal{I}^{k}_{ij} = (\partial_i h_0 | \partial_j h_0) \, ,
@@
     \caption{
-        Corner plot
+        Corner plot for the posteriors from an HMC (see text for details) on simulated noise with injected signal. 
+        Orange lines indicate the true values of the injection. 
+        Although not fully converged, it is clear that we find posteriors consistent with the injected parameters. 
     }
@@
 }
-Using a standard Gaussian likelihood, we then run the HMC sampler implemented in \flowMC for 200000 steps and the Gaussian random walk sampler for 1500000 steps (each with four randomly initialized independent chains).
+Using a standard Gaussian likelihood, we then run the HMC sampler implemented in \flowMC for 200000 steps and the Gaussian random walk (GRW) sampler for 1500000 steps (each with four randomly initialized independent chains).
 The number of steps and and mass matrix used for each example was hand tuned to give good performance for the specific sampler.
-\te{Describe more carefully what we do}
+The additional steps for the GRW sampler were required to achieve a similarly converged posterior.
 
-In Fig.~\ref{fig:corner}, the grey contours show the posterior recovered from our injected signal and the orange represents the injected parameters.
-From the one dimensional histograms along the diagonal, it is clear that we successfully recover all seven parameters.
+In Fig.~\ref{fig:corner}, the grey contours show the posterior recovered using the best chain (i.e. one that reached the highest log-likelihood values).
+The orange shows the true parameters of the injected signal.
+From the one dimensional histograms along the diagonal, it is clear that we consistently recover all seven parameters apart from $\phi_c$.
 This is expected since the injected binary is relatively nearby.
@@
 		\caption{
-			Autocorrelation
+			Autocorrelation as a function of lag for HMC and GRW on the same simulated data as discussed in Fig.~\ref{fig:corner}.
+            The smaller autocorrelation of the HMC leads to a larger number of independent samples and therefore a faster converging Monte Carlo.
 		}
@@
 
-To illustrate the benefit of a gradient based sampler, we run the same sampling as above but instead use a Gaussian random walk (GRW) sampler. 
-We sample for the same number of steps as used in the HMC with identical initializations to explore acceptance rates of new points.
-For the HMC example, find that the mean local acceptance rate across the 15 chains is around $\sim0.28$, where as the local acceptance rate for the equivalent GRW sampler is XXX.
-Since we are running the chains from the same initialization, this enhanced acceptance for HMC can be mainly associated with use of gradient information to guide sampling.
+% To illustrate the benefit of a gradient based sampler (such as HMC), we run the same sampling as above but instead use a Gaussian random walk (GRW) sampler. 
+% We sample for the same number of steps as used in the HMC with identical initializations to explore acceptance rates of new points.
+% For the HMC example, find that the mean local acceptance rate across the 15 chains is around $\sim0.28$, where as the local acceptance rate for the equivalent GRW sampler is XXX.
+% Since we are running the chains from the same initialization, this enhanced acceptance for HMC can be mainly associated with use of gradient information to guide sampling.
+Although further steps would be required to achieve a fully converged posterior, these chains are sufficient to illustrate the increased efficiency associated with HMC. 
+To illustrate this, in Fig.~\ref{fig:autocorrelation} we plot the autocorrelation as a function of lag for both the HMC and GRW samplers.
+The HMC autocorrelation is substantially lower than that of the GRW.
 We therefore expect gradient based samplers to converge significantly faster than typical samplers, especially in higher dimensions.
+In addition, we found that the effictive number of samples (a measure of the number of indepedent samples) is between 2 and 7 times larger for HMC across the different dimensions of the parameter space.
+
 In a follow up paper we will demonstrate that minute scale PE can be achieved by combining normalizing flows~\citep{2022arXiv221106397W, Gabrie:2021tlu}, GPU acceleration, and a derivative based sampler~\citep{PEpaper}.
+We therefore expect \jax waveforms to be crucial to future PE efforts in GW astronomy, especially for low-latency pipelines.
 

### 2023-01-05 src/scripts/autocorrelation.py

@@ -48,7 +48,7 @@ acl_HMC = az.autocorr(HMC_bestchain.T).T
 acl_gaussian = az.autocorr(gaussian_bestchain.T).T
 
 plt.plot(acl_HMC[:, 0], label="HMC")
-plt.plot(acl_gaussian[:, 0], label="Gaussian")
+plt.plot(acl_gaussian[:, 0], label="GRW")
 plt.xlim(0, 1000)
 plt.ylim(-0.1, 1.0)
 plt.xlabel("Lag")

### 2023-01-05 src/scripts/corner_plot.py

@@ -12,8 +12,8 @@ params = {
     "axes.titlesize": 18,
     # "axes.labelpad": 30,
     "axes.unicode_minus": False,
-    "xtick.labelsize": 10,
-    "ytick.labelsize": 10,
+    "xtick.labelsize": 12,
+    "ytick.labelsize": 12,
     "xtick.top": True,
     "ytick.right": True,
     "xtick.bottom": True,
@@ -75,15 +75,15 @@ axes = np.array(figure.axes).reshape((n_dim, n_dim))
 # Loop over the diagonal
 for i in range(n_dim):
     ax = axes[i, i]
-    ax.axvline(true_params[i], color="C1", alpha=0.5)
+    ax.axvline(true_params[i], color="C1", alpha=0.8)
 
 # Loop over the histograms
 for yi in range(n_dim):
     for xi in range(yi):
         ax = axes[yi, xi]
-        ax.axvline(true_params[xi], color="C1", alpha=0.5)
-        ax.axhline(true_params[yi], color="C1", alpha=0.5)
-        ax.plot(true_params[xi], true_params[yi], "s", color="C1", alpha=0.5)
+        ax.axvline(true_params[xi], color="C1", alpha=0.8)
+        ax.axhline(true_params[yi], color="C1", alpha=0.8)
+        ax.plot(true_params[xi], true_params[yi], "s", color="C1", alpha=0.8)
 
 figure.set_size_inches(10, 10)
 plt.savefig(paths.figures / "PE.pdf", bbox_inches="tight")

### 2023-01-05 src/scripts/mismatch_hist.py

@@ -2,6 +2,7 @@ import numpy as np
 import matplotlib.pyplot as plt
 import matplotlib as mpl
 import paths
+from scipy import stats
 
 params = {
     "font.size": 18,
@@ -38,14 +39,21 @@ mpl.rcParams.update(params)
 (catalog_num, ori_loss, opt_loss) = np.loadtxt(
     paths.data / "mismatch_hist.txt", unpack=True
 )
-bins = np.arange(-5, -1, 0.125)
+bins = np.arange(-5, -1, 0.1)
 
+# print(stats.mode(np.log10(ori_loss)[:]))
+# print(np.mean(np.log10(ori_loss)) / np.mean(np.log10(opt_loss)))
+# print(stats.mode(np.log10(ori_loss))[0] / stats.mode(np.log10(opt_loss))[0])
+
+plt.axvline(x=np.mean(np.log10(ori_loss)), color="C0", ls="--")
 plt.hist(np.log10(ori_loss), bins=bins, alpha=0.3, color="C0")
 plt.hist(np.log10(ori_loss), bins=bins, label="Original", histtype="step", color="C0")
+
+plt.axvline(x=np.mean(np.log10(opt_loss)), color="C1", ls="--")
 plt.hist(np.log10(opt_loss), bins=bins, alpha=0.3, color="C1")
 plt.hist(np.log10(opt_loss), bins=bins, label="Optimized", histtype="step", color="C1")
 plt.legend()
-plt.xlim(1e-5, 0.1)
+plt.xlim(-5, -1)
 plt.xlabel("$\log(\mathcal{M})$")
 plt.ylabel("Count")
 plt.savefig(paths.figures / "mismatch_hist.pdf", bbox_inches="tight")

### 2023-01-05 src/scripts/random_matches.py

@@ -84,7 +84,7 @@ sc = plt.scatter(
 )
 plt.colorbar(sc, label=r"$\log_{10}(1-\mathrm{Match})$")
 
-plt.scatter(Mtot[exactmatch_indices], chieff[exactmatch_indices], color="C0")
+# plt.scatter(Mtot[exactmatch_indices], chieff[exactmatch_indices], color="C0")
 # plt.xlabel(r"Total Mass, $M$")
 # plt.ylabel(r"Mass Ratio, $q$")
 plt.xlabel(r"Total Mass, $M$")

### 2023-01-06 src/tex/ms.tex
Finishing section 2

@@
 %     \item Finish intro \te{Done}
-%     \item Finish section 2 
+%     \item Finish section 2 \te{Done}
 %     \item Write captions for all plots \te{Done}
@@
 A variety of waveform families have been developed to accurately model the GW emission from COs. 
-When the COs are relatively well separated, the dynamics of the system can be well approximated with post-Newtonian expansion.
+When the COs are relatively well separated, the dynamics of the system can be well approximated by a post-Newtonian expansion.
 However, close to merger, numerical relatively simulations are required to accurately model the binary.
@@
 IMRPhenom waveforms are therefore ideally suited for AD, espeically using \jax. 
-In this paper we focus mainly on the aligned spin circular orbit model IMRPhenomD~\citep{Husa:2015iqa, Khan:2015jqa}.
+In this paper we focus on the aligned spin circular orbit model IMRPhenomD~\citep{Husa:2015iqa, Khan:2015jqa}.
 
@@
 \begin{equation}
+    \label{eq:inner_prod}
     \left(h_{\theta_1}|h_{\theta_2}\right) \equiv 4 \, \mathrm{Re} \int^{\infty}_{0} \mathrm{d} f \, \frac{ h_{\theta_1}(f) h^*_{\theta_2}(f)}{S_n(f)}\, ,
@@
 This is shown in Fig.~\ref{fig:match}, where we have calculated the $\mathrm{m}(h_{\theta_1}|\bar{h}_{\theta_1})$ across the entire parameter space.\footnote{
-    Specifically, we use $10^3$ points in the ranges $m_{1,2} = (1,100)\,M_{\odot}$ and $\chi_{1,2} = (-1,1)$.
+    Specifically, we use $10^4$ points in the ranges $m_{1,2} = (1,100)\,M_{\odot}$ and $\chi_{1,2} = (-1,1)$.
 In addition, we evaulated the waveforms on a frequency grid from $32\,$Hz to $1024\,$Hz with frequency spacing $\Delta f = 0.0125 \,$Hz.
@@
 Here, $h_{\theta_1}$ corresponds to the \ripple waveform implementation and $\bar{h}_{\theta_1}$ is the \lalsuite implementation. 
-From Fig.~\ref{fig:match}, it is clear that the \ripple waveform matches with the \lalsuite waveform to numerical precision across the majority of the parameter space.
-In particular, the blue points indicate parameter points where the $1-\mathrm{m}(h_{\theta_1}|h^{*}_{\theta_1})=0$ and would therefore not appear in the log scale colorbar.
+From Fig.~\ref{fig:match}, it is clear that the \ripple waveform matches with the \lalsuite waveform close to numerical precision across the entire parameter space.
+In particular, the reduced density of points at low masses is where the $1-\mathrm{m}(h_{\theta_1}|h^{*}_{\theta_1})=0$ causing points to not appear in the log scale colorbar.
+% In particular, the blue points indicate parameter points where the $1-\mathrm{m}(h_{\theta_1}|h^{*}_{\theta_1})=0$ and would therefore not appear in the log scale colorbar.
 \begin{figure}[t]
@@
             Match between the \ripple and \lalsuite implementations of the IMRPhenomD waveform as a function of total mass and effective spin. 
-            % The blue points indicate where the match is exactly 1.0 and would therefore not appear in the log scale. 
+            The lower density of points at low masses is caused by the match being exactly 1.0 and therefore not appearing in the log scale. 
             It is clear that the \ripple waveform matches the \lalsuite implementation many of orders of magnitude more than necessary for all data analysis tasks across the entire parameter space.
@@
 
-There remains some slight deviation between the two waveform implementations at low total mass. \te{Do I need to look more into this...? The discussion below looks odd with this plot}
-This is likely due to the fact that cubic interpolators, which are used within IMRPhenomD to calculate the ringdown and damping frequencies, are not currently supported in \jax.
-Instead, we initially use \texttt{scipy}'s cubic interpolator to create a fine grid of XXX values, which we then linearly interpolate during waveform generation.
+Note that in General Relativity (GR) the total mass of the system is simply an overall scaling of the waveforms frequency evolution and is therefore not a truly independent parameter.
+If we instead use a flat PSD in Eq.~\ref{eq:inner_prod} and scale the frequency grid to correctly, all dependence with $M$ seen in Fig.~\ref{fig:match} vanishes. 
+This figure instead illustrates a more realistic PSD where at low masses the waveform signal-to-noise is dominated by the inspiral but at high masses it is dominated by the merger.
+
+There remains some slight deviation between the two waveform implementations at high total mass. 
+This is partly due to the fact that cubic interpolators, which are used within IMRPhenomD to calculate the ringdown and damping frequencies, are not currently supported in \jax.
+Instead, we initially use \texttt{scipy}'s cubic interpolator to create a fine grid of $5\times10^5$ values, which we then linearly interpolate during waveform generation.
 Unfortunately, we cannot make the initial cubic interpolation arbitrarily fine as this would add additional computational overhead when loading the data during waveform evaluation.
-The remaining error in the waveform at high masses is therefore primarily down to the coarseness in this interpolation.
+% The remaining error in the waveform at high masses is therefore primarily down to the coarseness in this interpolation.
 Note however, that the differences are well below the accuracy requirements of the waveforms and will have no noticeable for realistic data analysis tasks.
@@
 }
-Since the parameter space of GWs in General Relativity (GR) has $\mathcal{O}(10)$ dimensons, the speed of derivative evaluation is less crucial than the stability.
+Since the parameter space of GWs in GR has $\mathcal{O}(10)$ dimensons, the speed of derivative evaluation is less crucial than the stability.
 However, this might change for waveforms in beyond GR models where many more parameters are added.
@@
 First, we demonstrated the speed and accuracy of our implementation of the aligned spin IMRPhenomD waveform.
-In particular, we showed that it matches the \lalsuite implementation to near machine precision and can be easily parallelized on a GPU, making waveform calls up to five orders of magnitude faster.
+In particular, we showed that it matches the \lalsuite implementation to near machine precision and can be easily parallelized on a GPU.
+Parallization on a GPU provides substantial speed increases; on a free google colab GPU we found that waveform evaluations are nearly 5 orders of magnitude faster than serial CPU evaluations.
 Second, we discussed three data analysis tasks which can all be substantially improved by utlizing derivative information of the waveform.

### 2023-01-06 src/tex/ms.tex
Finished first draft of discussion

@@
 
+In this paper, we have primarily focussed on the IMRPhenom family of waveforms as their closed form expression is perfectly suited for implementation in \jax.
+Two other waveform families are commonly used in GW data analysis: the effective-one-body (EOB) and numerical relativity surrogate (NRSurrogate). 
+Some progress has been made towards the construction of a differentiable NRsurrogate, but currently it seems difficult to implement EOB waveforms in \jax.
+In particular, the evolution of the Hamiltonian required to evaluate an EOB waveform is both inherently slow to differentiate and difficult to implement in \jax which requires fixed length arrays for its JIT compilation.
+Since EOB waveforms are often regarded as the state-of-the-art, more work is required to see if a fast, differentiable implementation is possible.
+
 Currently the biggest constraint to adopting differentiable waveforms is the need to re-write the most commonly used waveforms into \jax (or pure python).
@@
 This would ensure that the community can easily utilize differentiability and hardware acceleration in the future.
-
-In this paper, we have primarily focussed on the IMRPhenom family of waveforms as their closed form expression is perfectly suited for implementation in \jax.
-Two other waveform families are commonly used in GW data analysis: the effective-one-body (EOB) and numerical relativity surrogate (NRSurrogate). 
-Some progress has been made towards the construction of a differentiable NRsurrogate, currently it seems difficult to implement EOB waveforms in \jax.
-In particular, the evolution of the Hamiltonian required to evaluate an EOB waveform is both inherently slow to differentiate and difficult to implement in \jax which requires fixed length arrays for its JIT compilation.
-Since EOB waveforms are often regarded as the state-of-the-art, more work is required to see if a fast, differentiable implementation is possible.
-
-\te{End on a final good thing?}
 

### 2023-01-09 src/tex/ms.tex
Update ms.tex

@@
     Since their discovery in 2015, gravitaitonal waves (GWs) have developed into a unique observational probe used throughout fundamental and astro-phyics.
-    As more detections continue, data analysis taks are becoming increasingly computationally demanding.
+    As more detections continue, data analysis taks are becoming increasingly computationally demanding. \kw{I would cut everything before out.}
     Here, we propose the use of automatic differentiation through the programming framework \jax for accelerating a variety of these analsyis tasks.
@@
 In the most minimial example, this requires a Markov Chain Monte Carlo (MCMC) on an 11 dimensional parameter space for each event.
+\kw{This is not quite true.
+You can always marginalize over more parameters if you want, such as time of coalescence and phase.
+I would just mention the typical case of 15 parameters GR PE instead.}
 Beyond this simple scenario, more complex waveform models with many additional parameters may be used to test for deviations from General Relativity.
 Finally, using the results of PE, population synthesis models constrain the progenitors systems from which the black holes we see merging today began their journey~\citep{Wong:2022flg}.
-All of these tasks require significant compute.
+\kw{Need more reference and some rewording}
+All of these tasks require significant compute. \kw{This could be triggering to some people, since population analysis could be quite cheap}
 In this paper, we will argue that differentiable waveforms (and more generally differentiable pipelines) can play a significant role in alleaviating this computational demand.
@@
 Unfortunately, in the field of GW data analysis, analytic derivatives of the necessary quantities (such as the likelihood) have historically been difficult to obtain.
+\kw{This needs some rewording.
+For example, I think in IMRPhenomD they do have the analytical derivatives coded up, at least for a part of the waveforms.
+Instead of difficult to have the derivatives, I would rather mention it is clumpsy to get the derivatives, especially given the model complexity is growing.}
 Numerical derivatives also suffer from accuracy issues stemming from rounding or truncation errors.

### 2023-01-09 src/tex/ms.tex
Update ms.tex

@@
 \definecolor{pyGreen}{RGB}{44, 160, 44}
+\definecolor{Burnt}{RGB}{203,96,21}
 
@@
 \newcommand{\kw}[1]{{\color{rb4}[KW: #1 ]}}
+\newcommand{\AZ}[1]{{\color{Burnt}[AZ: #1]}}
+
+\definecolor{dfm}{HTML}{6C71C4}
+\newcommand{\dfm}[1]{{\color{dfm}[DFM: #1]}}
 
@@
 % Title
-\title{\ripple: Differentiable Waveforms for Gravitational Wave Data Analysis}
+\title{\ripple: Differentiable and Hardware-Accelerated Waveforms for Gravitational Wave Data Analysis}
 
@@
 % Author list
+\author{Thomas D.~P.~Edwards} \JHU \OKC \NORDITA
+\author{Kaze W.~K.~Wong} \CCA
+\author{Kelvin K.~H.~Lam} \CCA \CUHK
+
 \author{Adam Coogan} \UdeM \Mila
-\author{Thomas D.~P.~Edwards} \JHU \OKC \NORDITA
 \author{Daniel Foreman-Mackey} \CCA
 \author{Maximiliano Isi} \CCA
-\author{Kelvin K.~H.~Lam} \CCA \CUHK
-\author{Kaze W.~K.~Wong} \CCA
 \author{Aaron Zimmerman} \CGP
@@
 \begin{abstract}
-    Since their discovery in 2015, gravitaitonal waves (GWs) have developed into a unique observational probe used throughout fundamental and astro-phyics.
-    As more detections continue, data analysis taks are becoming increasingly computationally demanding. \kw{I would cut everything before out.}
-    Here, we propose the use of automatic differentiation through the programming framework \jax for accelerating a variety of these analsyis tasks.
-    Firstly, we demonstrate that highly accurate waveforms (IMRPhenomD) can be written in \jax and demonstrate that the serial evaluation speed of the waveform (and its derivative) is similar to the \lalsuite implementation in \CC.
-    Moreover, \jax allows for GPU accelerated waveform calls which can be over five orders of magnitude faster than serial evaluation on a CPU.
+    Since their discovery in 2015, gravitational waves (GWs) have developed into a unique observational probe used throughout fundamental and astro-phyics.
+    As more detections continue, data analysis tasks are becoming increasingly computationally demanding.
+    \kw{I would cut everything before out.}
+    Here, we propose the use of automatic differentiation through the programming framework \jax for accelerating a variety of these analysis tasks.
+    Firstly, we demonstrate that complete waveforms which cover the inspiral, merger, and ringdown of binary black holes (i.e.~IMRPhenomD) can be written in \jax and demonstrate that the serial evaluation speed of the waveform (and its derivative) is similar to the \lalsuite implementation in \CC.
+    Moreover, \jax allows for GPU-accelerated waveform calls which can be over five orders of magnitude faster than serial evaluation on a CPU.
     We then focus on three applications where efficient and differentiable waveforms are essential.
-    Firstly, we demonstrate how stochastic gradient descent can be used to optimize the $\sim 200$ coefficients that make up the waveform. 
-    In particular, we demonstrate that the typical \textit{match} with numerical relativiely waveforms can be improved by more than 10\% without any additional overhead.
-    Secondly, we show that Fisher forecasting calculations can be sped up by $\sim 100\times$ (on CPU) with no loss in accuracy.
-    This increased speed makes population forecasting substantially simpler, as discussed extensively in \citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.
-    Finally, we show that gradient based samplers like Hamiltonian Monte Carlo lead to significantly reduced autocorrelation values when compared to tradiaitional Monte Carlo methods.
-    Since differentiable waveforms appear to have substantial advantages for a variety of tasks throughtout GW science, we propose that waveform developers use \jax to build new waveforms moving forward.
+    Firstly, we demonstrate how stochastic gradient descent can be used to optimize the $\sim 200$ coefficients that are used to calibrate the waveform model. 
+    In particular, we demonstrate that the typical \textit{match} with numerical relatively waveforms can be improved by more than 10\% without any additional overhead.
+    Secondly, we show that Fisher forecasting calculations can be sped up by $\sim 100\times$ (on a CPU) with no loss in accuracy.
+    This increased speed makes population forecasting substantially simpler, as discussed extensively in \citet{Iacovelli:2022bbs, Iacovelli:2022mbg}.
+    Finally, we show that gradient-based samplers like Hamiltonian Monte Carlo lead to significantly reduced autocorrelation values when compared to traditional Monte Carlo methods.
+    Since differentiable waveforms appear to have substantial advantages for a variety of tasks throughout GW science, we propose that waveform developers use \jax to build new waveforms moving forward.
     Our waveform code, \ripple, can be found at \href{https://github.com/tedwards2412/ripple}{github.com/tedwards2412/ripple}, and will continue to be updated with new waveforms as they are formulated.
@@
 
-% \section{Todo}
-% \begin{itemize}
-%     \item Finish intro \te{Done}
-%     \item Finish section 2 \te{Done}
-%     \item Write captions for all plots \te{Done}
-%     \item Finish discussion
-% \end{itemize}
-
 \section{Introduction}
@@
 
-% Intro outline
-% - What are gravitational waves and what are data analysis tasks?
-% - What are derivatives, and how can they be used?
-% - What is automatic differentiation, and why has it become particularly useful now? JAX?
-% - Discussion of what waveforms are most ameanable to being differentiable
-% - In this paper
+The discovery of gravitational waves (GWs)~\citep{LIGOScientific:2016aoc} from inspiraling and merging compact objects (COs) has revolutionized our understanding of both fundamental physics and astronomy~\citep[e.g.][]{LIGOScientific:2021djp,LIGOScientific:2021sio,LIGOScientific:2021psn}.
+Although the data volumes from GW detectors are relatively small, analyzing the data is a computationally demanding task.
+In addition, this computational cost will substantially increase when next generation detectors come online~\citep{Maggiore:2019uih, Reitze:2019iox, Evans:2021gyd}.
 
-The discovery of gravitational waves (GWs) from inspiraling and mergering compact objects (COs) has revolutionized our understanding of both fundamental physics and astronomy~\citep{LIGOScientific:2021djp}.
-Although the data volumes from GW detectors are relatively small, analyzing the data is a computationally demanding task.
-In addition, this computational cost will substatially increase when next generation detectors come online.
+The complexity begins even before data taking, since GW searches using the matched-filtering technique~\citep{Owen:1998dk, Owen:1995tm} require the generation of large banks of template waveforms.
+% \sout{The complexity begins before data gathering where one is required to generate banks of template waveforms which will be used for a matched-filter search~\citep{Owen:1998dk, Owen:1995tm}. }
+Once potential candidates are found, parameter estimation (PE) is performed to extract the detailed source properties of each event~\citep{Christensen:2022bxb, 2020MNRAS.493.3132S, Ashton:2018jfp, Romero-Shaw:2020owr, Veitch:2014wba, Biwer:2018osg}.
+In the most minimal example of non-precessing binaries, this requires a Markov Chain Monte Carlo (MCMC) on an 11 dimensional parameter space for each event.
+\kw{This is not quite true. You can always marginalize over more parameters if you want, such as time of coalescence and phase. I would just mention the typical case of 15 parameters GR PE instead.}
+\AZ{Maybe modify; some other points are precessing PE requires 15 params, tidal prec 17 params, plus calibration marge and EoS inferences can incorporate even more}
+Beyond this simple scenario, more complex waveform models with additional parameters may be used to test for deviations from General Relativity~\citep{Krishnendu:2021fga, LIGOScientific:2020tif, LIGOScientific:2016lio, Yunes:2016jcc, Arun:2006yw, Agathos:2013upa}.
+% \AZ{Add PPE and maybe TIGER refs..?}
+Finally, using the results of PE, population synthesis models constrain the progenitors systems from which the black holes we see merging today began their journey~\citep{LIGOScientific:2020kqk, LIGOScientific:2021psn, Wong:2022flg}.
+\kw{Need more reference and some rewording}
+All of these tasks require significant computation.
+\kw{This could be triggering to some people, since population analysis could be quite cheap}
+In this paper, we will argue that differentiable waveforms (and more generally differentiable pipelines) can play a significant role in alleviating this computational demand.
 
-The complexity begins before data gathering where one is required to generate banks of template waveforms which will be used for a matched-filter search~\citep{Owen:1998dk, Owen:1995tm}. 
-Once potential candidates are found, parameter estimation (PE) is performed to extract the detailed source properties of each event.
-In the most minimial example, this requires a Markov Chain Monte Carlo (MCMC) on an 11 dimensional parameter space for each event.
-\kw{This is not quite true.
-You can always marginalize over more parameters if you want, such as time of coalescence and phase.
-I would just mention the typical case of 15 parameters GR PE instead.}
-Beyond this simple scenario, more complex waveform models with many additional parameters may be used to test for deviations from General Relativity.
-Finally, using the results of PE, population synthesis models constrain the progenitors systems from which the black holes we see merging today began their journey~\citep{Wong:2022flg}.
-\kw{Need more reference and some rewording}
-All of these tasks require significant compute. \kw{This could be triggering to some people, since population analysis could be quite cheap}
-In this paper, we will argue that differentiable waveforms (and more generally differentiable pipelines) can play a significant role in alleaviating this computational demand.
-
-Derivatives are ubiqitously useful throughout data analysis tasks.
-For instance, during PE, derivative information can be used to guide the sampling towards regions with higher likelihood values (e.g.
-in Hamiltonian Monte Carlo~\citep{2017arXiv170102434B} or Gradient Decsent).
+Derivatives are ubiquitously useful throughout data analysis tasks.
+For instance, during PE, gradients are used to improve performance of the samplers 
+% \dfm{This isn't correct. In HMC the gradients don't have the behavior. I'd just say something like ``gradients are used to improve performance''} 
+(e.g.~in Hamiltonian Monte Carlo~\citep{2017arXiv170102434B} or Gradient Descent~\citep{2016arXiv160904747R}).
 Gradients are particularly valuable for high dimensional spaces.
-Unfortunately, in the field of GW data analysis, analytic derivatives of the necessary quantities (such as the likelihood) have historically been difficult to obtain.
+Unfortunately, in the field of GW data analysis, analytic derivatives of the necessary quantities (such as the likelihood) have historically been difficult to obtain~\citep{Keppel:2013kia}.
 \kw{This needs some rewording.
@@
 Numerical derivatives also suffer from accuracy issues stemming from rounding or truncation errors.
-However, recent progress in automatic differentiation (AD) has shown promise in allowing general, fast derivative calcuations for gravitational waveforms~\citep{Coogan:2022qxs}.
+However, recent progress in automatic differentiation (AD) has shown promise in allowing general, fast derivative calculations for gravitational waveforms, with applications to constructing template banks~\citep{Coogan:2022qxs} and computing the Fisher information for forecasting~\citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.
 
-Automatic differentiaton is a family of methods used to compute machine precision derivatives with little computational ovearhead. 
+Automatic differentiation is a family of methods used to compute machine precision derivatives with little computational overhead. 
 AD's recent ascendance is primarily driven by its use in machine learning, particularly for derivative computations of neural networks which use gradient descent during training.
@@
 There are now a variety of AD implementations, most notably in deep learning frameworks such as \texttt{pytorch}~\citep{pytorch} and \texttt{tensorflow}~\citep{tensorflow2015-whitepaper}.
-More general frameworks exist in \texttt{julia}~\citep{zygote, forwarddiff}, although \texttt{julia}'s limited support in GW physics precludes its general use.
-Here we make use of \jax~\citep{jax2018github} due to its easy integration with \texttt{python} libraries and its just-in-time (JIT) compiler which allows the same code to run on a variety of hardware accelerators (i.e.
-GPUs).
+More general frameworks exist in \texttt{julia}~\citep{zygote, forwarddiff}, although \texttt{julia}'s limited use in GW analysis software precludes its general use.
+Here we make use of \jax~\citep{jax2018github} due to its easy integration with \texttt{python} libraries and its just-in-time (JIT) compiler which allows the same code to run on a variety of hardware accelerators (i.e.~GPUs).
 
+\AZ{Add refs here:}
 There are a variety of gravitational waveforms currently used in analysis pipelines.
-They are generally structured into different families, the most common of which are: the effective-one-body (EOB), the phenomenological inspiral-merger-ringerdown (IMRPhenom), and numerical relativity surrogate (NRsurrogate).
-Of these, the IMRPhenom family is most well suited for an AD implementation in \jax.
-In particular, their closed form expressions and fixed size frequency grids make a \jax implementation that complies with the constraints of JIT compilation simple.
-EOB waveforms on the other hand require dynamic frequency grids and are therefore tricky to implement in \jax. 
-In addition, EOB waveforms use hypergeometric functions which have limited support in \jax.
-NRsurrogate waveforms have no fundamental issue but require the user to download relatively large data files, making their use cubersome in practice.\te{Max/Kaze, please add to/change this if you think its incorrect}
+They are generally structured into different families, the most common of which are: the effective-one-body (EOB)~\citep{Damour:2008yg,Buonanno:2005xu, Buonanno:2000ef, Buonanno:1998gg, Damour:2000we}, the phenomenological inspiral-merger-ringerdown (IMRPhenom), and numerical relativity surrogate (NRsurrogate)~\citep{Varma:2018mmi,Varma:2019csw}.
+%Of these, the IMRPhenom family is most well suited for an AD implementation in \jax.
+Of these, the IMRPhenom family serves as a natural starting point for an AD implementation in \jax.
+Models like the non-precessing IMRPhenomD~\citep{Khan:2015jqa} model studied here and the precessing, higher-mode model IMRPhenomXPHM~\citep{Pratten:2020fqn, Pratten:2020ceb} are written in the frequency domain using closed-form expressions.
+This makes a
+% \sout{In particular, their closed form expressions and fixed size frequency grids make a }
+\jax implementation that complies with the constraints of JIT compilation simple.
+NRsurrogate models, which interpolate directly over waveforms produced by numerical relativity simulations, are in principle also straightfoward to implement in \jax.
+EOB waveforms on the other hand are produced by evolving the dynamics of an effective one body Hamiltonian, and are therefore more difficult to implement in \jax.
+For EOB waveforms, frequency-domain reduced-order models may be a convenient target~\citep[e.g.][]{Cotesta:2020qhw}.
+% to produce a waveform
+%require dynamic frequency grids and are therefore tricky to implement in \jax.
+%\AZ{What does this mean?
+I don't understand why EOB needs dynamic frequency grids.
+We should clarify, and we can also mention if \jax is well-suited to frequency-domain ROM implementations of EOB, and reference those.}
+%In addition, EOB waveforms use hypergeometric functions which have limited support in \jax.
+%\AZ{rephrase this in a more positive fashion:}
+%NRsurrogate waveforms have no fundamental issue but require the user to download relatively large data files, making their use cumbersome in practice.\te{Max/Kaze, please add to/change this if you think its incorrect}
 
 In this paper we argue that differentiable waveforms will be a vital component for the future of GW data analysis.
-In addition, we present \ripple, a small GW \texttt{python} package which can serve as a respository for differentiable implementations of some of the main waveforms currently used in LIGO and Virgo analyses.
+\AZ{I prefer to describe what is rather than what could be; can we say what elements exist, e.g.~
+we present \ripple... which includes \jax implementations of GW antenna patterns, the IMRPhenomD waveform, and...}
+In addition, we present \ripple, a small GW \texttt{python} package which can serve as a repository for differentiable implementations of some of the main waveforms currently used in LIGO and Virgo analyses.
 The remainder of this paper is structured as follows. 
-In Sec.~\ref{sec:waveforms} we discuss the differentiable IMRPhenomD waveform implemented in \ripple and perform some benchmarks to demonstrate its speed and accruacy. 
+In Sec.~\ref{sec:waveforms} we discuss the differentiable IMRPhenomD waveform implemented in \ripple and perform some benchmarks to demonstrate its speed and accuracy. 
 In Sec.~\ref{sec:applications} we discuss three distinct applications using differentiable waveforms. 
 Firstly, we illustrate how the fit coefficients that form part of the IMRPhenom waveform models could be improved by high dimensional fitting enabled by a differentiable waveform. 
-Secondly, we implement differnetiable detector response functions and show that the speed of Fisher matrix calculations can be substantially improved using AD.
+Secondly, we implement differentiable detector response functions and show that the speed of Fisher matrix calculations can be substantially improved using AD.
 Finally, we run an illustrative injection example using Hamiltonian Monte Carlo to demonstrate that the autocorrelation of derivative samplers is substantially smaller than that of traditional MCMCs.
@@
 
-A variety of waveform families have been developed to accurately model the GW emission from COs. 
+A variety of waveform families have been developed to accurately model the GW emission from COs~\citep{Schmidtreview}. 
 When the COs are relatively well separated, the dynamics of the system can be well approximated by a post-Newtonian expansion.
@@
 
-The implementation of IMRPhenomD in \lalsuite is in C, and therefore needs to be rewritten natively into \texttt{pyhton} to be compatible with \jax.
-We have re-written IMRPhenomD from scratch using a combination of pure \texttt{pyhton} and \jax derivatives.
-In addition, we have restructed the code for readability and evaluation speed as well as exposing the internal fitting coefficients to the user (which we will use later in Sec.~\ref{sec:applications}).
+The implementation of IMRPhenomD in \lalsuite is in C, and therefore needs to be rewritten natively into \texttt{python} to be compatible with \jax.
+We have re-written IMRPhenomD from scratch using a combination of pure \texttt{python} and \jax derivatives.
+In addition, we have restructured the code for readability and evaluation speed as well as exposing the internal fitting coefficients to the user (which we will use later in Sec.~\ref{sec:applications}).
 
-To demonstrate our implempation of IMRPhenomD is faithful to the \lalsuite implementation, we start by defining the noise weighted inner product:
+To demonstrate our implementation of IMRPhenomD is faithful to the \lalsuite implementation, we start by defining the noise weighted inner product:
 \begin{equation}
     \label{eq:inner_prod}
-    \left(h_{\theta_1}|h_{\theta_2}\right) \equiv 4 \, \mathrm{Re} \int^{\infty}_{0} \mathrm{d} f \, \frac{ h_{\theta_1}(f) h^*_{\theta_2}(f)}{S_n(f)}\, ,
+    \left(h_1|h_2\right) \equiv 4 \, \mathrm{Re} \int^{\infty}_{0} \mathrm{d} f \, \frac{ h^*_1(f) h_2(f)}{S_n(f)}\, ,
 \end{equation}
-where $S_n$ is the (one-sided) noise power spectral density (PSD) and $h_{\theta_1}$ is the frequency domain waveform evaluated at the instrinsic parameter point $\theta_1$.
+where $S_n$ is the (one-sided) noise power spectral density (PSD) and \AZ{$h_1$ and $h_2$ are the frequency domain waveforms which are to be compared.}
+\sout{ evaluated at the intrinsic parameter point $\theta_1$.}
 We can then normalize the inner product through
 \begin{equation}
-    \left[h_{\theta_1}|h_{\theta_2}\right] = \frac{\left(h_{\theta_1}|h_{\theta_2}\right)}{\sqrt{\left(h_{\theta_1}|h_{\theta_1}\right)\left(h_{\theta_2}|h_{\theta_2}\right)}}\, .
+    \left[h_1|h_2\right] = \frac{\left(h_1|h_2\right)}{\sqrt{\left(h_1|h_1\right)\left(h_2|h_2\right)}}\, .
 \end{equation}
@@
 \begin{equation}
-    \mathrm{m}(h_{\theta_1}|h_{\theta_2}) \equiv \max_{t_c,\, \phi_c} \left[h_{\theta_1} \, \middle| \, h_{\theta_2} \right]\,
+    \mathrm{m}(h_1|h_2) \equiv \max_{\Delta t_c,\, \Delta \phi_c} \left[h_1 \, \middle| \, h_2 \right]\,
 \end{equation}
-where $t_c$ and $\phi_c$ are, respectively, the time and phase of coalesence.
+where \AZ{$\Delta t_c$ and $\Delta \phi_c$ are, respectively, the differences in time and phase of coalescence between the two waveforms.}
 
 Since the match is a measure of the difference between two waveforms, we can use it to demonstrate that the implementation of IMRPhenomD in \ripple accurately matches the \lalsuite implementation.
-This is shown in Fig.~\ref{fig:match}, where we have calculated the $\mathrm{m}(h_{\theta_1}|\bar{h}_{\theta_1})$ across the entire parameter space.\footnote{
-    Specifically, we use $10^4$ points in the ranges $m_{1,2} = (1,100)\,M_{\odot}$ and $\chi_{1,2} = (-1,1)$.
+\AZ{Say what PSD was used and reference it.}
+This is shown in Fig.~\ref{fig:match}, where we have calculated the $\mathrm{m}(h_1|h_2)$ across the entire parameter space.\footnote{
+    Specifically, we use $10^4$ points \AZ{varying component masses $m_1$, $m_2$ and spin parameters $\chi_1$, $\chi_2$} in the ranges $m_{1,2} = (1,100)\,M_{\odot}$ and $\chi_{1,2} = (-1,1)$.
 In addition, we evaulated the waveforms on a frequency grid from $32\,$Hz to $1024\,$Hz with frequency spacing $\Delta f = 0.0125 \,$Hz.
 }
-Here, $h_{\theta_1}$ corresponds to the \ripple waveform implementation and $\bar{h}_{\theta_1}$ is the \lalsuite implementation. 
+\AZ{Here, $h_1$ corresponds to the \ripple waveform implementation and $h_2$ is the \lalsuite implementation, evaluated at the same point in parameter space.} 
 From Fig.~\ref{fig:match}, it is clear that the \ripple waveform matches with the \lalsuite waveform close to numerical precision across the entire parameter space.
-In particular, the reduced density of points at low masses is where the $1-\mathrm{m}(h_{\theta_1}|h^{*}_{\theta_1})=0$ causing points to not appear in the log scale colorbar.
+\AZ{strictly $=0$? That seems impossible. Fmla has wrong labels for $h$'s}
+In particular, the gray points at low masses are where $1-\mathrm{m}(h_1|h2)=0$ causing points to not appear in the log scale colorbar.
 % In particular, the blue points indicate parameter points where the $1-\mathrm{m}(h_{\theta_1}|h^{*}_{\theta_1})=0$ and would therefore not appear in the log scale colorbar.
@@
             Match between the \ripple and \lalsuite implementations of the IMRPhenomD waveform as a function of total mass and effective spin. 
-            The lower density of points at low masses is caused by the match being exactly 1.0 and therefore not appearing in the log scale. 
+            % The lower density of points at low masses is caused by the match being exactly 1.0 and therefore not appearing in the log scale. 
+            The gray points are where the match is exactly 1.0 and therefore would not appear in the log scale.
             It is clear that the \ripple waveform matches the \lalsuite implementation many of orders of magnitude more than necessary for all data analysis tasks across the entire parameter space.
+            \dfm{I think this figure could use some work.
+Maybe the points could be smaller to make them easier to resolve.
+And it could be interesting to clip the mismatch to something like $10^{-16}$ so that we don't lose the points with zero mismatch.
+Right now it looks less impressive!}
         }
@@
 
-Note that in General Relativity (GR) the total mass of the system is simply an overall scaling of the waveforms frequency evolution and is therefore not a truly independent parameter.
-If we instead use a flat PSD in Eq.~\ref{eq:inner_prod} and scale the frequency grid to correctly, all dependence with $M$ seen in Fig.~\ref{fig:match} vanishes. 
+Note that in General Relativity (GR) the total mass $M = m_1 + m_2$ of \AZ{a binary black hole simply serves as an overall scale for the system, so that the frequency evolution can be trivially rescaled.
+The total mass only impacts the match because the chosen PSD and frequency limits fix a reference scale.}
+\sout{the system is simply an overall scaling of the waveforms} 
+\sout{frequency evolution and is therefore not a} \sout{truly independent parameter.}
+If we instead use a flat PSD in Eq.~\ref{eq:inner_prod} and \AZ{rescale the frequency grid to units of $Mf$}, \sout{correctly,} all dependence with $M$ seen in Fig.~\ref{fig:match} vanishes. 
+\AZ{Revisit this after defining PSD above:} 
 This figure instead illustrates a more realistic PSD where at low masses the waveform signal-to-noise is dominated by the inspiral but at high masses it is dominated by the merger.
@@
 This is partly due to the fact that cubic interpolators, which are used within IMRPhenomD to calculate the ringdown and damping frequencies, are not currently supported in \jax.
+\dfm{What dimension of interpolation do you need?
+Maybe this is something we could implement...}
 Instead, we initially use \texttt{scipy}'s cubic interpolator to create a fine grid of $5\times10^5$ values, which we then linearly interpolate during waveform generation.
@@
 Since the \lalsuite implementation of IMRPhenomD is written in \texttt{C}, the evaluation of the waveform itself is faster than \ripple.
+\dfm{I'd soften this statement a little bit.
+In my own work, I often find that JAX implementations are just as fast or faster than C implementations, so I wouldn't state it that it's obvious that lalsuite should be faster!
+Where are these benchmarks implemented?
+Would it be easy for me to try them out?}
 Benchmarking on a MacBook Pro with an M1 Max Apple Silicon processor, we find that a single waveform evaluation takes $\sim 2-5\,\mathrm{ms}$ for \lalsuite (interfaced with python) and $\sim 5-8\,\mathrm{ms}$ for \ripple.\footnote{
@@
 Waveform calls can therefore be evaluated in parallel, massively improving performance.
-Performing the same benchmark as above on a google colab notebook with the freely provided GPU, we find that on average waveform evaluations take $\sim 0.00005\mathrm{ms}$, nearly five orders of magnitude faster.
+Performing the same benchmark as above on a Google Colab notebook with the freely provided GPU, we find that on average waveform evaluations take $\sim \sout{0.00005} 5 \times 10^{-5}\mathrm{ms}$, nearly five orders of magnitude faster.
 Generalizing \lalsuite waveforms to run on a GPU would be a significant undertaking.
 
-The primary arguement of this paper is that waveform derivatives will also be highly vauable to data analysis tasks. 
+The primary argument of this paper is that waveform derivatives will also be highly valuable to data analysis tasks. 
 AD provides two big advantages when it comes to evaluating derivatives compared to numerical differentiation.
-First, the accuracy of derivatives from AD are signficantly more stable than finite difference methods.
+First, the accuracy of derivatives from AD are significantly more stable than finite difference methods.
 In particular, finite differences suffer from both rounding and truncation errors, meaning that the user is required to \textit{tune} the width over which the difference is taken.
-On the other hand, AD produces machine precision derivatives with no tuninig.
-Second, AD scales favourably with the dimensionality of the function.
-In particular, for every input dimension, $D$, added one would need to evaluate the function at least $2D$ times to calcuate finite difference derivatives for all input parameters.
+On the other hand, AD produces machine precision derivatives with no tuning.
+Second, AD scales favorably with the dimensionality of the function.
+In particular, for every input dimension, $D$, added one would need to evaluate the function at least $2D$ times to calculate finite difference derivatives for all input parameters.
 For reverse mode AD, one only needs two function calls to evaluate the derivative of all input parameters, regardless of dimenison.\footnote{
@@
 }
-Since the parameter space of GWs in GR has $\mathcal{O}(10)$ dimensons, the speed of derivative evaluation is less crucial than the stability.
-However, this might change for waveforms in beyond GR models where many more parameters are added.
+Since the parameter space of GWs in GR has $\mathcal{O}(10)$ dimensions, the speed of derivative evaluation is less crucial than the stability.
+However, this might change for waveforms in beyond-GR models, \AZ{waveforms involving equation of state parameters for neutron stars, and models which account for calibration and waveform uncertainties.
+In these cases} many more parameters \AZ{can be} added.
 
@@
 
-Here, we will illustrate how three core tasks in GW science can be substantially improved through the use of differentiable waveforms.
-In the paper we will primarily look at toy examples, leaving a more careful analyses to future work. 
-The three tasks discussed here cover a wide range of GW science, starting with waveform development all the way to forecasting and parameter estimation.
+Here, we illustrate how three core tasks in GW science can be substantially improved through the use of differentiable waveforms.
+In the paper we primarily look at toy examples, leaving a more careful analyses to future work. 
+The three tasks discussed here cover a wide range of GW science, starting with waveform development all the way to Fisher forecasting and PE.
 
@@
 
+\AZ{Broad comments: need to cite which catalog the waveforms came from if any.
+The number of waveforms used seems much larger than the number originally used
+to calibrate PhenomD. 
+If that is true, we can't make a big deal about having a better mismatch compared to
+the original version, because of course optimizing over another set will improve 
+performance.
+Say how many waveforms we use to optimize, how many we compare to, where they came
+from.}
+
 Having an accurate waveform model is essential for many data analysis tasks in
 GW, such as searching for signals and estimating source parameters given data.
-While waveforms generated using numerical relativity simulations are commonly
-regarded as the most accurate, they are too computationally expensive to be used
+While waveforms generated using numerical relativity simulations are \sout{commonly
+regarded as the most accurate,} 
+\AZ{in principle the highest fidelity signal model,}
+they are too computationally expensive to be used
 in any practical data analysis tasks. To create waveform models that can be used
@@
 full numerical relativity waveforms, which are essentially simpler ansatzes that
-are only approximations but can be evaluated much faster.
+\sout{are only approximations but} can be evaluated much faster \AZ{and in regions not
+covered by numerical simulations, such as extremes of parameter space or earlier 
+phases of the binary inspiral}.
 
-Every waveform approximant has some fitting coefficients that are tuned during
-the process of calibrating the ansatz to NR waveforms. In the case of
+%\sout{Every waveform approximant has some}
+\AZ{Waveform approximants generally have free coefficients which are calibrated
+to NR waveforms to achieve high accuracy.}
+\sout{ fitting coefficients that are tuned during
+the process of calibrating the ansatz to NR waveforms.} In the case of
 IMRPhenomD, there are 209 fitting coefficients used to capture the separate
 behavior of the amplitude and phase as a function of the mass ratio and spins.
-The accuracy of fitting coefficients determines how well the approximant
-reproduces the NR waveform. The inaccuracy in obtaining the fitting coefficients
+\sout{The accuracy of fitting coefficients determines how well the approximant
+reproduces the NR waveform. 
+The}
+\AZ{Any} inaccuracy in obtaining the fitting coefficients
 means misrepresentation of the NR waveform, which can translate to systematic
@@
 coefficients are fitted in segments with amplitude and phase separated.
-Furthermore, IMRPhenomD is divided in three fitting segments, inpsiral, merger and ringdown.
+Furthermore, IMRPhenomD is divided in three fitting segments, inspiral, merger and ringdown.
 Each of these segments has their own set of fitting coefficients.
 After obtaining the fitting coefficient for individual segments, they are then
-"stitched" together such that they are continuous in the first derivative.
+``stitched'' together such that they are continuous in the first derivative.
 The process of stitching introduces some addition
@@
 fitting all coefficients, the correlation in fitting individual and connecting
-different segments can be accounted during the optimization. Therefore, there is
+different segments can be accounted for during the optimization. Therefore, there is
 potential improvement of the accuracy by fitting all coefficients at once.
+
+\AZ{Maybe comment on optimization method used for EOB waveforms?
+I think they do interesting Monte Carlo etc. Check what method is used 
+for IMRPhenomXPHM?}
 
@@
     \label{eq:mismatch}
-	\mathcal{M}(\lambda)=1-\mathrm{m}(h_{\mathrm{IMR}}(\lambda)|h_{\mathrm{NR}(\lambda)}),
+	\mathcal{M}(\lambda)=1-\mathrm{m}(h_{\mathrm{IMR}}(\lambda)|h_{\mathrm{NR}}(\lambda)),
 \end{equation}
@@
 We terminate the program after 30000 steps, in which $\lambda$ has converged to the optimum.
+\dfm{We can't possibly know that it has converged to ``the optimum''.
+I'd sat just leave it at ``We terminate the program after 30000 steps.''}
 % Add description of the optimization procedure.
@@
 ratio and spins. From Eq. \ref{eq:multi_loss}, we can see the averaging between
-waveforms with different intrinsic parameters implies there are trade-off in
+waveforms with different intrinsic parameters implies there are trade-offs in
 performance in different regions of the intrinsic parameter space.
@@
 
-In Fig. \ref{fig:mismatch_hist}, we show the distribution of log mismatches.
+In Fig.~\ref{fig:mismatch_hist}, we show the distribution of log mismatches.
 One can see the distribution of mismatch after optimization has smaller mismatch compared to before optimization,
@@
 % However, it is not improving for every waveforms. Say why briefly, and say there is another paper coming out.
-While many waveforms improved their accuracy, some remain to have similar mismatches.
+While many waveforms improved their accuracy, some retain similar mismatches. 
+\AZ{I don't understand this claim. The spin degeneracy is present in the physics of the waveform; $\chi_{\rm eff}$ is just one way of parametrizing the spins that tries to reduce the degeneracy.}
 This phenomenon is mainly due to the reduced spin approximation used in IMRPhenomD, where spin degeneracy is introduced, hence restricting further improvements. 
 In an upcoming paper, we will give a detailed discussion of how such an approximation scheme affects the optimization procedure. 
-Nevertheless, the peak of the original waveform distribution moved to the left end of the distribution, indicating that the optimization procedure improved the model. 
-Therefore, the new set of waveform coefficients can produce better waveforms for GW analyses.
+Nevertheless, the peak of the original waveform distribution has moved to the left end of the distribution, indicating that \sout{the} \AZ{our AD-assisted} optimization procedure \AZ{provides an improved implementation of the model}.
+\sout{Therefore, the new set of waveform coefficients can produce better waveforms for GW analyses. }
 
@@
         Automatic differentiation (AD) was used to compute the derivatives in Eq.~\ref{eq:fisher}.
-        As empahsized in the text, after JIT compilation, each error calculation was around 100 times faster than a similar forecasting code \texttt{GWbench}~\citep{Borhanian:2020ypi}.
+        As emphasized in the text, after JIT compilation, each error calculation was around 100 times faster than a similar forecasting code \texttt{GWbench}~\citep{Borhanian:2020ypi}.
     }
@@
 The Fisher matrix approach is built around the assumption of a Gaussian likelihood.
-Although in practice this assumption is often violated in real GW datasets, the results obtained using a Fisher analysis can provide quick and useful diagnotistics in evaluating sensitivities for a variety of models and detector configurations.
+Although in practice this assumption is often violated in real GW datasets, the results obtained using a Fisher analysis can provide quick and useful diagnostics in evaluating sensitivities for a variety of models and detector configurations.
 
@@
 Forecasting with Fisher matrices for third generation detectors has already been extensively explored in \citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.
-Here we purely want to illustrate the simplicitly and speed of AD for forecasting rather than providing new physics insights.
+Here we purely want to illustrate the simplicity and speed of AD for forecasting rather than providing new physics insights.
 We therefore consider a simple, three detector setup corresponding to the two LIGO detectors in addition to Virgo.
@@
     \label{eq:fisher}
-    \mathcal{I}^{k}_{ij} = (\partial_i h_0 | \partial_j h_0) \, ,
+    \mathcal{I}^{k}_{ij} = (\partial_i h^k | \partial_j h^k) \, ,
 \end{equation}
-where $k$ indicates the detector, $\partial_i = \partial/\partial \theta_i$, and $h_0$ is the strain measured by the detector which is given by,
+where $k$ indicates the detector, $\partial_i = \partial/\partial \theta_i$, and $h$ is the strain measured by the detector which is given by,
 \begin{equation}
-    h_0(\theta) = F_+(\lambda) h_{+}(\phi) + F_\times(\lambda) h_{\times}(\phi) \, .
+    h^k(\theta) = F_+(\phi) h^k_{+}(\Xi) + F_\times(\phi) h^k_{\times}(\Xi) \, .
 \end{equation}
-Note that here we have separated out the intrinsic ($\phi$) and extrinsic ($\lambda$) variables as well as introducing the detector projection functions for the plus and cross polarizations as $F_+$ and $F_\times$ respectively.
+Note that here we have separated out the extrinsic ($\phi$) and intrinsic ($\Xi$) variables as well as introducing the detector projection functions for the plus and cross polarizations as $F^k_+$ and $F^k_\times$ respectively.
 Since we are considering a three detector setup we simply add the Fisher matrices from the individual detectors to get the combined Fisher matrix:
@@
 \end{equation}
-Finally, we invert the Fisher matrix to calculate the covariance matrix.
+Finally, we invert the Fisher matrix to calculate the covariance matrix,
+\AZ{which proves forecasted measurement errors and parameter covariances for a signal with parameters $\theta$ observed by the given detector network}.
+\AZ{Again we need to state and cite the PSDs we use for the three detectors.}
 
@@
     \caption{Priors for the 11 dimensional parameter space used for the Fisher forecasting population analysis in Sec.~\ref{subsec:fisher}. 
-    $\boldsymbol{U}$ indicates a uniform distribution between the two variables in the brackets.}
+    $\boldsymbol{U}$ indicates a uniform distribution between the two variables in the brackets.
+    \AZ{$\cos \iota \in (-1,1)$, $\psi \in (0,2 \pi)$ (or $\pi$), RA is $\alpha \in (0, 2\pi)$, Dec is $\delta$, let $\vartheta = \pi/2 - \delta$, then $\cos (\vartheta) \in (-1,1)$ }
+    }
     \label{tab:priors}
@@
 
-To illustrate the computational speed of computing Fisher matrices with AD, we consider a population of binaries and compute the sky localtization error following Eq.~(28) in \citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.
+To illustrate the computational speed of computing Fisher matrices with AD, we consider a population of binaries and compute the sky localization error following Eq.~(28) in \cite{Iacovelli:2022bbs, Iacovelli:2022mbg}.
 Since the Fisher matrix approach is known to have both theoretical issues as well as numerical instabilities for low signal-to-noise events, we restrict our population to only nearby systems.
@@
 The distribution of sky localization errors from a population of $10^3$ binaries can be seen in Fig.~\ref{fig:sky_localization}.
-We have verified that our errors agree with a separate dedicated Fisher forecasting code~\citep{Borhanian:2020ypi} to within X percent.
-This demonstrates that AD can be used to accurately produce population level forecasts.
+We have verified that our errors agree with a separate dedicated Fisher forecasting code~\citep{Borhanian:2020ypi} to within {\color{red}{X}} percent.
+This demonstrates that AD can be used to accurately produce population-level forecasts.
 
-Moreover, each error calculation (including computing the Fisher matrices for each detector and the inversion procress) is substantially faster.
+Moreover, each error calculation (including computing the Fisher matrices for each detector and the inversion process) is substantially faster.
 In particular, we find that after compilation each error calculation takes approximately half a second on a single computing core.
 \texttt{GWbench}~\citep{Borhanian:2020ypi}, on the other hand, takes $\mathcal{O}$(minutes) for each Fisher calculation using the same detector setup and frequency grid.
-This factor of over 100 speed up is substantial considering the fact that a single core evaluation of the \ripple waveform is slower than the \lalsuite call.
-As discussed above, performance can be further improved by utilising hardware acceleration such as parallel GPU processing.
+This factor of over 100 speed up is substantial considering the fact that a single core evaluation of the \ripple waveform is slower than the \lalsuite call \AZ{used by \texttt{GWbench}}.
+As discussed above, performance can be further improved by utilizing hardware acceleration such as parallel GPU processing.
 AD therefore represents a fast and accurate way of performing population level analyses, and should be utilized for testing the capabilities of next generation detectors.
@@
 
-After the search algorithms have constructed a list of confidently detected binaries, the next step is to sample from the posterior of each sources parameters - so called parameter estimation (PE).
+After the search algorithms have constructed a list of confidently detected binaries, the next step is to sample from the posterior of each sources parameters - so called PE.
 To do this, one typically uses a Markov Chain Monte Carlo (MCMC) or nested sampler~\citep{multinest, dynesty}.
@@
 For example, sampling the 15 dimensional parameter space for a BBH system can take $\mathcal{O}(10)$ hours, while BNS systems can take up to weeks.
-Dedicated fast samplers have been designed to get approximate posteriors on the sky localization to faciliate follow-up electromagnetic observations.
+Dedicated fast samplers have been designed to get approximate posteriors on the sky localization to facilitate follow-up electromagnetic observations \citep[e.g.~\texttt{BAYESTAR}][]{Singer:2015ema}.
 Nevertheless, these do not present the whole picture; fast, general parameter estimation therefore remains a primary goal of GW data analysis.
 
-A primary issue with both MCMC and nested sampling is that neither utlizes information about the likelihood's derivative and must therefore randomly walk towards areas of highest likelihood.
+A primary issue with both MCMC and nested sampling is that neither utilizes information about the likelihood's derivative and must therefore randomly walk towards areas of highest likelihood.
 Derivative based samplers, on the other hand, have been shown to extrapolate well to higher dimensions although they sometimes come with their own drawbacks.
@@
 In particular, we will show that the autocorrelation of a Hamiltonian Monte Carlo (HMC) sampler is significantly lower than a traditional MCMC algorithm.
+\AZ{Add citations to other HMC papers in GW land}
 

### 2023-01-09 src/tex/ms.tex
Update ms.tex

--- 2023-01-09 src/tex/ms.tex
@@
 Performing the same benchmark as above on a google colab notebook with the freely provided GPU, we find that on average waveform evaluations take $\sim 0.00005\mathrm{ms}$, nearly five orders of magnitude faster.
+\kw{We should add some more detail related to this comparison.
+We should be clear that each individual waveform sample evaluation doesn't get us the performance by putting it on GPU (Jitting helps with that, not the GPU part).
+The GPU comes into play by allowing parallel evaluation of waveform sample, so for a tiny amount of time/frequency samples, we should expect GPU to lose to a CPU, but on any reasonable amount of time/frequency samples, GPU win a lot.}
 Generalizing \lalsuite waveforms to run on a GPU would be a significant undertaking.
@@
 }
+\kw{Maybe a bit more background on AD, the way we phrase it here sounds a bit too magical too me now.}
 Since the parameter space of GWs in GR has $\mathcal{O}(10)$ dimensons, the speed of derivative evaluation is less crucial than the stability.
@@
 
-Forecasting the sensitivity of future experiments is an essential task in GW science.
+Forecasting the sensitivity of future experiments is an essential task in GW science. \kw{LOL I won't say it is essential.}
 Due to its theoretical simplicity and evaluation speed, the Fisher matrix formalism is commonly deployed to estimate how well a binary system's parameters can be measured.
@@
 This factor of over 100 speed up is substantial considering the fact that a single core evaluation of the \ripple waveform is slower than the \lalsuite call.
+\kw{Want to pull a number on the total time it took on generating the population result? I imagine this should turn the theorists on and send them into a fishing frenzy.}
 As discussed above, performance can be further improved by utilising hardware acceleration such as parallel GPU processing.
@@
 We therefore expect \jax waveforms to be crucial to future PE efforts in GW astronomy, especially for low-latency pipelines.
+\kw{Not just low latency, but also high dimensional analysis, such as testing GR and lensing.}
 

### 2023-01-17 src/scripts/random_matches.py
Trying to update plot

@@ -90,7 +90,7 @@ plt.colorbar(sc, label=r"$\log_{10}(1-\mathrm{Match})$")
 plt.xlabel(r"Total Mass, $M$")
 plt.ylabel(r"Effective Spin, $\chi_{\rm eff}$")
 plt.xlim(0, 200)
-plt.ylim(-1, 1.0)
+plt.ylim(-1, 1.)
 # plt.ylim(1, 50)
 
 plt.savefig(paths.figures / "random_matches.pdf", bbox_inches="tight")

### 2023-01-18 src/tex/ms.tex
Adding citations to intro

--- 2023-01-09 src/tex/ms.tex
@@
     Since their discovery in 2015, gravitaitonal waves (GWs) have developed into a unique observational probe used throughout fundamental and astro-phyics.
-    As more detections continue, data analysis taks are becoming increasingly computationally demanding. \kw{I would cut everything before out.}
+    As more detections continue, data analysis taks are becoming increasingly computationally demanding.
     Here, we propose the use of automatic differentiation through the programming framework \jax for accelerating a variety of these analsyis tasks.
@@
 
-% Intro outline
-% - What are gravitational waves and what are data analysis tasks?
-% - What are derivatives, and how can they be used?
-% - What is automatic differentiation, and why has it become particularly useful now? JAX?
-% - Discussion of what waveforms are most ameanable to being differentiable
-% - In this paper
-
 The discovery of gravitational waves (GWs) from inspiraling and mergering compact objects (COs) has revolutionized our understanding of both fundamental physics and astronomy~\citep{LIGOScientific:2021djp}.
 Although the data volumes from GW detectors are relatively small, analyzing the data is a computationally demanding task.
-In addition, this computational cost will substatially increase when next generation detectors come online.
+In addition, this computational cost will substatially increase when next generation detectors come online~\citep{Maggiore:2019uih, Reitze:2019iox}.
 
 The complexity begins before data gathering where one is required to generate banks of template waveforms which will be used for a matched-filter search~\citep{Owen:1998dk, Owen:1995tm}. 
-Once potential candidates are found, parameter estimation (PE) is performed to extract the detailed source properties of each event.
+Once potential candidates are found, parameter estimation (PE) is performed to extract the detailed source properties of each event~\citep{Christensen:2022bxb, 2020MNRAS.493.3132S, Ashton:2018jfp, Romero-Shaw:2020owr, Veitch:2014wba, Biwer:2018osg}.
 In the most minimial example, this requires a Markov Chain Monte Carlo (MCMC) on an 11 dimensional parameter space for each event.
-\kw{This is not quite true.
-You can always marginalize over more parameters if you want, such as time of coalescence and phase.
-I would just mention the typical case of 15 parameters GR PE instead.}
-Beyond this simple scenario, more complex waveform models with many additional parameters may be used to test for deviations from General Relativity.
-Finally, using the results of PE, population synthesis models constrain the progenitors systems from which the black holes we see merging today began their journey~\citep{Wong:2022flg}.
-\kw{Need more reference and some rewording}
-All of these tasks require significant compute. \kw{This could be triggering to some people, since population analysis could be quite cheap}
+Beyond this simple scenario, more complex waveform models with many additional parameters may be used to test for deviations from General Relativity~\citep{Krishnendu:2021fga, LIGOScientific:2020tif, LIGOScientific:2016lio}.
+Finally, using the results of PE, population synthesis models constrain the progenitors systems from which the black holes we see merging today began their journey~\citep{LIGOScientific:2020kqk, Wong:2022flg}.
+All of these tasks require significant compute.
 In this paper, we will argue that differentiable waveforms (and more generally differentiable pipelines) can play a significant role in alleaviating this computational demand.
@@
 Unfortunately, in the field of GW data analysis, analytic derivatives of the necessary quantities (such as the likelihood) have historically been difficult to obtain.
-\kw{This needs some rewording.
-For example, I think in IMRPhenomD they do have the analytical derivatives coded up, at least for a part of the waveforms.
-Instead of difficult to have the derivatives, I would rather mention it is clumpsy to get the derivatives, especially given the model complexity is growing.}
 Numerical derivatives also suffer from accuracy issues stemming from rounding or truncation errors.

### 2023-01-18 src/tex/ms.tex
Updating fig 1

@@
 From Fig.~\ref{fig:match}, it is clear that the \ripple waveform matches with the \lalsuite waveform close to numerical precision across the entire parameter space.
-In particular, the reduced density of points at low masses is where the $1-\mathrm{m}(h_{\theta_1}|h^{*}_{\theta_1})=0$ causing points to not appear in the log scale colorbar.
+In particular, the gray points at low masses are where $1-\mathrm{m}(h_{\theta_1}|h^{*}_{\theta_1})=0$ causing points to not appear in the log scale colorbar.
 % In particular, the blue points indicate parameter points where the $1-\mathrm{m}(h_{\theta_1}|h^{*}_{\theta_1})=0$ and would therefore not appear in the log scale colorbar.
@@
             Match between the \ripple and \lalsuite implementations of the IMRPhenomD waveform as a function of total mass and effective spin. 
-            The lower density of points at low masses is caused by the match being exactly 1.0 and therefore not appearing in the log scale. 
+            % The lower density of points at low masses is caused by the match being exactly 1.0 and therefore not appearing in the log scale. 
+            The gray points are where the match is exactly 1.0 and therefore would not appear in the log scale.
             It is clear that the \ripple waveform matches the \lalsuite implementation many of orders of magnitude more than necessary for all data analysis tasks across the entire parameter space.

### 2023-01-18 src/scripts/random_matches.py

@@ -66,6 +66,9 @@ params = {
 mpl.rcParams.update(params)
 
 data = np.loadtxt(paths.data / "ripple_phenomD_matches.txt")
+data = data[data[:, -1].argsort()]
+data = data[::-1]
+
 exactmatch_indices = (1.0 - data[:, -1]) <= 0.0
 other_indices = (1.0 - data[:, -1]) > 0.0
 
@@ -75,6 +78,8 @@ Mtot = data[:, 1] + data[:, 0]
 
 # Plot and save
 plt.figure(figsize=(7, 5))
+plt.scatter(Mtot[exactmatch_indices], chieff[exactmatch_indices], color="gray", alpha=0.3)
+
 cm = plt.cm.get_cmap("inferno")
 sc = plt.scatter(
     Mtot[other_indices],
@@ -84,7 +89,6 @@ sc = plt.scatter(
 )
 plt.colorbar(sc, label=r"$\log_{10}(1-\mathrm{Match})$")
 
-# plt.scatter(Mtot[exactmatch_indices], chieff[exactmatch_indices], color="C0")
 # plt.xlabel(r"Total Mass, $M$")
 # plt.ylabel(r"Mass Ratio, $q$")
 plt.xlabel(r"Total Mass, $M$")

### 2023-01-18 src/tex/ms.tex
comments and minor edits

--- 2023-01-18 src/tex/ms.tex
@@
 \newcommand{\kw}[1]{{\color{rb4}[KW: #1 ]}}
+\newcommand{\amc}[1]{{\color{red}[AC: #1]}}
 
@@
 \newcommand{\NORDITA}{\affiliation{Nordic Institute for Theoretical Physics (NORDITA), 106 91 Stockholm, Sweden}}
+\newcommand{\Ciela}{\affiliation{Ciela -- Computation and Astrophysical Data Analysis Institute, Montréal, Quebec, Canada}}
 \newcommand{\UdeM}{\affiliation{Département de Physique, Université de Montréal, 1375 Avenue Thérèse-Lavoie-Roux, Montréal, QC H2V 0B3, Canada}} 
@@
 % Author list
-\author{Adam Coogan} \UdeM \Mila
+\author{Adam Coogan} \Ciela \UdeM \Mila
 \author{Thomas D.~P.~Edwards} \JHU \OKC \NORDITA
@@
 \begin{abstract}
-    Since their discovery in 2015, gravitaitonal waves (GWs) have developed into a unique observational probe used throughout fundamental and astro-phyics.
-    As more detections continue, data analysis taks are becoming increasingly computationally demanding.
+    Since their discovery in 2015, gravitational waves (GWs) have developed into a unique observational probe used throughout fundamental and astrophyics.
+    As detections accumulate, data analysis task are becoming increasingly computationally demanding.
     Here, we propose the use of automatic differentiation through the programming framework \jax for accelerating a variety of these analsyis tasks.
-    Firstly, we demonstrate that highly accurate waveforms (IMRPhenomD) can be written in \jax and demonstrate that the serial evaluation speed of the waveform (and its derivative) is similar to the \lalsuite implementation in \CC.
+    Firstly, we demonstrate that highly accurate waveforms (here, IMRPhenomD) can be written in \jax and demonstrate that the serial evaluation speed of the waveform (and its derivative) is similar to the \lalsuite implementation in \CC.
     Moreover, \jax allows for GPU accelerated waveform calls which can be over five orders of magnitude faster than serial evaluation on a CPU.
     We then focus on three applications where efficient and differentiable waveforms are essential.
-    Firstly, we demonstrate how stochastic gradient descent can be used to optimize the $\sim 200$ coefficients that make up the waveform. 
-    In particular, we demonstrate that the typical \textit{match} with numerical relativiely waveforms can be improved by more than 10\% without any additional overhead.
-    Secondly, we show that Fisher forecasting calculations can be sped up by $\sim 100\times$ (on CPU) with no loss in accuracy.
-    This increased speed makes population forecasting substantially simpler, as discussed extensively in \citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.
+    Firstly, we demonstrate how stochastic gradient descent can be used to optimize the $\sim 200$ coefficients that make up the waveform. \amc{what does ``make up the waveform'' mean?}
+    In particular, we demonstrate that the typical match with numerical relativity waveforms can be improved by more than 10\% without any additional overhead.
+    Secondly, we show that Fisher-forecasting calculations can be sped up by $\sim 100\times$ (on CPU) with no loss in accuracy.
+    This increased speed makes population forecasting substantially simpler, as discussed extensively in \citep{Iacovelli:2022bbs, Iacovelli:2022mbg}. \amc{I'd personally avoid refs in the abstract}
     Finally, we show that gradient based samplers like Hamiltonian Monte Carlo lead to significantly reduced autocorrelation values when compared to tradiaitional Monte Carlo methods.
     Since differentiable waveforms appear to have substantial advantages for a variety of tasks throughtout GW science, we propose that waveform developers use \jax to build new waveforms moving forward.
-    Our waveform code, \ripple, can be found at \href{https://github.com/tedwards2412/ripple}{github.com/tedwards2412/ripple}, and will continue to be updated with new waveforms as they are formulated.
+    Our waveform code, \ripple, can be found at \href{https://github.com/tedwards2412/ripple}{github.com/tedwards2412/ripple}, and will continue to be updated with new waveforms as they are implemented.
+\end{abstract}
 
-\end{abstract}
+% \section{Todo}
+% \begin{itemize}
+%     \item Finish intro \te{Done}
+%     \item Finish section 2 \te{Done}
+%     \item Write captions for all plots \te{Done}
+%     \item Finish discussion
+% \end{itemize}
 
@@
 
+% Intro outline
+% - What are gravitational waves and what are data analysis tasks?
+% - What are derivatives, and how can they be used?
+% - What is automatic differentiation, and why has it become particularly useful now? JAX?
+% - Discussion of what waveforms are most ameanable to being differentiable
+% - In this paper
+
 The discovery of gravitational waves (GWs) from inspiraling and mergering compact objects (COs) has revolutionized our understanding of both fundamental physics and astronomy~\citep{LIGOScientific:2021djp}.
 Although the data volumes from GW detectors are relatively small, analyzing the data is a computationally demanding task.
-In addition, this computational cost will substatially increase when next generation detectors come online~\citep{Maggiore:2019uih, Reitze:2019iox}.
+In addition, this computational cost will substantially increase when next generation detectors \amc{should reference some here} come online.
 
-The complexity begins before data gathering where one is required to generate banks of template waveforms which will be used for a matched-filter search~\citep{Owen:1998dk, Owen:1995tm}. 
-Once potential candidates are found, parameter estimation (PE) is performed to extract the detailed source properties of each event~\citep{Christensen:2022bxb, 2020MNRAS.493.3132S, Ashton:2018jfp, Romero-Shaw:2020owr, Veitch:2014wba, Biwer:2018osg}.
-In the most minimial example, this requires a Markov Chain Monte Carlo (MCMC) on an 11 dimensional parameter space for each event.
-Beyond this simple scenario, more complex waveform models with many additional parameters may be used to test for deviations from General Relativity~\citep{Krishnendu:2021fga, LIGOScientific:2020tif, LIGOScientific:2016lio}.
-Finally, using the results of PE, population synthesis models constrain the progenitors systems from which the black holes we see merging today began their journey~\citep{LIGOScientific:2020kqk, Wong:2022flg}.
-All of these tasks require significant compute.
+The complexity begins before data gathering, where one is required to generate banks of template waveforms which will be used for a matched-filter search~\citep{Owen:1998dk, Owen:1995tm}. 
+Once potential candidates are found, parameter estimation (PE) is performed to extract the detailed source properties of each event.
+In the most minimal example, this requires a Markov Chain Monte Carlo (MCMC) on an 11-dimensional parameter space for each event.
+Beyond this simple scenario, more complex waveform models with many additional parameters may be used to test for deviations from General Relativity.
+Finally, using the results of PE, population synthesis models constrain the progenitors systems from which the black holes we see merging today began their evolutionary journey~\citep{Wong:2022flg}.
+All of these tasks require significant computing resources.
 In this paper, we will argue that differentiable waveforms (and more generally differentiable pipelines) can play a significant role in alleaviating this computational demand.
 
-Derivatives are ubiqitously useful throughout data analysis tasks.
-For instance, during PE, derivative information can be used to guide the sampling towards regions with higher likelihood values (e.g.
-in Hamiltonian Monte Carlo~\citep{2017arXiv170102434B} or Gradient Decsent).
-Gradients are particularly valuable for high dimensional spaces.
+Derivatives are ubiqitously useful in data analysis tasks.
+For instance, during PE, derivative information can be used to guide an optimizer towards higher-likelihood values (e.g.\ using gradient descent) or allow a sampler to rapidly explore parameter space (e.g.\ using Hamiltonian Monte Carlo~\citep{2011hmcm.book..113N,2017arXiv170102434B} or gradient descent).
+\amc{previously said HMC was exploring high-likelihood regions, which isn't quite right.}
+Gradients are particularly helpful in high-dimensional spaces.
 Unfortunately, in the field of GW data analysis, analytic derivatives of the necessary quantities (such as the likelihood) have historically been difficult to obtain.
@@
 
-Automatic differentiaton is a family of methods used to compute machine precision derivatives with little computational ovearhead. 
+Automatic differentiation is a family of methods used to compute machine-precision derivatives with little computational overhead. 
 AD's recent ascendance is primarily driven by its use in machine learning, particularly for derivative computations of neural networks which use gradient descent during training.
@@
 More general frameworks exist in \texttt{julia}~\citep{zygote, forwarddiff}, although \texttt{julia}'s limited support in GW physics precludes its general use.
-Here we make use of \jax~\citep{jax2018github} due to its easy integration with \texttt{python} libraries and its just-in-time (JIT) compiler which allows the same code to run on a variety of hardware accelerators (i.e.
-GPUs).
+Here we make use of \jax~\citep{jax2018github} due to its easy integration with \texttt{python} libraries, seamless support for running code on different hardware accelerators (e.g.\ graphical processing units) and its just-in-time (JIT) compiler, which can substantially accelerate code.
+\amc{GPU support and JIT are separate features.}
 
@@
 Of these, the IMRPhenom family is most well suited for an AD implementation in \jax.
-In particular, their closed form expressions and fixed size frequency grids make a \jax implementation that complies with the constraints of JIT compilation simple.
+In particular, their closed-form expressions and fixed-size frequency grids make a \jax implementation that complies with the constraints of JIT compilation simple to implement.
 EOB waveforms on the other hand require dynamic frequency grids and are therefore tricky to implement in \jax. 
 In addition, EOB waveforms use hypergeometric functions which have limited support in \jax.
-NRsurrogate waveforms have no fundamental issue but require the user to download relatively large data files, making their use cubersome in practice.\te{Max/Kaze, please add to/change this if you think its incorrect}
+NRsurrogate waveforms have no fundamental issue, but require the user to download relatively large data files, making their use cubersome in practice.\te{Max/Kaze, please add to/change this if you think its incorrect}
 
@@
 In Sec.~\ref{sec:applications} we discuss three distinct applications using differentiable waveforms. 
-Firstly, we illustrate how the fit coefficients that form part of the IMRPhenom waveform models could be improved by high dimensional fitting enabled by a differentiable waveform. 
-Secondly, we implement differnetiable detector response functions and show that the speed of Fisher matrix calculations can be substantially improved using AD.
+Firstly, we illustrate how the fit coefficients that form part of the IMRPhenom waveform models can be better-optimized via the high-dimensional fitting methods enabled by differentiable waveforms. 
+Secondly, we implement differentiable detector response functions and show that the speed of Fisher matrix calculations for approximate PE can be substantially improved \amc{accelerated?} using AD.
 Finally, we run an illustrative injection example using Hamiltonian Monte Carlo to demonstrate that the autocorrelation of derivative samplers is substantially smaller than that of traditional MCMCs.
+Practically, this translates into much faster PE with no sacrifice in accuracy.
 The associated code can be found at \ripple~\citep{ripple}.
@@
 When the COs are relatively well separated, the dynamics of the system can be well approximated by a post-Newtonian expansion.
-However, close to merger, numerical relatively simulations are required to accurately model the binary.
+However, close to merger, numerical relatively (NR) simulations are required to accurately model the binary.
 Unfortunately, these numerical simulations are computationally expensive and cannot be run in conjunction with data analysis.
-Approximate, phenomenological waveforms have therefore been constructued to enable relatively fast waveform generation at sufficient accuracy.
+Approximate, phenomenological waveforms have therefore been constructed to enable relatively fast waveform generation at sufficient accuracy.
 
 As mentioned in the previous section, the three major waveform families that have been developed to date are:  EOB,  NRsurrogate, and IMRPhenom.
-Note that unlike NRsurrogate, both the EOB and PhenomIMR waveforms are calibrated to numerical relativity waveforms.
-EOB waveforms require one to model the binary system using a Hamiltonian and are typically slow to evaluate whereas IMRPhenom waveforms are constructed with simple closed-form expressions.
-IMRPhenom waveforms are therefore ideally suited for AD, espeically using \jax. 
-In this paper we focus on the aligned spin circular orbit model IMRPhenomD~\citep{Husa:2015iqa, Khan:2015jqa}.
+Note that unlike NRsurrogate, both the EOB and IMRPhenom waveforms must be calibrated to numerical relativity waveforms.
+EOB waveforms require one to model the binary system using a Hamiltonian and are typically slow to evaluate, whereas IMRPhenom waveforms are constructed with simple closed-form expressions.
+IMRPhenom waveforms are therefore ideally-suited for AD, especially using \jax. 
+In this paper we focus on the aligned-spin, circular-orbit model IMRPhenomD~\citep{Husa:2015iqa, Khan:2015jqa}.
 
-The implementation of IMRPhenomD in \lalsuite is in C, and therefore needs to be rewritten natively into \texttt{pyhton} to be compatible with \jax.
-We have re-written IMRPhenomD from scratch using a combination of pure \texttt{pyhton} and \jax derivatives.
+The implementation of IMRPhenomD in \lalsuite is in C, and therefore needs to be rewritten natively into \texttt{python} to be compatible with \jax.
+We have rewritten IMRPhenomD from scratch using a combination of pure \texttt{python} and \jax derivatives.
 In addition, we have restructed the code for readability and evaluation speed as well as exposing the internal fitting coefficients to the user (which we will use later in Sec.~\ref{sec:applications}).
@@
 Note that in General Relativity (GR) the total mass of the system is simply an overall scaling of the waveforms frequency evolution and is therefore not a truly independent parameter.
-If we instead use a flat PSD in Eq.~\ref{eq:inner_prod} and scale the frequency grid to correctly, all dependence with $M$ seen in Fig.~\ref{fig:match} vanishes. 
+If we instead use a flat PSD in Eq.~\ref{eq:inner_prod} and scale the frequency grid correctly \amc{what do you mean by that?}, all dependence with $M$ seen in Fig.~\ref{fig:match} vanishes. 
 This figure instead illustrates a more realistic PSD where at low masses the waveform signal-to-noise is dominated by the inspiral but at high masses it is dominated by the merger.
@@
 Although slower on a CPU benchmark, a key advantage of using \jax is its native ability to run on a GPU.
-Waveform calls can therefore be evaluated in parallel, massively improving performance.
-Performing the same benchmark as above on a google colab notebook with the freely provided GPU, we find that on average waveform evaluations take $\sim 0.00005\mathrm{ms}$, nearly five orders of magnitude faster.
+Waveform calls can therefore be evaluated in parallel, massively improving performance. \amc{mention \texttt{vmap} and/or \texttt{pmap} -- super trivial to parallelize! More work in C.}
+Performing the same benchmark as above on a Google Colab notebook with a freely-provided NVIDIA Tesla T4 GPU (far from the current state-of-the-art), we find that on average waveform evaluations take $\sim 0.00005\mathrm{ms}$, nearly five orders of magnitude faster.
 Generalizing \lalsuite waveforms to run on a GPU would be a significant undertaking.
 
-The primary arguement of this paper is that waveform derivatives will also be highly vauable to data analysis tasks. 
+The primary argument of this paper is that waveform derivatives will also be highly valuable to data analysis tasks. 
 AD provides two big advantages when it comes to evaluating derivatives compared to numerical differentiation.
-First, the accuracy of derivatives from AD are signficantly more stable than finite difference methods.
+First, derivatives from AD are significantly more stable than finite-difference methods.
 In particular, finite differences suffer from both rounding and truncation errors, meaning that the user is required to \textit{tune} the width over which the difference is taken.
-On the other hand, AD produces machine precision derivatives with no tuninig.
+On the other hand, AD produces machine-precision derivatives with no tuning.
 Second, AD scales favourably with the dimensionality of the function.
 In particular, for every input dimension, $D$, added one would need to evaluate the function at least $2D$ times to calcuate finite difference derivatives for all input parameters.
-For reverse mode AD, one only needs two function calls to evaluate the derivative of all input parameters, regardless of dimenison.\footnote{
-    Note that although the number of function calls is small, reverse mode AD does add memory overhead.
+For reverse-mode AD, one only needs two function calls to evaluate the derivative of all input parameters, regardless of dimension.\footnote{
+    Note that although the number of function calls is small, reverse-mode AD does add memory overhead.
     We've not found this to be limiting in any of the situations tested so far.
@@
 
-\subsection{Finetuning Waveform Coefficients}
+\subsection{Fine-Tuning Waveform Coefficients}
 \label{subsec:coeffs}
@@
 full numerical relativity waveforms, which are essentially simpler ansatzes that
-are only approximations but can be evaluated much faster.
+are only approximations but can be evaluated much faster. \amc{redundant -- remove or merge with earlier material.}
 
@@
 Previously in the construction of IMRPhenomD \citep{Khan:2015jqa}, waveform
-coefficients are fitted in segments with amplitude and phase separated.
+coefficients were fitted in segments with amplitude and phase separated.
 Furthermore, IMRPhenomD is divided in three fitting segments, inpsiral, merger and ringdown.
@@
 After obtaining the fitting coefficient for individual segments, they are then
-"stitched" together such that they are continuous in the first derivative.
-The process of stitching introduces some addition
+``stitched'' together such that they are continuous in the first derivative.
+The process of stitching introduces some additional
 inaccuracy in the waveform model, as the connections would affect the originally
 fitted segments.
-The coefficients of the current IMRPhenomD implementation are tuned in sub-sets
-of parameters instead of all together. This means the tuning process ignore the
-correlation between different sub-sets of parameters, so the best-fitted
-solution obtained by fitting sub-sets of parameters may not be the global
+
+The coefficients of the current IMRPhenomD implementation are tuned in subsets
+of parameters instead of all together. This means the tuning process ignores the
+correlation between different subsets of parameters, so the best-fit
+solution obtained by fitting subsets of parameters may not be the global
 optimum compared to fitting all coefficients at once. By jointly
@@
 measures the goodness-of-fit of the optimization procedure, which we choose it
-to be the mismatch between the NR waveform and the approximant waveform.
+to be the mismatch between the NR waveform and the approximant waveform:
 \begin{equation}
@@
 We terminate the program after 30000 steps, in which $\lambda$ has converged to the optimum.
-% Add description of the optimization procedure.
+\amc{say something to explain why this is working.
+I.e.: does the loss initially drop and then plateau?}
 
@@
 procedure, we took NR waveforms $h_{\mathrm{NR}}$ stated in \citep{Khan:2015jqa}
-as training waveforms to fit waveform coefficients.   
+as training waveforms to fit waveform coefficients. \amc{how many are there?}
 We can see the error of optimized waveform is lower than that of the original waveform for most of the domain.
 In particular, the error in the early inspiral region is decreased by half while other regions also show good improvement in accuracy.
+\amc{so this is a waveform in the training set?
+And I don't follow exactly what you did.
+Is this just fitting a single waveform?}
 
@@
 ratio and spins. From Eq. \ref{eq:multi_loss}, we can see the averaging between
-waveforms with different intrinsic parameters implies there are trade-off in
-performance in different regions of the intrinsic parameter space.
+waveforms with different intrinsic parameters implies there can be a trade-off in
+performance in different regions of the intrinsic parameter space. \amc{this means your fit generally will depend on how the training waveforms are distributed over parameter space.}
 
@@
 The Fisher matrix approach is built around the assumption of a Gaussian likelihood.
-Although in practice this assumption is often violated in real GW datasets, the results obtained using a Fisher analysis can provide quick and useful diagnotistics in evaluating sensitivities for a variety of models and detector configurations.
+Although in practice this assumption is often violated for realistic detector noise, the results obtained using a Fisher analysis can provide quick and useful diagnostics in evaluating sensitivities for a variety of models and detector configurations.
 
-Computing the Fisher matrix requires one to evaluate derivatives of the likelihood which in turn involves derivatives of the waveform model and detector projection functions.
+Computing the Fisher matrix requires one to evaluate derivatives of the likelihood, which in turn involves derivatives of the waveform model and detector projection functions.
 AD is therefore perfectly suited for computing Fisher matrices accurately and efficiently. 
@@
 Here we purely want to illustrate the simplicitly and speed of AD for forecasting rather than providing new physics insights.
-We therefore consider a simple, three detector setup corresponding to the two LIGO detectors in addition to Virgo.
+We therefore consider a simple, three-detector setup corresponding to the two LIGO detectors in addition to Virgo.
 
-The Fisher information matrix for a single detector is typically given by 
+The Fisher information matrix for a single detector is typically given by \amc{I'm guessing in practice you compute a Hessian, so maybe write that out?}
 \begin{equation}
@@
 \end{equation}
-Finally, we invert the Fisher matrix to calculate the covariance matrix.
+Finally, we invert the Fisher matrix to calculate the covariance matrix for our measurement error estimate.
 
@@
         Although not fully converged, it is clear that we find posteriors consistent with the injected parameters.
+\amc{could apply a mild Gaussian kernel to smooth this for the sake of presentation.
+Are the parameter ranges due to your priors?
+How did you handle the boundaries in the HMC?}
     }
@@
 
+\amc{most acronyms were already introduced.}
 After the search algorithms have constructed a list of confidently detected binaries, the next step is to sample from the posterior of each sources parameters - so called parameter estimation (PE).
-To do this, one typically uses a Markov Chain Monte Carlo (MCMC) or nested sampler~\citep{multinest, dynesty}.
+To do this, one typically uses a Markov Chain Monte Carlo (MCMC) or nested sampler~\citep{Skilling2004, multinest, dynesty}.
 Although robust, both MCMC and nested sampling are slow to converge and are known to perform poorly in high dimensional parameter spaces.
@@
 Dedicated fast samplers have been designed to get approximate posteriors on the sky localization to faciliate follow-up electromagnetic observations.
-Nevertheless, these do not present the whole picture; fast, general parameter estimation therefore remains a primary goal of GW data analysis.
+Nevertheless, these do not present the whole picture; fast, general parameter estimation therefore remains a key aim of GW data analysis.
 
 A primary issue with both MCMC and nested sampling is that neither utlizes information about the likelihood's derivative and must therefore randomly walk towards areas of highest likelihood.
-Derivative based samplers, on the other hand, have been shown to extrapolate well to higher dimensions although they sometimes come with their own drawbacks.
+Derivative-based samplers, on the other hand, have been shown to extrapolate well to higher dimensions, although they sometimes come with their own drawbacks.
 Here we simply aim to demonstrate the utility of a derivative based sampler and their efficiency on a small test problem.
-% In particular, we will show that proposals made using an Hamiltonian Monte Carlo (MC) sampler are accepted at a much higher rate than a traditional MCMC algorithm.
-In particular, we will show that the autocorrelation of a Hamiltonian Monte Carlo (HMC) sampler is significantly lower than a traditional MCMC algorithm.
+In particular, we will show that the autocorrelation time of a Hamiltonian Monte Carlo (HMC) sampler is significantly lower than a traditional MCMC algorithm.
 
-For our basic example we perform an injection recovery test on a seven dimensional parameter space with the same three detector network setup considered in \S~\ref{subsec:fisher}.
+For our basic example we perform an injection recovery test on a seven-dimensional parameter space with the same three detector setup considered in Sec.~\ref{subsec:fisher}.
 More specifically, we generate Gaussian noise consistent with the measured PSDs for each detector and then inject a BBH signal with parameters: $M_c = 23.82\,\mathrm{M_\odot}$,  $\eta = 0.248$, $\chi_1=0.3$, $\chi_1=-0.4$, $D = 1.6\,\mathrm{Gpc}$, $t_c = 0.0$, and $\phi_c = 0.0$.\footnote{
@@
 }
-Using a standard Gaussian likelihood, we then run the HMC sampler implemented in \flowMC for 200000 steps and the Gaussian random walk (GRW) sampler for 1500000 steps (each with four randomly initialized independent chains).
-The number of steps and and mass matrix used for each example was hand tuned to give good performance for the specific sampler.
-The additional steps for the GRW sampler were required to achieve a similarly converged posterior.
+Using a standard Gaussian likelihood, we then run the HMC sampler implemented in \flowMC for $2 \times 10^5$ steps and the gradient-free Gaussian random walk (GRW) sampler for $15 \times 10^5$ steps for comparison (each with four randomly-initialized independent chains).
+\amc{briefly summarize and give refs for flowMC and GRW.
+Is GRW Metropolis-Hastings with a Gaussian transition kernel?
+How'd you tune this?
+Why not use NUTS for HMC?}
+The number of steps and and mass matrix used for each example was hand-tuned to give good performance for the specific sampler.
+\amc{not a huge deal, but why hand tune?
+There are automated ways to do this in e.g.\ numpyro with a warm-up adaptation phase.}
+The additional steps for the GRW sampler were required to achieve a similarly converged posterior. \amc{Why not plot its posterior?}
 
@@
 % Since we are running the chains from the same initialization, this enhanced acceptance for HMC can be mainly associated with use of gradient information to guide sampling.
-Although further steps would be required to achieve a fully converged posterior, these chains are sufficient to illustrate the increased efficiency associated with HMC. 
-To illustrate this, in Fig.~\ref{fig:autocorrelation} we plot the autocorrelation as a function of lag for both the HMC and GRW samplers.
+Although further steps would be required to achieve a fully-converged posterior, these chains are sufficient to show the increased efficiency associated with HMC. 
+To further illustrate this, in Fig.~\ref{fig:autocorrelation} we plot the autocorrelation as a function of lag for both the HMC and GRW samplers.
 The HMC autocorrelation is substantially lower than that of the GRW.
-We therefore expect gradient based samplers to converge significantly faster than typical samplers, especially in higher dimensions.
-In addition, we found that the effictive number of samples (a measure of the number of indepedent samples) is between 2 and 7 times larger for HMC across the different dimensions of the parameter space.
+We therefore expect gradient-based samplers to converge significantly faster than typical samplers, especially in higher dimensions.
+In addition, we found that the effective number of samples \amc{provide ref?} (a measure of the number of indepedent samples) is between 2 and 7 times larger for HMC across the different dimensions of the parameter space, indicating its superior performance.
 
-In a follow up paper we will demonstrate that minute scale PE can be achieved by combining normalizing flows~\citep{2022arXiv221106397W, Gabrie:2021tlu}, GPU acceleration, and a derivative based sampler~\citep{PEpaper}.
+In a follow-up paper we will demonstrate that minute-scale PE can be achieved by combining normalizing flows~\citep{2022arXiv221106397W, Gabrie:2021tlu}, GPU acceleration, and a derivative based sampler~\citep{PEpaper}.
 We therefore expect \jax waveforms to be crucial to future PE efforts in GW astronomy, especially for low-latency pipelines.
@@
 In particular, we showed that it matches the \lalsuite implementation to near machine precision and can be easily parallelized on a GPU.
-Parallization on a GPU provides substantial speed increases; on a free google colab GPU we found that waveform evaluations are nearly 5 orders of magnitude faster than serial CPU evaluations.
-Second, we discussed three data analysis tasks which can all be substantially improved by utlizing derivative information of the waveform.
+Parallelization on a GPU provides substantial speed increases; on a free Google Colab GPU we found that waveform evaluations are nearly 5 orders of magnitude faster than serial CPU evaluations.
+Second, we discussed three data analysis tasks which can all be substantially improved by utilizing derivative information of the waveform.
 Although we primarily discuss toy examples in this paper, each can be extended to the full data analysis task, some of which will be shown in upcoming papers~\citep{PEpaper}.
@@
 Two other waveform families are commonly used in GW data analysis: the effective-one-body (EOB) and numerical relativity surrogate (NRSurrogate). 
-Some progress has been made towards the construction of a differentiable NRsurrogate, but currently it seems difficult to implement EOB waveforms in \jax.
-In particular, the evolution of the Hamiltonian required to evaluate an EOB waveform is both inherently slow to differentiate and difficult to implement in \jax which requires fixed length arrays for its JIT compilation.
+Some progress has been made towards the construction of a differentiable NRSurrogate, but currently it seems difficult to implement EOB waveforms in \jax.
+In particular, the evolution of the Hamiltonian required to evaluate an EOB waveform is both inherently slow to differentiate and difficult to implement in \jax, which requires fixed-length arrays for its JIT compilation.
 Since EOB waveforms are often regarded as the state-of-the-art, more work is required to see if a fast, differentiable implementation is possible.
 
-Currently the biggest constraint to adopting differentiable waveforms is the need to re-write the most commonly used waveforms into \jax (or pure python).
+Currently the biggest constraint to adopting differentiable waveforms is the need to rewrite the most commonly used waveforms into \jax (or pure python).
 In order to showcase the benefits of differentiable waveforms as quickly as possible, at the time of writing, we have only implemented an aligned spin GW model (IMRPhenomD).
 We plan on adding a variety of different waveforms to \ripple in the near future with the primary goal of reaching a \jax version of a fully precessing, higher order mode waveform such as IMRPhenomXPHM.
-Ideally, future waveforms should be implemented either in pure python or using \jax directly.
+Ideally, future waveforms should be implemented either in pure python or using \jax directly. \amc{if they're pure python, some translation will be required due to jax's restrictions.}
 This would ensure that the community can easily utilize differentiability and hardware acceleration in the future.
@@
 This work was supported by collaborative visits funded by the Cosmology and Astroparticle Student and Postdoc Exchange Network (CASPEN). 
-T.E. is supported by the Horizon Postdoctoral Fellowship.
+T.E.\ is supported by the Horizon Postdoctoral Fellowship.
+A.~C.\ acknowledges funding from the Schmidt Futures Foundation.
 

### 2023-01-18 src/tex/ms.tex
Edits

--- 2023-01-18 src/tex/ms.tex
@@
 \newcommand{\kw}[1]{{\color{rb4}[KW: #1 ]}}
+\newcommand{\AZ}[1]{{\color{Burnt}[AZ: #1]}}
 
@@
 \begin{abstract}
-    Since their discovery in 2015, gravitaitonal waves (GWs) have developed into a unique observational probe used throughout fundamental and astro-phyics.
-    As more detections continue, data analysis taks are becoming increasingly computationally demanding.
-    Here, we propose the use of automatic differentiation through the programming framework \jax for accelerating a variety of these analsyis tasks.
-    Firstly, we demonstrate that highly accurate waveforms (IMRPhenomD) can be written in \jax and demonstrate that the serial evaluation speed of the waveform (and its derivative) is similar to the \lalsuite implementation in \CC.
-    Moreover, \jax allows for GPU accelerated waveform calls which can be over five orders of magnitude faster than serial evaluation on a CPU.
+    Since their discovery in 2015, gravitational waves (GWs) have developed into a unique observational probe used throughout fundamental and astro-phyics.
+    As more detections continue, data analysis tasks are becoming increasingly computationally demanding.
+    Here, we propose the use of automatic differentiation through the programming framework \jax for accelerating a variety of these analysis tasks.
+    Firstly, we demonstrate that \sout{highly accurate waveforms  (IMRPhenomD)} \AZ{complete waveforms which cover the inspiral, merger, and ringdown of binary black holes (i.e.~IMRPhenomD)} can be written in \jax and demonstrate that the serial evaluation speed of the waveform (and its derivative) is similar to the \lalsuite implementation in \CC.
+    Moreover, \jax allows for GPU-accelerated waveform calls which can be over five orders of magnitude faster than serial evaluation on a CPU.
     We then focus on three applications where efficient and differentiable waveforms are essential.
-    Firstly, we demonstrate how stochastic gradient descent can be used to optimize the $\sim 200$ coefficients that make up the waveform. 
-    In particular, we demonstrate that the typical \textit{match} with numerical relativiely waveforms can be improved by more than 10\% without any additional overhead.
-    Secondly, we show that Fisher forecasting calculations can be sped up by $\sim 100\times$ (on CPU) with no loss in accuracy.
-    This increased speed makes population forecasting substantially simpler, as discussed extensively in \citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.
-    Finally, we show that gradient based samplers like Hamiltonian Monte Carlo lead to significantly reduced autocorrelation values when compared to tradiaitional Monte Carlo methods.
-    Since differentiable waveforms appear to have substantial advantages for a variety of tasks throughtout GW science, we propose that waveform developers use \jax to build new waveforms moving forward.
+    Firstly, we demonstrate how stochastic gradient descent can be used to optimize the $\sim 200$ coefficients that \sout{make up the waveform} \AZ{are used to calibrate the waveform model}. 
+    In particular, we demonstrate that the typical \textit{match} with numerical relatively waveforms can be improved by more than 10\% without any additional overhead.
+    Secondly, we show that Fisher forecasting calculations can be sped up by $\sim 100\times$ (on a CPU) with no loss in accuracy.
+    This increased speed makes population forecasting substantially simpler, as discussed extensively in \citet{Iacovelli:2022bbs, Iacovelli:2022mbg}.
+    Finally, we show that gradient-based samplers like Hamiltonian Monte Carlo lead to significantly reduced autocorrelation values when compared to traditional Monte Carlo methods.
+    Since differentiable waveforms appear to have substantial advantages for a variety of tasks throughout GW science, we propose that waveform developers use \jax to build new waveforms moving forward.
     Our waveform code, \ripple, can be found at \href{https://github.com/tedwards2412/ripple}{github.com/tedwards2412/ripple}, and will continue to be updated with new waveforms as they are formulated.
@@
 
-The discovery of gravitational waves (GWs) from inspiraling and mergering compact objects (COs) has revolutionized our understanding of both fundamental physics and astronomy~\citep{LIGOScientific:2021djp}.
+The discovery of gravitational waves (GWs) from inspiraling and merging compact objects (COs) has revolutionized our understanding of both fundamental physics and astronomy~\citep{LIGOScientific:2021djp}.
 Although the data volumes from GW detectors are relatively small, analyzing the data is a computationally demanding task.

### 2023-01-18 src/tex/ms.tex
Some edits and refs

@@
 
-The discovery of gravitational waves (GWs) from inspiraling and merging compact objects (COs) has revolutionized our understanding of both fundamental physics and astronomy~\citep{LIGOScientific:2021djp}.
+The discovery of gravitational waves (GWs)~\cite{LIGOScientific:2016aoc} from inspiraling and merging compact objects (COs) has revolutionized our understanding of both fundamental physics and astronomy~\citep[e.g.][]{LIGOScientific:2021djp,LIGOScientific:2021sio,LIGOScientific:2021psn}.
 Although the data volumes from GW detectors are relatively small, analyzing the data is a computationally demanding task.
-In addition, this computational cost will substatially increase when next generation detectors come online~\citep{Maggiore:2019uih, Reitze:2019iox}.
+In addition, this computational cost will substantially increase when next generation detectors come online~\citep{Maggiore:2019uih, Reitze:2019iox, Evans:2021gyd}.
 
-The complexity begins before data gathering where one is required to generate banks of template waveforms which will be used for a matched-filter search~\citep{Owen:1998dk, Owen:1995tm}. 
+\AZ{The complexity begins even before data taking, since GW searches using the matched-filtering technique~\citep{Owen:1998dk, Owen:1995tm} require the generation of large banks of template waveforms.}
+\sout{The complexity begins before data gathering where one is required to generate banks of template waveforms which will be used for a matched-filter search~\citep{Owen:1998dk, Owen:1995tm}. }
 Once potential candidates are found, parameter estimation (PE) is performed to extract the detailed source properties of each event~\citep{Christensen:2022bxb, 2020MNRAS.493.3132S, Ashton:2018jfp, Romero-Shaw:2020owr, Veitch:2014wba, Biwer:2018osg}.
-In the most minimial example, this requires a Markov Chain Monte Carlo (MCMC) on an 11 dimensional parameter space for each event.
-Beyond this simple scenario, more complex waveform models with many additional parameters may be used to test for deviations from General Relativity~\citep{Krishnendu:2021fga, LIGOScientific:2020tif, LIGOScientific:2016lio}.
-Finally, using the results of PE, population synthesis models constrain the progenitors systems from which the black holes we see merging today began their journey~\citep{LIGOScientific:2020kqk, Wong:2022flg}.
-All of these tasks require significant compute.
-In this paper, we will argue that differentiable waveforms (and more generally differentiable pipelines) can play a significant role in alleaviating this computational demand.
+In the most minimal example \AZ{of non-precessing binaries}, this requires a Markov Chain Monte Carlo (MCMC) on an 11 dimensional parameter space for each event.
+\AZ{Maybe modify; some other points are precessing PE requires 15 params, tidal prec 17 params, plus calibration marge and EoS inferences can incorporate even more}
+Beyond this simple scenario, more complex waveform models with \sout{many} additional parameters may be used to test for deviations from General Relativity~\citep{Krishnendu:2021fga, LIGOScientific:2020tif, LIGOScientific:2016lio}.
+\AZ{Add PPE and maybe TIGER refs..?}
+Finally, using the results of PE, population synthesis models constrain the progenitors systems from which the black holes we see merging today began their journey~\citep{LIGOScientific:2020kqk, LIGOScientific:2021psn, Wong:2022flg}.
+All of these tasks require significant computation.
+In this paper, we will argue that differentiable waveforms (and more generally differentiable pipelines) can play a significant role in alleviating this computational demand.
 
-Derivatives are ubiqitously useful throughout data analysis tasks.
+Derivatives are ubiquitously useful throughout data analysis tasks.
 For instance, during PE, derivative information can be used to guide the sampling towards regions with higher likelihood values (e.g.
@@
 
-Automatic differentiaton is a family of methods used to compute machine precision derivatives with little computational ovearhead. 
+Automatic differentiation is a family of methods used to compute machine precision derivatives with little computational overhead. 
 AD's recent ascendance is primarily driven by its use in machine learning, particularly for derivative computations of neural networks which use gradient descent during training.

### 2023-01-18 src/tex/ms.tex
Edits and comments

@@
 Derivatives are ubiquitously useful throughout data analysis tasks.
-For instance, during PE, derivative information can be used to guide the sampling towards regions with higher likelihood values (e.g.
-in Hamiltonian Monte Carlo~\citep{2017arXiv170102434B} or Gradient Decsent).
+For instance, during PE, derivative information can be used to guide the sampling towards regions with higher likelihood values (e.g.~in Hamiltonian Monte Carlo~\citep{2017arXiv170102434B} or Gradient Descent).
 Gradients are particularly valuable for high dimensional spaces.
@@
 There are now a variety of AD implementations, most notably in deep learning frameworks such as \texttt{pytorch}~\citep{pytorch} and \texttt{tensorflow}~\citep{tensorflow2015-whitepaper}.
-More general frameworks exist in \texttt{julia}~\citep{zygote, forwarddiff}, although \texttt{julia}'s limited support in GW physics precludes its general use.
-Here we make use of \jax~\citep{jax2018github} due to its easy integration with \texttt{python} libraries and its just-in-time (JIT) compiler which allows the same code to run on a variety of hardware accelerators (i.e.
-GPUs).
+More general frameworks exist in \texttt{julia}~\citep{zygote, forwarddiff}, although \texttt{julia}'s limited \sout{support} \AZ{use} in GW \sout{physics} \AZ{analysis software} precludes its general use.
+Here we make use of \jax~\citep{jax2018github} due to its easy integration with \texttt{python} libraries and its just-in-time (JIT) compiler which allows the same code to run on a variety of hardware accelerators (i.e.~GPUs).
 
@@
 EOB waveforms on the other hand require dynamic frequency grids and are therefore tricky to implement in \jax.
+\AZ{What does this mean?
+I don't understand why EOB needs dynamic frequency grids.
+We should clarify, and we can also mention if \jax is well-suited to frequency-domain ROM implementations of EOB, and reference those.}
 In addition, EOB waveforms use hypergeometric functions which have limited support in \jax.
-NRsurrogate waveforms have no fundamental issue but require the user to download relatively large data files, making their use cubersome in practice.\te{Max/Kaze, please add to/change this if you think its incorrect}
+NRsurrogate waveforms have no fundamental issue but require the user to download relatively large data files, making their use cumbersome in practice.\te{Max/Kaze, please add to/change this if you think its incorrect}
 
 In this paper we argue that differentiable waveforms will be a vital component for the future of GW data analysis.
-In addition, we present \ripple, a small GW \texttt{python} package which can serve as a respository for differentiable implementations of some of the main waveforms currently used in LIGO and Virgo analyses.
+\AZ{I prefer to describe what is rather than what could be; can we say what elements exist, e.g.~
+we present \ripple... which includes \jax implementations of GW antenna patterns, the IMRPhenomD waveform, and...}
+In addition, we present \ripple, a small GW \texttt{python} package which can serve as a repository for differentiable implementations of some of the main waveforms currently used in LIGO and Virgo analyses.
 The remainder of this paper is structured as follows. 
-In Sec.~\ref{sec:waveforms} we discuss the differentiable IMRPhenomD waveform implemented in \ripple and perform some benchmarks to demonstrate its speed and accruacy. 
+In Sec.~\ref{sec:waveforms} we discuss the differentiable IMRPhenomD waveform implemented in \ripple and perform some benchmarks to demonstrate its speed and accuracy. 
 In Sec.~\ref{sec:applications} we discuss three distinct applications using differentiable waveforms. 
 Firstly, we illustrate how the fit coefficients that form part of the IMRPhenom waveform models could be improved by high dimensional fitting enabled by a differentiable waveform. 
-Secondly, we implement differnetiable detector response functions and show that the speed of Fisher matrix calculations can be substantially improved using AD.
+Secondly, we implement differentiable detector response functions and show that the speed of Fisher matrix calculations can be substantially improved using AD.
 Finally, we run an illustrative injection example using Hamiltonian Monte Carlo to demonstrate that the autocorrelation of derivative samplers is substantially smaller than that of traditional MCMCs.

### 2023-01-18 src/tex/ms.tex
More comments/edits

@@
 Since the match is a measure of the difference between two waveforms, we can use it to demonstrate that the implementation of IMRPhenomD in \ripple accurately matches the \lalsuite implementation.
+\AZ{Say what PSD was used and reference it.}
 This is shown in Fig.~\ref{fig:match}, where we have calculated the $\mathrm{m}(h_{\theta_1}|\bar{h}_{\theta_1})$ across the entire parameter space.\footnote{
@@
 From Fig.~\ref{fig:match}, it is clear that the \ripple waveform matches with the \lalsuite waveform close to numerical precision across the entire parameter space.
+\AZ{strictly $=0$? That seems impossible. Fmla has wrong labels for $h$'s}
 In particular, the gray points at low masses are where $1-\mathrm{m}(h_{\theta_1}|h^{*}_{\theta_1})=0$ causing points to not appear in the log scale colorbar.
@@
 
-Note that in General Relativity (GR) the total mass of the system is simply an overall scaling of the waveforms frequency evolution and is therefore not a truly independent parameter.
-If we instead use a flat PSD in Eq.~\ref{eq:inner_prod} and scale the frequency grid to correctly, all dependence with $M$ seen in Fig.~\ref{fig:match} vanishes. 
+Note that in General Relativity (GR) the total mass $M$ of \AZ{a binary black hole simply serves as an overall scale for the system, so that the frequency evolution can be trivially rescaled.
+The total mass only impacts the match because the chosen PSD and frequency limits fix a reference scale.}
+\sout{the system is simply an overall scaling of the waveforms frequency evolution and is therefore not a truly independent parameter.}
+If we instead use a flat PSD in Eq.~\ref{eq:inner_prod} and \AZ{rescale the frequency grid to units of $Mf$}, \sout{correctly,} all dependence with $M$ seen in Fig.~\ref{fig:match} vanishes. 
+\AZ{Revisit this after defining PSD above:} 
 This figure instead illustrates a more realistic PSD where at low masses the waveform signal-to-noise is dominated by the inspiral but at high masses it is dominated by the merger.
@@
 Waveform calls can therefore be evaluated in parallel, massively improving performance.
-Performing the same benchmark as above on a google colab notebook with the freely provided GPU, we find that on average waveform evaluations take $\sim 0.00005\mathrm{ms}$, nearly five orders of magnitude faster.
+Performing the same benchmark as above on a google colab notebook with the freely provided GPU, we find that on average waveform evaluations take $\sim \sout{0.00005} 5 \times 10^{-5}\mathrm{ms}$, nearly five orders of magnitude faster.
 Generalizing \lalsuite waveforms to run on a GPU would be a significant undertaking.

### 2023-01-18 src/tex/ms.tex
Adding comments

@@
 
+\AZ{Add refs here:}
 There are a variety of gravitational waveforms currently used in analysis pipelines.
@@
 In particular, their closed form expressions and fixed size frequency grids make a \jax implementation that complies with the constraints of JIT compilation simple.
-EOB waveforms on the other hand require dynamic frequency grids and are therefore tricky to implement in \jax.
+\AZ{to fix:} EOB waveforms on the other hand require dynamic frequency grids and are therefore tricky to implement in \jax.
 \AZ{What does this mean?
@@
 In addition, EOB waveforms use hypergeometric functions which have limited support in \jax.
+\AZ{rephrase this in a more positive fashion:}
 NRsurrogate waveforms have no fundamental issue but require the user to download relatively large data files, making their use cumbersome in practice.\te{Max/Kaze, please add to/change this if you think its incorrect}
@@
 We have re-written IMRPhenomD from scratch using a combination of pure \texttt{pyhton} and \jax derivatives.
-In addition, we have restructed the code for readability and evaluation speed as well as exposing the internal fitting coefficients to the user (which we will use later in Sec.~\ref{sec:applications}).
+In addition, we have restructured the code for readability and evaluation speed as well as exposing the internal fitting coefficients to the user (which we will use later in Sec.~\ref{sec:applications}).
 
-To demonstrate our implempation of IMRPhenomD is faithful to the \lalsuite implementation, we start by defining the noise weighted inner product:
+To demonstrate our implementation of IMRPhenomD is faithful to the \lalsuite implementation, we start by defining the noise weighted inner product:
 \begin{equation}
     \label{eq:inner_prod}
-    \left(h_{\theta_1}|h_{\theta_2}\right) \equiv 4 \, \mathrm{Re} \int^{\infty}_{0} \mathrm{d} f \, \frac{ h_{\theta_1}(f) h^*_{\theta_2}(f)}{S_n(f)}\, ,
+    \left(h_1|h_2\right) \equiv 4 \, \mathrm{Re} \int^{\infty}_{0} \mathrm{d} f \, \frac{ h^*_1(f) h_2(f)}{S_n(f)}\, ,
 \end{equation}
-where $S_n$ is the (one-sided) noise power spectral density (PSD) and $h_{\theta_1}$ is the frequency domain waveform evaluated at the instrinsic parameter point $\theta_1$.
+where $S_n$ is the (one-sided) noise power spectral density (PSD) and $h_{\theta_1}$ is the frequency domain waveform evaluated at the intrinsic parameter point $\theta_1$.
 We can then normalize the inner product through
@@
 
+\AZ{Broad comments: need to cite which catalog the waveforms came from if any.
+The number of waveforms used seems much larger than the number originally used
+to calibrate PhenomD. 
+If that is true, we can't make a big deal about having a better mismatch compared to
+the original version, because of course optimizing over another set will improve 
+performance.
+Say how many waveforms we use to optimize, how many we compare to, where they came
+from.}
+
 Having an accurate waveform model is essential for many data analysis tasks in
@@
 potential improvement of the accuracy by fitting all coefficients at once.
+
+\AZ{Maybe comment on optimization method used for EOB waveforms?
+I think they do interesting Monte Carlo etc}
 
@@
     \caption{Priors for the 11 dimensional parameter space used for the Fisher forecasting population analysis in Sec.~\ref{subsec:fisher}. 
-    $\boldsymbol{U}$ indicates a uniform distribution between the two variables in the brackets.}
+    $\boldsymbol{U}$ indicates a uniform distribution between the two variables in the brackets.
+    \AZ{$\cos \iota \in (-1,1)$, $\psi \in (0,2 \pi)$ (or $\pi$), RA is $\alpha \in (0, 2\pi)$, Dec is $\cos \delta \in (-1,1)$ }
+    }
     \label{tab:priors}

### 2023-01-18 src/tex/ms.tex
comments

@@
     $\boldsymbol{U}$ indicates a uniform distribution between the two variables in the brackets.
-    \AZ{$\cos \iota \in (-1,1)$, $\psi \in (0,2 \pi)$ (or $\pi$), RA is $\alpha \in (0, 2\pi)$, Dec is $\cos \delta \in (-1,1)$ }
+    \AZ{$\cos \iota \in (-1,1)$, $\psi \in (0,2 \pi)$ (or $\pi$), RA is $\alpha \in (0, 2\pi)$, Dec is $\delta$, let $\vartheta = \pi/2 - \delta$, then $\cos (\vartheta) \in (-1,1)$ }
     }
@@
 
-To illustrate the computational speed of computing Fisher matrices with AD, we consider a population of binaries and compute the sky localtization error following Eq.~(28) in \citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.
+To illustrate the computational speed of computing Fisher matrices with AD, we consider a population of binaries and compute the sky localization error following Eq.~(28) in \citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.
 Since the Fisher matrix approach is known to have both theoretical issues as well as numerical instabilities for low signal-to-noise events, we restrict our population to only nearby systems.

### 2023-01-18 src/tex/ms.tex
Edits

@@
 They are generally structured into different families, the most common of which are: the effective-one-body (EOB), the phenomenological inspiral-merger-ringerdown (IMRPhenom), and numerical relativity surrogate (NRsurrogate).
-Of these, the IMRPhenom family is most well suited for an AD implementation in \jax.
-In particular, their closed form expressions and fixed size frequency grids make a \jax implementation that complies with the constraints of JIT compilation simple.
+%Of these, the IMRPhenom family is most well suited for an AD implementation in \jax.
+\AZ{Of these, the IMRPhenom family serves as a natural starting point for an AD implementation in \jax.
+Models like the non-precessing IMRPhenomD model studied here and the precessing, higher-mode model IMRPhenomXPHM~\citep{} are written in frequency domain using closed-form expressions.
+This makes a}
+\sout{In particular, their closed form expressions and fixed size frequency grids make a }
+\jax implementation that complies with the constraints of JIT compilation simple.
 \AZ{to fix:} EOB waveforms on the other hand require dynamic frequency grids and are therefore tricky to implement in \jax.
@@
 The total mass only impacts the match because the chosen PSD and frequency limits fix a reference scale.}
-\sout{the system is simply an overall scaling of the waveforms frequency evolution and is therefore not a truly independent parameter.}
+\sout{the system is simply an overall scaling of the waveforms} 
+\sout{frequency evolution and is therefore not a} \sout{truly independent parameter.}
 If we instead use a flat PSD in Eq.~\ref{eq:inner_prod} and \AZ{rescale the frequency grid to units of $Mf$}, \sout{correctly,} all dependence with $M$ seen in Fig.~\ref{fig:match} vanishes. 
@@
 
-The primary arguement of this paper is that waveform derivatives will also be highly vauable to data analysis tasks. 
+The primary argument of this paper is that waveform derivatives will also be highly valuable to data analysis tasks. 
 AD provides two big advantages when it comes to evaluating derivatives compared to numerical differentiation.
-First, the accuracy of derivatives from AD are signficantly more stable than finite difference methods.
+First, the accuracy of derivatives from AD are significantly more stable than finite difference methods.
 In particular, finite differences suffer from both rounding and truncation errors, meaning that the user is required to \textit{tune} the width over which the difference is taken.
-On the other hand, AD produces machine precision derivatives with no tuninig.
-Second, AD scales favourably with the dimensionality of the function.
-In particular, for every input dimension, $D$, added one would need to evaluate the function at least $2D$ times to calcuate finite difference derivatives for all input parameters.
+On the other hand, AD produces machine precision derivatives with no tuning.
+Second, AD scales favorably with the dimensionality of the function.
+In particular, for every input dimension, $D$, added one would need to evaluate the function at least $2D$ times to calculate finite difference derivatives for all input parameters.
 For reverse mode AD, one only needs two function calls to evaluate the derivative of all input parameters, regardless of dimenison.\footnote{
@@
 }
-Since the parameter space of GWs in GR has $\mathcal{O}(10)$ dimensons, the speed of derivative evaluation is less crucial than the stability.
+Since the parameter space of GWs in GR has $\mathcal{O}(10)$ dimensions, the speed of derivative evaluation is less crucial than the stability.
 However, this might change for waveforms in beyond GR models where many more parameters are added.
@@
 coefficients are fitted in segments with amplitude and phase separated.
-Furthermore, IMRPhenomD is divided in three fitting segments, inpsiral, merger and ringdown.
+Furthermore, IMRPhenomD is divided in three fitting segments, inspiral, merger and ringdown.
 Each of these segments has their own set of fitting coefficients.
@@
 
-Moreover, each error calculation (including computing the Fisher matrices for each detector and the inversion procress) is substantially faster.
+Moreover, each error calculation (including computing the Fisher matrices for each detector and the inversion process) is substantially faster.
 In particular, we find that after compilation each error calculation takes approximately half a second on a single computing core.

### 2023-01-18 src/tex/ms.tex
Edits

@@
 \jax implementation that complies with the constraints of JIT compilation simple.
-\AZ{to fix:} EOB waveforms on the other hand require dynamic frequency grids and are therefore tricky to implement in \jax.
-\AZ{What does this mean?
+NRsurrogate models, which interpolate directly over waveforms produced by numerical relativity simulations, are in principle also straightfoward to implement in \jax.
+EOB waveforms on the other hand are produced by evolving the dynamics of an effective one body Hamiltonian, and are therefore more difficult to implement in \jax.
+For EOB waveforms, frequency-domain reduced-order models may be a convenient target.
+% to produce a waveform
+%require dynamic frequency grids and are therefore tricky to implement in \jax.
+%\AZ{What does this mean?
 I don't understand why EOB needs dynamic frequency grids.
 We should clarify, and we can also mention if \jax is well-suited to frequency-domain ROM implementations of EOB, and reference those.}
-In addition, EOB waveforms use hypergeometric functions which have limited support in \jax.
-\AZ{rephrase this in a more positive fashion:}
-NRsurrogate waveforms have no fundamental issue but require the user to download relatively large data files, making their use cumbersome in practice.\te{Max/Kaze, please add to/change this if you think its incorrect}
+%In addition, EOB waveforms use hypergeometric functions which have limited support in \jax.
+%\AZ{rephrase this in a more positive fashion:}
+%NRsurrogate waveforms have no fundamental issue but require the user to download relatively large data files, making their use cumbersome in practice.\te{Max/Kaze, please add to/change this if you think its incorrect}
 

### 2023-01-19 src/tex/ms.tex
Comments and edits for the manuscript

@@
 
+\definecolor{dfm}{HTML}{6C71C4}
+\newcommand{\dfm}[1]{{\color{dfm}[DFM: #1]}}
+
 % Begin!
@@
 % Title
-\title{\ripple: Differentiable Waveforms for Gravitational Wave Data Analysis}
+\title{\ripple: Differentiable \dfm{and Hardware-Accelerated} Waveforms for Gravitational Wave Data Analysis}
 
@@
 Derivatives are ubiquitously useful throughout data analysis tasks.
-For instance, during PE, derivative information can be used to guide the sampling towards regions with higher likelihood values (e.g.~in Hamiltonian Monte Carlo~\citep{2017arXiv170102434B} or Gradient Descent).
+For instance, during PE, derivative information can be used to guide the sampling towards regions with higher likelihood values \dfm{This isn't correct.
+In HMC the gradients don't have the behavior.
+I'd just say something like ``gradients are used to improve performance''} (e.g.~in Hamiltonian Monte Carlo~\citealt{2017arXiv170102434B} or Gradient Descent).
 Gradients are particularly valuable for high dimensional spaces.
@@
             It is clear that the \ripple waveform matches the \lalsuite implementation many of orders of magnitude more than necessary for all data analysis tasks across the entire parameter space.
+            \dfm{I think this figure could use some work.
+Maybe the points could be smaller to make them easier to resolve.
+And it could be interesting to clip the mismatch to something like $10^{-16}$ so that we don't lose the points with zero mismatch.
+Right now it looks less impressive!}
         }
@@
 After obtaining the fitting coefficient for individual segments, they are then
-"stitched" together such that they are continuous in the first derivative.
+``stitched'' together such that they are continuous in the first derivative.
 The process of stitching introduces some addition

### 2023-01-19 src/tex/ms.tex
merging
Adding a bunch of references to intro

@@
 \AZ{Maybe modify; some other points are precessing PE requires 15 params, tidal prec 17 params, plus calibration marge and EoS inferences can incorporate even more}
-Beyond this simple scenario, more complex waveform models with \sout{many} additional parameters may be used to test for deviations from General Relativity~\citep{Krishnendu:2021fga, LIGOScientific:2020tif, LIGOScientific:2016lio}.
+Beyond this simple scenario, more complex waveform models with \sout{many} additional parameters may be used to test for deviations from General Relativity~\citep{Krishnendu:2021fga, LIGOScientific:2020tif, LIGOScientific:2016lio, Yunes:2016jcc, Arun:2006yw, Agathos:2013upa}.
 \AZ{Add PPE and maybe TIGER refs..?}
@@
 In HMC the gradients don't have the behavior.
-I'd just say something like ``gradients are used to improve performance''} (e.g.~in Hamiltonian Monte Carlo~\citealt{2017arXiv170102434B} or Gradient Descent).
+I'd just say something like ``gradients are used to improve performance''} (e.g.~in Hamiltonian Monte Carlo~\citep{2017arXiv170102434B} or Gradient Descent~\citep{2016arXiv160904747R}).
 Gradients are particularly valuable for high dimensional spaces.
-Unfortunately, in the field of GW data analysis, analytic derivatives of the necessary quantities (such as the likelihood) have historically been difficult to obtain.
+Unfortunately, in the field of GW data analysis, analytic derivatives of the necessary quantities (such as the likelihood) have historically been difficult to obtain~\citep{Keppel:2013kia}.
 Numerical derivatives also suffer from accuracy issues stemming from rounding or truncation errors.
@@
 \AZ{Of these, the IMRPhenom family serves as a natural starting point for an AD implementation in \jax.
-Models like the non-precessing IMRPhenomD model studied here and the precessing, higher-mode model IMRPhenomXPHM~\citep{} are written in frequency domain using closed-form expressions.
+Models like the non-precessing IMRPhenomD~\citep{Khan:2015jqa} model studied here and the precessing, higher-mode model IMRPhenomXPHM~\citep{Pratten:2020fqn, Pratten:2020ceb} are written in frequency domain using closed-form expressions.
 This makes a}

### 2023-01-20 src/tex/ms.tex
Testing cache issue from yesterday
Adding edit

@@
 
-The discovery of gravitational waves (GWs)~\cite{LIGOScientific:2016aoc} from inspiraling and merging compact objects (COs) has revolutionized our understanding of both fundamental physics and astronomy~\citep[e.g.][]{LIGOScientific:2021djp,LIGOScientific:2021sio,LIGOScientific:2021psn}.
+The discovery of gravitational waves (GWs)~\citep{LIGOScientific:2016aoc} from inspiraling and merging compact objects (COs) has revolutionized our understanding of both fundamental physics and astronomy~\citep[e.g.][]{LIGOScientific:2021djp,LIGOScientific:2021sio,LIGOScientific:2021psn}.
 Although the data volumes from GW detectors are relatively small, analyzing the data is a computationally demanding task.
@@
 Numerical derivatives also suffer from accuracy issues stemming from rounding or truncation errors.
-However, recent progress in automatic differentiation (AD) has shown promise in allowing general, fast derivative calcuations for gravitational waveforms~\citep{Coogan:2022qxs}.
+However, recent progress in automatic differentiation (AD) has shown promise in allowing general, fast derivative calculations for gravitational waveforms\AZ{, with applications to constructing template banks~\citep{Coogan:2022qxs} and computing the Fisher information for forecasting~\citep{Iacovelli:2022bbs, Iacovelli:2022mbg}}.
 

### 2023-01-20 src/tex/ms.tex
Adding ref
dealing with some comments

@@
     Here, we propose the use of automatic differentiation through the programming framework \jax for accelerating a variety of these analysis tasks.
-    Firstly, we demonstrate that \sout{highly accurate waveforms  (IMRPhenomD)} \AZ{complete waveforms which cover the inspiral, merger, and ringdown of binary black holes (i.e.~IMRPhenomD)} can be written in \jax and demonstrate that the serial evaluation speed of the waveform (and its derivative) is similar to the \lalsuite implementation in \CC.
+    Firstly, we demonstrate that complete waveforms which cover the inspiral, merger, and ringdown of binary black holes (i.e.~IMRPhenomD) can be written in \jax and demonstrate that the serial evaluation speed of the waveform (and its derivative) is similar to the \lalsuite implementation in \CC.
     Moreover, \jax allows for GPU-accelerated waveform calls which can be over five orders of magnitude faster than serial evaluation on a CPU.
     We then focus on three applications where efficient and differentiable waveforms are essential.
-    Firstly, we demonstrate how stochastic gradient descent can be used to optimize the $\sim 200$ coefficients that \sout{make up the waveform} \AZ{are used to calibrate the waveform model}. 
+    Firstly, we demonstrate how stochastic gradient descent can be used to optimize the $\sim 200$ coefficients that are used to calibrate the waveform model. 
     In particular, we demonstrate that the typical \textit{match} with numerical relatively waveforms can be improved by more than 10\% without any additional overhead.
@@
 EOB waveforms on the other hand are produced by evolving the dynamics of an effective one body Hamiltonian, and are therefore more difficult to implement in \jax.
-For EOB waveforms, frequency-domain reduced-order models may be a convenient target.
+For EOB waveforms, frequency-domain reduced-order models may be a convenient target~\citep[][e.g.]{Cotesta:2020qhw}.
 % to produce a waveform

### 2023-01-20 src/tex/ms.tex
Edits to sec 2

--- 2023-01-20 src/tex/ms.tex
@@
 \end{equation}
-where $S_n$ is the (one-sided) noise power spectral density (PSD) and $h_{\theta_1}$ is the frequency domain waveform evaluated at the intrinsic parameter point $\theta_1$.
+where $S_n$ is the (one-sided) noise power spectral density (PSD) and \AZ{$h_1$ and $h_2$ are the frequency domain waveforms which are to be compared.}
+\sout{ evaluated at the intrinsic parameter point $\theta_1$.}
 We can then normalize the inner product through
 \begin{equation}
-    \left[h_{\theta_1}|h_{\theta_2}\right] = \frac{\left(h_{\theta_1}|h_{\theta_2}\right)}{\sqrt{\left(h_{\theta_1}|h_{\theta_1}\right)\left(h_{\theta_2}|h_{\theta_2}\right)}}\, .
+    \left[h_1|h_2\right] = \frac{\left(h_1|h_2\right)}{\sqrt{\left(h_1|h_1\right)\left(h_2|h_2\right)}}\, .
 \end{equation}
@@
 \begin{equation}
-    \mathrm{m}(h_{\theta_1}|h_{\theta_2}) \equiv \max_{t_c,\, \phi_c} \left[h_{\theta_1} \, \middle| \, h_{\theta_2} \right]\,
+    \mathrm{m}(h_1|h_2) \equiv \max_{\Delta t_c,\, \Delta \phi_c} \left[h_1 \, \middle| \, h_2 \right]\,
 \end{equation}
-where $t_c$ and $\phi_c$ are, respectively, the time and phase of coalesence.
+where \AZ{$\Delta t_c$ and $\Delta \phi_c$ are, respectively, the differences in time and phase of coalescence between the two waveforms.}
 

### 2023-01-20 src/tex/ms.tex
Edits sectoion 2

@@
 \AZ{Say what PSD was used and reference it.}
-This is shown in Fig.~\ref{fig:match}, where we have calculated the $\mathrm{m}(h_{\theta_1}|\bar{h}_{\theta_1})$ across the entire parameter space.\footnote{
-    Specifically, we use $10^4$ points in the ranges $m_{1,2} = (1,100)\,M_{\odot}$ and $\chi_{1,2} = (-1,1)$.
+This is shown in Fig.~\ref{fig:match}, where we have calculated the $\mathrm{m}(h_1|h_2)$ across the entire parameter space.\footnote{
+    Specifically, we use $10^4$ points \AZ{varying component masses $m_1$, $m_2$ and spin parameters $\chi_1$, $\chi_2$} in the ranges $m_{1,2} = (1,100)\,M_{\odot}$ and $\chi_{1,2} = (-1,1)$.
 In addition, we evaulated the waveforms on a frequency grid from $32\,$Hz to $1024\,$Hz with frequency spacing $\Delta f = 0.0125 \,$Hz.
 }
-Here, $h_{\theta_1}$ corresponds to the \ripple waveform implementation and $\bar{h}_{\theta_1}$ is the \lalsuite implementation. 
+\AZ{Here, $h_1$ corresponds to the \ripple waveform implementation and $h_2$ is the \lalsuite implementation, evaluated at the same point in parameter space.} 
 From Fig.~\ref{fig:match}, it is clear that the \ripple waveform matches with the \lalsuite waveform close to numerical precision across the entire parameter space.
 \AZ{strictly $=0$? That seems impossible. Fmla has wrong labels for $h$'s}
-In particular, the gray points at low masses are where $1-\mathrm{m}(h_{\theta_1}|h^{*}_{\theta_1})=0$ causing points to not appear in the log scale colorbar.
+In particular, the gray points at low masses are where $1-\mathrm{m}(h_1|h2)=0$ causing points to not appear in the log scale colorbar.
 % In particular, the blue points indicate parameter points where the $1-\mathrm{m}(h_{\theta_1}|h^{*}_{\theta_1})=0$ and would therefore not appear in the log scale colorbar.
@@
 
-Note that in General Relativity (GR) the total mass $M$ of \AZ{a binary black hole simply serves as an overall scale for the system, so that the frequency evolution can be trivially rescaled.
+Note that in General Relativity (GR) the total mass $M = m_1 + m_2$ of \AZ{a binary black hole simply serves as an overall scale for the system, so that the frequency evolution can be trivially rescaled.
 The total mass only impacts the match because the chosen PSD and frequency limits fix a reference scale.}

### 2023-01-20 src/tex/ms.tex
Edits Sec 2

@@
 Waveform calls can therefore be evaluated in parallel, massively improving performance.
-Performing the same benchmark as above on a google colab notebook with the freely provided GPU, we find that on average waveform evaluations take $\sim \sout{0.00005} 5 \times 10^{-5}\mathrm{ms}$, nearly five orders of magnitude faster.
+Performing the same benchmark as above on a Google Colab notebook with the freely provided GPU, we find that on average waveform evaluations take $\sim \sout{0.00005} 5 \times 10^{-5}\mathrm{ms}$, nearly five orders of magnitude faster.
 Generalizing \lalsuite waveforms to run on a GPU would be a significant undertaking.
@@
 Since the parameter space of GWs in GR has $\mathcal{O}(10)$ dimensions, the speed of derivative evaluation is less crucial than the stability.
-However, this might change for waveforms in beyond GR models where many more parameters are added.
+However, this might change for waveforms in beyond-GR models, \AZ{waveforms involving equation of state parameters for neutron stars, and models which account for calibration and waveform uncertainties.
+In these cases} many more parameters \AZ{can be} added.
 

### 2023-01-20 src/tex/ms.tex
Edits Section 3

@@
 GW, such as searching for signals and estimating source parameters given data.
-While waveforms generated using numerical relativity simulations are commonly
-regarded as the most accurate, they are too computationally expensive to be used
+While waveforms generated using numerical relativity simulations are \sout{commonly
+regarded as the most accurate,} 
+\AZ{in principle the highest fidelity signal model,}
+they are too computationally expensive to be used
 in any practical data analysis tasks. To create waveform models that can be used
@@
 fitting all coefficients, the correlation in fitting individual and connecting
-different segments can be accounted during the optimization. Therefore, there is
+different segments can be accounted for during the optimization. Therefore, there is
 potential improvement of the accuracy by fitting all coefficients at once.
@@
 \AZ{Maybe comment on optimization method used for EOB waveforms?
-I think they do interesting Monte Carlo etc}
+I think they do interesting Monte Carlo etc. Check what method is used 
+for IMRPhenomXPHM?}
 

### 2023-01-20 src/tex/ms.tex
Working on comments

@@
 
-\AZ{The complexity begins even before data taking, since GW searches using the matched-filtering technique~\citep{Owen:1998dk, Owen:1995tm} require the generation of large banks of template waveforms.}
-\sout{The complexity begins before data gathering where one is required to generate banks of template waveforms which will be used for a matched-filter search~\citep{Owen:1998dk, Owen:1995tm}. }
+The complexity begins even before data taking, since GW searches using the matched-filtering technique~\citep{Owen:1998dk, Owen:1995tm} require the generation of large banks of template waveforms.
+% \sout{The complexity begins before data gathering where one is required to generate banks of template waveforms which will be used for a matched-filter search~\citep{Owen:1998dk, Owen:1995tm}. }
 Once potential candidates are found, parameter estimation (PE) is performed to extract the detailed source properties of each event~\citep{Christensen:2022bxb, 2020MNRAS.493.3132S, Ashton:2018jfp, Romero-Shaw:2020owr, Veitch:2014wba, Biwer:2018osg}.
-In the most minimal example \AZ{of non-precessing binaries}, this requires a Markov Chain Monte Carlo (MCMC) on an 11 dimensional parameter space for each event.
+In the most minimal example of non-precessing binaries, this requires a Markov Chain Monte Carlo (MCMC) on an 11 dimensional parameter space for each event.
 \AZ{Maybe modify; some other points are precessing PE requires 15 params, tidal prec 17 params, plus calibration marge and EoS inferences can incorporate even more}
-Beyond this simple scenario, more complex waveform models with \sout{many} additional parameters may be used to test for deviations from General Relativity~\citep{Krishnendu:2021fga, LIGOScientific:2020tif, LIGOScientific:2016lio, Yunes:2016jcc, Arun:2006yw, Agathos:2013upa}.
-\AZ{Add PPE and maybe TIGER refs..?}
+Beyond this simple scenario, more complex waveform models with additional parameters may be used to test for deviations from General Relativity~\citep{Krishnendu:2021fga, LIGOScientific:2020tif, LIGOScientific:2016lio, Yunes:2016jcc, Arun:2006yw, Agathos:2013upa}.
+% \AZ{Add PPE and maybe TIGER refs..?}
 Finally, using the results of PE, population synthesis models constrain the progenitors systems from which the black holes we see merging today began their journey~\citep{LIGOScientific:2020kqk, LIGOScientific:2021psn, Wong:2022flg}.
@@
 Derivatives are ubiquitously useful throughout data analysis tasks.
-For instance, during PE, derivative information can be used to guide the sampling towards regions with higher likelihood values \dfm{This isn't correct.
-In HMC the gradients don't have the behavior.
-I'd just say something like ``gradients are used to improve performance''} (e.g.~in Hamiltonian Monte Carlo~\citep{2017arXiv170102434B} or Gradient Descent~\citep{2016arXiv160904747R}).
+For instance, during PE, gradients are used to improve performance of the samplers 
+% \dfm{This isn't correct. In HMC the gradients don't have the behavior. I'd just say something like ``gradients are used to improve performance''} 
+(e.g.~in Hamiltonian Monte Carlo~\citep{2017arXiv170102434B} or Gradient Descent~\citep{2016arXiv160904747R}).
 Gradients are particularly valuable for high dimensional spaces.
@@
 Numerical derivatives also suffer from accuracy issues stemming from rounding or truncation errors.
-However, recent progress in automatic differentiation (AD) has shown promise in allowing general, fast derivative calculations for gravitational waveforms\AZ{, with applications to constructing template banks~\citep{Coogan:2022qxs} and computing the Fisher information for forecasting~\citep{Iacovelli:2022bbs, Iacovelli:2022mbg}}.
+However, recent progress in automatic differentiation (AD) has shown promise in allowing general, fast derivative calculations for gravitational waveforms, with applications to constructing template banks~\citep{Coogan:2022qxs} and computing the Fisher information for forecasting~\citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.
 
@@
     \label{eq:fisher}
-    \mathcal{I}^{k}_{ij} = (\partial_i h_0 | \partial_j h_0) \, ,
+    \mathcal{I}^{k}_{ij} = (\partial_i h^k | \partial_j h^k) \, ,
 \end{equation}
-where $k$ indicates the detector, $\partial_i = \partial/\partial \theta_i$, and $h_0$ is the strain measured by the detector which is given by,
+where $k$ indicates the detector, $\partial_i = \partial/\partial \theta_i$, and $h$ is the strain measured by the detector which is given by,
 \begin{equation}
-    h_0(\theta) = F_+(\lambda) h_{+}(\phi) + F_\times(\lambda) h_{\times}(\phi) \, .
+    h^k(\theta) = F_+(\phi) h^k_{+}(\Xi) + F_\times(\phi) h^k_{\times}(\Xi) \, .
 \end{equation}
-Note that here we have separated out the intrinsic ($\phi$) and extrinsic ($\lambda$) variables as well as introducing the detector projection functions for the plus and cross polarizations as $F_+$ and $F_\times$ respectively.
+Note that here we have separated out the intrinsic ($\phi$) and extrinsic ($\Xi$) variables as well as introducing the detector projection functions for the plus and cross polarizations as $F^k_+$ and $F^k_\times$ respectively.
 Since we are considering a three detector setup we simply add the Fisher matrices from the individual detectors to get the combined Fisher matrix:
@@
 
-To illustrate the computational speed of computing Fisher matrices with AD, we consider a population of binaries and compute the sky localization error following Eq.~(28) in \citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.
+To illustrate the computational speed of computing Fisher matrices with AD, we consider a population of binaries and compute the sky localization error following Eq.~(28) in \cite{Iacovelli:2022bbs, Iacovelli:2022mbg}.
 Since the Fisher matrix approach is known to have both theoretical issues as well as numerical instabilities for low signal-to-noise events, we restrict our population to only nearby systems.
@@
 \texttt{GWbench}~\citep{Borhanian:2020ypi}, on the other hand, takes $\mathcal{O}$(minutes) for each Fisher calculation using the same detector setup and frequency grid.
-This factor of over 100 speed up is substantial considering the fact that a single core evaluation of the \ripple waveform is slower than the \lalsuite call.
-As discussed above, performance can be further improved by utilising hardware acceleration such as parallel GPU processing.
+This factor of over 100 speed up is substantial considering the fact that a single core evaluation of the \ripple waveform is slower than the \lalsuite call \AZ{used by \texttt{GWbench}}.
+As discussed above, performance can be further improved by utilizing hardware acceleration such as parallel GPU processing.
 AD therefore represents a fast and accurate way of performing population level analyses, and should be utilized for testing the capabilities of next generation detectors.
@@
 
-A primary issue with both MCMC and nested sampling is that neither utlizes information about the likelihood's derivative and must therefore randomly walk towards areas of highest likelihood.
+A primary issue with both MCMC and nested sampling is that neither utilizes information about the likelihood's derivative and must therefore randomly walk towards areas of highest likelihood.
 Derivative based samplers, on the other hand, have been shown to extrapolate well to higher dimensions although they sometimes come with their own drawbacks.

### 2023-01-20 src/tex/ms.tex
edited most of intro

--- 2023-01-20 src/tex/ms.tex
@@
 % Title
-\title{\ripple: Differentiable \dfm{and Hardware-Accelerated} Waveforms for Gravitational Wave Data Analysis}
+\title{\ripple: Differentiable and Hardware-Accelerated Waveforms for Gravitational Wave Data Analysis}
 
@@
 There are now a variety of AD implementations, most notably in deep learning frameworks such as \texttt{pytorch}~\citep{pytorch} and \texttt{tensorflow}~\citep{tensorflow2015-whitepaper}.
-More general frameworks exist in \texttt{julia}~\citep{zygote, forwarddiff}, although \texttt{julia}'s limited \sout{support} \AZ{use} in GW \sout{physics} \AZ{analysis software} precludes its general use.
+More general frameworks exist in \texttt{julia}~\citep{zygote, forwarddiff}, although \texttt{julia}'s limited use in GW analysis software precludes its general use.
 Here we make use of \jax~\citep{jax2018github} due to its easy integration with \texttt{python} libraries and its just-in-time (JIT) compiler which allows the same code to run on a variety of hardware accelerators (i.e.~GPUs).
@@
 There are a variety of gravitational waveforms currently used in analysis pipelines.
-They are generally structured into different families, the most common of which are: the effective-one-body (EOB), the phenomenological inspiral-merger-ringerdown (IMRPhenom), and numerical relativity surrogate (NRsurrogate).
+They are generally structured into different families, the most common of which are: the effective-one-body (EOB)~\citep{Damour:2008yg,Buonanno:2005xu, Buonanno:2000ef, Buonanno:1998gg, Damour:2000we}, the phenomenological inspiral-merger-ringerdown (IMRPhenom), and numerical relativity surrogate (NRsurrogate)~\citep{Varma:2018mmi,Varma:2019csw}.
 %Of these, the IMRPhenom family is most well suited for an AD implementation in \jax.
-\AZ{Of these, the IMRPhenom family serves as a natural starting point for an AD implementation in \jax.
-Models like the non-precessing IMRPhenomD~\citep{Khan:2015jqa} model studied here and the precessing, higher-mode model IMRPhenomXPHM~\citep{Pratten:2020fqn, Pratten:2020ceb} are written in frequency domain using closed-form expressions.
-This makes a}
-\sout{In particular, their closed form expressions and fixed size frequency grids make a }
+Of these, the IMRPhenom family serves as a natural starting point for an AD implementation in \jax.
+Models like the non-precessing IMRPhenomD~\citep{Khan:2015jqa} model studied here and the precessing, higher-mode model IMRPhenomXPHM~\citep{Pratten:2020fqn, Pratten:2020ceb} are written in the frequency domain using closed-form expressions.
+This makes a
+% \sout{In particular, their closed form expressions and fixed size frequency grids make a }
 \jax implementation that complies with the constraints of JIT compilation simple.

### 2023-01-20 src/tex/ms.tex
Edits and ref

--- 2023-01-20 src/tex/ms.tex
@@
 EOB waveforms on the other hand are produced by evolving the dynamics of an effective one body Hamiltonian, and are therefore more difficult to implement in \jax.
-For EOB waveforms, frequency-domain reduced-order models may be a convenient target~\citep[][e.g.]{Cotesta:2020qhw}.
+For EOB waveforms, frequency-domain reduced-order models may be a convenient target~\citep[e.g.][]{Cotesta:2020qhw}.
 % to produce a waveform
@@
 
-The implementation of IMRPhenomD in \lalsuite is in C, and therefore needs to be rewritten natively into \texttt{pyhton} to be compatible with \jax.
-We have re-written IMRPhenomD from scratch using a combination of pure \texttt{pyhton} and \jax derivatives.
+The implementation of IMRPhenomD in \lalsuite is in C, and therefore needs to be rewritten natively into \texttt{python} to be compatible with \jax.
+We have re-written IMRPhenomD from scratch using a combination of pure \texttt{python} and \jax derivatives.
 In addition, we have restructured the code for readability and evaluation speed as well as exposing the internal fitting coefficients to the user (which we will use later in Sec.~\ref{sec:applications}).
@@
 \end{equation}
-Note that here we have separated out the intrinsic ($\phi$) and extrinsic ($\Xi$) variables as well as introducing the detector projection functions for the plus and cross polarizations as $F^k_+$ and $F^k_\times$ respectively.
+Note that here we have separated out the extrinsic ($\phi$) and intrinsic ($\Xi$) variables as well as introducing the detector projection functions for the plus and cross polarizations as $F^k_+$ and $F^k_\times$ respectively.
 Since we are considering a three detector setup we simply add the Fisher matrices from the individual detectors to get the combined Fisher matrix:
@@
 The distribution of sky localization errors from a population of $10^3$ binaries can be seen in Fig.~\ref{fig:sky_localization}.
-We have verified that our errors agree with a separate dedicated Fisher forecasting code~\citep{Borhanian:2020ypi} to within X percent.
-This demonstrates that AD can be used to accurately produce population level forecasts.
+We have verified that our errors agree with a separate dedicated Fisher forecasting code~\citep{Borhanian:2020ypi} to within {\color{red}{X}} percent.
+This demonstrates that AD can be used to accurately produce population-level forecasts.
 
@@
 For example, sampling the 15 dimensional parameter space for a BBH system can take $\mathcal{O}(10)$ hours, while BNS systems can take up to weeks.
-Dedicated fast samplers have been designed to get approximate posteriors on the sky localization to faciliate follow-up electromagnetic observations.
+Dedicated fast samplers have been designed to get approximate posteriors on the sky localization to facilitate follow-up electromagnetic observations \citep[e.g.~\texttt{BAYESTAR}][]{Singer:2015ema}.
 Nevertheless, these do not present the whole picture; fast, general parameter estimation therefore remains a primary goal of GW data analysis.

### 2023-01-20 src/tex/ms.tex
Comments and edits

@@
 The discovery of gravitational waves (GWs)~\citep{LIGOScientific:2016aoc} from inspiraling and merging compact objects (COs) has revolutionized our understanding of both fundamental physics and astronomy~\citep[e.g.][]{LIGOScientific:2021djp,LIGOScientific:2021sio,LIGOScientific:2021psn}.
-Although the data volumes from GW detectors are relatively small, analyzing the data is a computationally demanding task.
+Although the data volumes from GW detectors 
+\AZ{such as Advanced LIGO~\citep{LIGOScientific:2014pky} and Virgo~\citep{VIRGO:2014yos}} 
+are relatively small, analyzing the data is a computationally demanding task.
 In addition, this computational cost will substantially increase when next generation detectors come online~\citep{Maggiore:2019uih, Reitze:2019iox, Evans:2021gyd}.
@@
 Dedicated fast samplers have been designed to get approximate posteriors on the sky localization to facilitate follow-up electromagnetic observations \citep[e.g.~\texttt{BAYESTAR}][]{Singer:2015ema}.
+\AZ{We should be careful here.
+Many new algorithms have sped this a lot.
+Cogwheel can do fast PE, DINGO is a normalizing flow code that is very fast when trained.
+There are likelihood acceleration methods like ROQ, multibanding, relative binning/heterodyning that accelerate the process compared to the numbers quoted above.}
 Nevertheless, these do not present the whole picture; fast, general parameter estimation therefore remains a primary goal of GW data analysis.
@@
 For our basic example we perform an injection recovery test on a seven dimensional parameter space with the same three detector network setup considered in \S~\ref{subsec:fisher}.
-More specifically, we generate Gaussian noise consistent with the measured PSDs for each detector and then inject a BBH signal with parameters: $M_c = 23.82\,\mathrm{M_\odot}$,  $\eta = 0.248$, $\chi_1=0.3$, $\chi_1=-0.4$, $D = 1.6\,\mathrm{Gpc}$, $t_c = 0.0$, and $\phi_c = 0.0$.\footnote{
-    The remaining parameters are set to $\pi/3$.
+\AZ{As before, say what PSD and cite.}
+More specifically, we generate Gaussian noise consistent with the measured PSDs for each detector and then inject a BBH signal with parameters: chirp mass $M_c = 23.82\,\mathrm{M_\odot}$,  symmetric mass ratio $\eta = 0.248$, primary spin parameter $\chi_1=0.3$, secondary spin parameter \AZ{$\chi_2=-0.4$}, luminosity distance $D = 1.6\,\mathrm{Gpc}$, coalescence time $t_c = 0.0$, and coalescence phase $\phi_c = 0.0$.\footnote{
+    The remaining parameters are set to $\pi/3$. \AZ{Can we say what they are? $\iota$, $\psi$, $\alpha$, $\delta$?}
 }
-Using a standard Gaussian likelihood, we then run the HMC sampler implemented in \flowMC for 200000 steps and the Gaussian random walk (GRW) sampler for 1500000 steps (each with four randomly initialized independent chains).
+Using a standard Gaussian likelihood, we then run the HMC sampler implemented in \flowMC for \sout{200000} $2\times10^5$ steps and the Gaussian random walk (GRW) sampler for \sout{1500000} $1.5\times10^6$ steps (each with four randomly initialized independent chains).
+\AZ{What implementation of GRW? Can we cite something?}
 The number of steps and and mass matrix used for each example was hand tuned to give good performance for the specific sampler.
@@
 This is expected since the injected binary is relatively nearby.
+\AZ{Can we give the SNR?}
 
@@
 In a follow up paper we will demonstrate that minute scale PE can be achieved by combining normalizing flows~\citep{2022arXiv221106397W, Gabrie:2021tlu}, GPU acceleration, and a derivative based sampler~\citep{PEpaper}.
-We therefore expect \jax waveforms to be crucial to future PE efforts in GW astronomy, especially for low-latency pipelines.
+We therefore expect \jax waveforms to be \sout{crucial} \AZ{highly beneficial} to future PE efforts in GW astronomy, especially for low-latency pipelines.
 
@@
 Some progress has been made towards the construction of a differentiable NRsurrogate, but currently it seems difficult to implement EOB waveforms in \jax.
-In particular, the evolution of the Hamiltonian required to evaluate an EOB waveform is both inherently slow to differentiate and difficult to implement in \jax which requires fixed length arrays for its JIT compilation.
-Since EOB waveforms are often regarded as the state-of-the-art, more work is required to see if a fast, differentiable implementation is possible.
+In particular, the evolution of the Hamiltonian required to evaluate an EOB waveform is both inherently slow to differentiate and difficult to implement in \jax. 
+\sout{which requires fixed length arrays for its JIT compilation.}
+Since EOB methods are used to produce state-of-the-art waveforms for many applications,
+\sout{are often regarded as the state-of-the-art,} 
+more work is required to see if a fast, differentiable implementation is possible.
 
@@
 In order to showcase the benefits of differentiable waveforms as quickly as possible, at the time of writing, we have only implemented an aligned spin GW model (IMRPhenomD).
-We plan on adding a variety of different waveforms to \ripple in the near future with the primary goal of reaching a \jax version of a fully precessing, higher order mode waveform such as IMRPhenomXPHM.
+We plan on adding a variety of different waveforms to \ripple in the near future with the primary goal of reaching a \jax version of a fully precessing, higher order mode waveform such as IMRPhenomXPHM~\citep{Pratten:2020ceb}.
 Ideally, future waveforms should be implemented either in pure python or using \jax directly.
@@
 \section{Acknowledgments}
+This material is based upon work supported by NSF's LIGO Laboratory which is a major facility fully funded by the National Science Foundation.
 This work was supported by collaborative visits funded by the Cosmology and Astroparticle Student and Postdoc Exchange Network (CASPEN). 
 T.E. is supported by the Horizon Postdoctoral Fellowship.
+A.Z.~is supported by NSF Grant PHY-2207594.
 

### 2023-01-20 src/tex/ms.tex

@@
     As more detections continue, data analysis tasks are becoming increasingly computationally demanding.
+    \kw{I would cut everything before out.}
     Here, we propose the use of automatic differentiation through the programming framework \jax for accelerating a variety of these analysis tasks.
@@
 In the most minimal example of non-precessing binaries, this requires a Markov Chain Monte Carlo (MCMC) on an 11 dimensional parameter space for each event.
+\kw{This is not quite true. You can always marginalize over more parameters if you want, such as time of coalescence and phase. I would just mention the typical case of 15 parameters GR PE instead.}
 \AZ{Maybe modify; some other points are precessing PE requires 15 params, tidal prec 17 params, plus calibration marge and EoS inferences can incorporate even more}
@@
 Finally, using the results of PE, population synthesis models constrain the progenitors systems from which the black holes we see merging today began their journey~\citep{LIGOScientific:2020kqk, LIGOScientific:2021psn, Wong:2022flg}.
+\kw{Need more reference and some rewording}
 All of these tasks require significant computation.
+\kw{This could be triggering to some people, since population analysis could be quite cheap}
 In this paper, we will argue that differentiable waveforms (and more generally differentiable pipelines) can play a significant role in alleviating this computational demand.
@@
 Unfortunately, in the field of GW data analysis, analytic derivatives of the necessary quantities (such as the likelihood) have historically been difficult to obtain~\citep{Keppel:2013kia}.
+\kw{This needs some rewording.
+For example, I think in IMRPhenomD they do have the analytical derivatives coded up, at least for a part of the waveforms.
+Instead of difficult to have the derivatives, I would rather mention it is clumpsy to get the derivatives, especially given the model complexity is growing.}
 Numerical derivatives also suffer from accuracy issues stemming from rounding or truncation errors.
@@
 Performing the same benchmark as above on a Google Colab notebook with the freely provided GPU, we find that on average waveform evaluations take $\sim \sout{0.00005} 5 \times 10^{-5}\mathrm{ms}$, nearly five orders of magnitude faster.
+\kw{We should add some more detail related to this comparison.
+We should be clear that each individual waveform sample evaluation doesn't get us the performance by putting it on GPU (Jitting helps with that, not the GPU part).
+The GPU comes into play by allowing parallel evaluation of waveform sample, so for a tiny amount of time/frequency samples, we should expect GPU to lose to a CPU, but on any reasonable amount of time/frequency samples, GPU win a lot.}
 Generalizing \lalsuite waveforms to run on a GPU would be a significant undertaking.
@@
 }
+\kw{Maybe a bit more background on AD, the way we phrase it here sounds a bit too magical too me now.}
 Since the parameter space of GWs in GR has $\mathcal{O}(10)$ dimensions, the speed of derivative evaluation is less crucial than the stability.
@@
 This factor of over 100 speed up is substantial considering the fact that a single core evaluation of the \ripple waveform is slower than the \lalsuite call \AZ{used by \texttt{GWbench}}.
+\kw{Want to pull a number on the total time it took on generating the population result? I imagine this should turn the theorists on and send them into a fishing frenzy.}
 As discussed above, performance can be further improved by utilizing hardware acceleration such as parallel GPU processing.
@@
 We therefore expect \jax waveforms to be \sout{crucial} \AZ{highly beneficial} to future PE efforts in GW astronomy, especially for low-latency pipelines.
+\kw{Not just low latency, but also high dimensional analysis, such as testing GR and lensing.}
 

### 2023-01-20 src/tex/ms.tex
merging adams comments

@@
 \newcommand{\dfm}[1]{{\color{dfm}[DFM: #1]}}
+\newcommand{\amc}[1]{{\color{red}[AC: #1]}}
 
@@
     Secondly, we show that Fisher forecasting calculations can be sped up by $\sim 100\times$ (on a CPU) with no loss in accuracy.
-    This increased speed makes population forecasting substantially simpler, as discussed extensively in \citet{Iacovelli:2022bbs, Iacovelli:2022mbg}.
+    This increased speed makes population forecasting substantially simpler.
+    % , as discussed extensively in \citet{Iacovelli:2022bbs, Iacovelli:2022mbg}.
     Finally, we show that gradient-based samplers like Hamiltonian Monte Carlo lead to significantly reduced autocorrelation values when compared to traditional Monte Carlo methods.
     Since differentiable waveforms appear to have substantial advantages for a variety of tasks throughout GW science, we propose that waveform developers use \jax to build new waveforms moving forward.
-    Our waveform code, \ripple, can be found at \href{https://github.com/tedwards2412/ripple}{github.com/tedwards2412/ripple}, and will continue to be updated with new waveforms as they are formulated.
+    Our waveform code, \ripple, can be found at \href{https://github.com/tedwards2412/ripple}{github.com/tedwards2412/ripple}, and will continue to be updated with new waveforms as they are implemented.
 
@@
 Derivatives are ubiquitously useful throughout data analysis tasks.
-For instance, during PE, gradients are used to improve performance of the samplers 
-% \dfm{This isn't correct. In HMC the gradients don't have the behavior. I'd just say something like ``gradients are used to improve performance''} 
-(e.g.~in Hamiltonian Monte Carlo~\citep{2017arXiv170102434B} or Gradient Descent~\citep{2016arXiv160904747R}).
+% For instance, during PE, gradients are used to improve performance of the samplers 
+% % \dfm{This isn't correct. In HMC the gradients don't have the behavior. I'd just say something like ``gradients are used to improve performance''} 
+% (e.g.~in Hamiltonian Monte Carlo~\citep{2017arXiv170102434B} or Gradient Descent~\citep{2016arXiv160904747R}).
+For instance, during PE, derivative information can be used to guide an optimizer towards higher-likelihood values (e.g.\ using gradient descent~\citep{2016arXiv160904747R}) or allow a sampler to rapidly explore parameter space (e.g.\ using Hamiltonian Monte Carlo~\citep{2011hmcm.book..113N,2017arXiv170102434B}).
 Gradients are particularly valuable for high dimensional spaces.
@@
 
-Automatic differentiation is a family of methods used to compute machine precision derivatives with little computational overhead. 
+Automatic differentiation is a family of methods used to compute machine-precision derivatives with little computational overhead. 
 AD's recent ascendance is primarily driven by its use in machine learning, particularly for derivative computations of neural networks which use gradient descent during training.
@@
 More general frameworks exist in \texttt{julia}~\citep{zygote, forwarddiff}, although \texttt{julia}'s limited use in GW analysis software precludes its general use.
-Here we make use of \jax~\citep{jax2018github} due to its easy integration with \texttt{python} libraries and its just-in-time (JIT) compiler which allows the same code to run on a variety of hardware accelerators (i.e.~GPUs).
+Here we make use of \jax~\citep{jax2018github} due to its easy integration with \texttt{python} libraries, seamless support for running code on different hardware accelerators (e.g.\ graphical processing units) and its just-in-time (JIT) compiler, which can substantially accelerate code.
 
@@
 Firstly, we illustrate how the fit coefficients that form part of the IMRPhenom waveform models could be improved by high dimensional fitting enabled by a differentiable waveform. 
-Secondly, we implement differentiable detector response functions and show that the speed of Fisher matrix calculations can be substantially improved using AD.
+Secondly, we implement differentiable detector response functions and show that the speed of Fisher matrix calculations can be substantially accelerated using AD.
 Finally, we run an illustrative injection example using Hamiltonian Monte Carlo to demonstrate that the autocorrelation of derivative samplers is substantially smaller than that of traditional MCMCs.
+Practically, this translates into much faster PE with no sacrifice in accuracy.
 The associated code can be found at \ripple~\citep{ripple}.
@@
 When the COs are relatively well separated, the dynamics of the system can be well approximated by a post-Newtonian expansion.
-However, close to merger, numerical relatively simulations are required to accurately model the binary.
+However, close to merger, numerical relatively (NR) simulations are required to accurately model the binary.
 Unfortunately, these numerical simulations are computationally expensive and cannot be run in conjunction with data analysis.
-Approximate, phenomenological waveforms have therefore been constructued to enable relatively fast waveform generation at sufficient accuracy.
+Approximate, phenomenological waveforms have therefore been constructed to enable relatively fast waveform generation at sufficient accuracy.
 
 As mentioned in the previous section, the three major waveform families that have been developed to date are:  EOB,  NRsurrogate, and IMRPhenom.
-Note that unlike NRsurrogate, both the EOB and PhenomIMR waveforms are calibrated to numerical relativity waveforms.
-EOB waveforms require one to model the binary system using a Hamiltonian and are typically slow to evaluate whereas IMRPhenom waveforms are constructed with simple closed-form expressions.
-IMRPhenom waveforms are therefore ideally suited for AD, espeically using \jax. 
-In this paper we focus on the aligned spin circular orbit model IMRPhenomD~\citep{Husa:2015iqa, Khan:2015jqa}.
+Note that unlike NRsurrogate, both the EOB and IMRPhenom waveforms must be calibrated to numerical relativity waveforms.
+EOB waveforms require one to model the binary system using a Hamiltonian and are typically slow to evaluate, whereas IMRPhenom waveforms are constructed with simple closed-form expressions.
+IMRPhenom waveforms are therefore ideally-suited for AD, especially using \jax. 
+In this paper we focus on the aligned-spin, circular-orbit model IMRPhenomD~\citep{Husa:2015iqa, Khan:2015jqa}.
 
@@
 Waveform calls can therefore be evaluated in parallel, massively improving performance.
-Performing the same benchmark as above on a Google Colab notebook with the freely provided GPU, we find that on average waveform evaluations take $\sim \sout{0.00005} 5 \times 10^{-5}\mathrm{ms}$, nearly five orders of magnitude faster.
+Performing the same benchmark as above on a Google Colab notebook with a freely-provided NVIDIA Tesla T4 GPU (far from the current state-of-the-art), we find that on average waveform evaluations take $\sim \sout{0.00005} 5 \times 10^{-5}\mathrm{ms}$, nearly five orders of magnitude faster.
 \kw{We should add some more detail related to this comparison.
@@
 The GPU comes into play by allowing parallel evaluation of waveform sample, so for a tiny amount of time/frequency samples, we should expect GPU to lose to a CPU, but on any reasonable amount of time/frequency samples, GPU win a lot.}
+\amc{mention \texttt{vmap} and/or \texttt{pmap} -- super trivial to parallelize! More work in C.}
 Generalizing \lalsuite waveforms to run on a GPU would be a significant undertaking.
@@
 AD provides two big advantages when it comes to evaluating derivatives compared to numerical differentiation.
-First, the accuracy of derivatives from AD are significantly more stable than finite difference methods.
+First, the accuracy of derivatives from AD are significantly more stable than finite-difference methods.
 In particular, finite differences suffer from both rounding and truncation errors, meaning that the user is required to \textit{tune} the width over which the difference is taken.
-On the other hand, AD produces machine precision derivatives with no tuning.
+On the other hand, AD produces machine-precision derivatives with no tuning.
 Second, AD scales favorably with the dimensionality of the function.
 In particular, for every input dimension, $D$, added one would need to evaluate the function at least $2D$ times to calculate finite difference derivatives for all input parameters.
-For reverse mode AD, one only needs two function calls to evaluate the derivative of all input parameters, regardless of dimenison.\footnote{
-    Note that although the number of function calls is small, reverse mode AD does add memory overhead.
+For reverse-mode AD, one only needs two function calls to evaluate the derivative of all input parameters, regardless of dimension.\footnote{
+    Note that although the number of function calls is small, reverse-mode AD does add memory overhead.
     We've not found this to be limiting in any of the situations tested so far.
@@
 phases of the binary inspiral}. 
+\amc{redundant -- remove or merge with earlier material.}
 
@@
 Previously in the construction of IMRPhenomD \citep{Khan:2015jqa}, waveform
-coefficients are fitted in segments with amplitude and phase separated.
+coefficients were fitted in segments with amplitude and phase separated.
 Furthermore, IMRPhenomD is divided in three fitting segments, inspiral, merger and ringdown.
 Each of these segments has their own set of fitting coefficients.
-After obtaining the fitting coefficient for individual segments, they are then
-``stitched'' together such that they are continuous in the first derivative.
-The process of stitching introduces some addition
+After obtaining the fitting coefficient for individual segments, they are then ``stitched'' together such that they are continuous in the first derivative.
+The process of stitching introduces some additional
 inaccuracy in the waveform model, as the connections would affect the originally
 fitted segments.
-The coefficients of the current IMRPhenomD implementation are tuned in sub-sets
-of parameters instead of all together. This means the tuning process ignore the
-correlation between different sub-sets of parameters, so the best-fitted
-solution obtained by fitting sub-sets of parameters may not be the global
+
+The coefficients of the current IMRPhenomD implementation are tuned in subsets
+of parameters instead of all together. This means the tuning process ignores the
+correlation between different subsets of parameters, so the best-fit
+solution obtained by fitting subsets of parameters may not be the global
 optimum compared to fitting all coefficients at once. By jointly
@@
 I'd sat just leave it at ``We terminate the program after 30000 steps.''}
+\amc{say something to explain why this is working. I.e.: does the loss initially drop and then plateau?}
 % Add description of the optimization procedure.
@@
 performance in different regions of the intrinsic parameter space.
+\amc{this means your fit generally will depend on how the training waveforms are distributed over parameter space.}
 
@@
 The Fisher matrix approach is built around the assumption of a Gaussian likelihood.
-Although in practice this assumption is often violated in real GW datasets, the results obtained using a Fisher analysis can provide quick and useful diagnostics in evaluating sensitivities for a variety of models and detector configurations.
+Although in practice this assumption is often violated for realistic detector noise, the results obtained using a Fisher analysis can provide quick and useful diagnostics in evaluating sensitivities for a variety of models and detector configurations.
 
-Computing the Fisher matrix requires one to evaluate derivatives of the likelihood which in turn involves derivatives of the waveform model and detector projection functions.
+Computing the Fisher matrix requires one to evaluate derivatives of the likelihood, which in turn involves derivatives of the waveform model and detector projection functions.
 AD is therefore perfectly suited for computing Fisher matrices accurately and efficiently. 
@@
 Here we purely want to illustrate the simplicity and speed of AD for forecasting rather than providing new physics insights.
-We therefore consider a simple, three detector setup corresponding to the two LIGO detectors in addition to Virgo.
+We therefore consider a simple, three-detector setup corresponding to the two LIGO detectors in addition to Virgo.
 
-The Fisher information matrix for a single detector is typically given by 
+The Fisher information matrix for a single detector is typically given by \amc{I'm guessing in practice you compute a Hessian, so maybe write that out?}
 \begin{equation}
@@
 
+\amc{most acronyms were already introduced.}
 After the search algorithms have constructed a list of confidently detected binaries, the next step is to sample from the posterior of each sources parameters - so called PE.
-To do this, one typically uses a Markov Chain Monte Carlo (MCMC) or nested sampler~\citep{multinest, dynesty}.
+To do this, one typically uses a Markov Chain Monte Carlo (MCMC) or nested sampler~\citep{Skilling2004, multinest, dynesty}.
 Although robust, both MCMC and nested sampling are slow to converge and are known to perform poorly in high dimensional parameter spaces.
@@
 There are likelihood acceleration methods like ROQ, multibanding, relative binning/heterodyning that accelerate the process compared to the numbers quoted above.}
-Nevertheless, these do not present the whole picture; fast, general parameter estimation therefore remains a primary goal of GW data analysis.
+Nevertheless, these do not present the whole picture; fast, general PE therefore remains a key aim of GW data analysis
 
@@
 \AZ{What implementation of GRW? Can we cite something?}
+\amc{briefly summarize and give refs for flowMC and GRW. Is GRW Metropolis-Hastings with a Gaussian transition kernel? How'd you tune this? Why not use NUTS for HMC?}
 The number of steps and and mass matrix used for each example was hand tuned to give good performance for the specific sampler.
+\amc{not a huge deal, but why hand tune? There are automated ways to do this in e.g.\ numpyro with a warm-up adaptation phase.}
 The additional steps for the GRW sampler were required to achieve a similarly converged posterior.
+\amc{Why not plot its posterior?}
 
@@
 % Since we are running the chains from the same initialization, this enhanced acceptance for HMC can be mainly associated with use of gradient information to guide sampling.
-Although further steps would be required to achieve a fully converged posterior, these chains are sufficient to illustrate the increased efficiency associated with HMC. 
-To illustrate this, in Fig.~\ref{fig:autocorrelation} we plot the autocorrelation as a function of lag for both the HMC and GRW samplers.
+Although further steps would be required to achieve a fully-converged posterior, these chains are sufficient to show the increased efficiency associated with HMC. 
+To further illustrate this, in Fig.~\ref{fig:autocorrelation} we plot the autocorrelation as a function of lag for both the HMC and GRW samplers.
 The HMC autocorrelation is substantially lower than that of the GRW.
 We therefore expect gradient based samplers to converge significantly faster than typical samplers, especially in higher dimensions.
-In addition, we found that the effictive number of samples (a measure of the number of indepedent samples) is between 2 and 7 times larger for HMC across the different dimensions of the parameter space.
+In addition, we found that the effictive number of samples 
+\amc{provide ref?} 
+(a measure of the number of indepedent samples) is between 2 and 7 times larger for HMC across the different dimensions of the parameter space.
 
@@
 
-Currently the biggest constraint to adopting differentiable waveforms is the need to re-write the most commonly used waveforms into \jax (or pure python).
+Currently the biggest constraint to adopting differentiable waveforms is the need to rewrite the most commonly used waveforms into \jax (or pure python).
 In order to showcase the benefits of differentiable waveforms as quickly as possible, at the time of writing, we have only implemented an aligned spin GW model (IMRPhenomD).
@@
 Ideally, future waveforms should be implemented either in pure python or using \jax directly.
+\amc{if they're pure python, some translation will be required due to jax's restrictions.}
 This would ensure that the community can easily utilize differentiability and hardware acceleration in the future.
@@
 This work was supported by collaborative visits funded by the Cosmology and Astroparticle Student and Postdoc Exchange Network (CASPEN). 
-T.E. is supported by the Horizon Postdoctoral Fellowship.
+T.E.\ is supported by the Horizon Postdoctoral Fellowship.
 A.Z.~is supported by NSF Grant PHY-2207594.
+A.~C.\ acknowledges funding from the Schmidt Futures Foundation.
 

### 2023-01-20 src/tex/ms.tex
Dealing with comments in intro and abstract

@@
 \begin{abstract}
-    Since their discovery in 2015, gravitational waves (GWs) have developed into a unique observational probe used throughout fundamental and astro-phyics.
-    As more detections continue, data analysis tasks are becoming increasingly computationally demanding.
-    \kw{I would cut everything before out.}
-    Here, we propose the use of automatic differentiation through the programming framework \jax for accelerating a variety of these analysis tasks.
+    % Since their discovery in 2015, gravitational waves (GWs) have developed into a unique observational probe used throughout fundamental and astro-phyics.
+    % As more detections continue, data analysis tasks are becoming increasingly computationally demanding.
+    % \kw{I would cut everything before out.}
+    We propose the use of automatic differentiation through the programming framework \jax for accelerating a variety of analysis tasks throughout gravitational wave (GW) science.
     Firstly, we demonstrate that complete waveforms which cover the inspiral, merger, and ringdown of binary black holes (i.e.~IMRPhenomD) can be written in \jax and demonstrate that the serial evaluation speed of the waveform (and its derivative) is similar to the \lalsuite implementation in \CC.
@@
     We then focus on three applications where efficient and differentiable waveforms are essential.
-    Firstly, we demonstrate how stochastic gradient descent can be used to optimize the $\sim 200$ coefficients that are used to calibrate the waveform model. 
+    Firstly, we demonstrate how gradient descent can be used to optimize the $\sim 200$ coefficients that are used to calibrate the waveform model. 
     In particular, we demonstrate that the typical \textit{match} with numerical relatively waveforms can be improved by more than 10\% without any additional overhead.
@@
 The discovery of gravitational waves (GWs)~\citep{LIGOScientific:2016aoc} from inspiraling and merging compact objects (COs) has revolutionized our understanding of both fundamental physics and astronomy~\citep[e.g.][]{LIGOScientific:2021djp,LIGOScientific:2021sio,LIGOScientific:2021psn}.
-Although the data volumes from GW detectors 
-\AZ{such as Advanced LIGO~\citep{LIGOScientific:2014pky} and Virgo~\citep{VIRGO:2014yos}} 
-are relatively small, analyzing the data is a computationally demanding task.
+Although the data volumes from GW detectors such as Advanced LIGO~\citep{LIGOScientific:2014pky} and Virgo~\citep{VIRGO:2014yos} are relatively small, analyzing the data is a computationally demanding task.
 In addition, this computational cost will substantially increase when next generation detectors come online~\citep{Maggiore:2019uih, Reitze:2019iox, Evans:2021gyd}.
@@
 Once potential candidates are found, parameter estimation (PE) is performed to extract the detailed source properties of each event~\citep{Christensen:2022bxb, 2020MNRAS.493.3132S, Ashton:2018jfp, Romero-Shaw:2020owr, Veitch:2014wba, Biwer:2018osg}.
-In the most minimal example of non-precessing binaries, this requires a Markov Chain Monte Carlo (MCMC) on an 11 dimensional parameter space for each event.
-\kw{This is not quite true. You can always marginalize over more parameters if you want, such as time of coalescence and phase. I would just mention the typical case of 15 parameters GR PE instead.}
-\AZ{Maybe modify; some other points are precessing PE requires 15 params, tidal prec 17 params, plus calibration marge and EoS inferences can incorporate even more}
+For binary black holes with non-aligned spins, this requires a Markov Chain Monte Carlo (MCMC) on an 15 dimensional parameter space.
+More general examples can lead to a significant increase in dimension.\te{add citations}
+% In the most minimal example of non-precessing binaries, this requires a Markov Chain Monte Carlo (MCMC) on an 11 dimensional parameter space for each event.
+% \kw{This is not quite true. You can always marginalize over more parameters if you want, such as time of coalescence and phase. I would just mention the typical case of 15 parameters GR PE instead.}
+% \AZ{Maybe modify; some other points are precessing PE requires 15 params, tidal prec 17 params, plus calibration marge and EoS inferences can incorporate even more}
 Beyond this simple scenario, more complex waveform models with additional parameters may be used to test for deviations from General Relativity~\citep{Krishnendu:2021fga, LIGOScientific:2020tif, LIGOScientific:2016lio, Yunes:2016jcc, Arun:2006yw, Agathos:2013upa}.
@@
 Finally, using the results of PE, population synthesis models constrain the progenitors systems from which the black holes we see merging today began their journey~\citep{LIGOScientific:2020kqk, LIGOScientific:2021psn, Wong:2022flg}.
-\kw{Need more reference and some rewording}
-All of these tasks require significant computation.
-\kw{This could be triggering to some people, since population analysis could be quite cheap}
+\kw{Need more reference and some rewording}\te{added refs, happy?}
+% All of these tasks require significant computation.
+Overall, analyzing GWs therefore requires significant computation.
+% \kw{This could be triggering to some people, since population analysis could be quite cheap}
 In this paper, we will argue that differentiable waveforms (and more generally differentiable pipelines) can play a significant role in alleviating this computational demand.
@@
 Gradients are particularly valuable for high dimensional spaces.
-Unfortunately, in the field of GW data analysis, analytic derivatives of the necessary quantities (such as the likelihood) have historically been difficult to obtain~\citep{Keppel:2013kia}.
-\kw{This needs some rewording.
+Unfortunately, in the field of GW data analysis, analytic derivatives of the necessary quantities (such as the likelihood) have historically required a significant amount of work to obtain~\citep{Keppel:2013kia}.
+With waveforms only increasing in complexity, calculating analytic derivatives will become ever more tricky.
+% \kw{This needs some rewording.
 For example, I think in IMRPhenomD they do have the analytical derivatives coded up, at least for a part of the waveforms.
@@
 
-\AZ{Add refs here:}
+\AZ{Add refs here:}\te{Added a bunch, let me know if more are needed!}
 There are a variety of gravitational waveforms currently used in analysis pipelines.
@@
 In this paper we argue that differentiable waveforms will be a vital component for the future of GW data analysis.
-\AZ{I prefer to describe what is rather than what could be; can we say what elements exist, e.g.~
-we present \ripple... which includes \jax implementations of GW antenna patterns, the IMRPhenomD waveform, and...}
-In addition, we present \ripple, a small GW \texttt{python} package which can serve as a repository for differentiable implementations of some of the main waveforms currently used in LIGO and Virgo analyses.
+% \AZ{I prefer to describe what is rather than what could be; can we say what elements exist, e.g.~
+% we present \ripple... which includes \jax implementations of GW antenna patterns, the IMRPhenomD waveform, and...}
+% In addition, we present \ripple, a small GW \texttt{python} package which can serve as a repository for differentiable implementations of some of the main waveforms currently used in LIGO and Virgo analyses.
+In addition, we present \ripple, a small GW \texttt{python} package which, at the time of writing, includes a \jax implementation of the IMRPhenomD waveform and will be continually updated with new waveforms as they are implemented.
+% can serve as a repository for differentiable implementations of some of the main waveforms currently used in LIGO and Virgo analyses. 
 The remainder of this paper is structured as follows. 
@@
 
+\te{We use in total 11 waveforms for the optimization. Original PhenomD uses 19 @Kelvin add details of where to find this.}
+\te{Fig 2, plot was made after optimizing for ALL waveforms}
+\te{536 waveforms are subset of the SXS catalogue that are not precessing or eccentric. @Kelvin add details of where to find these things}
+\te{None of the 536 are part of the training set}
+\te{Used ALIGO design sensitivity for PSD}
+
 \AZ{Broad comments: need to cite which catalog the waveforms came from if any.
@@
 			Relative error between the numerical relatively and IMRPhenomD waveform amplitudes. 
-            The blue line shows the relative error of the IMRPhenomD model with the original model coefficients where as the orange line uses the IMRPhenomD model after optimization using stochastic gradient descent.
+            The blue line shows the relative error of the IMRPhenomD model with the original model coefficients where as the orange line uses the IMRPhenomD model after optimization using gradient descent.
             Interestingly, the error is reduced across the frequency range, demonstrating that high dimensional optimization is useful for all parts of the waveform.

### 2023-01-24 src/tex/ms.tex
Minor change
added MI comment cmd
some refs

@@
 \newcommand{\AZ}[1]{{\color{Burnt}[AZ: #1]}}
+\newcommand{\mi}[1]{\textbf{\color{teal}(MI: #1)}}
 
@@
     Moreover, \jax allows for GPU-accelerated waveform calls which can be over five orders of magnitude faster than serial evaluation on a CPU.
+    \te{Check and modify}
     We then focus on three applications where efficient and differentiable waveforms are essential.
@@
     Finally, we show that gradient-based samplers like Hamiltonian Monte Carlo lead to significantly reduced autocorrelation values when compared to traditional Monte Carlo methods.
-    Since differentiable waveforms appear to have substantial advantages for a variety of tasks throughout GW science, we propose that waveform developers use \jax to build new waveforms moving forward.
+    Since differentiable waveforms have substantial advantages for a variety of tasks throughout GW science, we propose that waveform developers use \jax to build new waveforms moving forward.
     Our waveform code, \ripple, can be found at \href{https://github.com/tedwards2412/ripple}{github.com/tedwards2412/ripple}, and will continue to be updated with new waveforms as they are implemented.
@@
 Once potential candidates are found, parameter estimation (PE) is performed to extract the detailed source properties of each event~\citep{Christensen:2022bxb, 2020MNRAS.493.3132S, Ashton:2018jfp, Romero-Shaw:2020owr, Veitch:2014wba, Biwer:2018osg}.
-For binary black holes with non-aligned spins, this requires a Markov Chain Monte Carlo (MCMC) on an 15 dimensional parameter space.
-More general examples can lead to a significant increase in dimension.\te{add citations}
+For binary black holes with non-aligned spins, this requires a Markov Chain Monte Carlo (MCMC) on a 15 dimensional parameter space.
+More general examples can lead to a significant increase in dimension.\te{add citations}  \mi{assuming you don't mean TGR, since it's mentioned below, do you mean calibration and tides?: for cal we can cite \cite{Farr:2014,Vitale:2020gvb}, and for tides maybe the BNS properties paper \cite{LIGOScientific:2018hze}, unless you want to cite a bunch of tidal waveform papers.}
 % In the most minimal example of non-precessing binaries, this requires a Markov Chain Monte Carlo (MCMC) on an 11 dimensional parameter space for each event.
@@
 Performing the same benchmark as above on a Google Colab notebook with a freely-provided NVIDIA Tesla T4 GPU (far from the current state-of-the-art), we find that on average waveform evaluations take $\sim \sout{0.00005} 5 \times 10^{-5}\mathrm{ms}$, nearly five orders of magnitude faster.
+\te{Check and modify}
 \kw{We should add some more detail related to this comparison.

### 2023-01-24 src/tex/ms.tex
bunch of updates

--- 2023-01-20 src/tex/ms.tex
@@
     We propose the use of automatic differentiation through the programming framework \jax for accelerating a variety of analysis tasks throughout gravitational wave (GW) science.
-    Firstly, we demonstrate that complete waveforms which cover the inspiral, merger, and ringdown of binary black holes (i.e.~IMRPhenomD) can be written in \jax and demonstrate that the serial evaluation speed of the waveform (and its derivative) is similar to the \lalsuite implementation in \CC.
-    Moreover, \jax allows for GPU-accelerated waveform calls which can be over five orders of magnitude faster than serial evaluation on a CPU.
+    Firstly, we demonstrate that complete waveforms which cover the inspiral, merger, and ringdown of binary black holes (i.e.~IMRPhenomD) can be written in \jax and demonstrate that the serial evaluation speed of the waveform (and its derivative) is similar to the \lalsuite implementation in \texttt{C}.
+    Moreover, \jax allows for GPU-accelerated waveform calls which can be over an order of magnitude faster than serial evaluation on a CPU.
     We then focus on three applications where efficient and differentiable waveforms are essential.
     Firstly, we demonstrate how gradient descent can be used to optimize the $\sim 200$ coefficients that are used to calibrate the waveform model. 
-    In particular, we demonstrate that the typical \textit{match} with numerical relatively waveforms can be improved by more than 10\% without any additional overhead.
+    In particular, we demonstrate that the typical \textit{match} with numerical relativity waveforms can be improved by more than 10\% without any additional overhead.
     Secondly, we show that Fisher forecasting calculations can be sped up by $\sim 100\times$ (on a CPU) with no loss in accuracy.
@@
 Once potential candidates are found, parameter estimation (PE) is performed to extract the detailed source properties of each event~\citep{Christensen:2022bxb, 2020MNRAS.493.3132S, Ashton:2018jfp, Romero-Shaw:2020owr, Veitch:2014wba, Biwer:2018osg}.
-For binary black holes with non-aligned spins, this requires a Markov Chain Monte Carlo (MCMC) on an 15 dimensional parameter space.
-More general examples can lead to a significant increase in dimension.\te{add citations}
+For binary black holes with non-aligned spins, this requires a Markov Chain Monte Carlo (MCMC) on a 15 dimensional parameter space.
+More general binary inspirals, such as those involving neutron stars, can lead to a significant increase in dimension.\te{add citations}
 % In the most minimal example of non-precessing binaries, this requires a Markov Chain Monte Carlo (MCMC) on an 11 dimensional parameter space for each event.
@@
 % \AZ{Maybe modify; some other points are precessing PE requires 15 params, tidal prec 17 params, plus calibration marge and EoS inferences can incorporate even more}
-Beyond this simple scenario, more complex waveform models with additional parameters may be used to test for deviations from General Relativity~\citep{Krishnendu:2021fga, LIGOScientific:2020tif, LIGOScientific:2016lio, Yunes:2016jcc, Arun:2006yw, Agathos:2013upa}.
+Beyond these simple scenarios, more complex waveform models with additional parameters may be used to test for deviations from General Relativity~\citep{Krishnendu:2021fga, LIGOScientific:2020tif, LIGOScientific:2016lio, Yunes:2016jcc, Arun:2006yw, Agathos:2013upa}.
 % \AZ{Add PPE and maybe TIGER refs..?}
@@
 % All of these tasks require significant computation.
-Overall, analyzing GWs therefore requires significant computation.
+Overall, GW data analysis therefore requires significant computation.
 % \kw{This could be triggering to some people, since population analysis could be quite cheap}
@@
 \jax implementation that complies with the constraints of JIT compilation simple.
-NRsurrogate models, which interpolate directly over waveforms produced by numerical relativity simulations, are in principle also straightfoward to implement in \jax.
+NRsurrogate models, which interpolate directly over waveforms produced by numerical relativity (NR) simulations, are in principle also straightfoward to implement in \jax.
 EOB waveforms on the other hand are produced by evolving the dynamics of an effective one body Hamiltonian, and are therefore more difficult to implement in \jax.
@@
 When the COs are relatively well separated, the dynamics of the system can be well approximated by a post-Newtonian expansion.
-However, close to merger, numerical relatively (NR) simulations are required to accurately model the binary.
+However, close to merger, NR simulations are required to accurately model the binary.
 Unfortunately, these numerical simulations are computationally expensive and cannot be run in conjunction with data analysis.
@@
 As mentioned in the previous section, the three major waveform families that have been developed to date are:  EOB,  NRsurrogate, and IMRPhenom.
-Note that unlike NRsurrogate, both the EOB and IMRPhenom waveforms must be calibrated to numerical relativity waveforms.
+Note that unlike NRsurrogate, both the EOB and IMRPhenom waveforms must be calibrated to NR waveforms.
 EOB waveforms require one to model the binary system using a Hamiltonian and are typically slow to evaluate, whereas IMRPhenom waveforms are constructed with simple closed-form expressions.
@@
 \end{equation}
-where $S_n$ is the (one-sided) noise power spectral density (PSD) and \AZ{$h_1$ and $h_2$ are the frequency domain waveforms which are to be compared.}
-\sout{ evaluated at the intrinsic parameter point $\theta_1$.}
+where $S_n$ is the (one-sided) noise power spectral density (PSD) and $h_1$ and $h_2$ are the frequency domain waveforms which are to be compared.
+% \sout{ evaluated at the intrinsic parameter point $\theta_1$.}
 We can then normalize the inner product through
@@
 \end{equation}
-where \AZ{$\Delta t_c$ and $\Delta \phi_c$ are, respectively, the differences in time and phase of coalescence between the two waveforms.}
+where $\Delta t_c$ and $\Delta \phi_c$ are, respectively, the differences in time and phase of coalescence between the two waveforms.
 
 Since the match is a measure of the difference between two waveforms, we can use it to demonstrate that the implementation of IMRPhenomD in \ripple accurately matches the \lalsuite implementation.
-\AZ{Say what PSD was used and reference it.}
+For this comparison, we use the $S_n$  presented in GWTC-2~\citep{LIGOScientific:2020ibl} from the Livingston detector (the most sensitive of the detectors).\footnote{
+    \url{https://dcc.ligo.org/LIGO-P2000251/public}.
+    }
 This is shown in Fig.~\ref{fig:match}, where we have calculated the $\mathrm{m}(h_1|h_2)$ across the entire parameter space.\footnote{
-    Specifically, we use $10^4$ points \AZ{varying component masses $m_1$, $m_2$ and spin parameters $\chi_1$, $\chi_2$} in the ranges $m_{1,2} = (1,100)\,M_{\odot}$ and $\chi_{1,2} = (-1,1)$.
+    Specifically, we use $10^4$ points varying component masses $m_1$, $m_2$ and spin parameters $\chi_1$, $\chi_2$ in the ranges $m_{1,2} = (1,100)\,M_{\odot}$ and $\chi_{1,2} = (-1,1)$.
 In addition, we evaulated the waveforms on a frequency grid from $32\,$Hz to $1024\,$Hz with frequency spacing $\Delta f = 0.0125 \,$Hz.
 }
-\AZ{Here, $h_1$ corresponds to the \ripple waveform implementation and $h_2$ is the \lalsuite implementation, evaluated at the same point in parameter space.} 
+Here, $h_1$ corresponds to the \ripple waveform implementation and $h_2$ is the \lalsuite implementation, evaluated at the same point in parameter space.
 From Fig.~\ref{fig:match}, it is clear that the \ripple waveform matches with the \lalsuite waveform close to numerical precision across the entire parameter space.
 \AZ{strictly $=0$? That seems impossible. Fmla has wrong labels for $h$'s}
-In particular, the gray points at low masses are where $1-\mathrm{m}(h_1|h2)=0$ causing points to not appear in the log scale colorbar.
+In particular, the gray points at low masses are where $1-\mathrm{m}(h_1|h_2)=0$ causing points to not appear in the log scale colorbar.
 % In particular, the blue points indicate parameter points where the $1-\mathrm{m}(h_{\theta_1}|h^{*}_{\theta_1})=0$ and would therefore not appear in the log scale colorbar.
@@
             It is clear that the \ripple waveform matches the \lalsuite implementation many of orders of magnitude more than necessary for all data analysis tasks across the entire parameter space.
+            \te{Current plot looking better?}
             \dfm{I think this figure could use some work.
@@
 
-Note that in General Relativity (GR) the total mass $M = m_1 + m_2$ of \AZ{a binary black hole simply serves as an overall scale for the system, so that the frequency evolution can be trivially rescaled.
-The total mass only impacts the match because the chosen PSD and frequency limits fix a reference scale.}
-\sout{the system is simply an overall scaling of the waveforms} 
-\sout{frequency evolution and is therefore not a} \sout{truly independent parameter.}
-If we instead use a flat PSD in Eq.~\ref{eq:inner_prod} and \AZ{rescale the frequency grid to units of $Mf$}, \sout{correctly,} all dependence with $M$ seen in Fig.~\ref{fig:match} vanishes. 
-\AZ{Revisit this after defining PSD above:} 
+Note that in General Relativity (GR) the total mass $M = m_1 + m_2$ of a binary black hole simply serves as an overall scale for the system, so that the frequency evolution can be trivially rescaled.
+The total mass only impacts the match because the chosen PSD and frequency limits fix a reference scale.
+% \sout{the system is simply an overall scaling of the waveforms} 
+% \sout{frequency evolution and is therefore not a} \sout{truly independent parameter.}
+If we instead use a flat PSD in Eq.~\ref{eq:inner_prod} and rescale the frequency grid to units of $Mf$, all dependence with $M$ seen in Fig.~\ref{fig:match} vanishes. 
+% \AZ{Revisit this after defining PSD above:} 
 This figure instead illustrates a more realistic PSD where at low masses the waveform signal-to-noise is dominated by the inspiral but at high masses it is dominated by the merger.
@@
 Maybe this is something we could implement...}
+\te{Its just one dimension, so is pretty easy.
+I tried the one from the cosmoJAX people and found that the match did slightly increase but it was a decent chunk slower (maybe a factor of 2 or so).
+So I just left it out.
+Would be happy to code up something and put it in if you think its necessary!}
 Instead, we initially use \texttt{scipy}'s cubic interpolator to create a fine grid of $5\times10^5$ values, which we then linearly interpolate during waveform generation.
@@
 % The remaining error in the waveform at high masses is therefore primarily down to the coarseness in this interpolation.
-Note however, that the differences are well below the accuracy requirements of the waveforms and will have no noticeable for realistic data analysis tasks.
+Note however, that the differences are well below the accuracy requirements of the waveforms and will have no noticeable effect for realistic data analysis tasks.
 
@@
 Fortunately, the IMRPhenom waveforms are constructed from simple closed-form expressions which are computationally efficient.
-Since the \lalsuite implementation of IMRPhenomD is written in \texttt{C}, the evaluation of the waveform itself is faster than \ripple.
-\dfm{I'd soften this statement a little bit.
+Even though the \lalsuite implementation of IMRPhenomD is written in \texttt{C}, the serial evaluation of the waveform in \ripple is comparably fast. 
+% \dfm{I'd soften this statement a little bit.
 In my own work, I often find that JAX implementations are just as fast or faster than C implementations, so I wouldn't state it that it's obvious that lalsuite should be faster!
@@
 Would it be easy for me to try them out?}
-Benchmarking on a MacBook Pro with an M1 Max Apple Silicon processor, we find that a single waveform evaluation takes $\sim 2-5\,\mathrm{ms}$ for \lalsuite (interfaced with python) and $\sim 5-8\,\mathrm{ms}$ for \ripple.\footnote{
-    For this benchmark we used $16\,\mathrm{Hz}$ and $1024\,\mathrm{Hz}$ for the lowest and highest frequencies respectively. 
-    In addition, we used a frequency spacing of $0.0125\,\mathrm{Hz}$. 
-    We performed this benchmark by evaluating the waveform $5000$ times and taking the average evaluation time.
+Benchmarking on a MacBook Pro with an M1 Max Apple Silicon processor, we find that a single waveform evaluation takes $\sim 0.5\,\mathrm{ms}$ for \lalsuite (interfaced with python) and $\sim 0.4\,\mathrm{ms}$ for \ripple.\footnote{
+    For this benchmark we used $24\,\mathrm{Hz}$ and $512\,\mathrm{Hz}$ for the lowest and highest frequencies respectively. 
+    In addition, we used a frequency spacing of $0.2\,\mathrm{Hz}$. 
+    We performed this benchmark by evaluating the waveform $10^{4}$ times and taking the average evaluation time.
 }
-Although slower on a CPU benchmark, a key advantage of using \jax is its native ability to run on a GPU.
-Waveform calls can therefore be evaluated in parallel, massively improving performance.
-Performing the same benchmark as above on a Google Colab notebook with a freely-provided NVIDIA Tesla T4 GPU (far from the current state-of-the-art), we find that on average waveform evaluations take $\sim \sout{0.00005} 5 \times 10^{-5}\mathrm{ms}$, nearly five orders of magnitude faster.
-\kw{We should add some more detail related to this comparison.
+Although similar on a CPU benchmark, \jax has a few key advantages. 
+First, its ability to JIT compile allows for significant performace gains (the above benchmark is already JIT compiled).
+Second, automatic vectorization can be achieved using the \texttt{vmap} function.
+Using \texttt{vmap} and performing the same benchmark as described above reduces the average evaluation speed to $\sim 0.14\,\mathrm{ms}$.
+Finally, \jax can natively on a GPU which allows for highly parallellized waveform evaluations.
+
+% Waveform calls can therefore be evaluated in parallel, massively improving performance.
+
+Again, performing the same benchmark as above on a NVIDIA Quadro 6000, we find that on average waveform evaluations take $\sim 0.02$, over an of magnitude faster than serial CPU evaluation.
+% \te{Check and modify}
+% \kw{We should add some more detail related to this comparison.
 We should be clear that each individual waveform sample evaluation doesn't get us the performance by putting it on GPU (Jitting helps with that, not the GPU part).
 The GPU comes into play by allowing parallel evaluation of waveform sample, so for a tiny amount of time/frequency samples, we should expect GPU to lose to a CPU, but on any reasonable amount of time/frequency samples, GPU win a lot.}
-\amc{mention \texttt{vmap} and/or \texttt{pmap} -- super trivial to parallelize! More work in C.}
+% \amc{mention \texttt{vmap} and/or \texttt{pmap} -- super trivial to parallelize! More work in C.}
 Generalizing \lalsuite waveforms to run on a GPU would be a significant undertaking.
 
-The primary argument of this paper is that waveform derivatives will also be highly valuable to data analysis tasks. 
+One of the primary argument of this paper is that waveform derivatives will also be highly valuable to data analysis tasks. 
 AD provides two big advantages when it comes to evaluating derivatives compared to numerical differentiation.
@@
 Since the parameter space of GWs in GR has $\mathcal{O}(10)$ dimensions, the speed of derivative evaluation is less crucial than the stability.
-However, this might change for waveforms in beyond-GR models, \AZ{waveforms involving equation of state parameters for neutron stars, and models which account for calibration and waveform uncertainties.
-In these cases} many more parameters \AZ{can be} added.
+However, this might change for waveforms in beyond-GR models, waveforms involving equation of state parameters for neutron stars, and models which account for calibration and waveform uncertainties.
+In these cases many more parameters can be added.
 
@@
 GW, such as searching for signals and estimating source parameters given data.
-While waveforms generated using numerical relativity simulations are \sout{commonly
+While waveforms generated using NR simulations are \sout{commonly
 regarded as the most accurate,} 
@@
 in data analysis, there are a number of groups creating ``approximants'' of the
-full numerical relativity waveforms, which are essentially simpler ansatzes that
+full NR waveforms, which are essentially simpler ansatzes that
 \sout{are only approximations but} can be evaluated much faster \AZ{and in regions not
@@
 the waveform generated by the IMRPhenomD and $h_{\mathrm{NR}}$ is the waveform
-generated by the numerical relativity simulation. 
+generated by the NR simulation. 
 Given the loss function, we use gradient descent to find the optimal fitting
@@
 		\caption{
-			Relative error between the numerical relatively and IMRPhenomD waveform amplitudes. 
+			Relative error between the numerical relativity and IMRPhenomD waveform amplitudes. 
             The blue line shows the relative error of the IMRPhenomD model with the original model coefficients where as the orange line uses the IMRPhenomD model after optimization using gradient descent.
@@
 In particular, we showed that it matches the \lalsuite implementation to near machine precision and can be easily parallelized on a GPU.
-Parallelization on a GPU provides substantial speed increases; on a free Google Colab GPU we found that waveform evaluations are nearly 5 orders of magnitude faster than serial CPU evaluations.
+Parallelization on a GPU provides substantial speed increases; on a NVIDIA Quadro 6000 GPU we found that waveform evaluations are over an order of magnitude faster than serial CPU evaluations.
 Second, we discussed three data analysis tasks which can all be substantially improved by utilizing derivative information of the waveform.
@@
 In particular, the evolution of the Hamiltonian required to evaluate an EOB waveform is both inherently slow to differentiate and difficult to implement in \jax. 
-\sout{which requires fixed length arrays for its JIT compilation.}
-Since EOB methods are used to produce state-of-the-art waveforms for many applications,
-\sout{are often regarded as the state-of-the-art,} 
-more work is required to see if a fast, differentiable implementation is possible.
+Since EOB methods are used to produce state-of-the-art waveforms for many applications, more work is required to see if a fast, differentiable implementation is possible.
 
@@
 We plan on adding a variety of different waveforms to \ripple in the near future with the primary goal of reaching a \jax version of a fully precessing, higher order mode waveform such as IMRPhenomXPHM~\citep{Pratten:2020ceb}.
-Ideally, future waveforms should be implemented either in pure python or using \jax directly.
-\amc{if they're pure python, some translation will be required due to jax's restrictions.}
+Ideally, future waveforms should be implemented either in \jax compatible python directly.
+% \amc{if they're pure python, some translation will be required due to jax's restrictions.}
 This would ensure that the community can easily utilize differentiability and hardware acceleration in the future.

### 2023-01-24 src/tex/ms.tex
comments throughout

--- 2023-01-24 src/tex/ms.tex
@@
 % \AZ{Maybe modify; some other points are precessing PE requires 15 params, tidal prec 17 params, plus calibration marge and EoS inferences can incorporate even more}
-Beyond this simple scenario, more complex waveform models with additional parameters may be used to test for deviations from General Relativity~\citep{Krishnendu:2021fga, LIGOScientific:2020tif, LIGOScientific:2016lio, Yunes:2016jcc, Arun:2006yw, Agathos:2013upa}.
+Beyond this simple scenario, more complex waveform models with additional parameters may be used to test for deviations from General Relativity~\citep{Arun:2006yw, Agathos:2013upa, Yunes:2016jcc, LIGOScientific:2016lio, LIGOScientific:2020tif, Krishnendu:2021fga}.
 % \AZ{Add PPE and maybe TIGER refs..?}
@@
 There are a variety of gravitational waveforms currently used in analysis pipelines.
-They are generally structured into different families, the most common of which are: the effective-one-body (EOB)~\citep{Damour:2008yg,Buonanno:2005xu, Buonanno:2000ef, Buonanno:1998gg, Damour:2000we}, the phenomenological inspiral-merger-ringerdown (IMRPhenom), and numerical relativity surrogate (NRsurrogate)~\citep{Varma:2018mmi,Varma:2019csw}.
+They are generally structured into different families, the most common of which are: the effective-one-body (EOB)~\citep{Damour:2008yg,Buonanno:2005xu, Buonanno:2000ef, Buonanno:1998gg, Damour:2000we}, the phenomenological inspiral-merger-ringerdown (IMRPhenom) \citep{Husa:2015iqa,Khan:2015jqa,Hannam:2013oca,Pratten:2020ceb}, and numerical relativity surrogate (NRsurrogate)~\citep{Blackman:2017pcm,Varma:2018mmi,Varma:2019csw}.
 %Of these, the IMRPhenom family is most well suited for an AD implementation in \jax.
@@
 Note that unlike NRsurrogate, both the EOB and IMRPhenom waveforms must be calibrated to numerical relativity waveforms.
+\mi{I think this is weird to say: the surrogate is fully trained on (``calibrated to'') NR}
 EOB waveforms require one to model the binary system using a Hamiltonian and are typically slow to evaluate, whereas IMRPhenom waveforms are constructed with simple closed-form expressions.
@@
 where \AZ{$\Delta t_c$ and $\Delta \phi_c$ are, respectively, the differences in time and phase of coalescence between the two waveforms.}
+\mi{I think it'd be good to define the mismatch $\mathcal{M} \equiv 1- \mathrm{m}$, since it's the standard quantity to discuss, and it's what we plot}
 
@@
 % In particular, the blue points indicate parameter points where the $1-\mathrm{m}(h_{\theta_1}|h^{*}_{\theta_1})=0$ and would therefore not appear in the log scale colorbar.
+\mi{to Aaron's point, maybe say ``0 up to numerical precision''?}
+
 \begin{figure}[t]
@@
 Right now it looks less impressive!}
+            \mi{I agree.
+Furthermore, it looks like there's a subpopulation of darker points hiding under the brighter ones up to and around $M \sim 100$?
+is there a dependence on the mass ratio?
+it'd be good to comment on this, and maybe show a corner plot adding $q$ to the mix.
+Also, $M$ is lacking units (I assume $M_\odot$, but it should be shown in the label)}
         }
@@
 This figure instead illustrates a more realistic PSD where at low masses the waveform signal-to-noise is dominated by the inspiral but at high masses it is dominated by the merger.
+\mi{it's important to say what PSD was used}
 
@@
 Second, AD scales favorably with the dimensionality of the function.
-In particular, for every input dimension, $D$, added one would need to evaluate the function at least $2D$ times to calculate finite difference derivatives for all input parameters.
+In particular, for every input dimension added, $D$, one would need to evaluate the function at least $2D$ times to calculate finite difference derivatives for all input parameters.
 For reverse-mode AD, one only needs two function calls to evaluate the derivative of all input parameters, regardless of dimension.\footnote{
@@
 
+\mi{this paragraph is redundant, I'd remove it}
 Having an accurate waveform model is essential for many data analysis tasks in
@@
 coefficients were fitted in segments with amplitude and phase separated.
+\mi{this previous sentence sounds odd: do we mean that amplitude and phase were fitted independently?}
 Furthermore, IMRPhenomD is divided in three fitting segments, inspiral, merger and ringdown.
@@
 for IMRPhenomXPHM?}
+\mi{I agree. it'd be good to comment on whether this is still done like this in
+modern waveforms.}
 
@@
 generated by the numerical relativity simulation. 
+\mi{I think the mismatch should be defined above, right after the match is defined. We should also state what PSD and sampling parameters are used}
 Given the loss function, we use gradient descent to find the optimal fitting
@@
 As a demonstration, Fig. \ref{fig:loss_compare} shows the relative error of
-original and optimized waveform at different frequencies. In this optimization
+the original and optimized waveforms at different frequencies \mi{for a particular set of parameters?}. In this optimization
 procedure, we took NR waveforms $h_{\mathrm{NR}}$ stated in \citep{Khan:2015jqa}
-as training waveforms to fit waveform coefficients. \amc{how many are there?}
-We can see the error of optimized waveform is lower than that of the original waveform for most of the domain.
+as training data to fit waveform coefficients. \amc{how many are there?}
+We can see the error of the optimized waveform is lower than that of the original waveform for most of the domain.
 In particular, the error in the early inspiral region is decreased by half while other regions also show good improvement in accuracy.
@@
 Is this just fitting a single waveform?}
+\mi{I was also confused by this}
 
@@
 coefficients to waveforms with different intrinsic parameters, such as mass
-ratio and spins. From Eq. \ref{eq:multi_loss}, we can see the averaging between
+ratio and spins. From Eq.~\eqref{eq:multi_loss}, we can see the averaging between
 waveforms with different intrinsic parameters implies there are trade-offs in
@@
 In Fig.~\ref{fig:mismatch_hist}, we show the distribution of log mismatches.
-One can see the distribution of mismatch after optimization has smaller mismatch compared to before optimization,
-showing the optimization improves the accuracy of the waveform model.
+\mi{for all parameters in the training set?}
+One can see the distribution of mismatch \mi{mean?} after optimization has smaller mismatch compared to before optimization,
+showing the optimization improves the accuracy of the waveform model \mi{on average?}.
 % However, it is not improving for every waveforms. Say why briefly, and say there is another paper coming out.
-While many waveforms improved their accuracy, some retain similar mismatches. 
+While many waveforms improved their accuracy, some retain similar mismatches \mi{``waveforms'' is overused and obfuscates this discussion: it's indistinctly used to refer to the NR data, the approximants in general, and specific sets of parameters---I think some rewriting is needed, e.g., here we could say ``While the mismatch generally improves for several parameter choices after training, it remains the same for others.''}.
 \AZ{I don't understand this claim. The spin degeneracy is present in the physics of the waveform; $\chi_{\rm eff}$ is just one way of parametrizing the spins that tries to reduce the degeneracy.}
@@
 
-Forecasting the sensitivity of future experiments is an essential task in GW science. \kw{LOL I won't say it is essential.}
-Due to its theoretical simplicity and evaluation speed, the Fisher matrix formalism is commonly deployed to estimate how well a binary system's parameters can be measured.
+Forecasting the sensitivity of future experiments is an essential task in GW science. \kw{LOL I won't say it is essential.} \mi{routine?}
+Due to its theoretical simplicity and evaluation speed, the Fisher matrix formalism is commonly deployed to estimate how well a binary system's parameters could be measured.
 The Fisher matrix approach is built around the assumption of a Gaussian likelihood.
@@
 \AZ{which proves forecasted measurement errors and parameter covariances for a signal with parameters $\theta$ observed by the given detector network}.
+\mi{I think Aaron means ``provides'' not ``proves''; also, we might add ``in the high SNR limit''}
 \AZ{Again we need to state and cite the PSDs we use for the three detectors.}
@@
 To illustrate the computational speed of computing Fisher matrices with AD, we consider a population of binaries and compute the sky localization error following Eq.~(28) in \cite{Iacovelli:2022bbs, Iacovelli:2022mbg}.
+\mi{should we say a word about computing FMs with multiple detectors?}
 Since the Fisher matrix approach is known to have both theoretical issues as well as numerical instabilities for low signal-to-noise events, we restrict our population to only nearby systems.
@@
 In Fig.~\ref{fig:corner}, the grey contours show the posterior recovered using the best chain (i.e. one that reached the highest log-likelihood values).
+\mi{why show a single chain?}
 The orange shows the true parameters of the injected signal.
@@
 In particular, we showed that it matches the \lalsuite implementation to near machine precision and can be easily parallelized on a GPU.
-Parallelization on a GPU provides substantial speed increases; on a free Google Colab GPU we found that waveform evaluations are nearly 5 orders of magnitude faster than serial CPU evaluations.
+Parallelization on a GPU provides substantial speed increases; on a free Google Colab GPU \mi{shouldn't we just say the name of the GPU (Tesla whatever)?
+why does it matter that it's accessed through google here?} we found that waveform evaluations are nearly 5 orders of magnitude faster than serial CPU evaluations.
 Second, we discussed three data analysis tasks which can all be substantially improved by utilizing derivative information of the waveform.
 Although we primarily discuss toy examples in this paper, each can be extended to the full data analysis task, some of which will be shown in upcoming papers~\citep{PEpaper}.
-Differentiable waveforms therefore represent a crucial tool to perform efficient GW science.
+Differentiable waveforms therefore represent a crucial advancement to perform efficient GW science.
 
@@
 Two other waveform families are commonly used in GW data analysis: the effective-one-body (EOB) and numerical relativity surrogate (NRSurrogate). 
-Some progress has been made towards the construction of a differentiable NRsurrogate, but currently it seems difficult to implement EOB waveforms in \jax.
+Some progress has been made towards the construction of a differentiable NRsurrogate \mi{cite our paper in prep?}, but currently it seems difficult to implement EOB waveforms in \jax.
 In particular, the evolution of the Hamiltonian required to evaluate an EOB waveform is both inherently slow to differentiate and difficult to implement in \jax. 
@@
 \amc{if they're pure python, some translation will be required due to jax's restrictions.}
+\mi{we could just say ``directly implemented under an autodifferentiation framework'' to be generic}
 This would ensure that the community can easily utilize differentiability and hardware acceleration in the future.

### 2023-01-24 src/tex/ms.tex

--- 2023-01-24 src/tex/ms.tex
@@
 For binary black holes with non-aligned spins, this requires a Markov Chain Monte Carlo (MCMC) on a 15 dimensional parameter space.
-More general binary inspirals, such as those involving neutron stars, can lead to a significant increase in dimension.\te{add citations}
+More general binary inspirals, such as those involving neutron stars, can lead to a significant increase in dimension.\te{add citations} \mi{assuming you don't mean TGR, since it's mentioned below, do you mean calibration and tides?: for cal we can cite \cite{Farr:2014,Vitale:2020gvb}, and for tides maybe the BNS properties paper \cite{LIGOScientific:2018hze}, unless you want to cite a bunch of tidal waveform papers.}
 % In the most minimal example of non-precessing binaries, this requires a Markov Chain Monte Carlo (MCMC) on an 11 dimensional parameter space for each event.
@@
 % \AZ{Maybe modify; some other points are precessing PE requires 15 params, tidal prec 17 params, plus calibration marge and EoS inferences can incorporate even more}
-Beyond these simple scenarios, more complex waveform models with additional parameters may be used to test for deviations from General Relativity~\citep{Krishnendu:2021fga, LIGOScientific:2020tif, LIGOScientific:2016lio, Yunes:2016jcc, Arun:2006yw, Agathos:2013upa}.
+Beyond these simple scenarios, more complex waveform models with additional parameters may be used to test for deviations from General Relativity~\citep{Arun:2006yw, Agathos:2013upa, Yunes:2016jcc, LIGOScientific:2016lio, LIGOScientific:2020tif, Krishnendu:2021fga}.
 % \AZ{Add PPE and maybe TIGER refs..?}
@@
 Note that unlike NRsurrogate, both the EOB and IMRPhenom waveforms must be calibrated to NR waveforms.
+\mi{I think this is weird to say: the surrogate is fully trained on (``calibrated to'') NR}
 EOB waveforms require one to model the binary system using a Hamiltonian and are typically slow to evaluate, whereas IMRPhenom waveforms are constructed with simple closed-form expressions.
@@
 where $\Delta t_c$ and $\Delta \phi_c$ are, respectively, the differences in time and phase of coalescence between the two waveforms.
+\mi{I think it'd be good to define the mismatch $\mathcal{M} \equiv 1- \mathrm{m}$, since it's the standard quantity to discuss, and it's what we plot}
 
@@
 generated by the NR simulation. 
+\mi{I think the mismatch should be defined above, right after the match is defined. We should also state what PSD and sampling parameters are used}
 Given the loss function, we use gradient descent to find the optimal fitting
@@
 Although we primarily discuss toy examples in this paper, each can be extended to the full data analysis task, some of which will be shown in upcoming papers~\citep{PEpaper}.
-Differentiable waveforms therefore represent a crucial tool to perform efficient GW science.
+Differentiable waveforms therefore represent a crucial advancement to perform efficient GW science.
 
@@
 % \amc{if they're pure python, some translation will be required due to jax's restrictions.}
+\mi{we could just say ``directly implemented under an autodifferentiation framework'' to be generic}
 This would ensure that the community can easily utilize differentiability and hardware acceleration in the future.

### 2023-01-24 src/tex/ms.tex
bunch of updates

@@
 \te{Used ALIGO design sensitivity for PSD}
+\te{@Kelvin, can you please add a citation to the 11 waveforms used for training as well as a citation to the 536 waveforms used from the SXS catalogue.
+I've indicated in the text where these should be added}
+\te{@Kelvin can you please also add a citation to the particular PSD used (again, its indicated in the text where to do this.)}
+\te{The thing we really want to compare is the mismatch with the rescaled frequency grid and a fixed flat PSD.}
+\te{Need to understand how the mass scale was introduced into the NR waveforms.}
 
-\AZ{Broad comments: need to cite which catalog the waveforms came from if any.
-The number of waveforms used seems much larger than the number originally used
-to calibrate PhenomD. 
-If that is true, we can't make a big deal about having a better mismatch compared to
-the original version, because of course optimizing over another set will improve 
-performance.
-Say how many waveforms we use to optimize, how many we compare to, where they came
-from.}
+% \AZ{Broad comments: need to cite which catalog the waveforms came from if any.
+% The number of waveforms used seems much larger than the number originally used
+% to calibrate PhenomD. 
+% If that is true, we can't make a big deal about having a better mismatch compared to
+% the original version, because of course optimizing over another set will improve 
+% performance.
+% Say how many waveforms we use to optimize, how many we compare to, where they came
+% from.}
 
-\mi{this paragraph is redundant, I'd remove it}
-Having an accurate waveform model is essential for many data analysis tasks in
-GW, such as searching for signals and estimating source parameters given data.
-While waveforms generated using NR simulations are \sout{commonly
-regarded as the most accurate,} 
-\AZ{in principle the highest fidelity signal model,}
-they are too computationally expensive to be used
-in any practical data analysis tasks. To create waveform models that can be used
-in data analysis, there are a number of groups creating ``approximants'' of the
-full NR waveforms, which are essentially simpler ansatzes that
-\sout{are only approximations but} can be evaluated much faster \AZ{and in regions not
-covered by numerical simulations, such as extremes of parameter space or earlier 
-phases of the binary inspiral}. 
-\amc{redundant -- remove or merge with earlier material.}
+Having an accurate waveform model is essential for many data analysis tasks throughout
+GW science.
+While waveforms generated using NR simulations are in principle the highest fidelity signal model, they are too computationally expensive to be used in any practical data analysis tasks.
+The community therefore utilizes waveform ``approximants'' (such as those discussed in the introduction) which can be evaluated much faster and in regions not covered by numerical simulations, such as extremes of parameter space or earlier phases of the binary inspiral.
+% \amc{redundant -- remove or merge with earlier material.}
 
 %\sout{Every waveform approximant has some}
-\AZ{Waveform approximants generally have free coefficients which are calibrated
-to NR waveforms to achieve high accuracy.}
-\sout{ fitting coefficients that are tuned during
-the process of calibrating the ansatz to NR waveforms.} In the case of
-IMRPhenomD, there are 209 fitting coefficients used to capture the separate
-behavior of the amplitude and phase as a function of the mass ratio and spins.
-\sout{The accuracy of fitting coefficients determines how well the approximant
-reproduces the NR waveform. 
-The}
-\AZ{Any} inaccuracy in obtaining the fitting coefficients
-means misrepresentation of the NR waveform, which can translate to systematic
-error in downstream data analysis task. For example, sufficiently large
+Waveform approximants generally have free coefficients which are calibrated
+to NR waveforms to achieve high accuracy.
+% \sout{ fitting coefficients that are tuned during
+% the process of calibrating the ansatz to NR waveforms.
+In the case of IMRPhenomD, there are 209 fitting coefficients used to capture the separate behavior of the amplitude and phase as a function of the mass ratio and spins.
+
+% \sout{The accuracy of fitting coefficients determines how well the approximant
+% reproduces the NR waveform. 
+% The}
+Any inaccuracy in obtaining the fitting coefficients
+leads to a misrepresentation of the NR waveform, which can translate to systematic
+error in downstream data analysis tasks. For example, sufficiently large
 systematic errors in the waveform would cause the recovered source parameters to be biased
@@
 Previously in the construction of IMRPhenomD \citep{Khan:2015jqa}, waveform
-coefficients were fitted in segments with amplitude and phase separated.
-\mi{this previous sentence sounds odd: do we mean that amplitude and phase were fitted independently?}
-Furthermore, IMRPhenomD is divided in three fitting segments, inspiral, merger and ringdown.
+coefficients for the amplitude and phase were fitted independently.
+Furthermore, IMRPhenomD is divided into three fitting segments: inspiral; merger; and ringdown.
 Each of these segments has their own set of fitting coefficients.
-After obtaining the fitting coefficient for individual segments, they are then ``stitched'' together such that they are continuous in the first derivative.
-The process of stitching introduces some additional
-inaccuracy in the waveform model, as the connections would affect the originally
-fitted segments.
+After obtaining the fitting coefficient for individual segments, they are then ``stitched'' together such that both the phase and amplitude are continuous in the first derivative.
+The process of stitching introduces some additional inaccuracy in the waveform model, as the connections affect the originally fitted segments.
 
-The coefficients of the current IMRPhenomD implementation are tuned in subsets
-of parameters instead of all together. This means the tuning process ignores the
-correlation between different subsets of parameters, so the best-fit
-solution obtained by fitting subsets of parameters may not be the global
-optimum compared to fitting all coefficients at once. By jointly
-fitting all coefficients, the correlation in fitting individual and connecting
-different segments can be accounted for during the optimization. Therefore, there is
-potential improvement of the accuracy by fitting all coefficients at once.
+The coefficients of the original IMRPhenomD implementation are tuned in subsets of parameters instead of all together. 
+This means the tuning process ignores the correlation between different subsets of parameters, so the provided best-fit solution may not be the global optimum. 
+We therefore aim to improve of the accuracy of the waveform by jointly fitting all coefficients at once.
 
@@
 
-In general, optimization problems in a high dimensional space benefit from
-having access to the gradient of the objective function. Since we can
-differentiate through the entire waveform model against the fitting parameters,
-one can use gradient descent to find the optimal fitting parameters all at once
-instead of fitting them in subsets.
+In general, optimization problems in a high dimensional space benefit from having access to the gradient of the objective function. 
+Since we can differentiate through the entire waveform model against the fitting parameters, one can use gradient descent to more efficiently find the local best fitting parameters.
 
-The first step in fitting all coefficients is to define a loss function that
-measures the goodness-of-fit of the optimization procedure, which we choose it
-to be the mismatch between the NR waveform and the approximant waveform:
+The first step is to define a loss function that measures the goodness-of-fit of the current waveform coefficients.
+Here, we choose it to be the mismatch between the NR waveform and the approximant waveform:
 \begin{equation}
     \label{eq:mismatch}
-	\mathcal{M}(\lambda)=1-\mathrm{m}(h_{\mathrm{IMR}}(\lambda)|h_{\mathrm{NR}}(\lambda)),
+	\mathcal{M}(\lambda)=1-\mathrm{m}(h_{\mathrm{IMR}}(\lambda)|h_{\mathrm{NR}}(\lambda))\,,
 \end{equation}
 where $\lambda$ is a vector of the fitting coefficients, $h_{\mathrm{IMR}}$ is
-the waveform generated by the IMRPhenomD and $h_{\mathrm{NR}}$ is the waveform
+the waveform generated by IMRPhenomD, and $h_{\mathrm{NR}}$ is the waveform
 generated by the NR simulation. 
 \mi{I think the mismatch should be defined above, right after the match is defined. We should also state what PSD and sampling parameters are used}
-Given the loss function, we use gradient descent to find the optimal fitting
-coefficients.
+\te{I think it makes it easier to define here cause we only now introduce the fitting coefficient $\lambda$.}
+Given the loss function, we use gradient descent to update the fitting
+coefficients:
 \begin{equation}
-	\lambda\leftarrow\lambda-\alpha\nabla\mathcal{M}, 
+	\lambda\leftarrow\lambda-\alpha\nabla\mathcal{M}\,, 
 \end{equation}
-where $\alpha$ is the learning rate.
-We set $\alpha$ to be a very small number, i.e.
-$10^{-6}$, since the initial guess already lies close to the desired optimum.
-We terminate the program after 30000 steps, in which $\lambda$ has converged to the optimum.
-\dfm{We can't possibly know that it has converged to ``the optimum''.
-I'd sat just leave it at ``We terminate the program after 30000 steps.''}
-\amc{say something to explain why this is working. I.e.: does the loss initially drop and then plateau?}
+where $\alpha$ is the learning rate. We set $\alpha$ to be a very small number, i.e. $10^{-6}$, since the initial guess already lies close to the desired optimum. 
+%   \dfm{We can't possibly know that it has converged to ``the optimum''. I'd sat just leave it at ``We terminate the program after 30000 steps.''}
+% \amc{say something to explain why this is working. I.e.: does the loss initially drop and then plateau?}
 % Add description of the optimization procedure.
 
-As a demonstration, Fig. \ref{fig:loss_compare} shows the relative error of
-the original and optimized waveforms at different frequencies \mi{for a particular set of parameters?}. In this optimization
-procedure, we took NR waveforms $h_{\mathrm{NR}}$ stated in \citep{Khan:2015jqa}
-as training data to fit waveform coefficients. \amc{how many are there?}
-We can see the error of the optimized waveform is lower than that of the original waveform for most of the domain.
-In particular, the error in the early inspiral region is decreased by half while other regions also show good improvement in accuracy.
-\amc{so this is a waveform in the training set?
-And I don't follow exactly what you did.
-Is this just fitting a single waveform?}
-\mi{I was also confused by this}
-
-We can generalize the fitting procedure described above to a collection of waveforms.
-In that case, the loss function is defined as the average of the mismatch of
-individual waveforms, given by
+To generalize the loss function to a collection of waveforms we use the average of the mismatch of individual waveforms, given by
 \begin{equation}
@@
 \end{equation}
-where $\mathcal{M}_i$ is the mismatch of individual training waveforms and $N$
-is the total number of training waveforms used in optimization. This
-optimization is more difficult since we are applying the same set of
-coefficients to waveforms with different intrinsic parameters, such as mass
-ratio and spins. From Eq.~\eqref{eq:multi_loss}, we can see the averaging between
-waveforms with different intrinsic parameters implies there are trade-offs in
-performance in different regions of the intrinsic parameter space.
-\amc{this means your fit generally will depend on how the training waveforms are distributed over parameter space.}
+where $\mathcal{M}_i$ is the mismatch of an individual training waveform and $N$ is the total number of training waveforms used in optimization.
+This optimization is more difficult since we are now applying the same set of coefficients to waveforms with different intrinsic parameters, such as mass ratio and spins. 
+From Eq.~\ref{eq:multi_loss}, we can see the averaging between waveforms with different intrinsic parameters implies there are trade-offs in performance for different regions of the intrinsic parameter space.
+Additionally, this means our best-fit points will generally depend on the distribution of training waveforms across the parameter space.
+% \amc{this means your fit generally will depend on how the training waveforms are distributed over parameter space.}
+
+For training, we use the publically available subset of the 19 waveforms used in the original IMRPhenomD paper (11 in total)\te{@kelvin add citation to waveforms here}. 
+We then run gradient descent, as described above, for $3\times 10^4$ steps. 
+Figure~\ref{fig:loss_compare} shows the relative error (against a particular NR waveform) of the original and optimized waveform at different frequencies.
+\te{@Kelvin is the NR waveform here part of the training set?}
+The vertical dashed lines indicate the stitching points for the phase i.e. when the inspiral is joined onto the merger.
+% In this optimization
+% procedure, we took NR waveforms $h_{\mathrm{NR}}$ stated in \citep{Khan:2015jqa}
+% as training waveforms to fit waveform coefficients. \amc{how many are there?}
+We can see the error of optimized waveform is lower than that of the original waveform for most of the domain.
+In particular, the error in the merger region is decreased by half while other regions also show good improvement in accuracy. 
+% \amc{so this is a waveform in the training set? And I don't follow exactly what you did. Is this just fitting a single waveform?}
+
+% \te{We terminate the program after 30000 steps.}
 
@@
 
-In Fig.~\ref{fig:mismatch_hist}, we show the distribution of log mismatches.
-\mi{for all parameters in the training set?}
-One can see the distribution of mismatch \mi{mean?} after optimization has smaller mismatch compared to before optimization,
-showing the optimization improves the accuracy of the waveform model \mi{on average?}.
+In Fig.~\ref{fig:mismatch_hist}, we show the distribution of log mismatches for 536 test waveforms.
+For testing we use 536 waveforms from the SXS catalogue (we simply choose all waveforms with aligned spins and eccentricity close to 0 \te{kelvin, please add specific eccentricity cut here})~\citep{Boyle:2019kee}.
+One can see that the distribution of mismatches after optimization is generally shifted to lower mismatch compared to the original waveform.
+% showing the optimization improves the accuracy of the waveform model.
 % However, it is not improving for every waveforms. Say why briefly, and say there is another paper coming out.
-While many waveforms improved their accuracy, some retain similar mismatches \mi{``waveforms'' is overused and obfuscates this discussion: it's indistinctly used to refer to the NR data, the approximants in general, and specific sets of parameters---I think some rewriting is needed, e.g., here we could say ``While the mismatch generally improves for several parameter choices after training, it remains the same for others.''}.
-\AZ{I don't understand this claim. The spin degeneracy is present in the physics of the waveform; $\chi_{\rm eff}$ is just one way of parametrizing the spins that tries to reduce the degeneracy.}
-This phenomenon is mainly due to the reduced spin approximation used in IMRPhenomD, where spin degeneracy is introduced, hence restricting further improvements. 
+% While many waveforms improved their accuracy, some retain similar mismatches. 
+% \AZ{I don't understand this claim. The spin degeneracy is present in the physics of the waveform; $\chi_{\rm eff}$ is just one way of parametrizing the spins that tries to reduce the degeneracy.}
+% This phenomenon is mainly due to the reduced spin approximation used in IMRPhenomD, where spin degeneracy is introduced, hence restricting further improvements. 
 In an upcoming paper, we will give a detailed discussion of how such an approximation scheme affects the optimization procedure. 
@@
 
-Forecasting the sensitivity of future experiments is an essential task in GW science. \kw{LOL I won't say it is essential.} \mi{routine?}
+Forecasting the sensitivity of future experiments is an routine task in GW science. 
 Due to its theoretical simplicity and evaluation speed, the Fisher matrix formalism is commonly deployed to estimate how well a binary system's parameters could be measured.

### 2023-01-24 src/scripts/loss_compare.py

@@ -2,6 +2,7 @@ import numpy as np
 import matplotlib.pyplot as plt
 import matplotlib as mpl
 import paths
+# from ripple.waveforms import IMRPhenomD_utils
 
 params = {
     "font.size": 18,
@@ -39,13 +40,29 @@ mpl.rcParams.update(params)
 
 
 plt.axvline(x=0.018, color="k", alpha=0.5, ls="--")
-# fRD =
-# 0.5fRD
-plt.axvline(x=0.08, color="k", alpha=0.4, ls="--")
+
+# theta = np.array([25.0, 25.0, -0.8, -0.8])
+
+# # Calculate the frequency regions
+# coeffs = IMRPhenomD_utils.get_coeffs(theta)
+# f1, f2, _, _, _, _ = IMRPhenomD_utils.get_transition_frequencies(
+#     theta, coeffs[5], coeffs[6]
+# )
+
+# print(f2)
+MSUN = 1.988409902147041637325262574352366540e30  # kg
+G = 6.67430e-11  # m^3 / kg / s^2
+C = 299792458.0  # m / s
+gt = G * MSUN / (C ** 3.0)
+
+f2 =  149.1018563859192 * (25. + 25.) * gt
+plt.axvline(x=f2, color="k", alpha=0.4, ls="--")
+
 plt.plot(freq, ori_error, label="Original")
 plt.plot(freq, opt_error, label="Optimized")
 plt.legend()
 plt.xlabel(r"$Mf$")
 plt.ylabel(r"$\Delta|\tilde{h}(f)|$")
-plt.xlim(0.01, 0.1)
+plt.xlim(0.01, 0.09)
+plt.ylim(0.0, 0.05)
 plt.savefig(paths.figures / "loss_compare.pdf", bbox_inches="tight")

### 2023-01-25 src/tex/ms.tex
fixing more comments

@@
 More general frameworks exist in \texttt{julia}~\citep{zygote, forwarddiff}, although \texttt{julia}'s limited use in GW analysis software precludes its general use.
-Here we make use of \jax~\citep{jax2018github} due to its easy integration with \texttt{python} libraries, seamless support for running code on different hardware accelerators (e.g.\ graphical processing units) and its just-in-time (JIT) compiler, which can substantially accelerate code.
+Here we make use of \jax~\citep{jax2018github} due to its easy integration with \texttt{python} libraries, seamless support for running code on different hardware accelerators,such as graphical processing units (GPUs), and its just-in-time (JIT) compiler, which can substantially accelerate code.
 
@@
 EOB waveforms on the other hand are produced by evolving the dynamics of an effective one body Hamiltonian, and are therefore more difficult to implement in \jax.
-For EOB waveforms, frequency-domain reduced-order models may be a convenient target~\citep[e.g.][]{Cotesta:2020qhw}.
+For EOB waveforms, frequency-domain reduced-order models may be a convenient target~\citep[e.g.]{Cotesta:2020qhw}.
 % to produce a waveform
@@
 This figure instead illustrates a more realistic PSD where at low masses the waveform signal-to-noise is dominated by the inspiral but at high masses it is dominated by the merger.
-\mi{it's important to say what PSD was used}
 
@@
 % This phenomenon is mainly due to the reduced spin approximation used in IMRPhenomD, where spin degeneracy is introduced, hence restricting further improvements. 
-In an upcoming paper, we will give a detailed discussion of how such an approximation scheme affects the optimization procedure. 
-Nevertheless, the peak of the original waveform distribution has moved to the left end of the distribution, indicating that \sout{the} \AZ{our AD-assisted} optimization procedure \AZ{provides an improved implementation of the model}.
-\sout{Therefore, the new set of waveform coefficients can produce better waveforms for GW analyses. }
+% In an upcoming paper, we will give a detailed discussion of how such an approximation scheme affects the optimization procedure. 
+In particular, the peak of the original waveform distribution has moved by nearly an order of magnitude, indicating that our AD-assisted optimization procedure provides an improved implementation of the model.
+% \sout{Therefore, the new set of waveform coefficients can produce better waveforms for GW analyses. }
 
@@
 Due to its theoretical simplicity and evaluation speed, the Fisher matrix formalism is commonly deployed to estimate how well a binary system's parameters could be measured.
-The Fisher matrix approach is built around the assumption of a Gaussian likelihood.
+The Fisher matrix approach is built around the assumption of a Gaussian likelihood~\citep{Vallisneri:2007ev}.
 Although in practice this assumption is often violated for realistic detector noise, the results obtained using a Fisher analysis can provide quick and useful diagnostics in evaluating sensitivities for a variety of models and detector configurations.
@@
 
-The Fisher information matrix for a single detector is typically given by \amc{I'm guessing in practice you compute a Hessian, so maybe write that out?}
+The Fisher information matrix for a single detector is typically given by 
+% \amc{I'm guessing in practice you compute a Hessian, so maybe write that out?}
 \begin{equation}
@@
 Finally, we invert the Fisher matrix to calculate the covariance matrix,
-\AZ{which proves forecasted measurement errors and parameter covariances for a signal with parameters $\theta$ observed by the given detector network}.
-\mi{I think Aaron means ``provides'' not ``proves''; also, we might add ``in the high SNR limit''}
+which provides forecasted measurement errors and parameter covariances for a signal with parameters $\theta$ observed by the given detector network in the high signal-to-noise limit.
+% \mi{I think Aaron means ``provides'' not ``proves''; also, we might add ``in the high SNR limit''}
 \AZ{Again we need to state and cite the PSDs we use for the three detectors.}
@@
     \hline\hline
-    $m_1, m_2$ & $\boldsymbol{U}[1.0, 50]\, \mathrm{M}_\odot$ \\ \hline
-    $\chi_1, \chi_2$ & [-0.99, 0.99] \\ \hline
-    $D$ & [500, 1000]\, Mpc \\ \hline
+
+    $m_1, m_2$ & $\boldsymbol{U}[20, 60]\, \mathrm{M}_\odot$ \\ \hline
+    $\chi_1, \chi_2$ & $\boldsymbol{U}[-0.8, 0.8]$ \\ \hline
+    $D$ & $\left(400 (\boldsymbol{U}[0,1]^{1/3}) + 600\right)$\, Mpc \\ \hline
     $t_c$ & 0.0 \\ \hline
     $\phi_c$ & 0.0 \\ \hline
-    Inclincation Angle, $\iota$ & $\boldsymbol{U}[0, 1]$ \\ \hline
-    Polarization Angle, $\psi$ & $\boldsymbol{U}[0, 1]$ \\ \hline
-    Right Ascension, $\delta$ & $\boldsymbol{U}[0, 1]$ \\ \hline
-    Declination, $\alpha$ & $\boldsymbol{U}[0, 1]$ \\ 
+    Inclincation Angle, $\iota$ & $\cos^{-1}(\boldsymbol{U}[-1, 1])$ \\ \hline
+    Polarization Angle, $\psi$ & $\boldsymbol{U}[0, 2\pi]$ \\ \hline
+    Right Ascension, $\alpha$ & $\boldsymbol{U}[0, 2\pi]$ \\ \hline
+    Declination, $\delta$ & $\pi/2 - \cos^{-1}(\boldsymbol{U}[-1, 1])$ \\ 
     \hline\hline
@@
     $\boldsymbol{U}$ indicates a uniform distribution between the two variables in the brackets. 
-    \AZ{$\cos \iota \in (-1,1)$, $\psi \in (0,2 \pi)$ (or $\pi$), RA is $\alpha \in (0, 2\pi)$, Dec is $\delta$, let $\vartheta = \pi/2 - \delta$, then $\cos (\vartheta) \in (-1,1)$ }
+    The distance prior ensures that the binaries are uniformly distirbuted in volume.
     }
@@
 To illustrate the computational speed of computing Fisher matrices with AD, we consider a population of binaries and compute the sky localization error following Eq.~(28) in \cite{Iacovelli:2022bbs, Iacovelli:2022mbg}.
-\mi{should we say a word about computing FMs with multiple detectors?}
 Since the Fisher matrix approach is known to have both theoretical issues as well as numerical instabilities for low signal-to-noise events, we restrict our population to only nearby systems.
@@
 The distribution of sky localization errors from a population of $10^3$ binaries can be seen in Fig.~\ref{fig:sky_localization}.
-We have verified that our errors agree with a separate dedicated Fisher forecasting code~\citep{Borhanian:2020ypi} to within {\color{red}{X}} percent.
+We have verified that our errors agree with a separate dedicated Fisher forecasting code~\citep{Borhanian:2020ypi} to within 20 percent.\te{Check number}
 This demonstrates that AD can be used to accurately produce population-level forecasts.
@@
 Moreover, each error calculation (including computing the Fisher matrices for each detector and the inversion process) is substantially faster.
-In particular, we find that after compilation each error calculation takes approximately half a second on a single computing core.
+In particular, we find that after compilation, each error calculation takes approximately half a second on a single computing core.
 \texttt{GWbench}~\citep{Borhanian:2020ypi}, on the other hand, takes $\mathcal{O}$(minutes) for each Fisher calculation using the same detector setup and frequency grid.
-This factor of over 100 speed up is substantial considering the fact that a single core evaluation of the \ripple waveform is slower than the \lalsuite call \AZ{used by \texttt{GWbench}}.
-\kw{Want to pull a number on the total time it took on generating the population result? I imagine this should turn the theorists on and send them into a fishing frenzy.}
+This factor of over 100 speed up is substantial considering the fact that a single core evaluation of the \ripple waveform is similar to \lalsuite which is used by \texttt{GWbench}.
+On a MacBook Pro with an M1 Max Apple Silicon processor, JIT compilation takes $\sim 270\,s$ and the full population analysis takes less $\sim 8$ minutes.
+% \kw{Want to pull a number on the total time it took on generating the population result? I imagine this should turn the theorists on and send them into a fishing frenzy.}
 As discussed above, performance can be further improved by utilizing hardware acceleration such as parallel GPU processing.

### 2023-01-25 src/scripts/random_matches.py

@@ -113,7 +113,7 @@ plt.colorbar(sc, label=r"$\log_{10}(1-\mathrm{Match})$")
 
 # plt.xlabel(r"Total Mass, $M$")
 # plt.ylabel(r"Mass Ratio, $q$")
-plt.xlabel(r"Total Mass, $M$")
+plt.xlabel(r"Total Mass, $M\,[\mathrm{M}_\odot]$ ")
 plt.ylabel(r"Effective Spin, $\chi_{\rm eff}$")
 plt.xlim(0, 200)
 plt.ylim(-1, 1.)

### 2023-01-25 src/tex/ms.tex
Adding EOB refs

@@
 There are a variety of gravitational waveforms currently used in analysis pipelines.
-They are generally structured into different families, the most common of which are: the effective-one-body (EOB)~\citep{Damour:2008yg,Buonanno:2005xu, Buonanno:2000ef, Buonanno:1998gg, Damour:2000we}, the phenomenological inspiral-merger-ringerdown (IMRPhenom) \citep{Husa:2015iqa,Khan:2015jqa,Hannam:2013oca,Pratten:2020ceb}, and numerical relativity surrogate (NRsurrogate)~\citep{Blackman:2017pcm,Varma:2018mmi,Varma:2019csw}.
+They are generally structured into different families, the most common of which are: the effective-one-body (EOB)~\citep{Damour:2008yg,Buonanno:2005xu, Buonanno:2000ef, Buonanno:1998gg, Damour:2000we,Ossokine:2020kjp,Nagar:2021gss,Ramos-Buades:2021adz,Albertini:2021tbt}, the phenomenological inspiral-merger-ringerdown (IMRPhenom) \citep{Husa:2015iqa,Khan:2015jqa,Hannam:2013oca,Pratten:2020ceb}, and numerical relativity surrogate (NRsurrogate)~\citep{Blackman:2017pcm,Varma:2018mmi,Varma:2019csw}.
 %Of these, the IMRPhenom family is most well suited for an AD implementation in \jax.
@@
 AD is therefore perfectly suited for computing Fisher matrices accurately and efficiently. 
-Forecasting with Fisher matrices for third generation detectors has already been extensively explored in \citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.
+Forecasting with Fisher matrices for third generation detectors has already been extensively explored in~\citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.
 Here we purely want to illustrate the simplicity and speed of AD for forecasting rather than providing new physics insights.

### 2023-01-25 src/tex/ms.tex
fixing more comments

@@
 For binary black holes with non-aligned spins, this requires a Markov Chain Monte Carlo (MCMC) on a 15 dimensional parameter space.
-More general binary inspirals, such as those involving neutron stars, can lead to a significant increase in dimension.\te{add citations} \mi{assuming you don't mean TGR, since it's mentioned below, do you mean calibration and tides?: for cal we can cite \cite{Farr:2014,Vitale:2020gvb}, and for tides maybe the BNS properties paper \cite{LIGOScientific:2018hze}, unless you want to cite a bunch of tidal waveform papers.}
+More general binary inspirals, such as those involving neutron stars, can lead to a significant increase in dimension.
+% \te{add citations} \mi{assuming you don't mean TGR, since it's mentioned below, do you mean calibration and tides?: for cal we can cite \cite{Farr:2014,Vitale:2020gvb}, and for tides maybe the BNS properties paper \cite{LIGOScientific:2018hze}, unless you want to cite a bunch of tidal waveform papers.}
 % In the most minimal example of non-precessing binaries, this requires a Markov Chain Monte Carlo (MCMC) on an 11 dimensional parameter space for each event.
@@
 % \AZ{Maybe modify; some other points are precessing PE requires 15 params, tidal prec 17 params, plus calibration marge and EoS inferences can incorporate even more}
-Beyond these simple scenarios, more complex waveform models with additional parameters may be used to test for deviations from General Relativity~\citep{Arun:2006yw, Agathos:2013upa, Yunes:2016jcc, LIGOScientific:2016lio, LIGOScientific:2020tif, Krishnendu:2021fga}.
+Beyond these simple scenarios, more complex waveform models with additional parameters may be used to include calibration errors~\citep{Farr:2014,Vitale:2020gvb}, model the tides of neutron stars~\citep{LIGOScientific:2018hze}, and test for deviations from General Relativity~\citep{Arun:2006yw, Agathos:2013upa, Yunes:2016jcc, LIGOScientific:2016lio, LIGOScientific:2020tif, Krishnendu:2021fga}..
 % \AZ{Add PPE and maybe TIGER refs..?}
 Finally, using the results of PE, population synthesis models constrain the progenitors systems from which the black holes we see merging today began their journey~\citep{LIGOScientific:2020kqk, LIGOScientific:2021psn, Wong:2022flg}.
-\kw{Need more reference and some rewording}\te{added refs, happy?}
+% \kw{Need more reference and some rewording}\te{added refs, happy?}
 % All of these tasks require significant computation.
@@
 
-\AZ{Add refs here:}\te{Added a bunch, let me know if more are needed!}
+% \AZ{Add refs here:}\te{Added a bunch, let me know if more are needed!}
 There are a variety of gravitational waveforms currently used in analysis pipelines.
@@
 As mentioned in the previous section, the three major waveform families that have been developed to date are:  EOB,  NRsurrogate, and IMRPhenom.
-Note that unlike NRsurrogate, both the EOB and IMRPhenom waveforms must be calibrated to NR waveforms.
-\mi{I think this is weird to say: the surrogate is fully trained on (``calibrated to'') NR}
+% Note that unlike NRsurrogate, both the EOB and IMRPhenom waveforms must be calibrated to NR waveforms.
+% \mi{I think this is weird to say: the surrogate is fully trained on (``calibrated to'') NR}
 EOB waveforms require one to model the binary system using a Hamiltonian and are typically slow to evaluate, whereas IMRPhenom waveforms are constructed with simple closed-form expressions.
@@
 From Fig.~\ref{fig:match}, it is clear that the \ripple waveform matches with the \lalsuite waveform close to numerical precision across the entire parameter space.
-\AZ{strictly $=0$? That seems impossible. Fmla has wrong labels for $h$'s}
-In particular, the gray points at low masses are where $1-\mathrm{m}(h_1|h_2)=0$ causing points to not appear in the log scale colorbar.
+% \AZ{strictly $=0$? That seems impossible. Fmla has wrong labels for $h$'s}
+In particular, the gray points at low masses are where $1-\mathrm{m}(h_1|h_2)=0$ (up to numerical precision) causing points to not appear in the log scale colorbar.\te{Edit according to updated plot.}
 % In particular, the blue points indicate parameter points where the $1-\mathrm{m}(h_{\theta_1}|h^{*}_{\theta_1})=0$ and would therefore not appear in the log scale colorbar.
-\mi{to Aaron's point, maybe say ``0 up to numerical precision''?}
+% \mi{to Aaron's point, maybe say ``0 up to numerical precision''?}
 
@@
 % \mi{I think Aaron means ``provides'' not ``proves''; also, we might add ``in the high SNR limit''}
-\AZ{Again we need to state and cite the PSDs we use for the three detectors.}
 
@@
     $\chi_1, \chi_2$ & $\boldsymbol{U}[-0.8, 0.8]$ \\ \hline
-    $D$ & $\left(400 (\boldsymbol{U}[0,1]^{1/3}) + 600\right)$\, Mpc \\ \hline
+    $D$ (Uniform in volume) & $[600, 900]$\,Mpc \\ \hline
     $t_c$ & 0.0 \\ \hline
     $\phi_c$ & 0.0 \\ \hline
-    Inclincation Angle, $\iota$ & $\cos^{-1}(\boldsymbol{U}[-1, 1])$ \\ \hline
+    Inclincation Angle, $\cos(\iota)$ & $\boldsymbol{U}[-1, 1]$ \\ \hline
     Polarization Angle, $\psi$ & $\boldsymbol{U}[0, 2\pi]$ \\ \hline
     Right Ascension, $\alpha$ & $\boldsymbol{U}[0, 2\pi]$ \\ \hline
-    Declination, $\delta$ & $\pi/2 - \cos^{-1}(\boldsymbol{U}[-1, 1])$ \\ 
+    Declination, $\sin(\delta)$ & $\boldsymbol{U}[-1, 1]$ \\ 
     \hline\hline
@@
 A full list of the distributions used to generate the various parameters in our population are given in Tab.~\ref{tab:priors}. 
+Additionally, we use $20\,$Hz, $1024\,$Hz, and $16\,$s for the minimum frequency, sampling frequency, and sample length.
+Our noise curves correspond to the design PSDs for LIGO Hanford, LIGO Livingston (\textsc{SimNoisePSDaLIGOZeroDetHighPower}) and Virgo (\textsc{SimNoisePSDAdvVirgo}).\footnote{\url{https://lscsoft.docs.ligo.org/lalsuite/lalinspiral/psds_8py_source.html}.}
 The resulting population produces binaries with signal-to-noise ratios ranging from $\mathcal{O}(10-10^2)$.
@@
 The distribution of sky localization errors from a population of $10^3$ binaries can be seen in Fig.~\ref{fig:sky_localization}.
-We have verified that our errors agree with a separate dedicated Fisher forecasting code~\citep{Borhanian:2020ypi} to within 20 percent.\te{Check number}
+We have verified that our errors agree with a separate dedicated Fisher forecasting code~\citep{Borhanian:2020ypi} to within 20\%.\te{Check number}
 This demonstrates that AD can be used to accurately produce population-level forecasts.
@@
 This factor of over 100 speed up is substantial considering the fact that a single core evaluation of the \ripple waveform is similar to \lalsuite which is used by \texttt{GWbench}.
-On a MacBook Pro with an M1 Max Apple Silicon processor, JIT compilation takes $\sim 270\,s$ and the full population analysis takes less $\sim 8$ minutes.
+On a MacBook Pro with an M1 Max Apple Silicon processor, JIT compilation takes $\sim 270\,s$ and the full population analysis takes less than $\sim 9$ minutes.
 % \kw{Want to pull a number on the total time it took on generating the population result? I imagine this should turn the theorists on and send them into a fishing frenzy.}
@@
 We plan on adding a variety of different waveforms to \ripple in the near future with the primary goal of reaching a \jax version of a fully precessing, higher order mode waveform such as IMRPhenomXPHM~\citep{Pratten:2020ceb}.
-Ideally, future waveforms should be implemented either in \jax compatible python directly.
+Ideally, future waveforms should be implemented under an AD framework such as \jax.
 % \amc{if they're pure python, some translation will be required due to jax's restrictions.}
-\mi{we could just say ``directly implemented under an autodifferentiation framework'' to be generic}
+% \mi{we could just say ``directly implemented under an autodifferentiation framework'' to be generic}
 This would ensure that the community can easily utilize differentiability and hardware acceleration in the future.

### 2023-01-25 src/scripts/random_matches.py

@@ -69,8 +69,8 @@ params = {
 mpl.rcParams.update(params)
 
 data = np.loadtxt(paths.data / "ripple_phenomD_matches.txt")
-data = data[data[:, -1].argsort()]
-data = data[::-1]
+# data = data[data[:, -1].argsort()]
+# data = data[::-1]
 
 exactmatch_indices = (1.0 - data[:, -1]) <= 0.0
 other_indices = (1.0 - data[:, -1]) > 0.0

### 2023-01-25 src/tex/ms.tex
Adding comment on EOB fitting

@@
 % (e.g.~in Hamiltonian Monte Carlo~\citep{2017arXiv170102434B} or Gradient Descent~\citep{2016arXiv160904747R}).
-For instance, during PE, derivative information can be used to guide an optimizer towards higher-likelihood values (e.g.\ using gradient descent~\citep{2016arXiv160904747R}) or allow a sampler to rapidly explore parameter space (e.g.\ using Hamiltonian Monte Carlo~\citep{2011hmcm.book..113N,2017arXiv170102434B}).
+For instance, during PE, derivative information can be used to guide an optimizer towards higher-likelihood values (e.g.\ using gradient descent~\citep{2016arXiv160904747R}) or allow a sampler to rapidly explore parameter space (e.g.\ using Hamiltonian Monte Carlo (HMC)~\citep{2011hmcm.book..113N,2017arXiv170102434B}).
 Gradients are particularly valuable for high dimensional spaces.
@@
 
-\AZ{Maybe comment on optimization method used for EOB waveforms?
-I think they do interesting Monte Carlo etc. Check what method is used 
-for IMRPhenomXPHM?}
-\mi{I agree. it'd be good to comment on whether this is still done like this in
-modern waveforms.}
+%AZ: Addressed below
+%\AZ{Maybe comment on optimization method used for EOB waveforms?
+%I think they do interesting Monte Carlo etc. Check what method is used 
+%for IMRPhenomXPHM?}
+%\mi{I agree. it'd be good to comment on whether this is still done like this in
+%modern waveforms.}
 
@@
 % \sout{Therefore, the new set of waveform coefficients can produce better waveforms for GW analyses. }
+
+\AZ{While we focused on \texttt{IMRPhenomD} here, the ability to apply AD to the calibration parameters may assist in other approaches to calibration, such as that used for the aligned-spin EOB model in~\citet{Bohe:2016gbl}.
+There MCMC methods were used in a two-step procedure to optimize the calibration parameters. 
+Derivative information, if implemented for EOB waveforms, could allow for the application of other sampling methods such as HMC, or possibly optimization over the entire set of NR waveforms at once.}
 

### 2023-01-25 src/tex/ms.tex
adding some comments

@@
             It is clear that the \ripple waveform matches the \lalsuite implementation many of orders of magnitude more than necessary for all data analysis tasks across the entire parameter space.
-            \te{Current plot looking better?}
+            \te{Ok so some of the remaining error was coming from a weird thing at the cutoff of the waveform.
+I've now fixed this, so error is further down.
+Let me know if you still think it needs editing.}
             \dfm{I think this figure could use some work.
@@
 \te{None of the 536 are part of the training set}
-\te{Used ALIGO design sensitivity for PSD}
+\te{Used flat PSD}
 \te{@Kelvin, can you please add a citation to the 11 waveforms used for training as well as a citation to the 536 waveforms used from the SXS catalogue.
 I've indicated in the text where these should be added}
-\te{@Kelvin can you please also add a citation to the particular PSD used (again, its indicated in the text where to do this.)}
 \te{The thing we really want to compare is the mismatch with the rescaled frequency grid and a fixed flat PSD.}
@@
 
+To evaluate Eq.~\ref{eq:multi_loss} we use a flat PSD and a frequency array which is scaled by the total mass.\footnote{
+    In particular, we use $M f_l = $, $M f_u = $, and $M\Delta f_l = $ for the dimensionless lower, upper, and frequency spacing respectively.
+We fix M to be $50\,[\mathrm{M}_\odot]$ throughout this section.\te{@Kelvin to fill in.}
+}
+Since the NR waveforms do not have a total mass associated with them, we use assign a fixed total mass to compare with the \ripple waveform.
+
 For training, we use the publically available subset of the 19 waveforms used in the original IMRPhenomD paper (11 in total)\te{@kelvin add citation to waveforms here}. 
 We then run gradient descent, as described above, for $3\times 10^4$ steps. 
-Figure~\ref{fig:loss_compare} shows the relative error (against a particular NR waveform) of the original and optimized waveform at different frequencies.
-\te{@Kelvin is the NR waveform here part of the training set?}
+Figure~\ref{fig:loss_compare} shows the relative error (against a test NR waveform; see below) of the original and optimized waveform as a function of dimensionless frequency.
 The vertical dashed lines indicate the stitching points for the phase i.e. when the inspiral is joined onto the merger.
@@
             Interestingly, the error is reduced across the frequency range, demonstrating that high dimensional optimization is useful for all parts of the waveform.
+\te{@Kelvin, change data to flat PSD.}
 		}
@@
             Overall, the optimized distribution shifts to lower mismatches and therefore a better waveform model. 
-            More quantitatively, the mean mismatch (shown as vertical dashed lines) is reduced by $\sim11$\%.
+            More quantitatively, the mean mismatch (shown as vertical dashed lines) is reduced by $\sim11$\%. \te{@Kelvin, change data to flat PSD.}
 		}

### 2023-01-25 src/tex/ms.tex
Fixed everything apart from 3.3

@@
 where $\Delta t_c$ and $\Delta \phi_c$ are, respectively, the differences in time and phase of coalescence between the two waveforms.
-\mi{I think it'd be good to define the mismatch $\mathcal{M} \equiv 1- \mathrm{m}$, since it's the standard quantity to discuss, and it's what we plot}
+Finally, we can define the mismatch as 
+\begin{equation}
+    \label{eq:mismatch_1}
+	\mathcal{M}(h_1|h_2) \equiv 1- \mathrm{m}(h_1|h_2)\,,
+\end{equation}
+
+% \mi{I think it'd be good to define the mismatch $\mathcal{M} \equiv 1- \mathrm{m}$, since it's the standard quantity to discuss, and it's what we plot}
 
@@
 % \AZ{strictly $=0$? That seems impossible. Fmla has wrong labels for $h$'s}
-In particular, the gray points at low masses are where $1-\mathrm{m}(h_1|h_2)=0$ (up to numerical precision) causing points to not appear in the log scale colorbar.\te{Edit according to updated plot.}
+In fact, the scale in Fig.~\ref{fig:match} is clipped such that points which give $1-\mathrm{m}(h_1|h_2)=0$ (up to numerical precision) are plotted as the lowest value in the colorbar.
 % In particular, the blue points indicate parameter points where the $1-\mathrm{m}(h_{\theta_1}|h^{*}_{\theta_1})=0$ and would therefore not appear in the log scale colorbar.
@@
             % The lower density of points at low masses is caused by the match being exactly 1.0 and therefore not appearing in the log scale. 
-            The gray points are where the match is exactly 1.0 and therefore would not appear in the log scale.
+            % The gray points are where the match is exactly 1.0 and therefore would not appear in the log scale.
+            Points which give $\mathcal{M}=0$ (up to numerical precision) are plotted as the lowest value in the colorbar
             It is clear that the \ripple waveform matches the \lalsuite implementation many of orders of magnitude more than necessary for all data analysis tasks across the entire parameter space.
@@
 There remains some slight deviation between the two waveform implementations at high total mass. 
-This is partly due to the fact that cubic interpolators, which are used within IMRPhenomD to calculate the ringdown and damping frequencies, are not currently supported in \jax.
+% This is partly due to the fact that cubic interpolators, which are used within IMRPhenomD to calculate the ringdown and damping frequencies, are not currently supported in \jax.
 \dfm{What dimension of interpolation do you need?
 Maybe this is something we could implement...}
-\te{Its just one dimension, so is pretty easy.
+% \te{Its just one dimension, so is pretty easy.
 I tried the one from the cosmoJAX people and found that the match did slightly increase but it was a decent chunk slower (maybe a factor of 2 or so).
@@
 }
-\kw{Maybe a bit more background on AD, the way we phrase it here sounds a bit too magical too me now.}
+For a review of AD methods see~\cite{2018arXiv181105031M}.
+% \kw{Maybe a bit more background on AD, the way we phrase it here sounds a bit too magical too me now.}
 Since the parameter space of GWs in GR has $\mathcal{O}(10)$ dimensions, the speed of derivative evaluation is less crucial than the stability.
@@
 generated by the NR simulation. 
-\mi{I think the mismatch should be defined above, right after the match is defined. We should also state what PSD and sampling parameters are used}
-\te{I think it makes it easier to define here cause we only now introduce the fitting coefficient $\lambda$.}
+% \mi{I think the mismatch should be defined above, right after the match is defined. We should also state what PSD and sampling parameters are used}
+% \te{I think it makes it easier to define here cause we only now introduce the fitting coefficient $\lambda$.}
 Given the loss function, we use gradient descent to update the fitting
@@
 
-\AZ{While we focused on \texttt{IMRPhenomD} here, the ability to apply AD to the calibration parameters may assist in other approaches to calibration, such as that used for the aligned-spin EOB model in~\citet{Bohe:2016gbl}.
-There MCMC methods were used in a two-step procedure to optimize the calibration parameters. 
-Derivative information, if implemented for EOB waveforms, could allow for the application of other sampling methods such as HMC, or possibly optimization over the entire set of NR waveforms at once.}
+While we focused on \texttt{IMRPhenomD} here, the ability to apply AD to the calibration parameters may assist in other approaches to calibration, such as that used for the aligned-spin EOB model in~\citet{Bohe:2016gbl}.
+There, MCMC methods were used in a two-step procedure to optimize the calibration parameters. 
+Derivative information, if implemented for EOB waveforms, could allow for the application of other sampling methods such as HMC, or possibly optimization over the entire set of NR waveforms at once.
 
@@
     $\chi_1, \chi_2$ & $\boldsymbol{U}[-0.8, 0.8]$ \\ \hline
-    $D$ (Uniform in volume) & $[600, 900]$\,Mpc \\ \hline
+    $D$ (Uniform in Volume) & $[600, 900]$\,Mpc \\ \hline
     $t_c$ & 0.0 \\ \hline

### 2023-01-25 src/scripts/random_matches.py

@@ -93,7 +93,7 @@ sc = plt.scatter(
     vmax=-12,
     s=20
 )
-plt.colorbar(sc, label=r"$\log_{10}(1-\mathrm{Match})$")
+plt.colorbar(sc, label=r"Mismatch, $\log_{10}(\mathcal{M})$")
 
 # thetas = np.array([Mtot,chieff]).T
 # vor = Voronoi(thetas)

### 2023-01-25 src/tex/ms.tex
Fixing typo

@@
     \label{eq:mismatch_1}
-	\mathcal{M}(h_1|h_2) \equiv 1- \mathrm{m}(h_1|h_2)\,,
+	\mathcal{M}(h_1|h_2) \equiv 1- \mathrm{m}(h_1|h_2)\,.
 \end{equation}
@@
 % \AZ{strictly $=0$? That seems impossible. Fmla has wrong labels for $h$'s}
-In fact, the scale in Fig.~\ref{fig:match} is clipped such that points which give $1-\mathrm{m}(h_1|h_2)=0$ (up to numerical precision) are plotted as the lowest value in the colorbar.
+In fact, the scale in Fig.~\ref{fig:match} is clipped such that points which give $\mathcal{M}=0$ (up to numerical precision) are plotted as the lowest value in the colorbar.
 % In particular, the blue points indicate parameter points where the $1-\mathrm{m}(h_{\theta_1}|h^{*}_{\theta_1})=0$ and would therefore not appear in the log scale colorbar.
@@
 There remains some slight deviation between the two waveform implementations at high total mass. 
-% This is partly due to the fact that cubic interpolators, which are used within IMRPhenomD to calculate the ringdown and damping frequencies, are not currently supported in \jax.
-\dfm{What dimension of interpolation do you need?
-Maybe this is something we could implement...}
+This is partly due to the fact that cubic interpolators, which are used within IMRPhenomD to calculate the ringdown and damping frequencies, are not currently supported in \jax. 
+% \dfm{What dimension of interpolation do you need? Maybe this is something we could implement...}
 % \te{Its just one dimension, so is pretty easy.

### 2023-01-25 src/tex/ms.tex
fix comments

@@
 
-\te{We use in total 11 waveforms for the optimization. Original PhenomD uses 19 @Kelvin add details of where to find this.}
+% \te{We use in total 11 waveforms for the optimization. Original PhenomD uses 19 @Kelvin add details of where to find this.}
 \te{Fig 2, plot was made after optimizing for ALL waveforms}
-\te{536 waveforms are subset of the SXS catalogue that are not precessing or eccentric. @Kelvin add details of where to find these things}
+% \te{536 waveforms are subset of the SXS catalogue that are not precessing or eccentric. @Kelvin add details of where to find these things}
 \te{None of the 536 are part of the training set}
 \te{Used flat PSD}
-\te{@Kelvin, can you please add a citation to the 11 waveforms used for training as well as a citation to the 536 waveforms used from the SXS catalogue.
+% \te{@Kelvin, can you please add a citation to the 11 waveforms used for training as well as a citation to the 536 waveforms used from the SXS catalogue.
 I've indicated in the text where these should be added}
@@
     \label{eq:mismatch}
-	\mathcal{M}(\lambda)=1-\mathrm{m}(h_{\mathrm{IMR}}(\lambda)|h_{\mathrm{NR}}(\lambda))\,,
+	\mathcal{M}(\lambda)=1-\mathrm{m}(h_{\theta}^{\mathrm{IMR}}(\lambda)|h_{\theta}^{\mathrm{NR}}(\lambda))\,,
 \end{equation}
-where $\lambda$ is a vector of the fitting coefficients, $h_{\mathrm{IMR}}$ is
-the waveform generated by IMRPhenomD, and $h_{\mathrm{NR}}$ is the waveform
+where $\lambda$ is a vector of the fitting coefficients, $h_{\theta}^{\mathrm{IMR}}$ is
+the waveform generated by IMRPhenomD, and $h_{\theta}^{\mathrm{NR}}$ is the waveform
 generated by the NR simulation. 
@@
 \end{equation}
-where $\alpha$ is the learning rate. We set $\alpha$ to be a very small number, i.e. $10^{-6}$, since the initial guess already lies close to the desired optimum. 
+where $\alpha$ is the learning rate. We set $\alpha$ to be $10^{-6}$. 
 %   \dfm{We can't possibly know that it has converged to ``the optimum''. I'd sat just leave it at ``We terminate the program after 30000 steps.''}
@@
 To evaluate Eq.~\ref{eq:multi_loss} we use a flat PSD and a frequency array which is scaled by the total mass.\footnote{
-    In particular, we use $M f_l = $, $M f_u = $, and $M\Delta f_l = $ for the dimensionless lower, upper, and frequency spacing respectively.
-We fix M to be $50\,[\mathrm{M}_\odot]$ throughout this section.\te{@Kelvin to fill in.}
+    In particular, we use $M f_l = 2.5 \times 10^{-3}$, $M f_u = 0.2$, and $M\Delta f = 2.5 \times 10^{-6}$ for the dimensionless lower, upper, and frequency spacing respectively.
+We fix $M$ to be $50\,[\mathrm{M}_\odot]$ throughout this section.
 }
@@
 
-For training, we use the publically available subset of the 19 waveforms used in the original IMRPhenomD paper (11 in total)\te{@kelvin add citation to waveforms here}. 
+For training, we use the publically available subset (11 waveforms) of the 19 waveforms used in the original IMRPhenomD paper~\citep{Khan:2015jqa}. 
+These 11 waveforms are taken from the SXS catalog~\citep{Boyle:2019kee}.
 We then run gradient descent, as described above, for $3\times 10^4$ steps. 
@@
             Interestingly, the error is reduced across the frequency range, demonstrating that high dimensional optimization is useful for all parts of the waveform.
-\te{@Kelvin, change data to flat PSD.}
-		}
 		\label{fig:loss_compare}
@@
             Overall, the optimized distribution shifts to lower mismatches and therefore a better waveform model. 
-            More quantitatively, the mean mismatch (shown as vertical dashed lines) is reduced by $\sim11$\%. \te{@Kelvin, change data to flat PSD.}
+            More quantitatively, the mean mismatch (shown as vertical dashed lines) is reduced by $\sim36$\%. 
 		}
@@
 In Fig.~\ref{fig:mismatch_hist}, we show the distribution of log mismatches for 536 test waveforms.
-For testing we use 536 waveforms from the SXS catalogue (we simply choose all waveforms with aligned spins and eccentricity close to 0 \te{kelvin, please add specific eccentricity cut here})~\citep{Boyle:2019kee}.
+For testing, we use 536 waveforms from the SXS catalog (we simply choose all waveforms with aligned spins, $\chi_{x,y}<5\times 10^{-3}$, and eccentricity $<2\times 10^{-3}$)~\citep{Boyle:2019kee}.
 One can see that the distribution of mismatches after optimization is generally shifted to lower mismatch compared to the original waveform.

### 2023-01-26 src/tex/ms.tex
minor build pdf fix
fixing stuff in sec 3

@@
 % (e.g.~in Hamiltonian Monte Carlo~\citep{2017arXiv170102434B} or Gradient Descent~\citep{2016arXiv160904747R}).
-For instance, during PE, derivative information can be used to guide an optimizer towards higher-likelihood values (e.g.\ using gradient descent~\citep{2016arXiv160904747R}) or allow a sampler to rapidly explore parameter space (e.g.\ using Hamiltonian Monte Carlo (HMC)~\citep{2011hmcm.book..113N,2017arXiv170102434B}).
+For instance, during PE, derivative information can be used to guide an optimizer towards higher-likelihood values (e.g.\,using gradient descent~\citep{2016arXiv160904747R}) or allow a sampler to rapidly explore parameter space (e.g.\, using Hamiltonian Monte Carlo (HMC)~\citep{2011hmcm.book..113N,2017arXiv170102434B}).
 Gradients are particularly valuable for high dimensional spaces.
@@
             Interestingly, the error is reduced across the frequency range, demonstrating that high dimensional optimization is useful for all parts of the waveform.
+        }
 		\label{fig:loss_compare}
@@
 The distribution of sky localization errors from a population of $10^3$ binaries can be seen in Fig.~\ref{fig:sky_localization}.
-We have verified that our errors agree with a separate dedicated Fisher forecasting code~\citep{Borhanian:2020ypi} to within 20\%.\te{Check number}
+We have verified that our errors agree with a separate dedicated Fisher forecasting code~\citep{Borhanian:2020ypi} to within ~30\%.
 This demonstrates that AD can be used to accurately produce population-level forecasts.
@@
 
-\amc{most acronyms were already introduced.}
+% \amc{most acronyms were already introduced.}
 After the search algorithms have constructed a list of confidently detected binaries, the next step is to sample from the posterior of each sources parameters - so called PE.
-To do this, one typically uses a Markov Chain Monte Carlo (MCMC) or nested sampler~\citep{Skilling2004, multinest, dynesty}.
+To do this, one typically uses an MCMC or nested sampler~\citep{Skilling2004, multinest, dynesty}.
 Although robust, both MCMC and nested sampling are slow to converge and are known to perform poorly in high dimensional parameter spaces.
@@
 Dedicated fast samplers have been designed to get approximate posteriors on the sky localization to facilitate follow-up electromagnetic observations \citep[e.g.~\texttt{BAYESTAR}][]{Singer:2015ema}. 
-\AZ{We should be careful here.
+Moreover, a number of methods have been developed to speed up PE well below the numbers quoted above~\citep{Dax:2021tsq, Islam:2022afg,Roulet:2022kot, Zackay:2018qdy, Cornish:2021lje, Canizares:2013ywa}.
+% \AZ{We should be careful here.
 Many new algorithms have sped this a lot.
@@
 % In particular, we will show that proposals made using an Hamiltonian Monte Carlo (MC) sampler are accepted at a much higher rate than a traditional MCMC algorithm.
-In particular, we will show that the autocorrelation of a Hamiltonian Monte Carlo (HMC) sampler is significantly lower than a traditional MCMC algorithm.
-\AZ{Add citations to other HMC papers in GW land}
+In particular, we will show that the autocorrelation of a Hamiltonian Monte Carlo (HMC) sampler is significantly lower than a traditional MCMC algorithm~\citep{Bouffanais:2018hoz, 2014CQGra..31n5004P}.
 
 For our basic example we perform an injection recovery test on a seven dimensional parameter space with the same three detector network setup considered in \S~\ref{subsec:fisher}.
-\AZ{As before, say what PSD and cite.}
-More specifically, we generate Gaussian noise consistent with the measured PSDs for each detector and then inject a BBH signal with parameters: chirp mass $M_c = 23.82\,\mathrm{M_\odot}$,  symmetric mass ratio $\eta = 0.248$, primary spin parameter $\chi_1=0.3$, secondary spin parameter \AZ{$\chi_2=-0.4$}, luminosity distance $D = 1.6\,\mathrm{Gpc}$, coalescence time $t_c = 0.0$, and coalescence phase $\phi_c = 0.0$.\footnote{
-    The remaining parameters are set to $\pi/3$. \AZ{Can we say what they are? $\iota$, $\psi$, $\alpha$, $\delta$?}
+Again, our noise curves correspond to the design PSDs for LIGO Hanford, LIGO Livingston (\textsc{SimNoisePSDaLIGOZeroDetHighPower}), and Virgo (\textsc{SimNoisePSDAdvVirgo})
+We use $20\,$Hz, $1024\,$Hz, and $16\,$s for the minimum frequency, sampling frequency, and sample length.
+More specifically, we generate Gaussian noise consistent with the measured PSDs for each detector and then inject a BBH signal with parameters: chirp mass $M_c = 23.82\,\mathrm{M_\odot}$,  symmetric mass ratio $\eta = 0.248$, primary spin parameter $\chi_1=0.3$, secondary spin parameter $\chi_2=-0.4$, luminosity distance $D = 1.6\,\mathrm{Gpc}$, coalescence time $t_c = 0.0$, and coalescence phase $\phi_c = 0.0$.\footnote{
+    The remaining parameters (inclination angle $\iota$, polarization angle $\psi$, right ascension $\alpha$, declination $\delta$) are set to $\pi/3$.
 }
@@
 We therefore expect gradient based samplers to converge significantly faster than typical samplers, especially in higher dimensions.
-In addition, we found that the effictive number of samples 
-\amc{provide ref?} 
+In addition, we found that the effictive number of samples~\citep{arviz_2019},\footnote{
+    Computed using \textsc{arviz}~\url{https://python.arviz.org/en/stable/api/generated/arviz.ess.html}
+    }
 (a measure of the number of indepedent samples) is between 2 and 7 times larger for HMC across the different dimensions of the parameter space.
@@
 In a follow up paper we will demonstrate that minute scale PE can be achieved by combining normalizing flows~\citep{2022arXiv221106397W, Gabrie:2021tlu}, GPU acceleration, and a derivative based sampler~\citep{PEpaper}.
-We therefore expect \jax waveforms to be \sout{crucial} \AZ{highly beneficial} to future PE efforts in GW astronomy, especially for low-latency pipelines.
-\kw{Not just low latency, but also high dimensional analysis, such as testing GR and lensing.}
+We therefore expect \jax waveforms to be highly beneficial to future PE efforts in GW astronomy, particular for low-latency pipelines and higher dimensional analysese.
+% \kw{Not just low latency, but also high dimensional analysis, such as testing GR and lensing.}
 

### 2023-01-26 src/scripts/mismatch_hist.py
calculating change in mismatch

@@ -41,6 +41,10 @@ mpl.rcParams.update(params)
 )
 bins = np.arange(-5, -1, 0.1)
 
+print(np.mean(ori_loss))
+print(np.mean(opt_loss))
+# print((np.mean(ori_loss) - np.mean(opt_loss))*100/np.mean(ori_loss))
+
 # print(stats.mode(np.log10(ori_loss)[:]))
 # print(np.mean(np.log10(ori_loss)) / np.mean(np.log10(opt_loss)))
 # print(stats.mode(np.log10(ori_loss))[0] / stats.mode(np.log10(opt_loss))[0])

### 2023-01-26 src/tex/ms.tex
Getting ready for P&P

@@
 
-Again, performing the same benchmark as above on a NVIDIA Quadro 6000, we find that on average waveform evaluations take $\sim 0.02$, over an of magnitude faster than serial CPU evaluation.
+Again, performing the same benchmark as above on a NVIDIA Quadro 6000, we find that on average waveform evaluations take $\sim 0.02\,\mathrm{ms}$, over an of magnitude faster than serial CPU evaluation.
 % \te{Check and modify}
@@
 % \te{We use in total 11 waveforms for the optimization. Original PhenomD uses 19 @Kelvin add details of where to find this.}
-\te{Fig 2, plot was made after optimizing for ALL waveforms}
+% \te{Fig 2, plot was made after optimizing for ALL waveforms}
 % \te{536 waveforms are subset of the SXS catalogue that are not precessing or eccentric. @Kelvin add details of where to find these things}
-\te{None of the 536 are part of the training set}
-\te{Used flat PSD}
+% \te{None of the 536 are part of the training set}
+% \te{Used flat PSD}
 % \te{@Kelvin, can you please add a citation to the 11 waveforms used for training as well as a citation to the 536 waveforms used from the SXS catalogue.
 I've indicated in the text where these should be added}
-\te{The thing we really want to compare is the mismatch with the rescaled frequency grid and a fixed flat PSD.}
-\te{Need to understand how the mass scale was introduced into the NR waveforms.}
+% \te{The thing we really want to compare is the mismatch with the rescaled frequency grid and a fixed flat PSD.}
+% \te{Need to understand how the mass scale was introduced into the NR waveforms.}
 
@@
     In particular, we use $M f_l = 2.5 \times 10^{-3}$, $M f_u = 0.2$, and $M\Delta f = 2.5 \times 10^{-6}$ for the dimensionless lower, upper, and frequency spacing respectively.
-We fix $M$ to be $50\,[\mathrm{M}_\odot]$ throughout this section.
+We fix $M$ to be $50\,\mathrm{M}_\odot$ throughout this section.
 }
@@
             Overall, the optimized distribution shifts to lower mismatches and therefore a better waveform model. 
-            More quantitatively, the mean mismatch (shown as vertical dashed lines) is reduced by $\sim36$\%. 
+            More quantitatively, the median mismatch (shown as vertical dashed lines) is reduced by $\sim50$\%. 
 		}

### 2023-01-26 src/scripts/mismatch_hist.py

@@ -39,25 +39,23 @@ mpl.rcParams.update(params)
 (catalog_num, ori_loss, opt_loss) = np.loadtxt(
     paths.data / "mismatch_hist.txt", unpack=True
 )
+
+print("Median Change is:", (np.median(opt_loss) - np.median(ori_loss))*100/np.median(ori_loss))
+
 bins = np.arange(-5, -1, 0.1)
-
-print(np.mean(ori_loss))
-print(np.mean(opt_loss))
-# print((np.mean(ori_loss) - np.mean(opt_loss))*100/np.mean(ori_loss))
-
 # print(stats.mode(np.log10(ori_loss)[:]))
 # print(np.mean(np.log10(ori_loss)) / np.mean(np.log10(opt_loss)))
 # print(stats.mode(np.log10(ori_loss))[0] / stats.mode(np.log10(opt_loss))[0])
 
-plt.axvline(x=np.mean(np.log10(ori_loss)), color="C0", ls="--")
+plt.axvline(x=np.log10(np.median(ori_loss)), color="C0", ls="--")
 plt.hist(np.log10(ori_loss), bins=bins, alpha=0.3, color="C0")
 plt.hist(np.log10(ori_loss), bins=bins, label="Original", histtype="step", color="C0")
 
-plt.axvline(x=np.mean(np.log10(opt_loss)), color="C1", ls="--")
+plt.axvline(x=np.log10(np.median(opt_loss)), color="C1", ls="--")
 plt.hist(np.log10(opt_loss), bins=bins, alpha=0.3, color="C1")
 plt.hist(np.log10(opt_loss), bins=bins, label="Optimized", histtype="step", color="C1")
 plt.legend()
 plt.xlim(-5, -1)
-plt.xlabel("$\log(\mathcal{M})$")
+plt.xlabel("$\log_{10}(\mathcal{M})$")
 plt.ylabel("Count")
 plt.savefig(paths.figures / "mismatch_hist.pdf", bbox_inches="tight")

### 2023-01-26 src/tex/ms.tex
NRsur in prep citation

@@
 Two other waveform families are commonly used in GW data analysis: the effective-one-body (EOB) and numerical relativity surrogate (NRSurrogate). 
-Some progress has been made towards the construction of a differentiable NRsurrogate \mi{cite our paper in prep?}, but currently it seems difficult to implement EOB waveforms in \jax.
+A differentiable NRsurrogate implementation is currently under development \citep{NRSurAD}, but currently it seems difficult to implement EOB waveforms in \jax.
 In particular, the evolution of the Hamiltonian required to evaluate an EOB waveform is both inherently slow to differentiate and difficult to implement in \jax. 

### 2023-01-26 src/tex/ms.tex
Ready for P&P

@@
             It is clear that the \ripple waveform matches the \lalsuite implementation many of orders of magnitude more than necessary for all data analysis tasks across the entire parameter space.
-            \te{Ok so some of the remaining error was coming from a weird thing at the cutoff of the waveform.
+            % \te{Ok so some of the remaining error was coming from a weird thing at the cutoff of the waveform.
 I've now fixed this, so error is further down.
 Let me know if you still think it needs editing.}
-            \dfm{I think this figure could use some work.
+            % \dfm{I think this figure could use some work.
 Maybe the points could be smaller to make them easier to resolve.
@@
 Right now it looks less impressive!}
-            \mi{I agree.
+            % \mi{I agree.
 Furthermore, it looks like there's a subpopulation of darker points hiding under the brighter ones up to and around $M \sim 100$?
@@
         Although not fully converged, it is clear that we find posteriors consistent with the injected parameters. 
-\amc{could apply a mild Gaussian kernel to smooth this for the sake of presentation.
-Are the parameter ranges due to your priors?
-How did you handle the boundaries in the HMC?}
+        % \amc{could apply a mild Gaussian kernel to smooth this for the sake of presentation. 
+        % \amc{Are the parameter ranges due to your priors? How did you handle the boundaries in the HMC?}
     }
@@
 
-For our basic example we perform an injection recovery test on a seven dimensional parameter space with the same three detector network setup considered in \S~\ref{subsec:fisher}.
-Again, our noise curves correspond to the design PSDs for LIGO Hanford, LIGO Livingston (\textsc{SimNoisePSDaLIGOZeroDetHighPower}), and Virgo (\textsc{SimNoisePSDAdvVirgo})
+For our basic example we perform an injection recovery test on a seven dimensional parameter space with the two LIGO detectors in our network.
+Our noise curves correspond to the design PSDs for LIGO Hanford and Livingston (\textsc{SimNoisePSDaLIGOZeroDetHighPower}).
 We use $20\,$Hz, $1024\,$Hz, and $16\,$s for the minimum frequency, sampling frequency, and sample length.
@@
 }
-Using a standard Gaussian likelihood, we then run the HMC sampler implemented in \flowMC for \sout{200000} $2\times10^5$ steps and the Gaussian random walk (GRW) sampler for \sout{1500000} $1.5\times10^6$ steps (each with four randomly initialized independent chains).
-\AZ{What implementation of GRW? Can we cite something?}
-\amc{briefly summarize and give refs for flowMC and GRW. Is GRW Metropolis-Hastings with a Gaussian transition kernel? How'd you tune this? Why not use NUTS for HMC?}
+Using a standard Gaussian likelihood, we then run the HMC sampler implemented in \flowMC~\citep{2022arXiv221106397W} for $2\times10^5$ steps and the random walk Metropolis Hastings (RWMH) sampler~\citep{1953JChPh..21.1087M} for $1.5\times10^6$ steps (each with four randomly initialized independent chains).
+% \AZ{What implementation of GRW? Can we cite something?}
+% \amc{briefly summarize and give refs for flowMC and GRW. Is GRW Metropolis-Hastings with a Gaussian transition kernel? How'd you tune this? Why not use NUTS for HMC?}
 The number of steps and and mass matrix used for each example was hand tuned to give good performance for the specific sampler.
-\amc{not a huge deal, but why hand tune? There are automated ways to do this in e.g.\ numpyro with a warm-up adaptation phase.}
+% \amc{not a huge deal, but why hand tune? There are automated ways to do this in e.g.\ numpyro with a warm-up adaptation phase.}
 The additional steps for the GRW sampler were required to achieve a similarly converged posterior.
-\amc{Why not plot its posterior?}
+% \amc{Why not plot its posterior?}
+
+We note at this point that neither pure HMC nor RWMH are the most modern versions of gradient and non-gradient based samplers.
+For example, for gradient based samplers one could use a No-U-Turn sampler~\citep{2011arXiv1111.4246H} or the Metropolis-adjusted Langevin algorithm~\citep{2013arXiv1309.2983X}.
+Traditional MCMC methods such as nested sampling~\citep{Skilling2004, multinest, dynesty} or the Affine Invariant MCMC Ensemble sampler implemented in \texttt{emcee}~\citep{2013PASP..125..306F} will also lead to more efficient sampling of the posterior than basic RWMH.
+Here we instead seek to demonstrate the simplicity with which HMC can be implemented within a differentiable pipeline.
+In addition, as we discuss below, we find that a basic HMC algorithm will produce significantly more efficient sampling that RWMH, motivating further exploration of gradient-based samplers for GW PE~\citep{PEpaper}.
+
+% \te{Add paragraph with caveats.}
+% \te{Change GRW to random walk metrpolis hastings (RWMH)}
 
 In Fig.~\ref{fig:corner}, the grey contours show the posterior recovered using the best chain (i.e. one that reached the highest log-likelihood values).
-\mi{why show a single chain?}
+% \mi{why show a single chain?}
 The orange shows the true parameters of the injected signal.
 From the one dimensional histograms along the diagonal, it is clear that we consistently recover all seven parameters apart from $\phi_c$.
-This is expected since the injected binary is relatively nearby.
-\AZ{Can we give the SNR?}
+This is expected since the injected binary is relatively nearby with an SNR of $\sim13$.
+% \AZ{Can we give the SNR?}
 

### 2023-01-26 src/scripts/autocorrelation.py

@@ -48,7 +48,7 @@ acl_HMC = az.autocorr(HMC_bestchain.T).T
 acl_gaussian = az.autocorr(gaussian_bestchain.T).T
 
 plt.plot(acl_HMC[:, 0], label="HMC")
-plt.plot(acl_gaussian[:, 0], label="GRW")
+plt.plot(acl_gaussian[:, 0], label="RWMH")
 plt.xlim(0, 1000)
 plt.ylim(-0.1, 1.0)
 plt.xlabel("Lag")

### 2023-01-26 src/scripts/corner_plot.py

@@ -40,6 +40,11 @@ mpl.rcParams.update(params)
 data_HMC = np.load(paths.data / "chains_HMC.npz")
 HMC_bestchain = data_HMC["chains"][np.argmax(data_HMC["log_prob"][:, -1])]
 
+data_gaussian = np.load(paths.data / "chains_gaussian.npz")
+gaussian_bestchain = data_gaussian["chains"][
+    np.argmax(data_gaussian["log_prob"][:, -1])
+]
+
 
 def ms_to_Mc_eta(m):
     r"""
@@ -67,24 +72,28 @@ true_params = np.array(
 )
 
 # Plot all chains
-figure = corner.corner(HMC_bestchain, labels=labels, labelpad=0.1)
+# figure = corner.corner(gaussian_bestchain, labels=labels, labelpad=0.1, smooth=True, color="C1", alpha=0.3, label="GRW")
+
+figure = corner.corner(HMC_bestchain, labels=labels, labelpad=0.1, smooth=True, truths=[Mc, eta, 0.3, -0.4, distance, 0.0, 0.0])
+
 
 # Extract the axes
-axes = np.array(figure.axes).reshape((n_dim, n_dim))
+# axes = np.array(figure.axes).reshape((n_dim, n_dim))
 
 # Loop over the diagonal
-for i in range(n_dim):
-    ax = axes[i, i]
-    ax.axvline(true_params[i], color="C1", alpha=0.8)
+# for i in range(n_dim):
+#     ax = axes[i, i]
+#     ax.axvline(true_params[i], color="C1", alpha=0.8)
 
-# Loop over the histograms
-for yi in range(n_dim):
-    for xi in range(yi):
-        ax = axes[yi, xi]
-        ax.axvline(true_params[xi], color="C1", alpha=0.8)
-        ax.axhline(true_params[yi], color="C1", alpha=0.8)
-        ax.plot(true_params[xi], true_params[yi], "s", color="C1", alpha=0.8)
+# # Loop over the histograms
+# for yi in range(n_dim):
+#     for xi in range(yi):
+#         ax = axes[yi, xi]
+#         ax.axvline(true_params[xi], color="C1", alpha=0.8)
+#         ax.axhline(true_params[yi], color="C1", alpha=0.8)
+#         ax.plot(true_params[xi], true_params[yi], "s", color="C1", alpha=0.8)
 
 figure.set_size_inches(10, 10)
+plt.legend()
 plt.savefig(paths.figures / "PE.pdf", bbox_inches="tight")
 

### 2023-01-26 src/scripts/random_matches.py

@@ -75,7 +75,7 @@ data = np.loadtxt(paths.data / "ripple_phenomD_matches.txt")
 exactmatch_indices = (1.0 - data[:, -1]) <= 0.0
 other_indices = (1.0 - data[:, -1]) > 0.0
 
-q = data[:, 1] / data[:, 0]
+# q = data[:, 1] / data[:, 0]
 chieff = (data[:, 0] * data[:, 2] + data[:, 1] * data[:, 3]) / (data[:, 0] + data[:, 1])
 Mtot = data[:, 1] + data[:, 0]
 
@@ -91,7 +91,7 @@ sc = plt.scatter(
     cmap=cm,
     vmin=-16,
     vmax=-12,
-    s=20
+    s=5
 )
 plt.colorbar(sc, label=r"Mismatch, $\log_{10}(\mathcal{M})$")
 

### 2023-01-26 src/scripts/sky_localization.py
Force updating sky localization plot

@@ -88,7 +88,7 @@ plt.hist(sky_error_list, bins=bins, density=True, alpha=0.3, color="C0")
 plt.hist(sky_error_list, bins=bins, density=True, histtype="step", color="C0")
 
 plt.xscale("log")
-plt.xlim(0.1, 300)
+plt.xlim(0.1, 301)
 plt.xlabel("Sky Localization Error [deg$^2$]")
 plt.ylabel("Probability")
 plt.savefig(paths.figures / "sky_localization.pdf", bbox_inches="tight")

### 2023-02-06 src/tex/ms.tex
Fixing small issues and dealing with PnP comments

@@
     Firstly, we demonstrate how gradient descent can be used to optimize the $\sim 200$ coefficients that are used to calibrate the waveform model. 
-    In particular, we demonstrate that the typical \textit{match} with numerical relativity waveforms can be improved by more than 10\% without any additional overhead.
+    In particular, we demonstrate that the typical \textit{match} with numerical relativity waveforms can be improved by more than 50\% without any additional overhead.
     Secondly, we show that Fisher forecasting calculations can be sped up by $\sim 100\times$ (on a CPU) with no loss in accuracy.
@@
 % \sout{The complexity begins before data gathering where one is required to generate banks of template waveforms which will be used for a matched-filter search~\citep{Owen:1998dk, Owen:1995tm}. }
-Once potential candidates are found, parameter estimation (PE) is performed to extract the detailed source properties of each event~\citep{Christensen:2022bxb, 2020MNRAS.493.3132S, Ashton:2018jfp, Romero-Shaw:2020owr, Veitch:2014wba, Biwer:2018osg}.
+Once potential candidates are found, parameter estimation (PE) is performed to extract the detailed source properties of each event~\citep{Christensen:2022bxb, 2020MNRAS.493.3132S, Ashton:2018jfp, Romero-Shaw:2020owr, Veitch:2014wba, Biwer:2018osg, kombine}.
 For binary black holes with non-aligned spins, this requires a Markov Chain Monte Carlo (MCMC) on a 15 dimensional parameter space.
@@
 % \AZ{Maybe modify; some other points are precessing PE requires 15 params, tidal prec 17 params, plus calibration marge and EoS inferences can incorporate even more}
-Beyond these simple scenarios, more complex waveform models with additional parameters may be used to include calibration errors~\citep{Farr:2014,Vitale:2020gvb}, model the tides of neutron stars~\citep{LIGOScientific:2018hze}, and test for deviations from General Relativity~\citep{Arun:2006yw, Agathos:2013upa, Yunes:2016jcc, LIGOScientific:2016lio, LIGOScientific:2020tif, Krishnendu:2021fga}..
+Beyond these simple scenarios, more complex waveform models with additional parameters may be used to include calibration errors~\citep{Farr:2014,Vitale:2020gvb}, model the tides of neutron stars~\citep{LIGOScientific:2018hze}, and test for deviations from General Relativity~\citep{Arun:2006yw, Agathos:2013upa, Yunes:2016jcc, LIGOScientific:2016lio, LIGOScientific:2020tif, Krishnendu:2021fga}.
 % \AZ{Add PPE and maybe TIGER refs..?}
@@
 More general frameworks exist in \texttt{julia}~\citep{zygote, forwarddiff}, although \texttt{julia}'s limited use in GW analysis software precludes its general use.
-Here we make use of \jax~\citep{jax2018github} due to its easy integration with \texttt{python} libraries, seamless support for running code on different hardware accelerators,such as graphical processing units (GPUs), and its just-in-time (JIT) compiler, which can substantially accelerate code.
+Here we make use of \jax~\citep{jax2018github} due to its easy integration with \texttt{python} libraries, seamless support for running code on different hardware accelerators, such as graphical processing units (GPUs), and its just-in-time (JIT) compiler, which can substantially accelerate code.
 
@@
 There are a variety of gravitational waveforms currently used in analysis pipelines.
-They are generally structured into different families, the most common of which are: the effective-one-body (EOB)~\citep{Damour:2008yg,Buonanno:2005xu, Buonanno:2000ef, Buonanno:1998gg, Damour:2000we,Ossokine:2020kjp,Nagar:2021gss,Ramos-Buades:2021adz,Albertini:2021tbt}, the phenomenological inspiral-merger-ringerdown (IMRPhenom) \citep{Husa:2015iqa,Khan:2015jqa,Hannam:2013oca,Pratten:2020ceb}, and numerical relativity surrogate (NRsurrogate)~\citep{Blackman:2017pcm,Varma:2018mmi,Varma:2019csw}.
+They are generally structured into different families, the most common of which are: the effective-one-body (EOB)~\citep{Damour:2008yg,Buonanno:2005xu, Buonanno:2000ef, Buonanno:1998gg, Damour:2000we,Ossokine:2020kjp,Nagar:2021gss,Ramos-Buades:2021adz,Albertini:2021tbt, Cotesta:2020qhw, Bohe:2016gbl}, the phenomenological inspiral-merger-ringerdown (IMRPhenom) \citep{Husa:2015iqa,Khan:2015jqa,Hannam:2013oca,Pratten:2020ceb, Smith:2016qas, Pratten:2020fqn, Garcia-Quiros:2020qpx}, and numerical relativity surrogate (NRsurrogate)~\citep{Blackman:2017pcm,Varma:2018mmi,Varma:2019csw}.
 %Of these, the IMRPhenom family is most well suited for an AD implementation in \jax.
@@
 
-One of the primary argument of this paper is that waveform derivatives will also be highly valuable to data analysis tasks. 
+One of the primary s of this paper is that waveform derivatives will also be highly valuable to data analysis tasks. 
 AD provides two big advantages when it comes to evaluating derivatives compared to numerical differentiation.
@@
 Forecasting the sensitivity of future experiments is an routine task in GW science. 
-Due to its theoretical simplicity and evaluation speed, the Fisher matrix formalism is commonly deployed to estimate how well a binary system's parameters could be measured.
+Due to its theoretical simplicity and evaluation speed, the Fisher matrix formalism~\citep{Cornish:2010kf} is commonly deployed to estimate how well a binary system's parameters could be measured.
 The Fisher matrix approach is built around the assumption of a Gaussian likelihood~\citep{Vallisneri:2007ev}.
@@
 Dedicated fast samplers have been designed to get approximate posteriors on the sky localization to facilitate follow-up electromagnetic observations \citep[e.g.~\texttt{BAYESTAR}][]{Singer:2015ema}. 
-Moreover, a number of methods have been developed to speed up PE well below the numbers quoted above~\citep{Dax:2021tsq, Islam:2022afg,Roulet:2022kot, Zackay:2018qdy, Cornish:2021lje, Canizares:2013ywa}.
+Moreover, a number of methods have been developed to speed up PE well below the numbers quoted above~\citep{Dax:2021tsq, Islam:2022afg,Roulet:2022kot, Zackay:2018qdy, Cornish:2021lje, Canizares:2013ywa, Leslie:2021ssu}.
 % \AZ{We should be careful here.
@@
 There are likelihood acceleration methods like ROQ, multibanding, relative binning/heterodyning that accelerate the process compared to the numbers quoted above.}
-Nevertheless, these do not present the whole picture; fast, general PE therefore remains a key aim of GW data analysis
+Nevertheless, these do not present the whole picture; fast, general PE therefore remains a key aim of GW data analysis.
 

### 2023-02-06 src/tex/ms.tex
fixing rest of pnp

@@
 % \sout{The complexity begins before data gathering where one is required to generate banks of template waveforms which will be used for a matched-filter search~\citep{Owen:1998dk, Owen:1995tm}. }
-Once potential candidates are found, parameter estimation (PE) is performed to extract the detailed source properties of each event~\citep{Christensen:2022bxb, 2020MNRAS.493.3132S, Ashton:2018jfp, Romero-Shaw:2020owr, Veitch:2014wba, Biwer:2018osg, kombine}.
+Once potential candidates are found, parameter estimation (PE) is performed to extract the detailed source properties of each event~\citep{Christensen:2022bxb, 2020MNRAS.493.3132S, Ashton:2018jfp, Romero-Shaw:2020owr, Veitch:2014wba, Biwer:2018osg, kombine, 10.1093/mnras/stv2422}.
 For binary black holes with non-aligned spins, this requires a Markov Chain Monte Carlo (MCMC) on a 15 dimensional parameter space.
@@
 Although we primarily discuss toy examples in this paper, each can be extended to the full data analysis task, some of which will be shown in upcoming papers~\citep{PEpaper}.
-Differentiable waveforms therefore represent a crucial advancement to perform efficient GW science.
+Differentiable waveforms therefore represent a crucial advancement towards efficient GW science.
 
-In this paper, we have primarily focussed on the IMRPhenom family of waveforms as their closed form expression is perfectly suited for implementation in \jax.
+In this paper, we have primarily focussed on the IMRPhenom family of waveforms as their closed form expression is perfectly suited for a \jax implementation.
 Two other waveform families are commonly used in GW data analysis: the effective-one-body (EOB) and numerical relativity surrogate (NRSurrogate). 
-A differentiable NRsurrogate implementation is currently under development \citep{NRSurAD}, but currently it seems difficult to implement EOB waveforms in \jax.
+A differentiable NRsurrogate implementation is under development~\citep{NRSurAD}, but it currently seems difficult to implement EOB waveforms in \jax.
 In particular, the evolution of the Hamiltonian required to evaluate an EOB waveform is both inherently slow to differentiate and difficult to implement in \jax. 

### 2023-02-06 src/tex/ms.tex
fixing rest of pnp

@@
 Here, we illustrate how three core tasks in GW science can be substantially improved through the use of differentiable waveforms.
-In the paper we primarily look at toy examples, leaving a more careful analyses to future work. 
+In this paper we primarily look at toy examples, leaving more careful analyses to future work. 
 The three tasks discussed here cover a wide range of GW science, starting with waveform development all the way to Fisher forecasting and PE.
@@
 \end{equation}
-where $\mathcal{M}_i$ is the mismatch of an individual training waveform and $N$ is the total number of training waveforms used in optimization.
+where $\mathcal{M}_i$ is the mismatch of an individual training waveform and $N$ is the total number of training waveforms used in the optimization.
 This optimization is more difficult since we are now applying the same set of coefficients to waveforms with different intrinsic parameters, such as mass ratio and spins. 
@@
 }
-Since the NR waveforms do not have a total mass associated with them, we use assign a fixed total mass to compare with the \ripple waveform.
+Since the NR waveforms do not have a total mass associated with them, we assign a fixed total mass to compare with the \ripple waveform.
 
@@
 
-In Fig.~\ref{fig:mismatch_hist}, we show the distribution of log mismatches for 536 test waveforms.
-For testing, we use 536 waveforms from the SXS catalog (we simply choose all waveforms with aligned spins, $\chi_{x,y}<5\times 10^{-3}$, and eccentricity $<2\times 10^{-3}$)~\citep{Boyle:2019kee}.
+In Fig.~\ref{fig:mismatch_hist}, we show the distribution of log mismatches for a set of test waveforms.
+In particular, we use 536 waveforms from the SXS catalog (we simply choose all waveforms with aligned spins, $\chi_{x,y}<5\times 10^{-3}$, and eccentricity $<2\times 10^{-3}$)~\citep{Boyle:2019kee}.
 One can see that the distribution of mismatches after optimization is generally shifted to lower mismatch compared to the original waveform.
@@
 
-Forecasting the sensitivity of future experiments is an routine task in GW science. 
+Forecasting the sensitivity of future experiments is a routine task in GW science. 
 Due to its theoretical simplicity and evaluation speed, the Fisher matrix formalism~\citep{Cornish:2010kf} is commonly deployed to estimate how well a binary system's parameters could be measured.
@@
 % \amc{most acronyms were already introduced.}
-After the search algorithms have constructed a list of confidently detected binaries, the next step is to sample from the posterior of each sources parameters - so called PE.
+After the search algorithms have constructed a list of confidently detected binaries, the next step is to sample from the posterior of each sources parameter's - so called PE.
 To do this, one typically uses an MCMC or nested sampler~\citep{Skilling2004, multinest, dynesty}.
@@
 Derivative based samplers, on the other hand, have been shown to extrapolate well to higher dimensions although they sometimes come with their own drawbacks.
-Here we simply aim to demonstrate the utility of a derivative based sampler and their efficiency on a small test problem.
+Here we simply aim to demonstrate the utility of a derivative based sampler and its efficiency on a small test problem.
 % In particular, we will show that proposals made using an Hamiltonian Monte Carlo (MC) sampler are accepted at a much higher rate than a traditional MCMC algorithm.
@@
 In a follow up paper we will demonstrate that minute scale PE can be achieved by combining normalizing flows~\citep{2022arXiv221106397W, Gabrie:2021tlu}, GPU acceleration, and a derivative based sampler~\citep{PEpaper}.
-We therefore expect \jax waveforms to be highly beneficial to future PE efforts in GW astronomy, particular for low-latency pipelines and higher dimensional analysese.
+We therefore expect \jax waveforms to be highly beneficial to future PE efforts in GW astronomy, particular for low-latency pipelines and higher dimensional analyses.
 % \kw{Not just low latency, but also high dimensional analysis, such as testing GR and lensing.}

### 2023-02-07 src/tex/ms.tex
fix adam's ciela affiliation
typos, affiliation, preprint

@@
 \newcommand{\Mila}{\affiliation{Mila -- Quebec AI Institute, 6666 St-Urbain, \#200, Montreal, QC, H2S 3H1}} 
+\newcommand{\WI}{\affiliation{Weinberg Institute, University of Texas at Austin, Austin, TX 78712, USA}}
 \newcommand{\CGP}{\affiliation{Center for Gravitational Physics, University of Texas at Austin, Austin, TX 78712, USA}}
@@
 
-\author{Adam Coogan} \UdeM \Mila
+\author{Adam Coogan} \Ciela \UdeM \Mila
 \author{Daniel Foreman-Mackey} \CCA
 \author{Maximiliano Isi} \CCA
-\author{Aaron Zimmerman} \CGP
+\author{Aaron Zimmerman} \WI
 
@@
             % The gray points are where the match is exactly 1.0 and therefore would not appear in the log scale.
-            Points which give $\mathcal{M}=0$ (up to numerical precision) are plotted as the lowest value in the colorbar
+            Points which give $\mathcal{M}=0$ (up to numerical precision) are plotted as the lowest value in the colorbar.
             It is clear that the \ripple waveform matches the \lalsuite implementation many of orders of magnitude more than necessary for all data analysis tasks across the entire parameter space.
@@
 
-One of the primary s of this paper is that waveform derivatives will also be highly valuable to data analysis tasks. 
+One of the primary aims of this paper is that waveform derivatives will also be highly valuable to data analysis tasks. 
 AD provides two big advantages when it comes to evaluating derivatives compared to numerical differentiation.
@@
 A.~C.\ acknowledges funding from the Schmidt Futures Foundation.
+This document has been assigned preprint numbers UTWI-4-2023 and LIGO-P2300025.
 

### 2023-02-07 src/tex/ms.tex
Changes to Eq 9

@@
 \end{equation}
-where $k$ indicates the detector, $\partial_i = \partial/\partial \theta_i$, and $h$ is the strain measured by the detector which is given by,
+where $k$ indicates the detector, $\partial_i = \partial/\partial \theta_i$, and $h^k$ is the strain measured by the detector which is given by,
 \begin{equation}
-    h^k(\theta) = F_+(\phi) h^k_{+}(\Xi) + F_\times(\phi) h^k_{\times}(\Xi) \, .
+    h^k(\theta) = F^k_+(\phi) h_{+}(\Xi) + F^k_\times(\phi) h_{\times}(\Xi) \, .
 \end{equation}

### 2023-02-07 src/tex/ms.tex
Fixing a space

@@
 
-In Fig.~\ref{fig:corner}, the grey contours show the posterior recovered using the best chain (i.e. one that reached the highest log-likelihood values).
+In Fig.~\ref{fig:corner}, the grey contours show the posterior recovered using the best chain (i.e.~one that reached the highest log-likelihood values).
 % \mi{why show a single chain?}

### 2023-02-07 src/tex/ms.tex
changing to eqref

@@
 % \sout{frequency evolution and is therefore not a} \sout{truly independent parameter.}
-If we instead use a flat PSD in Eq.~\ref{eq:inner_prod} and rescale the frequency grid to units of $Mf$, all dependence with $M$ seen in Fig.~\ref{fig:match} vanishes. 
+If we instead use a flat PSD in Eq.~\eqref{eq:inner_prod} and rescale the frequency grid to units of $Mf$, all dependence with $M$ seen in Fig.~\ref{fig:match} vanishes. 
 % \AZ{Revisit this after defining PSD above:} 
@@
 This optimization is more difficult since we are now applying the same set of coefficients to waveforms with different intrinsic parameters, such as mass ratio and spins. 
-From Eq.~\ref{eq:multi_loss}, we can see the averaging between waveforms with different intrinsic parameters implies there are trade-offs in performance for different regions of the intrinsic parameter space.
+From Eq.~\eqref{eq:multi_loss}, we can see the averaging between waveforms with different intrinsic parameters implies there are trade-offs in performance for different regions of the intrinsic parameter space.
 Additionally, this means our best-fit points will generally depend on the distribution of training waveforms across the parameter space.
@@
 
-To evaluate Eq.~\ref{eq:multi_loss} we use a flat PSD and a frequency array which is scaled by the total mass.\footnote{
+To evaluate Eq.~\eqref{eq:multi_loss} we use a flat PSD and a frequency array which is scaled by the total mass.\footnote{
     In particular, we use $M f_l = 2.5 \times 10^{-3}$, $M f_u = 0.2$, and $M\Delta f = 2.5 \times 10^{-6}$ for the dimensionless lower, upper, and frequency spacing respectively.
@@
 		\caption{
-			Distribution of 536 log mismatches (see Eq.~\ref{eq:mismatch}) for the original (blue) and optimized (orange) IMRPhenomD. 
+			Distribution of 536 log mismatches (see Eq.~\eqref{eq:mismatch}) for the original (blue) and optimized (orange) IMRPhenomD. 
             Overall, the optimized distribution shifts to lower mismatches and therefore a better waveform model. 
@@
         Distribution of Fisher information sky localization error for a population of nearby binaries.
-        Automatic differentiation (AD) was used to compute the derivatives in Eq.~\ref{eq:fisher}.
+        Automatic differentiation (AD) was used to compute the derivatives in Eq.~\eqref{eq:fisher}.
         As emphasized in the text, after JIT compilation, each error calculation was around 100 times faster than a similar forecasting code \texttt{GWbench}~\citep{Borhanian:2020ypi}.

### 2023-02-08 src/tex/ms.tex
preparing for arxiv submission
removing comments for arxiv submission

@@
 \definecolor{Burnt}{RGB}{203,96,21}
-
-% Link setup
-% \newcommand\myshade{80}
-% \hypersetup{
-%   linkcolor  = pyGreen!\myshade!black,
-%   citecolor  = pyBlue!\myshade!black,
-%   urlcolor   = pyRed!\myshade!black,
-%   colorlinks = true
-% }
 
@@
 \begin{abstract}
-    % Since their discovery in 2015, gravitational waves (GWs) have developed into a unique observational probe used throughout fundamental and astro-phyics.
-    % As more detections continue, data analysis tasks are becoming increasingly computationally demanding.
-    % \kw{I would cut everything before out.}
     We propose the use of automatic differentiation through the programming framework \jax for accelerating a variety of analysis tasks throughout gravitational wave (GW) science.
@@
     This increased speed makes population forecasting substantially simpler.
-    % , as discussed extensively in \citet{Iacovelli:2022bbs, Iacovelli:2022mbg}.
     Finally, we show that gradient-based samplers like Hamiltonian Monte Carlo lead to significantly reduced autocorrelation values when compared to traditional Monte Carlo methods.
@@
 The complexity begins even before data taking, since GW searches using the matched-filtering technique~\citep{Owen:1998dk, Owen:1995tm} require the generation of large banks of template waveforms.
-% \sout{The complexity begins before data gathering where one is required to generate banks of template waveforms which will be used for a matched-filter search~\citep{Owen:1998dk, Owen:1995tm}. }
 Once potential candidates are found, parameter estimation (PE) is performed to extract the detailed source properties of each event~\citep{Christensen:2022bxb, 2020MNRAS.493.3132S, Ashton:2018jfp, Romero-Shaw:2020owr, Veitch:2014wba, Biwer:2018osg, kombine, 10.1093/mnras/stv2422}.
@@
 More general binary inspirals, such as those involving neutron stars, can lead to a significant increase in dimension.
-% \te{add citations} \mi{assuming you don't mean TGR, since it's mentioned below, do you mean calibration and tides?: for cal we can cite \cite{Farr:2014,Vitale:2020gvb}, and for tides maybe the BNS properties paper \cite{LIGOScientific:2018hze}, unless you want to cite a bunch of tidal waveform papers.}
-% In the most minimal example of non-precessing binaries, this requires a Markov Chain Monte Carlo (MCMC) on an 11 dimensional parameter space for each event.
-% \kw{This is not quite true. You can always marginalize over more parameters if you want, such as time of coalescence and phase. I would just mention the typical case of 15 parameters GR PE instead.}
-% \AZ{Maybe modify; some other points are precessing PE requires 15 params, tidal prec 17 params, plus calibration marge and EoS inferences can incorporate even more}
 Beyond these simple scenarios, more complex waveform models with additional parameters may be used to include calibration errors~\citep{Farr:2014,Vitale:2020gvb}, model the tides of neutron stars~\citep{LIGOScientific:2018hze}, and test for deviations from General Relativity~\citep{Arun:2006yw, Agathos:2013upa, Yunes:2016jcc, LIGOScientific:2016lio, LIGOScientific:2020tif, Krishnendu:2021fga}.
-% \AZ{Add PPE and maybe TIGER refs..?}
 Finally, using the results of PE, population synthesis models constrain the progenitors systems from which the black holes we see merging today began their journey~\citep{LIGOScientific:2020kqk, LIGOScientific:2021psn, Wong:2022flg}.
-% \kw{Need more reference and some rewording}\te{added refs, happy?}
-% All of these tasks require significant computation.
 Overall, GW data analysis therefore requires significant computation.
-% \kw{This could be triggering to some people, since population analysis could be quite cheap}
 In this paper, we will argue that differentiable waveforms (and more generally differentiable pipelines) can play a significant role in alleviating this computational demand.
@@
 Derivatives are ubiquitously useful throughout data analysis tasks.
-% For instance, during PE, gradients are used to improve performance of the samplers 
-% % \dfm{This isn't correct. In HMC the gradients don't have the behavior. I'd just say something like ``gradients are used to improve performance''} 
-% (e.g.~in Hamiltonian Monte Carlo~\citep{2017arXiv170102434B} or Gradient Descent~\citep{2016arXiv160904747R}).
 For instance, during PE, derivative information can be used to guide an optimizer towards higher-likelihood values (e.g.\,using gradient descent~\citep{2016arXiv160904747R}) or allow a sampler to rapidly explore parameter space (e.g.\, using Hamiltonian Monte Carlo (HMC)~\citep{2011hmcm.book..113N,2017arXiv170102434B}).
@@
 With waveforms only increasing in complexity, calculating analytic derivatives will become ever more tricky.
-% \kw{This needs some rewording.
-For example, I think in IMRPhenomD they do have the analytical derivatives coded up, at least for a part of the waveforms.
-Instead of difficult to have the derivatives, I would rather mention it is clumpsy to get the derivatives, especially given the model complexity is growing.}
 Numerical derivatives also suffer from accuracy issues stemming from rounding or truncation errors.
@@
 
-% \AZ{Add refs here:}\te{Added a bunch, let me know if more are needed!}
 There are a variety of gravitational waveforms currently used in analysis pipelines.
 They are generally structured into different families, the most common of which are: the effective-one-body (EOB)~\citep{Damour:2008yg,Buonanno:2005xu, Buonanno:2000ef, Buonanno:1998gg, Damour:2000we,Ossokine:2020kjp,Nagar:2021gss,Ramos-Buades:2021adz,Albertini:2021tbt, Cotesta:2020qhw, Bohe:2016gbl}, the phenomenological inspiral-merger-ringerdown (IMRPhenom) \citep{Husa:2015iqa,Khan:2015jqa,Hannam:2013oca,Pratten:2020ceb, Smith:2016qas, Pratten:2020fqn, Garcia-Quiros:2020qpx}, and numerical relativity surrogate (NRsurrogate)~\citep{Blackman:2017pcm,Varma:2018mmi,Varma:2019csw}.
-%Of these, the IMRPhenom family is most well suited for an AD implementation in \jax.
 Of these, the IMRPhenom family serves as a natural starting point for an AD implementation in \jax.
 Models like the non-precessing IMRPhenomD~\citep{Khan:2015jqa} model studied here and the precessing, higher-mode model IMRPhenomXPHM~\citep{Pratten:2020fqn, Pratten:2020ceb} are written in the frequency domain using closed-form expressions.
-This makes a
-% \sout{In particular, their closed form expressions and fixed size frequency grids make a }
-\jax implementation that complies with the constraints of JIT compilation simple.
+This makes a \jax implementation that complies with the constraints of JIT compilation simple.
 NRsurrogate models, which interpolate directly over waveforms produced by numerical relativity (NR) simulations, are in principle also straightfoward to implement in \jax.
@@
 For EOB waveforms, frequency-domain reduced-order models may be a convenient target~\citep[e.g.]{Cotesta:2020qhw}.
-% to produce a waveform
-%require dynamic frequency grids and are therefore tricky to implement in \jax.
-%\AZ{What does this mean?
-I don't understand why EOB needs dynamic frequency grids.
-We should clarify, and we can also mention if \jax is well-suited to frequency-domain ROM implementations of EOB, and reference those.}
-%In addition, EOB waveforms use hypergeometric functions which have limited support in \jax.
-%\AZ{rephrase this in a more positive fashion:}
-%NRsurrogate waveforms have no fundamental issue but require the user to download relatively large data files, making their use cumbersome in practice.\te{Max/Kaze, please add to/change this if you think its incorrect}
 
 In this paper we argue that differentiable waveforms will be a vital component for the future of GW data analysis.
-% \AZ{I prefer to describe what is rather than what could be; can we say what elements exist, e.g.~
-% we present \ripple... which includes \jax implementations of GW antenna patterns, the IMRPhenomD waveform, and...}
-% In addition, we present \ripple, a small GW \texttt{python} package which can serve as a repository for differentiable implementations of some of the main waveforms currently used in LIGO and Virgo analyses.
 In addition, we present \ripple, a small GW \texttt{python} package which, at the time of writing, includes a \jax implementation of the IMRPhenomD waveform and will be continually updated with new waveforms as they are implemented.
-% can serve as a repository for differentiable implementations of some of the main waveforms currently used in LIGO and Virgo analyses. 
 The remainder of this paper is structured as follows. 
@@
 As mentioned in the previous section, the three major waveform families that have been developed to date are:  EOB,  NRsurrogate, and IMRPhenom.
-% Note that unlike NRsurrogate, both the EOB and IMRPhenom waveforms must be calibrated to NR waveforms.
-% \mi{I think this is weird to say: the surrogate is fully trained on (``calibrated to'') NR}
 EOB waveforms require one to model the binary system using a Hamiltonian and are typically slow to evaluate, whereas IMRPhenom waveforms are constructed with simple closed-form expressions.
@@
 where $S_n$ is the (one-sided) noise power spectral density (PSD) and $h_1$ and $h_2$ are the frequency domain waveforms which are to be compared.
-% \sout{ evaluated at the intrinsic parameter point $\theta_1$.}
 We can then normalize the inner product through
@@
 
-% \mi{I think it'd be good to define the mismatch $\mathcal{M} \equiv 1- \mathrm{m}$, since it's the standard quantity to discuss, and it's what we plot}
-
 Since the match is a measure of the difference between two waveforms, we can use it to demonstrate that the implementation of IMRPhenomD in \ripple accurately matches the \lalsuite implementation.
@@
 From Fig.~\ref{fig:match}, it is clear that the \ripple waveform matches with the \lalsuite waveform close to numerical precision across the entire parameter space.
-% \AZ{strictly $=0$? That seems impossible. Fmla has wrong labels for $h$'s}
 In fact, the scale in Fig.~\ref{fig:match} is clipped such that points which give $\mathcal{M}=0$ (up to numerical precision) are plotted as the lowest value in the colorbar.
-% In particular, the blue points indicate parameter points where the $1-\mathrm{m}(h_{\theta_1}|h^{*}_{\theta_1})=0$ and would therefore not appear in the log scale colorbar.
-% \mi{to Aaron's point, maybe say ``0 up to numerical precision''?}
 
@@
             Match between the \ripple and \lalsuite implementations of the IMRPhenomD waveform as a function of total mass and effective spin. 
-            % The lower density of points at low masses is caused by the match being exactly 1.0 and therefore not appearing in the log scale. 
-            % The gray points are where the match is exactly 1.0 and therefore would not appear in the log scale.
             Points which give $\mathcal{M}=0$ (up to numerical precision) are plotted as the lowest value in the colorbar.
             It is clear that the \ripple waveform matches the \lalsuite implementation many of orders of magnitude more than necessary for all data analysis tasks across the entire parameter space.
-            % \te{Ok so some of the remaining error was coming from a weird thing at the cutoff of the waveform.
-I've now fixed this, so error is further down.
-Let me know if you still think it needs editing.}
-            % \dfm{I think this figure could use some work.
-Maybe the points could be smaller to make them easier to resolve.
-And it could be interesting to clip the mismatch to something like $10^{-16}$ so that we don't lose the points with zero mismatch.
-Right now it looks less impressive!}
-            % \mi{I agree.
-Furthermore, it looks like there's a subpopulation of darker points hiding under the brighter ones up to and around $M \sim 100$?
-is there a dependence on the mass ratio?
-it'd be good to comment on this, and maybe show a corner plot adding $q$ to the mix.
-Also, $M$ is lacking units (I assume $M_\odot$, but it should be shown in the label)}
         }
@@
 The total mass only impacts the match because the chosen PSD and frequency limits fix a reference scale.
-% \sout{the system is simply an overall scaling of the waveforms} 
-% \sout{frequency evolution and is therefore not a} \sout{truly independent parameter.}
 If we instead use a flat PSD in Eq.~\eqref{eq:inner_prod} and rescale the frequency grid to units of $Mf$, all dependence with $M$ seen in Fig.~\ref{fig:match} vanishes. 
-% \AZ{Revisit this after defining PSD above:} 
 This figure instead illustrates a more realistic PSD where at low masses the waveform signal-to-noise is dominated by the inspiral but at high masses it is dominated by the merger.
@@
 This is partly due to the fact that cubic interpolators, which are used within IMRPhenomD to calculate the ringdown and damping frequencies, are not currently supported in \jax. 
-% \dfm{What dimension of interpolation do you need? Maybe this is something we could implement...}
-% \te{Its just one dimension, so is pretty easy.
-I tried the one from the cosmoJAX people and found that the match did slightly increase but it was a decent chunk slower (maybe a factor of 2 or so).
-So I just left it out.
-Would be happy to code up something and put it in if you think its necessary!}
 Instead, we initially use \texttt{scipy}'s cubic interpolator to create a fine grid of $5\times10^5$ values, which we then linearly interpolate during waveform generation.
 Unfortunately, we cannot make the initial cubic interpolation arbitrarily fine as this would add additional computational overhead when loading the data during waveform evaluation.
-% The remaining error in the waveform at high masses is therefore primarily down to the coarseness in this interpolation.
+Additionally, we use a Fourier transform to efficiently maximize over $\Delta t_c$ which will inherently come with some inaccuracy.
+This inaccuracy will primarily affect the mismatch at higher total masses where the waveform is dominated by the merger.
 Note however, that the differences are well below the accuracy requirements of the waveforms and will have no noticeable effect for realistic data analysis tasks.
@@
 Even though the \lalsuite implementation of IMRPhenomD is written in \texttt{C}, the serial evaluation of the waveform in \ripple is comparably fast. 
-% \dfm{I'd soften this statement a little bit.
-In my own work, I often find that JAX implementations are just as fast or faster than C implementations, so I wouldn't state it that it's obvious that lalsuite should be faster!
-Where are these benchmarks implemented?
-Would it be easy for me to try them out?}
 Benchmarking on a MacBook Pro with an M1 Max Apple Silicon processor, we find that a single waveform evaluation takes $\sim 0.5\,\mathrm{ms}$ for \lalsuite (interfaced with python) and $\sim 0.4\,\mathrm{ms}$ for \ripple.\footnote{
@@
 
-% Waveform calls can therefore be evaluated in parallel, massively improving performance.
-
 Again, performing the same benchmark as above on a NVIDIA Quadro 6000, we find that on average waveform evaluations take $\sim 0.02\,\mathrm{ms}$, over an of magnitude faster than serial CPU evaluation.
-% \te{Check and modify}
-% \kw{We should add some more detail related to this comparison.
-We should be clear that each individual waveform sample evaluation doesn't get us the performance by putting it on GPU (Jitting helps with that, not the GPU part).
-The GPU comes into play by allowing parallel evaluation of waveform sample, so for a tiny amount of time/frequency samples, we should expect GPU to lose to a CPU, but on any reasonable amount of time/frequency samples, GPU win a lot.}
-% \amc{mention \texttt{vmap} and/or \texttt{pmap} -- super trivial to parallelize! More work in C.}
 Generalizing \lalsuite waveforms to run on a GPU would be a significant undertaking.
@@
 For a review of AD methods see~\cite{2018arXiv181105031M}.
-% \kw{Maybe a bit more background on AD, the way we phrase it here sounds a bit too magical too me now.}
 Since the parameter space of GWs in GR has $\mathcal{O}(10)$ dimensions, the speed of derivative evaluation is less crucial than the stability.
@@
 In the next section, we will discuss a variety of potential use cases of differentiable waveforms.
-
-% Emphasise the stability of gradients. $Done$
-% Not just a faster way of evaluating PhenomD. $Done$
-% Change plot to be q vs chieff. $Done$
-% Note that lalsuite cannot be made massively parallel on GPUs. $Done$
-% Don't need a full section on Fisher, but make sure to mention. People love Fisher.
-% Use acceptance rate as a good explanation of speed ups
-% Wall time a messy comparison. Based too much off of hardware
-% Emphasise that gradients are beneficial
 
@@
 
-% \te{We use in total 11 waveforms for the optimization. Original PhenomD uses 19 @Kelvin add details of where to find this.}
-% \te{Fig 2, plot was made after optimizing for ALL waveforms}
-% \te{536 waveforms are subset of the SXS catalogue that are not precessing or eccentric. @Kelvin add details of where to find these things}
-% \te{None of the 536 are part of the training set}
-% \te{Used flat PSD}
-% \te{@Kelvin, can you please add a citation to the 11 waveforms used for training as well as a citation to the 536 waveforms used from the SXS catalogue.
-I've indicated in the text where these should be added}
-% \te{The thing we really want to compare is the mismatch with the rescaled frequency grid and a fixed flat PSD.}
-% \te{Need to understand how the mass scale was introduced into the NR waveforms.}
-
-% \AZ{Broad comments: need to cite which catalog the waveforms came from if any.
-% The number of waveforms used seems much larger than the number originally used
-% to calibrate PhenomD. 
-% If that is true, we can't make a big deal about having a better mismatch compared to
-% the original version, because of course optimizing over another set will improve 
-% performance.
-% Say how many waveforms we use to optimize, how many we compare to, where they came
-% from.}
-
 Having an accurate waveform model is essential for many data analysis tasks throughout
@@
 The community therefore utilizes waveform ``approximants'' (such as those discussed in the introduction) which can be evaluated much faster and in regions not covered by numerical simulations, such as extremes of parameter space or earlier phases of the binary inspiral.
-% \amc{redundant -- remove or merge with earlier material.}
 
-%\sout{Every waveform approximant has some}
 Waveform approximants generally have free coefficients which are calibrated
 to NR waveforms to achieve high accuracy.
-% \sout{ fitting coefficients that are tuned during
-% the process of calibrating the ansatz to NR waveforms.
 In the case of IMRPhenomD, there are 209 fitting coefficients used to capture the separate behavior of the amplitude and phase as a function of the mass ratio and spins.
 
-% \sout{The accuracy of fitting coefficients determines how well the approximant
-% reproduces the NR waveform. 
-% The}
 Any inaccuracy in obtaining the fitting coefficients
@@
 
-%AZ: Addressed below
-%\AZ{Maybe comment on optimization method used for EOB waveforms?
-%I think they do interesting Monte Carlo etc. Check what method is used 
-%for IMRPhenomXPHM?}
-%\mi{I agree. it'd be good to comment on whether this is still done like this in
-%modern waveforms.}
-
 In general, optimization problems in a high dimensional space benefit from having access to the gradient of the objective function. 
@@
 generated by the NR simulation. 
-% \mi{I think the mismatch should be defined above, right after the match is defined. We should also state what PSD and sampling parameters are used}
-% \te{I think it makes it easier to define here cause we only now introduce the fitting coefficient $\lambda$.}
 Given the loss function, we use gradient descent to update the fitting
@@
 where $\alpha$ is the learning rate. We set $\alpha$ to be $10^{-6}$.
-%   \dfm{We can't possibly know that it has converged to ``the optimum''. I'd sat just leave it at ``We terminate the program after 30000 steps.''}
-% \amc{say something to explain why this is working. I.e.: does the loss initially drop and then plateau?}
-% Add description of the optimization procedure.
 
@@
 Additionally, this means our best-fit points will generally depend on the distribution of training waveforms across the parameter space.
-% \amc{this means your fit generally will depend on how the training waveforms are distributed over parameter space.}
 
@@
 The vertical dashed lines indicate the stitching points for the phase i.e. when the inspiral is joined onto the merger.
-% In this optimization
-% procedure, we took NR waveforms $h_{\mathrm{NR}}$ stated in \citep{Khan:2015jqa}
-% as training waveforms to fit waveform coefficients. \amc{how many are there?}
 We can see the error of optimized waveform is lower than that of the original waveform for most of the domain.
 In particular, the error in the merger region is decreased by half while other regions also show good improvement in accuracy.
-% \amc{so this is a waveform in the training set? And I don't follow exactly what you did. Is this just fitting a single waveform?}
-
-% \te{We terminate the program after 30000 steps.}
 
@@
 One can see that the distribution of mismatches after optimization is generally shifted to lower mismatch compared to the original waveform.
-% showing the optimization improves the accuracy of the waveform model.
-% However, it is not improving for every waveforms. Say why briefly, and say there is another paper coming out.
-% While many waveforms improved their accuracy, some retain similar mismatches. 
-% \AZ{I don't understand this claim. The spin degeneracy is present in the physics of the waveform; $\chi_{\rm eff}$ is just one way of parametrizing the spins that tries to reduce the degeneracy.}
-% This phenomenon is mainly due to the reduced spin approximation used in IMRPhenomD, where spin degeneracy is introduced, hence restricting further improvements. 
-% In an upcoming paper, we will give a detailed discussion of how such an approximation scheme affects the optimization procedure. 
 In particular, the peak of the original waveform distribution has moved by nearly an order of magnitude, indicating that our AD-assisted optimization procedure provides an improved implementation of the model.
-% \sout{Therefore, the new set of waveform coefficients can produce better waveforms for GW analyses. }
 
@@
 The Fisher information matrix for a single detector is typically given by 
-% \amc{I'm guessing in practice you compute a Hessian, so maybe write that out?}
 \begin{equation}
@@
 which provides forecasted measurement errors and parameter covariances for a signal with parameters $\theta$ observed by the given detector network in the high signal-to-noise limit.
-% \mi{I think Aaron means ``provides'' not ``proves''; also, we might add ``in the high SNR limit''}
-
-% \begin{table*}[ht!]
-%     \begin{center}
-%     \begin{tabular}{c|c|c|c|c|c|c|c|c}
-%     \hline\hline
-%     \hfil $\boldsymbol{m_1, m_2}$ &
-%     \hfil $\boldsymbol{\chi_1, \chi_2}$ &
-%     \hfil $\boldsymbol{D}$ &
-%     \hfil $\boldsymbol{t_c}$ &
-%     \hfil $\boldsymbol{\phi_c}$ &
-%     \hfil $\boldsymbol{\iota}$ &
-%     \hfil $\boldsymbol{\psi}$ &
-%     \hfil $\boldsymbol{RA}$ &
-%     \hfil $\boldsymbol{DEC}$
-%     \\
-%     \hline\hline
-%     \hfil [0.1, 3] &
-%     \hfil [0, 1] &
-%     \hfil [0, 1] &
-%     \hfil [0.7, 1.3] &
-%     \hfil [-3, 0] &
-%     \hfil [0, 3] &
-%     \hfil [-3, 3] &
-%     \hfil [-3, 3] &
-%     \hfil [-3, 3]
-%     \\
-%     \hline\hline
-%     \end{tabular}
-%     \end{center}
-%     \caption{caption here}
-%     \label{tab:priors}
-% \end{table*}
 
 \begin{table}[t]
-    % \begin{center}
     \centering
@@
     \end{tabular}
-    % \end{center}
     \caption{Priors for the 11 dimensional parameter space used for the Fisher forecasting population analysis in Sec.~\ref{subsec:fisher}. 
@@
 On a MacBook Pro with an M1 Max Apple Silicon processor, JIT compilation takes $\sim 270\,s$ and the full population analysis takes less than $\sim 9$ minutes.
-% \kw{Want to pull a number on the total time it took on generating the population result? I imagine this should turn the theorists on and send them into a fishing frenzy.}
 As discussed above, performance can be further improved by utilizing hardware acceleration such as parallel GPU processing.
@@
         Although not fully converged, it is clear that we find posteriors consistent with the injected parameters. 
-        % \amc{could apply a mild Gaussian kernel to smooth this for the sake of presentation. 
-        % \amc{Are the parameter ranges due to your priors? How did you handle the boundaries in the HMC?}
     }
@@
 
-% \amc{most acronyms were already introduced.}
 After the search algorithms have constructed a list of confidently detected binaries, the next step is to sample from the posterior of each sources parameter's - so called PE.
@@
 Moreover, a number of methods have been developed to speed up PE well below the numbers quoted above~\citep{Dax:2021tsq, Islam:2022afg,Roulet:2022kot, Zackay:2018qdy, Cornish:2021lje, Canizares:2013ywa, Leslie:2021ssu}.
-% \AZ{We should be careful here.
-Many new algorithms have sped this a lot.
-Cogwheel can do fast PE, DINGO is a normalizing flow code that is very fast when trained.
-There are likelihood acceleration methods like ROQ, multibanding, relative binning/heterodyning that accelerate the process compared to the numbers quoted above.}
 Nevertheless, these do not present the whole picture; fast, general PE therefore remains a key aim of GW data analysis.
@@
 Here we simply aim to demonstrate the utility of a derivative based sampler and its efficiency on a small test problem.
-% In particular, we will show that proposals made using an Hamiltonian Monte Carlo (MC) sampler are accepted at a much higher rate than a traditional MCMC algorithm.
 In particular, we will show that the autocorrelation of a Hamiltonian Monte Carlo (HMC) sampler is significantly lower than a traditional MCMC algorithm~\citep{Bouffanais:2018hoz, 2014CQGra..31n5004P}.
@@
 Using a standard Gaussian likelihood, we then run the HMC sampler implemented in \flowMC~\citep{2022arXiv221106397W} for $2\times10^5$ steps and the random walk Metropolis Hastings (RWMH) sampler~\citep{1953JChPh..21.1087M} for $1.5\times10^6$ steps (each with four randomly initialized independent chains).
-% \AZ{What implementation of GRW? Can we cite something?}
-% \amc{briefly summarize and give refs for flowMC and GRW. Is GRW Metropolis-Hastings with a Gaussian transition kernel? How'd you tune this? Why not use NUTS for HMC?}
 The number of steps and and mass matrix used for each example was hand tuned to give good performance for the specific sampler.
-% \amc{not a huge deal, but why hand tune? There are automated ways to do this in e.g.\ numpyro with a warm-up adaptation phase.}
 The additional steps for the GRW sampler were required to achieve a similarly converged posterior.
-% \amc{Why not plot its posterior?}
 
@@
 
-% \te{Add paragraph with caveats.}
-% \te{Change GRW to random walk metrpolis hastings (RWMH)}
-
 In Fig.~\ref{fig:corner}, the grey contours show the posterior recovered using the best chain (i.e.~one that reached the highest log-likelihood values).
-% \mi{why show a single chain?}
 The orange shows the true parameters of the injected signal.
@@
 This is expected since the injected binary is relatively nearby with an SNR of $\sim13$.
-% \AZ{Can we give the SNR?}
 
@@
 
-% To illustrate the benefit of a gradient based sampler (such as HMC), we run the same sampling as above but instead use a Gaussian random walk (GRW) sampler. 
-% We sample for the same number of steps as used in the HMC with identical initializations to explore acceptance rates of new points.
-% For the HMC example, find that the mean local acceptance rate across the 15 chains is around $\sim0.28$, where as the local acceptance rate for the equivalent GRW sampler is XXX.
-% Since we are running the chains from the same initialization, this enhanced acceptance for HMC can be mainly associated with use of gradient information to guide sampling.
 Although further steps would be required to achieve a fully-converged posterior, these chains are sufficient to show the increased efficiency associated with HMC. 
@@
 We therefore expect \jax waveforms to be highly beneficial to future PE efforts in GW astronomy, particular for low-latency pipelines and higher dimensional analyses.
-% \kw{Not just low latency, but also high dimensional analysis, such as testing GR and lensing.}
 
@@
 Ideally, future waveforms should be implemented under an AD framework such as \jax.
-% \amc{if they're pure python, some translation will be required due to jax's restrictions.}
-% \mi{we could just say ``directly implemented under an autodifferentiation framework'' to be generic}
 This would ensure that the community can easily utilize differentiability and hardware acceleration in the future.

### 2023-10-04 src/tex/ms.tex
Starting updates for review

@@
-% Define document class
+git a% Define document class
 \documentclass[twocolumn]{aastex631}
@@
 
-The discovery of gravitational waves (GWs)~\citep{LIGOScientific:2016aoc} from inspiraling and merging compact objects (COs) has revolutionized our understanding of both fundamental physics and astronomy~\citep[e.g.][]{LIGOScientific:2021djp,LIGOScientific:2021sio,LIGOScientific:2021psn}.
+The first detection of gravitational waves (GWs)~\citep{LIGOScientific:2016aoc} from inspiraling and merging compact objects (COs) has revolutionized our understanding of both fundamental physics and astronomy~\citep[e.g.][]{LIGOScientific:2021djp,LIGOScientific:2021sio,LIGOScientific:2021psn}.
 Although the data volumes from GW detectors such as Advanced LIGO~\citep{LIGOScientific:2014pky} and Virgo~\citep{VIRGO:2014yos} are relatively small, analyzing the data is a computationally demanding task.
@@
 Unfortunately, in the field of GW data analysis, analytic derivatives of the necessary quantities (such as the likelihood) have historically required a significant amount of work to obtain~\citep{Keppel:2013kia}.
-With waveforms only increasing in complexity, calculating analytic derivatives will become ever more tricky.
+With waveforms only increasing in complexity, calculating analytic derivatives will become even trickier.
 Numerical derivatives also suffer from accuracy issues stemming from rounding or truncation errors.
@@
 
-In this paper we argue that differentiable waveforms will be a vital component for the future of GW data analysis.
+In this paper we argue that automatically differentiable waveforms may be a useful tool for the future of GW data analysis.
 In addition, we present \ripple, a small GW \texttt{python} package which, at the time of writing, includes a \jax implementation of the IMRPhenomD waveform and will be continually updated with new waveforms as they are implemented.
 The remainder of this paper is structured as follows. 
-In Sec.~\ref{sec:waveforms} we discuss the differentiable IMRPhenomD waveform implemented in \ripple and perform some benchmarks to demonstrate its speed and accuracy. 
+In Sec.~\ref{sec:waveforms} we discuss the automatically differentiable IMRPhenomD waveform implemented in \ripple and perform some benchmarks to demonstrate its speed and accuracy. 
 In Sec.~\ref{sec:applications} we discuss three distinct applications using differentiable waveforms. 
-Firstly, we illustrate how the fit coefficients that form part of the IMRPhenom waveform models could be improved by high dimensional fitting enabled by a differentiable waveform. 
+Firstly, we illustrate how the coefficients that form part of the IMRPhenomD waveform model could be improved by high dimensional fitting enabled by a automatically differentiable waveform. 
 Secondly, we implement differentiable detector response functions and show that the speed of Fisher matrix calculations can be substantially accelerated using AD.
-Finally, we run an illustrative injection example using Hamiltonian Monte Carlo to demonstrate that the autocorrelation of derivative samplers is substantially smaller than that of traditional MCMCs.
-Practically, this translates into much faster PE with no sacrifice in accuracy.
+Finally, we run an illustrative injection example using HMC to demonstrate that the autocorrelation of derivative samplers is substantially smaller than that of traditional MCMCs.
+% Practically, this translates into much faster PE with no sacrifice in accuracy.
 The associated code can be found at \ripple~\citep{ripple}.
 
-\section{Differentiable Waveforms}
+\section{Implementation and Benchmarking}
 \label{sec:waveforms}
@@
 
-The implementation of IMRPhenomD in \lalsuite is in C, and therefore needs to be rewritten natively into \texttt{python} to be compatible with \jax.
+The implementation of IMRPhenomD in \lalsuite~\citep{lalsuite} is in C, and therefore needs to be rewritten natively into \texttt{python} to be compatible with \jax.
 We have re-written IMRPhenomD from scratch using a combination of pure \texttt{python} and \jax derivatives.

### 2023-10-05 src/tex/ms.tex
Starting updates for review
More updates for the review

@@
-git a% Define document class
+% Define document class
 \documentclass[twocolumn]{aastex631}
@@
 Note however, that the differences are well below the accuracy requirements of the waveforms and will have no noticeable effect for realistic data analysis tasks.
+This is demonstrated in Fig.~\ref{fig:mismatch_hist} where the mismatch between NR waveforms and IMRPhenomD is always above $10^-5$.
 
@@
 Using \texttt{vmap} and performing the same benchmark as described above reduces the average evaluation speed to $\sim 0.14\,\mathrm{ms}$.
-Finally, \jax can natively on a GPU which allows for highly parallellized waveform evaluations.
+Finally, \jax can natively run on a GPU which allows for highly parallellized waveform evaluations.
 
@@
 Generalizing \lalsuite waveforms to run on a GPU would be a significant undertaking.
+Note that jit for both CPU and GPU evaluation adds an additional overhead to the first evaluation of the waveform which is not included in the above benchmarks. 
+Note also that the GPU benchmark does not include the time taken to move the evaluated waveform from the GPU to the CPU.
 
@@
 In particular, for every input dimension added, $D$, one would need to evaluate the function at least $2D$ times to calculate finite difference derivatives for all input parameters.
-For reverse-mode AD, one only needs two function calls to evaluate the derivative of all input parameters, regardless of dimension.\footnote{
+For reverse-mode AD, one only needs two function calls to evaluate the derivative of all input parameters, regardless of the number of dimensions.\footnote{
     Note that although the number of function calls is small, reverse-mode AD does add memory overhead.
@@
 }
+In gravitational wave physics one typically wants to compute derivatives of a scalar quantity (e.g., the likelihood) with respect to the intrinsic parameters of the binary (e.g., the masses and spins).
+Reverse mode AD is therefore ideal for this task.
 For a review of AD methods see~\cite{2018arXiv181105031M}.
@@
 
-Overall, we have demonstrated that the IMRPhenom waveform family is ideally suited for AD.
+Overall, we argued that the IMRPhenom waveform family is well suited for AD, showing it explicity for IMRPhenomD.
 Moreover, we have shown that our implementation of IMRPhenomD in \ripple is accurate and quick to evaluate, especially when hardware acceleration is available.
-In the next section, we will discuss a variety of potential use cases of differentiable waveforms.
+In the next section, we will discuss a variety of potential use cases of automatically differentiable waveforms.
 

### 2023-10-11 src/tex/ms.tex
More updates for the review
mostly finished editing section

@@
 Note however, that the differences are well below the accuracy requirements of the waveforms and will have no noticeable effect for realistic data analysis tasks.
-This is demonstrated in Fig.~\ref{fig:mismatch_hist} where the mismatch between NR waveforms and IMRPhenomD is always above $10^-5$.
+This is demonstrated in Fig.~\ref{fig:mismatch_hist} where the mismatch between NR waveforms and IMRPhenomD is always above $10^{-5}$.
 
@@
 Even though the \lalsuite implementation of IMRPhenomD is written in \texttt{C}, the serial evaluation of the waveform in \ripple is comparably fast. 
-Benchmarking on a MacBook Pro with an M1 Max Apple Silicon processor, we find that a single waveform evaluation takes $\sim 0.5\,\mathrm{ms}$ for \lalsuite (interfaced with python) and $\sim 0.4\,\mathrm{ms}$ for \ripple.\footnote{
-    For this benchmark we used $24\,\mathrm{Hz}$ and $512\,\mathrm{Hz}$ for the lowest and highest frequencies respectively. 
-    In addition, we used a frequency spacing of $0.2\,\mathrm{Hz}$. 
-    We performed this benchmark by evaluating the waveform $10^{4}$ times and taking the average evaluation time.
-}
+The evaluation time of a single waveform (averaged over $10^3$ calls) is shown in Fig.~\ref{fig:timing} as a function of the total mass. 
+In particular, we use $16\,\mathrm{Hz}$ and $1024\,\mathrm{Hz}$ for the lowest and highest frequencies respectively.
+To calculate the frequency spacing, we first estimate the length of time taken for the binary to merge using Eq.~(29) from~\cite{Yu:2023lml}. 
+Following LIGO convention, we take the closest factor of two higher than this time estimate as $T$. 
+The frequency spacing is then given as $\Delta f = 1/T$.
+
+% Benchmarking on a MacBook Pro with an M1 Max Apple Silicon processor, we find that a single waveform evaluation takes $\sim 0.5\,\mathrm{ms}$ for \lalsuite (interfaced with python) and $\sim 0.4\,\mathrm{ms}$ for \ripple.\footnote{
+%     For this benchmark we used $24\,\mathrm{Hz}$ and $512\,\mathrm{Hz}$ for the lowest and highest frequencies respectively. 
+%     In addition, we used a frequency spacing of $0.2\,\mathrm{Hz}$. 
+%     We performed this benchmark by evaluating the waveform $10^{4}$ times and taking the average evaluation time.
+% }
+For the CPU benchmarks we use a MacBook Pro with an M1 Max Apple Silicon processor. 
 Although similar on a CPU benchmark, \jax has a few key advantages. 
 First, its ability to JIT compile allows for significant performace gains (the above benchmark is already JIT compiled).
-Second, automatic vectorization can be achieved using the \texttt{vmap} function.
-Using \texttt{vmap} and performing the same benchmark as described above reduces the average evaluation speed to $\sim 0.14\,\mathrm{ms}$.
+Second, automatic vectorization can be achieved using the \texttt{vmap} function which leads to a noticeable speedup when evaluating a batch of waveforms.
 Finally, \jax can natively run on a GPU which allows for highly parallellized waveform evaluations.
 
-Again, performing the same benchmark as above on a NVIDIA Quadro 6000, we find that on average waveform evaluations take $\sim 0.02\,\mathrm{ms}$, over an of magnitude faster than serial CPU evaluation.
+Again, performing the same benchmark as above on an NVIDIA A100 with 40GB of memory, we find that waveform evaluations are over an of magnitude faster than serial CPU evaluation.
 Generalizing \lalsuite waveforms to run on a GPU would be a significant undertaking.
@@
 Note also that the GPU benchmark does not include the time taken to move the evaluated waveform from the GPU to the CPU.
+
+\begin{figure}[t]
+	\script{timing.py}
+	\begin{centering}
+		\includegraphics[width=\linewidth]{figures/timing.pdf}
+		\caption{
+			Benchmarking the evaluation time of a single IMRPhenomD waveform as a function of the total mass.
+As the total mass increases, the time to merger decreases leading to a coarser frequency grid and shorter evaluation times.
+The details of the benchmarks are given in the text.
+        }
+		\label{fig:timing}
+	\end{centering}
+\end{figure}
 
@@
 For a review of AD methods see~\cite{2018arXiv181105031M}.
-Since the parameter space of GWs in GR has $\mathcal{O}(10)$ dimensions, the speed of derivative evaluation is less crucial than the stability.
+Since the parameter space of GWs in GR has $\mathcal{O}(10)$ dimensions, the speed of derivative evaluation is less crucial than numerical stability.
 However, this might change for waveforms in beyond-GR models, waveforms involving equation of state parameters for neutron stars, and models which account for calibration and waveform uncertainties.

### 2023-10-11 src/scripts/timing.py

@@ -1,0 +1,59 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import matplotlib as mpl
+import paths
+from scipy import stats
+
+params = {
+    "font.size": 18,
+    "legend.fontsize": 18,
+    "legend.frameon": False,
+    "axes.labelsize": 18,
+    "axes.titlesize": 18,
+    "xtick.labelsize": 18,
+    "ytick.labelsize": 18,
+    "figure.figsize": (7, 5),
+    "xtick.top": True,
+    "axes.unicode_minus": False,
+    "ytick.right": True,
+    "xtick.bottom": True,
+    "ytick.left": True,
+    "xtick.major.pad": 8,
+    "xtick.major.size": 8,
+    "xtick.minor.size": 4,
+    "ytick.major.size": 8,
+    "ytick.minor.size": 4,
+    "xtick.direction": "in",
+    "ytick.direction": "in",
+    "axes.linewidth": 1.5,
+    "text.usetex": False,
+    "font.family": "serif",
+    "font.serif": "cmr10",
+    "mathtext.fontset": "cm",
+    "axes.formatter.use_mathtext": True,  # needed when using cm=cmr10 for normal text
+}
+
+mpl.rcParams.update(params)
+
+# palette = ['#2e4854', '#557b82','#bab2a9', '#c98769','#a1553a']
+palette = ['#306B37', 'darkgoldenrod', '#3F7BB6', '#BF4145', "#cf630a"]
+
+plt.rcParams['axes.prop_cycle'] = plt.cycler(color=palette)
+
+M_list = np.linspace(10, 100, 10)
+loop_time_list_lalsuite = np.loadtxt(paths.data / "loop_time_lalsuite.txt")
+loop_time_list = np.loadtxt(paths.data / "loop_time.txt")
+vmapped_time_list = np.loadtxt(paths.data / "vmapped_time.txt")
+gpu_time_list = np.loadtxt(paths.data / "vmapped_time_gpu.txt")
+
+plt.plot(M_list, loop_time_list_lalsuite * 1000, label="Lalsuite", lw = 3, color="k")
+plt.plot(M_list, loop_time_list * 1000, label="CPU (loop)", lw = 2, ls="-.")
+plt.plot(M_list, vmapped_time_list * 1000, label="CPU (vmap)", ls="--", lw = 2)
[…]

### 2023-10-18 src/tex/ms.tex
Adding more for the referee

@@
 
-Here, we illustrate how three core tasks in GW science can be substantially improved through the use of differentiable waveforms.
+Here, we illustrate how differentiable waveforms can contribute to the improvement of three core tasks in GW science.
 In this paper we primarily look at toy examples, leaving more careful analyses to future work. 
@@
 
-The coefficients of the original IMRPhenomD implementation are tuned in subsets of parameters instead of all together. 
-This means the tuning process ignores the correlation between different subsets of parameters, so the provided best-fit solution may not be the global optimum. 
+% The coefficients of the original IMRPhenomD implementation are tuned in subsets of parameters instead of all together. 
+Since the coefficients for each segment are tuned individually, the correlation between different parameters are ignored, meaning that the provided best-fit solution may not be the global optimum. 
 We therefore aim to improve of the accuracy of the waveform by jointly fitting all coefficients at once.
@@
     \label{eq:mismatch}
-	\mathcal{M}(\lambda)=1-\mathrm{m}(h_{\theta}^{\mathrm{IMR}}(\lambda)|h_{\theta}^{\mathrm{NR}}(\lambda))\,,
+	\mathcal{M}(\lambda)=1-\mathrm{m}(h_{\theta}^{\mathrm{IMR}}(\lambda)|h_{\theta}^{\mathrm{NR}})\,,
 \end{equation}
@@
 In Fig.~\ref{fig:mismatch_hist}, we show the distribution of log mismatches for a set of test waveforms.
-In particular, we use 536 waveforms from the SXS catalog (we simply choose all waveforms with aligned spins, $\chi_{x,y}<5\times 10^{-3}$, and eccentricity $<2\times 10^{-3}$)~\citep{Boyle:2019kee}.
+In particular, we use 536 waveforms from the SXS catalog (we     choose all waveforms with aligned spins, $\chi_{x,y}<5\times 10^{-3}$, and eccentricity $<2\times 10^{-3}$)~\citep{Boyle:2019kee}.
 One can see that the distribution of mismatches after optimization is generally shifted to lower mismatch compared to the original waveform.
@@
 While we focused on \texttt{IMRPhenomD} here, the ability to apply AD to the calibration parameters may assist in other approaches to calibration, such as that used for the aligned-spin EOB model in~\citet{Bohe:2016gbl}.
-There, MCMC methods were used in a two-step procedure to optimize the calibration parameters. 
-Derivative information, if implemented for EOB waveforms, could allow for the application of other sampling methods such as HMC, or possibly optimization over the entire set of NR waveforms at once.
+Automative derivatives, if implemented for EOB waveforms, could allow for the application of other sampling methods such as HMC, or possibly optimization over the entire set of NR waveforms at once.
 
@@
 We have verified that our errors agree with a separate dedicated Fisher forecasting code~\citep{Borhanian:2020ypi} to within ~30\%.
-This demonstrates that AD can be used to accurately produce population-level forecasts.
+This demonstrates that AD can be used to accurately do Fisher forecasting for a large population of events.
 
 Moreover, each error calculation (including computing the Fisher matrices for each detector and the inversion process) is substantially faster.
-In particular, we find that after compilation, each error calculation takes approximately half a second on a single computing core.
+In particular, we find that after compilation, each Fisher calculation takes approximately half a second on a single computing core.
 \texttt{GWbench}~\citep{Borhanian:2020ypi}, on the other hand, takes $\mathcal{O}$(minutes) for each Fisher calculation using the same detector setup and frequency grid.
 This factor of over 100 speed up is substantial considering the fact that a single core evaluation of the \ripple waveform is similar to \lalsuite which is used by \texttt{GWbench}.
-On a MacBook Pro with an M1 Max Apple Silicon processor, JIT compilation takes $\sim 270\,s$ and the full population analysis takes less than $\sim 9$ minutes.
+On a MacBook Pro with an M1 Max Apple Silicon processor, JIT compilation takes $\sim 270\,s$ and computing sky localization errors for the full population takes less than $\sim 9$ minutes.
 As discussed above, performance can be further improved by utilizing hardware acceleration such as parallel GPU processing.
@@
 Here we simply aim to demonstrate the utility of a derivative based sampler and its efficiency on a small test problem.
-In particular, we will show that the autocorrelation of a Hamiltonian Monte Carlo (HMC) sampler is significantly lower than a traditional MCMC algorithm~\citep{Bouffanais:2018hoz, 2014CQGra..31n5004P}.
+In particular, we will show that the autocorrelation of a HMC sampler is significantly lower than a traditional MCMC algorithm~\citep{Bouffanais:2018hoz, 2014CQGra..31n5004P}.
 
@@
 The number of steps and and mass matrix used for each example was hand tuned to give good performance for the specific sampler.
-The additional steps for the GRW sampler were required to achieve a similarly converged posterior.
+The additional steps for the RWMH sampler were required to achieve a similarly converged posterior.
 
@@
 		\caption{
-			Autocorrelation as a function of lag for HMC and GRW on the same simulated data as discussed in Fig.~\ref{fig:corner}.
+			Autocorrelation as a function of lag for HMC and RWMH on the same simulated data as discussed in Fig.~\ref{fig:corner}.
             The smaller autocorrelation of the HMC leads to a larger number of independent samples and therefore a faster converging Monte Carlo.
@@
 Although further steps would be required to achieve a fully-converged posterior, these chains are sufficient to show the increased efficiency associated with HMC. 
-To further illustrate this, in Fig.~\ref{fig:autocorrelation} we plot the autocorrelation as a function of lag for both the HMC and GRW samplers.
-The HMC autocorrelation is substantially lower than that of the GRW.
+To further illustrate this, in Fig.~\ref{fig:autocorrelation} we plot the autocorrelation as a function of lag for both the HMC and RWMH samplers.
+The HMC autocorrelation is substantially lower than that of the RWMH.
 We therefore expect gradient based samplers to converge significantly faster than typical samplers, especially in higher dimensions.
@@
 In particular, we showed that it matches the \lalsuite implementation to near machine precision and can be easily parallelized on a GPU.
-Parallelization on a GPU provides substantial speed increases; on a NVIDIA Quadro 6000 GPU we found that waveform evaluations are over an order of magnitude faster than serial CPU evaluations.
+Parallelization on a GPU provides substantial speed increases; on an A100 40GB GPU we found that waveform evaluations are over an order of magnitude faster than serial CPU evaluations.
 Second, we discussed three data analysis tasks which can all be substantially improved by utilizing derivative information of the waveform.
+\te{Add a couple of sentences}
+
 Although we primarily discuss toy examples in this paper, each can be extended to the full data analysis task, some of which will be shown in upcoming papers~\citep{PEpaper}.
@@
 In this paper, we have primarily focussed on the IMRPhenom family of waveforms as their closed form expression is perfectly suited for a \jax implementation.
-Two other waveform families are commonly used in GW data analysis: the effective-one-body (EOB) and numerical relativity surrogate (NRSurrogate). 
+% Two other waveform families are commonly used in GW data analysis: the effective-one-body (EOB) and numerical relativity surrogate (NRSurrogate). 
 A differentiable NRsurrogate implementation is under development~\citep{NRSurAD}, but it currently seems difficult to implement EOB waveforms in \jax.

### 2023-10-18 src/scripts/loss_compare.py

@@ -63,6 +63,7 @@ plt.plot(freq, opt_error, label="Optimized")
 plt.legend()
 plt.xlabel(r"$Mf$")
 plt.ylabel(r"$\Delta|\tilde{h}(f)|$")
-plt.xlim(0.01, 0.09)
+# plt.xlim(0.01, 0.2)
 plt.ylim(0.0, 0.05)
+plt.xscale("log")
 plt.savefig(paths.figures / "loss_compare.pdf", bbox_inches="tight")

### 2023-11-06 src/tex/ms.tex
Slight updates

@@
         Corner plot for the posteriors from an HMC (see text for details) on simulated noise with injected signal. 
-        Orange lines indicate the true values of the injection. 
+        Blue lines indicate the true values of the injection. 
         Although not fully converged, it is clear that we find posteriors consistent with the injected parameters. 
@@
 In Fig.~\ref{fig:corner}, the grey contours show the posterior recovered using the best chain (i.e.~one that reached the highest log-likelihood values).
-The orange shows the true parameters of the injected signal.
+The blue shows the true parameters of the injected signal.
 From the one dimensional histograms along the diagonal, it is clear that we consistently recover all seven parameters apart from $\phi_c$.
@@
 
-In a follow up paper we will demonstrate that minute scale PE can be achieved by combining normalizing flows~\citep{2022arXiv221106397W, Gabrie:2021tlu}, GPU acceleration, and a derivative based sampler~\citep{PEpaper}.
+In our companion paper we demonstrate that minute scale PE can be achieved by combining normalizing flows~\citep{2022arXiv221106397W, Gabrie:2021tlu}, GPU acceleration, and a derivative based sampler~\citep{PEpaper}.
 We therefore expect \jax waveforms to be highly beneficial to future PE efforts in GW astronomy, particular for low-latency pipelines and higher dimensional analyses.
@@
 Second, we discussed three data analysis tasks which can all be substantially improved by utilizing derivative information of the waveform.
-\te{Add a couple of sentences}
+In particular, we showed that AD can be used to fine-tune waveform coefficients, perform efficient and accurate Fisher forecasting, and enhance the efficiency of sampling algorithms during parameter estimation.
 

### 2023-12-05 src/tex/ms.tex
Finished edited 3.1

@@
 NRsurrogate models, which interpolate directly over waveforms produced by numerical relativity (NR) simulations, are in principle also straightfoward to implement in \jax.
-EOB waveforms on the other hand are produced by evolving the dynamics of an effective one body Hamiltonian, and are therefore more difficult to implement in \jax.
+EOB waveforms on the other hand are produced by evolving the dynamics of an effective one body Hamiltonian, and are therefore more difficult to implement in \jax.\footnote{
+    In particular, the dynamically chosen time step for solving ODEs accurately makes JIT compilation in JAX difficult.
+However, the \texttt{diffrax} package \citep{diffrax} has shown good performance for ODE solving in a variety of settings and could potentially be used to implement EOB waveforms in \jax.
+    }
 For EOB waveforms, frequency-domain reduced-order models may be a convenient target~\citep[e.g.]{Cotesta:2020qhw}.
@@
 generated by the NR simulation. 
-Given the loss function, we use gradient descent to update the fitting
-coefficients:
+Given the loss function, we use gradient descent to update the fitting coefficients:
 \begin{equation}
@@
 \end{equation}
-where $\alpha$ is the learning rate. We set $\alpha$ to be $10^{-6}$.
+where $\alpha$ is the learning rate and we can compute $\nabla\mathcal{M}$ using AD. We set $\alpha$ to be $10^{-6}$.
 
@@
 To evaluate Eq.~\eqref{eq:multi_loss} we use a flat PSD and a frequency array which is scaled by the total mass.\footnote{
-    In particular, we use $M f_l = 2.5 \times 10^{-3}$, $M f_u = 0.2$, and $M\Delta f = 2.5 \times 10^{-6}$ for the dimensionless lower, upper, and frequency spacing respectively.
+    In particular, we use $M f_{\mathrm{l}} = 2.5 \times 10^{-3}$, $M f_{\mathrm{u}} = 0.2$, and $M\Delta f = 2.5 \times 10^{-6}$ for the dimensionless lower, upper, and frequency spacing respectively.
 We fix $M$ to be $50\,\mathrm{M}_\odot$ throughout this section.
@@
 These 11 waveforms are taken from the SXS catalog~\citep{Boyle:2019kee}.
-We then run gradient descent, as described above, for $3\times 10^4$ steps. 
+We then run gradient descent, as described above, until the validation loss stops decreasing (the validation is loss is calculated using the NR waveforms discussed below). 
 Figure~\ref{fig:loss_compare} shows the relative error (against a test NR waveform; see below) of the original and optimized waveform as a function of dimensionless frequency.
@@
 We can see the error of optimized waveform is lower than that of the original waveform for most of the domain.
-In particular, the error in the merger region is decreased by half while other regions also show good improvement in accuracy.
+In particular, the relative error in the merger region (in between the two vertical dashed lines) is decreased by half while other regions also show good improvement in accuracy.
 
@@
 A differentiable NRsurrogate implementation is under development~\citep{NRSurAD}, but it currently seems difficult to implement EOB waveforms in \jax.
-In particular, the evolution of the Hamiltonian required to evaluate an EOB waveform is both inherently slow to differentiate and difficult to implement in \jax. 
+As mentioned in the introduction, the evolution of the Hamiltonian required to evaluate an EOB waveform is both inherently slow to differentiate and difficult to implement in \jax although some progress has been made~\citep{diffrax}.
 Since EOB methods are used to produce state-of-the-art waveforms for many applications, more work is required to see if a fast, differentiable implementation is possible.

### 2023-12-05 src/scripts/loss_compare.py

@@ -55,15 +55,46 @@ G = 6.67430e-11  # m^3 / kg / s^2
 C = 299792458.0  # m / s
 gt = G * MSUN / (C ** 3.0)
 
-f2 =  149.1018563859192 * (25. + 25.) * gt
-plt.axvline(x=f2, color="k", alpha=0.4, ls="--")
+c1 = "#102F68"  # blue
+c2 = "#B02423"  # red
 
-plt.plot(freq, ori_error, label="Original")
-plt.plot(freq, opt_error, label="Optimized")
+# f2 =  149.1018563859192 * (25. + 25.) * gt
+# plt.axvline(x=f2, color="k", alpha=0.4, ls="--")
+
+f_ins = 0.018
+f_rd = 149.1018563859192 * (25.0 + 25.0) * gt
+
+data = np.loadtxt(paths.data / "0154_compare.dat")
+
+f_uniform = data[:, 0]
+NR_amp = data[:, 1]
+NR_angle = data[:, 2]
+IMR_amp = data[:, 3]
+IMR_angle = data[:, 4]
+IMR_opt_amp = data[:, 5]
+IMR_opt_angle = data[:, 6]
+
+plt.semilogx(
+    f_uniform,
+    (NR_amp - IMR_amp)/ NR_amp,
+    label="IMR original",
+    linestyle=(0, (3, 1)),
+    color=c1,
+    lw=2,
+)
+plt.semilogx(
+    f_uniform, (NR_amp - IMR_opt_amp)/ NR_amp, label="IMR optimized", color=c2, lw=2
+)
+plt.axvline(x=f_ins, color="k", alpha=0.5, linestyle=(0, (1, 1.05)))
+plt.axvline(x=f_rd, color="k", alpha=0.5, linestyle=(0, (1, 1.05)))
+
+# plt.plot(freq, ori_error, label="Original")
+# plt.plot(freq, opt_error, label="Optimized")
 plt.legend()
 plt.xlabel(r"$Mf$")
-plt.ylabel(r"$\Delta|\tilde{h}(f)|$")
+# plt.ylabel(r"$\Delta|\tilde{h}(f)|$")
+plt.ylabel(r"$(\tilde{h}-\tilde{h}_{\mathrm{NR}})/\tilde{h}_{\mathrm{NR}}$")
 # plt.xlim(0.01, 0.2)
-plt.ylim(0.0, 0.05)
+# plt.ylim(0.0, 0.05)
 plt.xscale("log")
[…]

### 2023-12-05 src/scripts/mismatch_hist.py

@@ -5,16 +5,16 @@ import paths
 from scipy import stats
 
 params = {
-    "font.size": 18,
-    "legend.fontsize": 18,
+    "font.size": 15,
+    "legend.fontsize": 12,
     "legend.frameon": False,
-    "axes.labelsize": 18,
-    "axes.titlesize": 18,
-    "xtick.labelsize": 18,
-    "ytick.labelsize": 18,
-    "figure.figsize": (7, 5),
-    "xtick.top": True,
+    "axes.labelsize": 15,
+    "axes.titlesize": 15,
+    "xtick.labelsize": 15,
+    "ytick.labelsize": 15,
     "axes.unicode_minus": False,
+    "figure.figsize": (6, 5),
+    "xtick.top": True,
     "ytick.right": True,
     "xtick.bottom": True,
     "ytick.left": True,
@@ -36,26 +36,62 @@ params = {
 
 mpl.rcParams.update(params)
 
-(catalog_num, ori_loss, opt_loss) = np.loadtxt(
-    paths.data / "mismatch_hist.txt", unpack=True
+# (catalog_num, ori_loss, opt_loss) = np.loadtxt(
+#     paths.data / "mismatch_hist.txt", unpack=True
+# )
+q148 = np.loadtxt(paths.data / "q148.dat")
+
+
+c1 = "#102F68"
+c2 = "#B02423"
+
+print("Median Change is:", (np.median(q148[:, 5]) - np.median(q148[:, 4]))*100/np.median(q148[:, 4]))
+
+q148 = np.loadtxt(paths.data / "q148.dat")
+bins = np.linspace(-5, -1, 15)
+
+plt.hist(np.log10(q148[:, 4]), bins=bins, alpha=0.2, color=c1)
+plt.hist(
+    np.log10(q148[:, 4]),
+    bins=bins,
+    color=c1,
+    histtype="step",
+    linewidth=2,
+    label="Original",
+)
+plt.axvline(
+    x=np.log10(np.median(q148[:, 4])),
+    color=c1,
+    zorder=-20,
+    alpha=1.0,
+    lw=2,
+    linestyle=(0, (1, 1.3)),
 )
 
-print("Median Change is:", (np.median(opt_loss) - np.median(ori_loss))*100/np.median(ori_loss))
+plt.hist(
+    np.log10(q148[:, 5]),
+    bins=bins,
+    histtype="step",
[…]

### 2023-12-15 src/tex/ms.tex
nearly done

@@
 To evaluate Eq.~\eqref{eq:multi_loss} we use a flat PSD and a frequency array which is scaled by the total mass.\footnote{
-    In particular, we use $M f_{\mathrm{l}} = 2.5 \times 10^{-3}$, $M f_{\mathrm{u}} = 0.2$, and $M\Delta f = 2.5 \times 10^{-6}$ for the dimensionless lower, upper, and frequency spacing respectively.
+    In particular, we use $M f_{\mathrm{l}} = 2.5 \times 10^{-3}$, $M f_{\mathrm{u}} = 1.2 f_{\textrm{RD}}$ ($f_{\textrm{RD}}$ is defined in \cite{Khan:2015jqa}), and $M\Delta f = 2.5 \times 10^{-6}$ for the dimensionless lower, upper, and frequency spacing respectively.
 We fix $M$ to be $50\,\mathrm{M}_\odot$ throughout this section.
@@
     \caption{
-        Corner plot for the posteriors from an HMC (see text for details) on simulated noise with injected signal. 
+        Corner plot for the posteriors (see text for details) on simulated noise with injected signal using HMC (blue) and RWMH (black). 
         Blue lines indicate the true values of the injection. 
@@
 Although further steps would be required to achieve a fully-converged posterior, these chains are sufficient to show the increased efficiency associated with HMC. 
-To further illustrate this, in Fig.~\ref{fig:autocorrelation} we plot the autocorrelation as a function of lag for both the HMC and RWMH samplers.
+To further illustrate this, in Fig.~\ref{fig:autocorrelation} we plot the autocorrelation as a function of lag\footnote{
+    Lag here corresponds to the number of steps between samples.
+} for both the HMC and RWMH samplers.
 The HMC autocorrelation is substantially lower than that of the RWMH.
+In particular, we see that the autocorrelation reaches close to zero at around 300 lag, meaning that samples become efficiently uncorrelated.
+This is in contrast to RWHM which remains highly correlated even at 1000 lag.
 We therefore expect gradient based samplers to converge significantly faster than typical samplers, especially in higher dimensions.
-In addition, we found that the effictive number of samples~\citep{arviz_2019},\footnote{
-    Computed using \textsc{arviz}~\url{https://python.arviz.org/en/stable/api/generated/arviz.ess.html}
+In addition, we found that the effictive number of samples~\citep{arviz_2019}, (a measure of the number of indepedent samples)\footnote{
+    Computed using \textsc{arviz}~\url{https://python.arviz.org/en/stable/api/generated/arviz.ess.html}.
     }
-(a measure of the number of indepedent samples) is between 2 and 7 times larger for HMC across the different dimensions of the parameter space.
+is between 2 and 7 times larger for HMC across the different dimensions of the parameter space.
 

### 2023-12-15 src/scripts/corner_plot.py

@@ -74,7 +74,9 @@ true_params = np.array(
 # Plot all chains
 # figure = corner.corner(gaussian_bestchain, labels=labels, labelpad=0.1, smooth=True, color="C1", alpha=0.3, label="GRW")
 
-figure = corner.corner(HMC_bestchain, labels=labels, labelpad=0.1, smooth=True, truths=[Mc, eta, 0.3, -0.4, distance, 0.0, 0.0])
+figure = corner.corner(HMC_bestchain, labels=labels, labelpad=0.1, smooth=True, truths=[Mc, eta, 0.3, -0.4, distance, 0.0, 0.0], label="HMC", color="C0", hist_kwargs={"density": True})
+
+corner.corner(gaussian_bestchain, labels=labels, labelpad=0.1, smooth=True, color="black", fig=figure, alpha=0.1, label="RWMH", hist_kwargs={"density": True})
 
 
 # Extract the axes

### 2023-12-19 src/tex/ms.tex
so close

@@
     We propose the use of automatic differentiation through the programming framework \jax for accelerating a variety of analysis tasks throughout gravitational wave (GW) science.
-    Firstly, we demonstrate that complete waveforms which cover the inspiral, merger, and ringdown of binary black holes (i.e.~IMRPhenomD) can be written in \jax and demonstrate that the serial evaluation speed of the waveform (and its derivative) is similar to the \lalsuite implementation in \texttt{C}.
+    Firstly, we demonstrate that complete waveforms which cover the inspiral, merger, and ringdown of binary black holes (i.e.,~IMRPhenomD) can be written in \jax and demonstrate that the serial evaluation speed of the waveform (and its derivative) is similar to the \lalsuite implementation in \texttt{C}.
     Moreover, \jax allows for GPU-accelerated waveform calls which can be over an order of magnitude faster than serial evaluation on a CPU.
@@
     Firstly, we demonstrate how gradient descent can be used to optimize the $\sim 200$ coefficients that are used to calibrate the waveform model. 
-    In particular, we demonstrate that the typical \textit{match} with numerical relativity waveforms can be improved by more than 50\% without any additional overhead.
+    In particular, we demonstrate that the typical \textit{match} with numerical relativity waveforms can be improved by more than 50\%.
     Secondly, we show that Fisher forecasting calculations can be sped up by $\sim 100\times$ (on a CPU) with no loss in accuracy.
-    This increased speed makes population forecasting substantially simpler.
+    This increased speed makes Fisher forecasting for a population of events substantially simpler.
     Finally, we show that gradient-based samplers like Hamiltonian Monte Carlo lead to significantly reduced autocorrelation values when compared to traditional Monte Carlo methods.
@@
 
-The first detection of gravitational waves (GWs)~\citep{LIGOScientific:2016aoc} from inspiraling and merging compact objects (COs) has revolutionized our understanding of both fundamental physics and astronomy~\citep[e.g.][]{LIGOScientific:2021djp,LIGOScientific:2021sio,LIGOScientific:2021psn}.
+The first detection of gravitational waves (GWs)~\citep{LIGOScientific:2016aoc} from inspiraling and merging compact objects (COs) has revolutionized our understanding of both fundamental physics and astronomy~\citep[e.g.,][]{LIGOScientific:2021djp,LIGOScientific:2021sio,LIGOScientific:2021psn}.
 Although the data volumes from GW detectors such as Advanced LIGO~\citep{LIGOScientific:2014pky} and Virgo~\citep{VIRGO:2014yos} are relatively small, analyzing the data is a computationally demanding task.
@@
 Beyond these simple scenarios, more complex waveform models with additional parameters may be used to include calibration errors~\citep{Farr:2014,Vitale:2020gvb}, model the tides of neutron stars~\citep{LIGOScientific:2018hze}, and test for deviations from General Relativity~\citep{Arun:2006yw, Agathos:2013upa, Yunes:2016jcc, LIGOScientific:2016lio, LIGOScientific:2020tif, Krishnendu:2021fga}.
-Finally, using the results of PE, population synthesis models constrain the progenitors systems from which the black holes we see merging today began their journey~\citep{LIGOScientific:2020kqk, LIGOScientific:2021psn, Wong:2022flg}.
+Finally, using the results of PE, population synthesis models constrain the progenitor systems from which the black holes we see merging today began their journey~\citep{LIGOScientific:2020kqk, LIGOScientific:2021psn, Wong:2022flg}.
 Overall, GW data analysis therefore requires significant computation.
-In this paper, we will argue that differentiable waveforms (and more generally differentiable pipelines) can play a significant role in alleviating this computational demand.
+In this paper, we will argue that differentiable waveforms (and more generally differentiable pipelines) can play a useful role in alleviating this computational demand.
 
 Derivatives are ubiquitously useful throughout data analysis tasks.
-For instance, during PE, derivative information can be used to guide an optimizer towards higher-likelihood values (e.g.\,using gradient descent~\citep{2016arXiv160904747R}) or allow a sampler to rapidly explore parameter space (e.g.\, using Hamiltonian Monte Carlo (HMC)~\citep{2011hmcm.book..113N,2017arXiv170102434B}).
+For instance, during PE, derivative information can be used to guide an optimizer towards higher-likelihood values (e.g.,\,using gradient descent~\citep{2016arXiv160904747R}) or allow a sampler to rapidly explore parameter space (e.g.,\, using Hamiltonian Monte Carlo (HMC)~\citep{2011hmcm.book..113N,2017arXiv170102434B}).
 Gradients are particularly valuable for high dimensional spaces.
@@
     }
-The full derivative can then be constructed using the chain rule.
+The full derivative can then be constructed using the chain rule~\citep{2018arXiv181105031M}.
 There are now a variety of AD implementations, most notably in deep learning frameworks such as \texttt{pytorch}~\citep{pytorch} and \texttt{tensorflow}~\citep{tensorflow2015-whitepaper}.
@@
 This makes a \jax implementation that complies with the constraints of JIT compilation simple.
-NRsurrogate models, which interpolate directly over waveforms produced by numerical relativity (NR) simulations, are in principle also straightfoward to implement in \jax.
+NRsurrogate models, which interpolate directly over waveforms produced by numerical relativity (NR) simulations, are in principle also straightfoward to implement in \jax and will be explored int he future.
 EOB waveforms on the other hand are produced by evolving the dynamics of an effective one body Hamiltonian, and are therefore more difficult to implement in \jax.\footnote{
-    In particular, the dynamically chosen time step for solving ODEs accurately makes JIT compilation in JAX difficult.
+    In particular, the dynamically chosen time step for solving ODEs accurately makes JIT compilation in \jax difficult.
 However, the \texttt{diffrax} package \citep{diffrax} has shown good performance for ODE solving in a variety of settings and could potentially be used to implement EOB waveforms in \jax.
     }
-For EOB waveforms, frequency-domain reduced-order models may be a convenient target~\citep[e.g.]{Cotesta:2020qhw}.
+For EOB waveforms, frequency-domain reduced-order models may therefore be a more convenient target~\citep[e.g.,][]{Cotesta:2020qhw}.
 
@@
     }
-This is shown in Fig.~\ref{fig:match}, where we have calculated the $\mathrm{m}(h_1|h_2)$ across the entire parameter space.\footnote{
-    Specifically, we use $10^4$ points varying component masses $m_1$, $m_2$ and spin parameters $\chi_1$, $\chi_2$ in the ranges $m_{1,2} = (1,100)\,M_{\odot}$ and $\chi_{1,2} = (-1,1)$.
+This is shown in Fig.~\ref{fig:match}, where we have calculated the $\mathcal{m}(h_1|h_2)$ across the entire parameter space.\footnote{
+    Specifically, we use $10^4$ points with varying component masses $m_1$, $m_2$ and spin parameters $\chi_1$, $\chi_2$ in the ranges $m_{1,2} = (1,100)\,M_{\odot}$ and $\chi_{1,2} = (-1,1)$.
 In addition, we evaulated the waveforms on a frequency grid from $32\,$Hz to $1024\,$Hz with frequency spacing $\Delta f = 0.0125 \,$Hz.
@@
 Unfortunately, we cannot make the initial cubic interpolation arbitrarily fine as this would add additional computational overhead when loading the data during waveform evaluation.
-Additionally, we use a Fourier transform to efficiently maximize over $\Delta t_c$ which will inherently come with some inaccuracy.
-This inaccuracy will primarily affect the mismatch at higher total masses where the waveform is dominated by the merger.
+% Additionally, we use a Fourier transform to efficiently maximize over $\Delta t_c$ which will inherently come with some inaccuracy.
+% This inaccuracy will primarily affect the mismatch at higher total masses where the waveform is dominated by the merger.
 Note however, that the differences are well below the accuracy requirements of the waveforms and will have no noticeable effect for realistic data analysis tasks.
@@
 In particular, for every input dimension added, $D$, one would need to evaluate the function at least $2D$ times to calculate finite difference derivatives for all input parameters.
-For reverse-mode AD, one only needs two function calls to evaluate the derivative of all input parameters, regardless of the number of dimensions.\footnote{
+For reverse-mode AD~\citep{2018arXiv181105031M}, one only needs two function calls to evaluate the derivative of all input parameters, regardless of the number of dimensions.\footnote{
     Note that although the number of function calls is small, reverse-mode AD does add memory overhead.
@@
 \begin{equation}
+    \label{eq:grad_descent}
 	\lambda\leftarrow\lambda-\alpha\nabla\mathcal{M}\,, 
@@
 \end{equation}
-where $\mathcal{M}_i$ is the mismatch of an individual training waveform and $N$ is the total number of training waveforms used in the optimization.
+where $\mathcal{M}_i$ is the mismatch of an individual training waveform and $N$ is the total number of training waveforms used in the optimization.\footnote{
+    Of course, now we replace $\mathcal{M}$ with $\mathcal{L}$ in the gradient descent update~\eqref{eq:grad_descent}.
+}
 This optimization is more difficult since we are now applying the same set of coefficients to waveforms with different intrinsic parameters, such as mass ratio and spins. 
@@
 Figure~\ref{fig:loss_compare} shows the relative error (against a test NR waveform; see below) of the original and optimized waveform as a function of dimensionless frequency.
-The vertical dashed lines indicate the stitching points for the phase i.e. when the inspiral is joined onto the merger.
+The vertical dashed lines indicate the stitching points for the phase i.e., when the inspiral is joined onto the merger.
 We can see the error of optimized waveform is lower than that of the original waveform for most of the domain.
@@
 For example, sampling the 15 dimensional parameter space for a BBH system can take $\mathcal{O}(10)$ hours, while BNS systems can take up to weeks.
-Dedicated fast samplers have been designed to get approximate posteriors on the sky localization to facilitate follow-up electromagnetic observations \citep[e.g.~\texttt{BAYESTAR}][]{Singer:2015ema}. 
+Dedicated fast samplers have been designed to get approximate posteriors on the sky localization to facilitate follow-up electromagnetic observations \citep[e.g.,~\texttt{BAYESTAR}][]{Singer:2015ema}.
 Moreover, a number of methods have been developed to speed up PE well below the numbers quoted above~\citep{Dax:2021tsq, Islam:2022afg,Roulet:2022kot, Zackay:2018qdy, Cornish:2021lje, Canizares:2013ywa, Leslie:2021ssu}.
@@
 Here we simply aim to demonstrate the utility of a derivative based sampler and its efficiency on a small test problem.
-In particular, we will show that the autocorrelation of a HMC sampler is significantly lower than a traditional MCMC algorithm~\citep{Bouffanais:2018hoz, 2014CQGra..31n5004P}.
+In particular, we will show that the autocorrelation of an HMC sampler is significantly lower than a traditional MCMC algorithm~\citep{Bouffanais:2018hoz, 2014CQGra..31n5004P}.
 
@@
 Using a standard Gaussian likelihood, we then run the HMC sampler implemented in \flowMC~\citep{2022arXiv221106397W} for $2\times10^5$ steps and the random walk Metropolis Hastings (RWMH) sampler~\citep{1953JChPh..21.1087M} for $1.5\times10^6$ steps (each with four randomly initialized independent chains).
-The number of steps and and mass matrix used for each example was hand tuned to give good performance for the specific sampler.
+The number of steps and mass matrix used for each example was hand tuned to give good performance for the specific sampler.
 The additional steps for the RWMH sampler were required to achieve a similarly converged posterior.
@@
 
-In Fig.~\ref{fig:corner}, the grey contours show the posterior recovered using the best chain (i.e.~one that reached the highest log-likelihood values).
-The blue shows the true parameters of the injected signal.
-From the one dimensional histograms along the diagonal, it is clear that we consistently recover all seven parameters apart from $\phi_c$.
+In Fig.~\ref{fig:corner}, the blue contours show the posterior recovered by HMC using the best chain (i.e.,~one that reached the highest log-likelihood values).
+The blue lines show the true parameters of the injected signal.
+From the one dimensional histograms along the diagonal, it is clear that we consistently recover all seven parameters apart from $\phi_c$ (which is known to be measured poorly).
 This is expected since the injected binary is relatively nearby with an SNR of $\sim13$.
@@
 In our companion paper we demonstrate that minute scale PE can be achieved by combining normalizing flows~\citep{2022arXiv221106397W, Gabrie:2021tlu}, GPU acceleration, and a derivative based sampler~\citep{PEpaper}.
-We therefore expect \jax waveforms to be highly beneficial to future PE efforts in GW astronomy, particular for low-latency pipelines and higher dimensional analyses.
+We therefore expect \jax waveforms to be beneficial to future PE efforts in GW astronomy, particular for low-latency pipelines and higher dimensional analyses.
 
@@
 Although we primarily discuss toy examples in this paper, each can be extended to the full data analysis task, some of which will be shown in upcoming papers~\citep{PEpaper}.
-Differentiable waveforms therefore represent a crucial advancement towards efficient GW science.
+Differentiable waveforms therefore represent a useful advancement towards efficient GW science.
 
@@
 A differentiable NRsurrogate implementation is under development~\citep{NRSurAD}, but it currently seems difficult to implement EOB waveforms in \jax.
-As mentioned in the introduction, the evolution of the Hamiltonian required to evaluate an EOB waveform is both inherently slow to differentiate and difficult to implement in \jax although some progress has been made~\citep{diffrax}.
+As mentioned in the introduction, the evolution of the Hamiltonian required to evaluate an EOB waveform is both inherently slow to differentiate and difficult to implement in \jax, although some progress has been made~\citep{diffrax}.
 Since EOB methods are used to produce state-of-the-art waveforms for many applications, more work is required to see if a fast, differentiable implementation is possible.

### 2023-12-20 src/tex/ms.tex
penultemate commit

@@
 
-To illustrate the computational speed of computing Fisher matrices with AD, we consider a population of binaries and compute the sky localization error following Eq.~(28) in \cite{Iacovelli:2022bbs, Iacovelli:2022mbg}.
+To illustrate the computational speed of computing Fisher matrices with AD, we consider a population of binaries and compute the sky localization error following Eq.~(28) in \cite{Iacovelli:2022bbs, Iacovelli:2022mbg}.\footnote{
+    Note that we include all 11 parameters when computing the covariance matrix, and only plot the sky localization error for illustrative purposes.
+}
 Since the Fisher matrix approach is known to have both theoretical issues as well as numerical instabilities for low signal-to-noise events, we restrict our population to only nearby systems.
@@
 \section{Acknowledgments}
+T.E.
+would like to thanks Luca Reali extensively for his help with running GWBench.
 This material is based upon work supported by NSF's LIGO Laboratory which is a major facility fully funded by the National Science Foundation.

### 2024-01-11 src/tex/ms.tex
finished version ready for reply

@@
     In particular, we demonstrate that the typical \textit{match} with numerical relativity waveforms can be improved by more than 50\%.
-    Secondly, we show that Fisher forecasting calculations can be sped up by $\sim 100\times$ (on a CPU) with no loss in accuracy.
+    Secondly, we show that Fisher forecasting calculations can be sped up by $\sim 3-5\times$ (on a CPU) with no loss in accuracy.
     This increased speed makes Fisher forecasting for a population of events substantially simpler.
@@
 
+\begin{figure}[t!]
+	\script{loss_compare.py}
+	\begin{centering}
+		\includegraphics[width=\linewidth]{figures/loss_compare.pdf}
+		\caption{
+			Relative error between the numerical relativity and IMRPhenomD waveform amplitudes. 
+            The blue line shows the relative error of the IMRPhenomD model with the original model coefficients where as the orange line uses the IMRPhenomD model after optimization using gradient descent.
+            Interestingly, the error is reduced across the frequency range, demonstrating that high dimensional optimization is useful for all parts of the waveform.
+        }
+		\label{fig:loss_compare}
+	\end{centering}
+\end{figure}
+
+\begin{figure}[t!]
+	\script{mismatch_hist.py}
+	\begin{centering}
+		\includegraphics[width=\linewidth]{figures/mismatch_hist.pdf}
+		\caption{
+			Distribution of 536 log mismatches (see Eq.~\eqref{eq:mismatch}) for the original (blue) and optimized (orange) IMRPhenomD. 
+            Overall, the optimized distribution shifts to lower mismatches and therefore a better waveform model. 
+            More quantitatively, the median mismatch (shown as vertical dashed lines) is reduced by $\sim50$\%. 
+		}
+		\label{fig:mismatch_hist}
+	\end{centering}
+\end{figure}
+
 The first step is to define a loss function that measures the goodness-of-fit of the current waveform coefficients.
@@
 
-\begin{figure}[t]
-	\script{loss_compare.py}
-	\begin{centering}
-		\includegraphics[width=\linewidth]{figures/loss_compare.pdf}
-		\caption{
-			Relative error between the numerical relativity and IMRPhenomD waveform amplitudes. 
-            The blue line shows the relative error of the IMRPhenomD model with the original model coefficients where as the orange line uses the IMRPhenomD model after optimization using gradient descent.
-            Interestingly, the error is reduced across the frequency range, demonstrating that high dimensional optimization is useful for all parts of the waveform.
-        }
-		\label{fig:loss_compare}
-	\end{centering}
-\end{figure}
-
-\begin{figure}[t]
-	\script{mismatch_hist.py}
-	\begin{centering}
-		\includegraphics[width=\linewidth]{figures/mismatch_hist.pdf}
-		\caption{
-			Distribution of 536 log mismatches (see Eq.~\eqref{eq:mismatch}) for the original (blue) and optimized (orange) IMRPhenomD. 
-            Overall, the optimized distribution shifts to lower mismatches and therefore a better waveform model. 
-            More quantitatively, the median mismatch (shown as vertical dashed lines) is reduced by $\sim50$\%. 
-		}
-		\label{fig:mismatch_hist}
-	\end{centering}
-\end{figure}
-
 In Fig.~\ref{fig:mismatch_hist}, we show the distribution of log mismatches for a set of test waveforms.
@@
 
+% \begin{figure}[t!]
+%     \script{sky_localization.py}
+%     \centering
+%     \includegraphics[width=\linewidth]{figures/sky_localization.pdf}
+%     \caption{
+%         Distribution of Fisher information sky localization error for a population of nearby binaries.
+%         Automatic differentiation (AD) was used to compute the derivatives in Eq.~\eqref{eq:fisher}.
+%         As emphasized in the text, after JIT compilation, each error calculation was around 100 times faster than a similar forecasting code \texttt{GWbench}~\citep{Borhanian:2020ypi}.
+%     }
+%     \label{fig:sky_localization}
+% \end{figure}
+
 \begin{figure}[t!]
-    \script{sky_localization.py}
+    \script{chirp_mass_error.py}
     \centering
-    \includegraphics[width=\linewidth]{figures/sky_localization.pdf}
+    \includegraphics[width=\linewidth]{figures/chirp_mass_error.pdf}
     \caption{
-        Distribution of Fisher information sky localization error for a population of nearby binaries.
+        Distribution of Fisher information chirp mass error for a population of nearby binaries.
         Automatic differentiation (AD) was used to compute the derivatives in Eq.~\eqref{eq:fisher}.
-        As emphasized in the text, after JIT compilation, each error calculation was around 100 times faster than a similar forecasting code \texttt{GWbench}~\citep{Borhanian:2020ypi}.
+        As emphasized in the text, after JIT compilation, each error calculation was between three and five times faster than a similar forecasting code \texttt{GWbench}~\citep{Borhanian:2020ypi}.
     }
-    \label{fig:sky_localization}
+    \label{fig:chirp_mass_error}
 \end{figure}
@@
 
-To illustrate the computational speed of computing Fisher matrices with AD, we consider a population of binaries and compute the sky localization error following Eq.~(28) in \cite{Iacovelli:2022bbs, Iacovelli:2022mbg}.\footnote{
-    Note that we include all 11 parameters when computing the covariance matrix, and only plot the sky localization error for illustrative purposes.
+To illustrate the computational speed of computing Fisher matrices with AD, we consider a population of binaries and compute the $1\,\sigma$ chirp mass error\cite{Iacovelli:2022bbs, Iacovelli:2022mbg}.\footnote{
+    Note that we include all 11 parameters when computing the covariance matrix, and only plot the chirp mass error for illustrative purposes.
 }
-Since the Fisher matrix approach is known to have both theoretical issues as well as numerical instabilities for low signal-to-noise events, we restrict our population to only nearby systems.
+Since the Fisher matrix approach is known to have both theoretical issues as well as numerical instabilities for low signal-to-noise events, we restrict our population to only nearby (high SNR) systems.
 A full list of the distributions used to generate the various parameters in our population are given in Tab.~\ref{tab:priors}. 
-Additionally, we use $20\,$Hz, $1024\,$Hz, and $16\,$s for the minimum frequency, sampling frequency, and sample length.
+Additionally, we use $20\,$Hz, $2048\,$Hz, and $16\,$s for the minimum frequency, sampling frequency, and sample length.
 Our noise curves correspond to the design PSDs for LIGO Hanford, LIGO Livingston (\textsc{SimNoisePSDaLIGOZeroDetHighPower}) and Virgo (\textsc{SimNoisePSDAdvVirgo}).\footnote{\url{https://lscsoft.docs.ligo.org/lalsuite/lalinspiral/psds_8py_source.html}.}
@@
 
-The distribution of sky localization errors from a population of $10^3$ binaries can be seen in Fig.~\ref{fig:sky_localization}.
+The distribution of chirp mass errors from a population of $5 \times 10^3$ binaries can be seen in Fig.~\ref{fig:chirp_mass_error}.
 We have verified that our errors agree with a separate dedicated Fisher forecasting code~\citep{Borhanian:2020ypi} to within ~30\%.
@@
 Moreover, each error calculation (including computing the Fisher matrices for each detector and the inversion process) is substantially faster.
-In particular, we find that after compilation, each Fisher calculation takes approximately half a second on a single computing core.
-\texttt{GWbench}~\citep{Borhanian:2020ypi}, on the other hand, takes $\mathcal{O}$(minutes) for each Fisher calculation using the same detector setup and frequency grid.
-This factor of over 100 speed up is substantial considering the fact that a single core evaluation of the \ripple waveform is similar to \lalsuite which is used by \texttt{GWbench}.
-On a MacBook Pro with an M1 Max Apple Silicon processor, JIT compilation takes $\sim 270\,s$ and computing sky localization errors for the full population takes less than $\sim 9$ minutes.
+In particular, we find that after compilation, each Fisher calculation takes approximately one second on a single computing core.
+\texttt{GWbench}~\citep{Borhanian:2020ypi}, on the other hand, takes $\mathcal{O}$(3-5 seconds) for each Fisher calculation using the same detector setup and frequency grid.
+This factor of 3-5 speed up is substantial considering the fact that a single core evaluation of the \ripple waveform is similar to \lalsuite which is used by \texttt{GWbench}.
+On a MacBook Pro with an M1 Max Apple Silicon processor, JIT compilation takes $\sim 70\,s$ and computing chirp mass errors for the full population takes less than $\mathcal{O}$(2 hours).
 As discussed above, performance can be further improved by utilizing hardware acceleration such as parallel GPU processing.

### 2024-01-11 src/scripts/chirp_mass_error.py

@@ -1,0 +1,77 @@
+import matplotlib.pyplot as plt
+import numpy as np
+import paths
+import matplotlib as mpl
+
+params = {
+    "font.size": 18,
+    "legend.fontsize": 18,
+    "legend.frameon": False,
+    "axes.labelsize": 18,
+    "axes.titlesize": 18,
+    "xtick.labelsize": 18,
+    "ytick.labelsize": 18,
+    "figure.figsize": (7, 5),
+    "xtick.top": True,
+    "axes.unicode_minus": False,
+    "ytick.right": True,
+    "xtick.bottom": True,
+    "ytick.left": True,
+    "xtick.major.pad": 8,
+    "xtick.major.size": 8,
+    "xtick.minor.size": 4,
+    "ytick.major.size": 8,
+    "ytick.minor.size": 4,
+    "xtick.direction": "in",
+    "ytick.direction": "in",
+    "axes.linewidth": 1.5,
+    "text.usetex": False,
+    "font.family": "serif",
+    "font.serif": "cmr10",
+    "mathtext.fontset": "cm",
+    "axes.formatter.use_mathtext": True,  # needed when using cm=cmr10 for normal text
+}
+
+
+mpl.rcParams.update(params)
+
+(
+    Mc,
+    eta,
+    chi1,
+    chi2,
+    dist,
+    tc,
+    phic,
+    inclination,
+    polarization_angle,
+    ra,
+    dec,
+    sky_error_list,
+    SNR_list,
+    Mchirp_err_list,
+) = np.loadtxt(paths.data / "sky_localization.txt", unpack=True)
+
+
+# Load GWbench data
+GWbench_Mcs = np.loadtxt(paths.data / "GWbench_Mcs.txt")
+
+plt.figure(figsize=(7, 5))
+bins = np.geomspace(0.1, np.max(Mchirp_err_list), 30)
+plt.hist(Mchirp_err_list, bins=bins, density=True, alpha=0.3, color="C0")
+plt.hist(Mchirp_err_list, bins=bins, density=True, histtype="step", color="C0", lw=2, label="Ripple")
+
+plt.hist(GWbench_Mcs, bins=bins, density=True, alpha=0.3, color="C3")
+plt.hist(GWbench_Mcs, bins=bins, density=True, histtype="step", color="C3", lw=2, label="GWBench")
+
[…]

### 2024-01-23 src/tex/ms.tex
fixed typo (RWHM -> RWMH)
minor typo corrections

@@
 This makes a \jax implementation that complies with the constraints of JIT compilation simple.
-NRsurrogate models, which interpolate directly over waveforms produced by numerical relativity (NR) simulations, are in principle also straightfoward to implement in \jax and will be explored int he future.
+NRsurrogate models, which interpolate directly over waveforms produced by numerical relativity (NR) simulations, are in principle also straightfoward to implement in \jax and will be explored in the future.
 EOB waveforms on the other hand are produced by evolving the dynamics of an effective one body Hamiltonian, and are therefore more difficult to implement in \jax.\footnote{
@@
 In Sec.~\ref{sec:applications} we discuss three distinct applications using differentiable waveforms. 
-Firstly, we illustrate how the coefficients that form part of the IMRPhenomD waveform model could be improved by high dimensional fitting enabled by a automatically differentiable waveform. 
+Firstly, we illustrate how the coefficients that form part of the IMRPhenomD waveform model could be improved by high dimensional fitting enabled by an automatically differentiable waveform. 
 Secondly, we implement differentiable detector response functions and show that the speed of Fisher matrix calculations can be substantially accelerated using AD.
@@
     }
-This is shown in Fig.~\ref{fig:match}, where we have calculated the $\mathcal{m}(h_1|h_2)$ across the entire parameter space.\footnote{
+This is shown in Fig.~\ref{fig:match}, where we have calculated the $\mathcal{M}(h_1|h_2)$ across the entire parameter space.\footnote{
     Specifically, we use $10^4$ points with varying component masses $m_1$, $m_2$ and spin parameters $\chi_1$, $\chi_2$ in the ranges $m_{1,2} = (1,100)\,M_{\odot}$ and $\chi_{1,2} = (-1,1)$.
@@
 
-One of the primary aims of this paper is that waveform derivatives will also be highly valuable to data analysis tasks. 
+One of the primary aims of this paper is to show that waveform derivatives will also be highly valuable to data analysis tasks. 
 AD provides two big advantages when it comes to evaluating derivatives compared to numerical differentiation.
@@
 			Relative error between the numerical relativity and IMRPhenomD waveform amplitudes. 
-            The blue line shows the relative error of the IMRPhenomD model with the original model coefficients where as the orange line uses the IMRPhenomD model after optimization using gradient descent.
+            The blue line shows the relative error of the IMRPhenomD model with the original model coefficients where as the red line uses the IMRPhenomD model after optimization using gradient descent.
             Interestingly, the error is reduced across the frequency range, demonstrating that high dimensional optimization is useful for all parts of the waveform.
@@
 		\caption{
-			Distribution of 536 log mismatches (see Eq.~\eqref{eq:mismatch}) for the original (blue) and optimized (orange) IMRPhenomD. 
+			Distribution of 536 log mismatches (see Eq.~\eqref{eq:mismatch}) for the original (blue) and optimized (red) IMRPhenomD. 
             Overall, the optimized distribution shifts to lower mismatches and therefore a better waveform model. 
@@
 
-To illustrate the computational speed of computing Fisher matrices with AD, we consider a population of binaries and compute the $1\,\sigma$ chirp mass error\cite{Iacovelli:2022bbs, Iacovelli:2022mbg}.\footnote{
+To illustrate the computational speed of computing Fisher matrices with AD, we consider a population of binaries and compute the $1\,\sigma$ chirp mass error.\footnote{
     Note that we include all 11 parameters when computing the covariance matrix, and only plot the chirp mass error for illustrative purposes.
@@
 In particular, we see that the autocorrelation reaches close to zero at around 300 lag, meaning that samples become efficiently uncorrelated.
-This is in contrast to RWHM which remains highly correlated even at 1000 lag.
+This is in contrast to RWMH which remains highly correlated even at 1000 lag.
 We therefore expect gradient based samplers to converge significantly faster than typical samplers, especially in higher dimensions.
@@
 T.E.
-would like to thanks Luca Reali extensively for his help with running GWBench.
+would like to thank Luca Reali extensively for his help with running GWBench.
 This material is based upon work supported by NSF's LIGO Laboratory which is a major facility fully funded by the National Science Foundation.

### 2024-01-25 src/tex/ms.tex
updating showyourwork
typo correction

@@
 While we focused on \texttt{IMRPhenomD} here, the ability to apply AD to the calibration parameters may assist in other approaches to calibration, such as that used for the aligned-spin EOB model in~\citet{Bohe:2016gbl}.
-Automative derivatives, if implemented for EOB waveforms, could allow for the application of other sampling methods such as HMC, or possibly optimization over the entire set of NR waveforms at once.
+Automatic derivatives, if implemented for EOB waveforms, could allow for the application of other sampling methods such as HMC, or possibly optimization over the entire set of NR waveforms at once.
 

### 2024-06-06 src/scripts/timing.py
reply updates

@@ -45,11 +45,13 @@ loop_time_list_lalsuite = np.loadtxt(paths.data / "loop_time_lalsuite.txt")
 loop_time_list = np.loadtxt(paths.data / "loop_time.txt")
 vmapped_time_list = np.loadtxt(paths.data / "vmapped_time.txt")
 gpu_time_list = np.loadtxt(paths.data / "vmapped_time_gpu.txt")
+gpu_time_list_float32 = np.loadtxt(paths.data / "vmapped_time_gpu_float32.txt")
 
 plt.plot(M_list, loop_time_list_lalsuite * 1000, label="Lalsuite", lw = 3, color="k")
 plt.plot(M_list, loop_time_list * 1000, label="CPU (loop)", lw = 2, ls="-.")
 plt.plot(M_list, vmapped_time_list * 1000, label="CPU (vmap)", ls="--", lw = 2)
 plt.plot(M_list, gpu_time_list * 1000, label="GPU", ls=":", lw = 2 )
+plt.plot(M_list, gpu_time_list_float32 * 1000, label="GPU float32", ls=":", lw = 2 )
 plt.yscale("log")
 plt.xlabel("Total mass, $M$ [M$_\odot$]")
 plt.ylabel("Time [ms]")

### 2024-06-06 src/tex/ms.tex
reply updates

@@
 
-Again, performing the same benchmark as above on an NVIDIA A100 with 40GB of memory, we find that waveform evaluations are over an of magnitude faster than serial CPU evaluation.
+Again, performing the same benchmark as above on an NVIDIA A100 with 40GB of memory, we find that waveform evaluations are over an order of magnitude faster than serial CPU evaluation.
 Generalizing \lalsuite waveforms to run on a GPU would be a significant undertaking.
@@
 AD provides two big advantages when it comes to evaluating derivatives compared to numerical differentiation.
-First, the accuracy of derivatives from AD are significantly more stable than finite-difference methods.
+First, the accuracy of derivatives from AD is significantly more stable than finite-difference methods.
 In particular, finite differences suffer from both rounding and truncation errors, meaning that the user is required to \textit{tune} the width over which the difference is taken.
@@
 
-Having an accurate waveform model is essential for many data analysis tasks throughout
-GW science.
+Having an accurate waveform model is essential for many data analysis tasks throughout GW science.
 While waveforms generated using NR simulations are in principle the highest fidelity signal model, they are too computationally expensive to be used in any practical data analysis tasks.
@@
 Since the coefficients for each segment are tuned individually, the correlation between different parameters are ignored, meaning that the provided best-fit solution may not be the global optimum. 
-We therefore aim to improve of the accuracy of the waveform by jointly fitting all coefficients at once.
+We therefore aim to improve the accuracy of the waveform by jointly fitting all coefficients at once.\footnote{
+    For a further breakdown of how the optimization changes across the parameter space, see our accompanying paper~y\citep{Lam:2023oga}.
+}
 
@@
 These 11 waveforms are taken from the SXS catalog~\citep{Boyle:2019kee}.
-We then run gradient descent, as described above, until the validation loss stops decreasing (the validation is loss is calculated using the NR waveforms discussed below). 
+We then run gradient descent, as described above, until the validation loss stops decreasing (the validation loss is calculated using the NR waveforms discussed below). 
 Figure~\ref{fig:loss_compare} shows the relative error (against a test NR waveform; see below) of the original and optimized waveform as a function of dimensionless frequency.
@@
 In Fig.~\ref{fig:mismatch_hist}, we show the distribution of log mismatches for a set of test waveforms.
-In particular, we use 536 waveforms from the SXS catalog (we choose all waveforms with aligned spins, $\chi_{x,y}<5\times 10^{-3}$, and eccentricity $<2\times 10^{-3}$)~\citep{Boyle:2019kee}.
+In particular, we use 536 waveforms from the SXS catalog i.e., all waveforms with aligned spins, $|\chi_{x,y}|<5\times 10^{-3}$, and eccentricity $<2\times 10^{-3}$~\citep{Boyle:2019kee}.
 One can see that the distribution of mismatches after optimization is generally shifted to lower mismatch compared to the original waveform.
-In particular, the peak of the original waveform distribution has moved by nearly an order of magnitude, indicating that our AD-assisted optimization procedure provides an improved implementation of the model.
+In particular, the median mismatch (shown as vertical dashed lines in Fig.~\ref{fig:mismatch_hist}) is reduced by $\sim50$\%, indicating that our AD-assisted optimization procedure provides an improved implementation of the model.
 
@@
         Automatic differentiation (AD) was used to compute the derivatives in Eq.~\eqref{eq:fisher}.
-        As emphasized in the text, after JIT compilation, each error calculation was between three and five times faster than a similar forecasting code \texttt{GWbench}~\citep{Borhanian:2020ypi}.
+        As emphasized in the text, after JIT compilation, each error calculation was between three and five times faster than a numerical derivative forecasting code \texttt{GWbench}~\citep{Borhanian:2020ypi}.
     }
@@
 Forecasting the sensitivity of future experiments is a routine task in GW science. 
-Due to its theoretical simplicity and evaluation speed, the Fisher matrix formalism~\citep{Cornish:2010kf} is commonly deployed to estimate how well a binary system's parameters could be measured.
+Due to its theoretical simplicity and evaluation speed, the Fisher matrix formalism (see e.g.,~\citep{Cornish:2010kf}) is commonly deployed to estimate how well a binary system's parameters could be measured.
 The Fisher matrix approach is built around the assumption of a Gaussian likelihood~\citep{Vallisneri:2007ev}.
@@
 \end{equation}
-Note that here we have separated out the extrinsic ($\phi$) and intrinsic ($\Xi$) variables as well as introducing the detector projection functions for the plus and cross polarizations as $F^k_+$ and $F^k_\times$ respectively.
+Note that here we have separated out the extrinsic ($\phi$) and intrinsic ($\Xi$) variables as well as introduced the detector projection functions for the plus and cross polarizations as $F^k_+$ and $F^k_\times$ respectively.
 Since we are considering a three detector setup we simply add the Fisher matrices from the individual detectors to get the combined Fisher matrix:
@@
 The distribution of chirp mass errors from a population of $5 \times 10^3$ binaries can be seen in Fig.~\ref{fig:chirp_mass_error}.
-We have verified that our errors agree with a separate dedicated Fisher forecasting code~\citep{Borhanian:2020ypi} to within ~30\%.
+We have verified that our errors agree with a separate dedicated Fisher forecasting code~\citep{Borhanian:2020ypi} to within ~30\%.\footnote{
+    Note that since Ref.~\citep{Borhanian:2020ypi} is based on numerical derivatives, the agreement is not expected to be close to machine precision.
+}
 This demonstrates that AD can be used to accurately do Fisher forecasting for a large population of events.
@@
 
-After the search algorithms have constructed a list of confidently detected binaries, the next step is to sample from the posterior of each sources parameter's - so called PE.
+After the search algorithms have constructed a list of confidently detected binaries, the next step is to sample from the posterior of each source parameter - so called PE.
 To do this, one typically uses an MCMC or nested sampler~\citep{Skilling2004, multinest, dynesty}.
@@
 
-Although we primarily discuss toy examples in this paper, each can be extended to the full data analysis task, some of which will be shown in upcoming papers~\citep{PEpaper}.
+Although we primarily discuss toy examples in this paper, each can be extended to the full data analysis task, some of which will be shown in upcoming papers~\citep{PEpaper, Lam:2023oga}.
 Differentiable waveforms therefore represent a useful advancement towards efficient GW science.

### 2024-06-07 src/tex/ms.tex
reply updates

@@
 We therefore aim to improve the accuracy of the waveform by jointly fitting all coefficients at once.\footnote{
-    For a further breakdown of how the optimization changes across the parameter space, see our accompanying paper~y\citep{Lam:2023oga}.
+    For a further breakdown of how the optimization changes across the parameter space, see our accompanying paper~\citep{Lam:2023oga}.
 }
